### 2024-10-29 paper.tex
Start working on a paper

\documentclass[sigconf]{acmart}
%\documentclass[sigconf,review,anonymous]{acmart}

\usepackage{enumitem}
\setlist{leftmargin=1.5em}
\setlength\itemsep{0.5em}

\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title[Denicek: Programming Substrate for Concrete, Collaborative, Interactive
  Programming]{{\scshape Denicek}: Programming Substrate for Concrete, \\Collaborative, Interactive Programming}

\author{Tomas Petricek}
\email{tomas@tomasp.net}
\orcid{0000-0002-7242-2208}
\affiliation{%
  \institution{Faculty of Mathematics and Physics, Charles University}
  \city{Prague}
  \country{Czech Republic}
}

\begin{abstract}
Research on interactive programming systems gave rise to a range of programming experiences,
including programming by demonstration, local-first collaborative editing, structure
editing, schema and code co-evolution, provenance tracking and output invalidation. Those
experiences are compelling, but they are hard to implement on the basis of existing programming
languages and systems.

We contribute the Denicek computational substrate. Denicek represents a program as a series of edits
that construct or transform a document consisting of data and formulas. Denicek provides two
primitive operations on series of edits, merging and conflict checking, that form the backbone
of the implementation of the aforementioned programming experiences.

We discuss the architecture of Denicek, document notable design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of a simple innovative data exploration environment.
The case study shows that the Denicek computational substrate provides a pathway to the design
of richer and more accessible interactive programming systems.
\end{abstract}

\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
  Your, Paper}

%\begin{teaserfigure}
%  \includegraphics[width=\textwidth]{sampleteaser}
%  \caption{Seattle Mariners at Spring Training, 2010.}
%  \Description{Enjoying the baseball game from the third-base
%  seats. Ichiro Suzuki preparing to bat.}
%  \label{fig:teaser}
%\end{teaserfigure}
\maketitle

\newpage
~
\newpage
\section{Introduction}
intro

\subsection{Programming Experiences}
list

\subsection{Substrate}
how it works very roughly

\subsection{Contributions}
one main thing - substrate - with other things

\section{Background}
all the references

\section{Walkthrough}
conference organizer

1) PBD and interaction

2) Merging and schema evolution

3) Evaluation and maybe provenance

\section{Architecture}

also NAIVE REALISM

\section{Implementation}
(how each of the features is achieved)

1) PBD (create textbox but not button)

2) Merging (refactoring)

3) Interaction (add button)

4) Evaluation and invalidation

5) Schema evolution

\section{Design Discussion}
\section{Case study}
(data science environment)

\section{Discussion}
\subsection{Heuristic evaluation}
\subsection{Limitations}
maybe

DESIGN PROCESS = formative examples + evaluation case study

xx
\newpage
xx

~\\
introduction \\
background \\
related work \\
design process / goals \\
case study \\
formative research \\
formative study \\
analysis \\
system \\
implementation \\
evaluation / heuristic evaluation \\
discussion and limitations \\

\newpage

\section{Introduction}

The computational substrate using which software is built determines the capabilities that the
software can provide. An imperative substrate that views programs as instructions modifying
bytes in memory makes it almost impossible to allow end-user inspection or reprogramming of
running software.

A computational substrate defines what software is built from. This may be objects as in
Smalltalk, lists as in Lisp, or memory with data and code as in UNIX/C.
The different substrates enable different kinds of programming experiences.
For example, object-oriented programming has historically been linked to the development
of graphical user interfaces (where objects can correspond to elements on the screen).
It has also enabled the development of visual programming environments such as the Alternate
Reality Kit, based on message sending between objects.

In principle, any computational substrate can be used to develop any programming experience,
but the greater the impedance mismatch between the substrate and the desired experience,
the more difficult it will be to provide the experience and combine it with the rest of the
system and other programming experiences developed for the system. (One can implement support
for programming-by-demonstration using C/C++, for example as part of a game scripting engine,
but it will not work with the rest of the ordinary C/C++ ecosystem.)

\subsection{Substrate}

The question asked in this paper is, what would be the ideal programming substrate
for supporting a range of programming experiences that make programs more
collaborative, transparent and allows for a gradual transition from non-programmer
to a programmer. We want a programming substrate that makes it easy to develop
programming experiences such as:

\begin{itemize}
\item \emph{Programming by demonstration} --
  Allow non-programmers to construct simple programs by performing examples of the expected behaviour. \cite{leiva-2021-rapido}.
\item \emph{Local-first collaboration} --
  Multiple users should be able to use and modify a single program, preferrably without requiring a central server. \cite{kleppmann-2019-local}
\item \emph{Provenance tracking} --
  The execution of the program should leave an understandable trace that lets the user understand why program resulted in a particular result.
\item \emph{Schema evolution [extra-ish]} --
  When the user evolves the structure of the program, data and code should co-evolve automatically to match the new structure.
\item \emph{Notational freedom [extra-ish?]} --
  Allow users to adapt the program using a notation that suits them and is appropriate for the programming task at hand. [Joel]
\item \emph{Concrete programming [extra?]} --
  It should be possible to reuse parts of program or program logic without constructing abstractions, for example by managed copy \& paste.\cite{edwards-2006-copypaste,edwards-2022-copypaste}
\end{itemize}

substrate as defined by \cite{jakubovic-2022-ladder}

% pbd
\cite{leiva-2021-rapido,cypher-1993-pbd}
\cite{chen-2023-miwa}

% lcoal first
\cite{kleppmann-2019-local,klokmose-2024-mywebstrates}
% provenance
\cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
\cite{ricciotti-2017-imperative,perera-2012-functional}
\cite{perera-2022-linked}

% visualizations of results or execution
% https://www.dcs.warwick.ac.uk/pvw04/p01.pdf
% https://dl.acm.org/doi/10.1145/3313831.3376494

% previews - livelits

% Schema evolution
% our paper; see https://openproceedings.org/2023/conf/edbt/paper-160.pdf - good refs in 2.2

% https://x.com/jonathoda/status/1185888711210389504

% also cool: https://arxiv.org/pdf/2303.06777

% Hazel family of things
% structure editing - sandblocks

Joel's definition of substrate in Onward!
Bret Victor talk
\url{https://www.youtube.com/watch?v=ef2jpjTEB5U}

In what ways is a substrate "natural"?

thinglab - create line by cloning, it sticks to mouse pointer, clicking sticks it to something else
squeak - has all the browsers (method search...)

computational substrate
how it differs from computational media?
more low-level - media suggests that there it comes

\section{The whatever system}
\subsection{Document + Edits}
defines
\begin{itemize}
  \item selectors
  \item nodes
  \item edits
\end{itemize}

\subsection{Walkthrough}
* todo list? (or counter, but that is a bit boring)

\section{Themes}
* programming by demonstration
  - binding interactions to gui elements (event handlers)
* provenance tracking
  - Amy Ko's whyline, Probe Log by HPI, enables linked visualizations
* merging of edit histories / collaborative editing
  - bonus - can share restricted link to allow users fill out
    forms (allow partial edits only / def by selector?)
* scehma change - change data \& code accordingly
* everything is an edit
  - interaction with the GUI
  - evaluation? tbd
* copy \& paste abstraction
    (requires finishing new approach to formulas!)
  - edit before copy to propagate edit to other places
    (or edit after copy to make it specific to a case)
  - higher order copying from https://tomasp.net/academic/papers/copy-paste/paint22.pdf
* augmenters
  - cf. bonnie nardi (calls them something else - Jonathan says)
  - add programming by demonstration data wrangling gui to table (trigger interactions)
    cf. lorgnette

\section{Applications}
* todo list / counter / maybe too simple
* (if used in the walkthrough, maybe something else? board game as in varv - tic tac toe? or 7guis?)
* conference organizer
* data exploration (ala histogram)
* linked charts

\section{Extras}
* metablocks?
* self-sustainability
* some non-browser implementation of this (as in Varv?)

explicit structure
self-sustainability
notational freedom

\newpage
~

Maybe have 'enabled' for edits afterall?
(we can merge with conflicts and disable some edits, but keep them in history for info)

NOTES
type Edit =
  { Kind : EditKind
    Dependencies : Selectors list }  -- only needed for evaluated edits

VALUE vs STRUCTURE distinction
* good in theory, nice for implementation
* tricky to use! needs some assistance tools

TODO - things to work on
* "represent" edits somewhere in document as "library of functions"
  and then call those from buttons (rather than embedding them directly)
  allow some kind of abstraction (as in Histogram) to make them reusable
* figure out how to do evaluation better
  (based on the stored abstractions? but need to store provenance...)

SEMANTIC CONDITIONS
\url{https://www.youtube.com/watch?v=NBnc2ToS_j0}
(has a section on this in background)

SUBSTRATE DESIGN PROBLEMS
* selectors - all for structure / index for data
  (but it is useful to allow others...)
  (multiselect also bad for checks!)
* groups/conditions/preconditions
  (c.f. email to jonathan)
  tried conditions on edits; trying groups with check edits
* what to do with "disabled edits"? for example when we remove all checked
  (before, this created edit groups with "check" but if the check was false,
  the group was ignored and this messed up merging - because we wouldn't know if the
  edit had any effect or not)

Evaluation
* evaluated edits have to be migrated to the end
  (if there are conflicts, they are dropped)
  Think of this as maintaining a tree:

  e3
  |
  e2   evaluated
  |  /
  e1
  |
  e0

  this has to be serialized as e0 -> e1 -> e2 -> e3 -> evaluated

  evaluated edits do not became part of the main history
  but hang on the side

ISSUES
* if we merge a thing with saved-interactions with something, hashes will change!

NOTES
* ListAppendFrom - we need this, because we cannot encode this.
* for records, we can RecordAdd(sel, fld, ..) @ Copy(sel @ [Field fld], src) but
  this does not work for lists - because we do not know the index!
  (and we cannot look into current document, because it will differ for saved-interactions)

TODO
* many things with <tag> selectors currently do not work
  (e.g. `matches` for highlighting) because if we collect path of a current node,
  we collect indices and get /some/2/another - and cannot tell if this matches
  /some/<li>/another - we'd have to collect more detailed path info!

INTERACTION
* replay stored event handlers against the old version?
  (this way, adding an item to a speakers list gets migrated \& adds a new table row!)
* simlarly!! we need merge in order to apply edits to multiple targets
  (when you remove all items in a list, the indices change)
  (but I guess we should do this against version at the time of saving too....)

[this \& evaluation = the unreasonable effectiveness of merging]

Notes on storing and reusing edits
* references need to be represented as references so that they get updated
  (NO! not if we reply them against old version, which seems better - but there are 2 design choices)
* how to apply them to multiple targets? use Move to update the selectors instead of
  replacing the prefix manually

IDEA: Type check edit groups to ensure they preserve structure but not individual edits eg when adding list item

CONDITIONALS
https://toby.li/files/p311-radensky.pdf

REMAINING IMPLEMENTATION TODOs:

* Some kind of provenance visualization
* Some kind of matchers/transformers mechanism (ideally to add interactive buttons to tables)
* Apply to all (remove completed in TODO)

\newpage
~

\bibliographystyle{plain}
\bibliography{paper}

\end{document}

### 2024-10-30 paper.tex
System design - work in progress

@@
 
+\usepackage{hhline,xcolor,colortbl}
+\usepackage{fontawesome}
 \usepackage{enumitem}
+\usepackage{fancyvrb}
 \setlist{leftmargin=1.5em}
@@
 
-We discuss the architecture of Denicek, document notable design considerations and elaborate
+We discuss the architecture of Denicek, document key design considerations and elaborate
 the implementation of the programming experiences listed above. To evaluate the proposed
@@
 \newpage
-~
-\newpage
+
 \section{Introduction}
-intro
+
+Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+
+TODO: Maybe do not talk about ``programs'' because this is more documents with formulas - not that fancy programs.
+Computational documents?
 
@@
 
-\section{Architecture}
+\newpage
 
-also NAIVE REALISM
+% ==================================================================================================
+
+\arrayrulecolor{lightgray}
+\definecolor{ekgray}{gray}{0.90}
+
+\begin{figure}
+\newcommand{\seltablecol}[3]{
+\sffamily\small{\bfseries #2} & {\footnotesize #1} & \footnotesize #3\\
+}
+\newcommand{\ndtablecol}[4]{
+\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize #3}\\[-0.2em]
+&\sffamily\footnotesize #4\\[0.3em]
+}
+
+\begin{tabular}{|clp{18.08em}|}
+\hline
+\rowcolor{ekgray}
+&&\\[-1em]
+\rowcolor{ekgray}
+\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
+\hline
+&&\\[-1em]
+\seltablecol{\Verb|field|}{Field}{Refers to record field of a given name}
+\seltablecol{\Verb|number|}{Index}{Refers to list element at a given index}
+\seltablecol{\Verb|*|}{Any}{Refers to all children of a list node}
+\hline
+\end{tabular}
+
+~\\[0.5em]
+
+\begin{tabular}{|cl|}
+\hline
+\rowcolor{ekgray}
+&\\[-1em]
+\rowcolor{ekgray}
+& \sffamily\small{\bfseries Node}\;\,\textit{\footnotesize arguments} \\[0.2em]
+\hline
+&\\[-1em]
+\ndtablecol{\faListUl}{List}{tag, child$_1$, $\ldots$, child$_n$}
+  {Ordered list of nodes, addressable by index. Renders as \Verb|<tag>| with children}
+\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
+  {Record with children addressable by \textit{field} name. Renders as \Verb|<tag>|.}
+\ndtablecol{\faExternalLink}{Reference}{selectors}
+  {Reference to another document location. Displays the \Verb|/selectors| as a link.}
+\ndtablecol{\faFont}{Primitivie}{string \textnormal{or} number}
+  {Numerical or textual primitive value. Renders as an HTML text node.}
+\hline
+\end{tabular}
+\vspace{-0.5em}
+\caption{Structure of selectors and document nodes.}
+\label{fig:doc}
+\end{figure}
+
+% ==================================================================================================
+
+\section{The Denicek Substrate}
+\label{sec:system}
+Denicek represents programs as a sequence of edits construct a computational document.
+In this section, we describe the structure of documents and edits, as well as the
+operations that form the backbone of the system and are used to implement a range of
+interesting programming experiences, discussed in \S\ref{sec:impl}.
+
+\subsection{Selectors, Documents and Edits}
+Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
+computational document visible to the user, although parts can be collapsible or hidden using CSS.
+Records and lists are rendered as HTML elements of a specified tag with children becoming child
+elements. Field names are hidden in the rendered document.
+
+References to document location are used in both document itself (reference nodes) and in edits
+(target and a source). They are represented as a sequence of selectors (Fig.~\ref{fig:doc}).
+The document model assumes that lists are homogeneous and records heterogeneous. Thus the
+\textit{Any} selector makes it possible to refer to all children of a list, but there is no
+way to refer to all children of a record.
+
+\subsubsection*{Document Edits}
+The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
+require \emph{target} to which they are applied. The target is specified through selectors.
+Most edits can only be applied when the target is a node of a particular kind.
+
+The edit operations (Fig.~\ref{fig:edits}) are designed to allow any transformation of a document
+through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
+Denicek updates selectors when document structure changes. Fig.~\ref{fig:edits} distinguishes
+between edits that keep existing references in a document unchanged (above) and edits that affect
+references (below). When a selector is \emph{invalidated}, e.g., when deleting a field or a
+list item to which there is a reference, Denicek prevents the edit. Copying also invalidates
+selectors because it is ambiguous whether selectors referring to the original location should
+refer to the source or the target of the copying after the edit (and since reference can only
+refer to one location, duplicating it is not possible). Updating a field, wrapping or reordering
+requires changing existing references in document correspondingly.
+
+% ==================================================================================================
+
+\begin{figure}
+\newcommand{\ektablecol}[6]{
+\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 & \sffamily\footnotesize #5 \\[-0.2em]
+&\multicolumn{3}{l|}{\sffamily\footnotesize #6}\\[0.3em]
+}
+\begin{tabular}{|c|p{13em}rr|}
+\hline
+\rowcolor{ekgray}
+&&&\\[-1em]
+\rowcolor{ekgray}
+ & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target & \sffamily\footnotesize\bfseries Selectors \\[0.2em]
+\hline
+&&&\\[-1em]
+\ektablecol{\faPlus}{Add}{field, node}{Record}{Unchanged}
+  {Add \textit{node} as a \textit{field} to the specified record.}
+\ektablecol{\faAt}{Append}{node}{List}{Unchanged}
+  {Append \textit{node} to the end of the specified list.}
+\ektablecol{\faPaperclip}{AppendFrom}{selectors}{List}{Unchanged}
+  {Append node from \emph{selectors} to the end of the specified list.}
+\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}{Unchanged}
+  {Apply primitive \textit{transform} to the specified primitive.}
+\ektablecol{\faCode}{UpdateTag}{old tag, new tag}{Tagged}{Unchanged}
+  {Change the tag of a specified list or record from \textit{old} to \textit{new}.}
+\hline
+&&&\\[-1em]
+\ektablecol{\faFont}{UpdateField}{old field, new field}{Record}{Update}
+  {Rename the field of a specified record from \textit{old} to \textit{new}.}
+\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}{Invalidate}
+  {Delete the field \textit{field} of a specified record.}
+\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Invalidate}
+  {Delete the item at a given \textit{index} of a specified list.}
+\ektablecol{\faSort}{Reorder}{permutation}{List}{Update}
+  {Reorder items of a specified list according to a \textit{permutation}.}
+\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}{Update}
+  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
+\ektablecol{\faListUl}{WrapList}{tag}{Any}{Update}
+  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
+\ektablecol{\faCopy}{Copy}{selectors}{Any}{Invalidate}
+  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
+\hline
+\end{tabular}
+\vspace{-0.5em}
+\caption{Types of document edits.}
+\label{fig:edits}
+\end{figure}
+
+% ==================================================================================================
+
+\subsubsection*{Automatic Reference Update}
+There are three situations in which automatic update of references is undesirable. First, if
+an edit is not applied to all elements of a list (reference contains the \emph{index} selector),
+it should not affect references that refer into the list. Such edits may turn document into an
+inconsistent state, but they typically do so temporarily during document construction
+(checking such edits is discussed in \S\ref{sec:discuss-future}).
+
+Second, documents can contain values that can be of multiple different kinds (i.e., a union type).
+In such case, references should not be updated when the kind of the value changes. For example,
+a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
+involves wrapping, but this should not affect references to the formula.
+To support those cases, it is possible to specify that edits that normally affect selectors
+(Fig.~\ref{fig:edits}, below) should not do so. Moreover, this setting is required when the
+target reference contains the \emph{index} selector.
+
+\subsubsection*{Non-Conditionality of Edits.}
+NON-CONDITIONALITY effect does not depend on document state
+
+\subsection{Merging histories}
+is tricky in various ways
+
+\subsubsection*{Updating structure for structural edits}
+yay
+
+\subsection{Conflict detection}
+is pretty easy
+
+\subsection{Properties}
+
+PROPERTIES
+- can change any document to any other - sure, via remove add - but also more semantically
+(if it contained all values, can we do it without removing and adding them?)
 
 \section{Implementation}
+\label{sec:impl}
 (how each of the features is achieved)
@@
 
-4) Evaluation and invalidation
+\subsection{Evaluation and invalidation}
+\label{sec:impl-eval}
 
@@
 \section{Design Discussion}
+
+Although Denicek does not explicitly track document structure (or schema, or type), all documents
+have an implicit structure.
+
+c.f. notes
+
+memory mapped graphics etc.
+
+unifying lists and records?
+
 \section{Case study}
@@
 \subsection{Limitations}
+\subsection{Future work}
+\label{sec:discuss-future}
+use type system to check temporarily invalid state
+
+ohter formal things
+* non-conditionality of edits
+* show that they can transform document from any to any without removing/readding values
+
 maybe

### 2024-11-01 paper.tex
Explain primitive operations, first draft

@@
 
+JUSTIFICATION - this is a technical paper but belongs here!
+
 \subsection{Programming Experiences}
@@
 how it works very roughly
+
+The coolest trick - almost everything (PBD, collaboration, evaluation) is done through merge!
 
@@
 
-\begin{tabular}{|clp{18.08em}|}
+\begin{tabular}{|llp{18.08em}|}
 \hline
@@
 \rowcolor{ekgray}
-& \sffamily\small{\bfseries Node}\;\,\textit{\footnotesize arguments} \\[0.2em]
+& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
 \hline
@@
 \vspace{-0.5em}
-\caption{Structure of selectors and document nodes.}
+\caption{Structure of selectors and document nodes}
 \label{fig:doc}
+\vspace{-1em}
 \end{figure}
@@
 \label{sec:system}
-Denicek represents programs as a sequence of edits construct a computational document.
-In this section, we describe the structure of documents and edits, as well as the
-operations that form the backbone of the system and are used to implement a range of
-interesting programming experiences, discussed in \S\ref{sec:impl}.
+Denicek represents programs as sequences of edits that construct and transform a computational
+document. In this section, we describe the structure of documents and edits, as well as the
+operations that form the backbone of the system and are used to implement a wide range of
+programming experiences, discussed in \S\ref{sec:impl}.
 
 \subsection{Selectors, Documents and Edits}
+A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
 Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
-computational document visible to the user, although parts can be collapsible or hidden using CSS.
+document visible to the user, although parts can be collapsible or hidden using CSS.
 Records and lists are rendered as HTML elements of a specified tag with children becoming child
@@
 References to document location are used in both document itself (reference nodes) and in edits
-(target and a source). They are represented as a sequence of selectors (Fig.~\ref{fig:doc}).
-The document model assumes that lists are homogeneous and records heterogeneous. Thus the
+(target of the edit). They are represented as a sequence of selectors (Fig.~\ref{fig:doc}).
+The document model assumes that lists are homogeneous and records heterogeneous, and so the
 \textit{Any} selector makes it possible to refer to all children of a list, but there is no
@@
 The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
-require \emph{target} to which they are applied. The target is specified through selectors.
-Most edits can only be applied when the target is a node of a particular kind.
+require \emph{target} to which they are applied. Target is a reference and can contain the
+\textit{Any} selector, in which case the edit is applied to multiple nodes simultaneously.
+Most edits can only be applied to target node(s) of a particular kind.
 
-The edit operations (Fig.~\ref{fig:edits}) are designed to allow any transformation of a document
+The edits are designed to allow any transformation of a document
 through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
-Denicek updates selectors when document structure changes. Fig.~\ref{fig:edits} distinguishes
+Denicek updates references when document structure changes. Fig.~\ref{fig:edits} distinguishes
 between edits that keep existing references in a document unchanged (above) and edits that affect
 references (below). When a selector is \emph{invalidated}, e.g., when deleting a field or a
-list item to which there is a reference, Denicek prevents the edit. Copying also invalidates
+list item to which there is a reference, Denicek rejects the edit. Copying also invalidates
 selectors because it is ambiguous whether selectors referring to the original location should
-refer to the source or the target of the copying after the edit (and since reference can only
-refer to one location, duplicating it is not possible). Updating a field, wrapping or reordering
-requires changing existing references in document correspondingly.
+refer to the source or the target of the copying after the edit (and references cannot refer to
+multiple structurally incompatible locations). Updating a field, wrapping, reordering or deleting
+a list item requires updating references in the document correspondingly.
 
@@
 }
-\begin{tabular}{|c|p{13em}rr|}
+\begin{tabular}{|cp{13em}rr|}
 \hline
@@
   {Delete the field \textit{field} of a specified record.}
-\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Invalidate}
+\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Update}
   {Delete the item at a given \textit{index} of a specified list.}
@@
 \vspace{-0.5em}
-\caption{Types of document edits.}
+\caption{Summary of document edit types in Denicek}
 \label{fig:edits}
+\vspace{-1em}
 \end{figure}
@@
 \subsubsection*{Automatic Reference Update}
-There are three situations in which automatic update of references is undesirable. First, if
-an edit is not applied to all elements of a list (reference contains the \emph{index} selector),
-it should not affect references that refer into the list. Such edits may turn document into an
-inconsistent state, but they typically do so temporarily during document construction
-(checking such edits is discussed in \S\ref{sec:discuss-future}).
+There are two situations in which automatic update of references is undesirable. If an edit is
+applied to a singular element of a list (reference contains the \textit{Index} selector),
+references that refer into any element of the list should be unchanged.  Such edits may turn
+document into an inconsistent state, but they typically do so temporarily during document
+construction (checking such edits is discussed in \S\ref{sec:discuss-future}).
 
-Second, documents can contain values that can be of multiple different kinds (i.e., a union type).
+Documents can also contain values that can be of multiple different kinds (i.e., a union type).
 In such case, references should not be updated when the kind of the value changes. For example,
@@
 involves wrapping, but this should not affect references to the formula.
-To support those cases, it is possible to specify that edits that normally affect selectors
-(Fig.~\ref{fig:edits}, below) should not do so. Moreover, this setting is required when the
-target reference contains the \emph{index} selector.
+To support those cases, it is possible to annotate edits that normally affect selectors
+(Fig.~\ref{fig:edits}, below) as non-structural. This annotation is required when the
+target reference contains the \emph{Index} selector.
 
-\subsubsection*{Non-Conditionality of Edits.}
-NON-CONDITIONALITY effect does not depend on document state
+\subsubsection*{No Conditional Edits.}
+An important aspect of the design is that the effect an edit has on references inside
+the document does not depend on the current value of the document. This makes it possible to
+define merging solely in terms of edits, without reference to current document state.
+(The effect of \textit{Copy} depends only on the existing document structure, but not on its value;
+while \textit{AppendFrom} affects only single list element and is thus labelled as a
+non-structural edit.)
 
-\subsection{Merging histories}
-is tricky in various ways
+This design choice makes it impossible to encode computational logic directly in the edits
+(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
+has to be provided as an additional  mechanism on top of the underlying Denicek substrate.
 
-\subsubsection*{Updating structure for structural edits}
-yay
+\subsection{Primitive Operations}
+Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
+sequences of edits can be checked for conflicts or merged. Denicek identifies edit histories by
+a (git-like) hash, computed from the hash of the parent and the current edit, which is used to identify
+common shared part of the history during conflict checking and merging.
 
-\subsection{Conflict detection}
-is pretty easy
+\subsubsection*{Applying Edits}
+When applying an edit, Denicek locates the target node and transform it according to the
+edit. If the edit may affect references in the document (Fig~\ref{fig:edits}, below), Denicek
+updates matching references in the document according to the rules shown in Fig.~\ref{fig:updates}
+provided that (i) the edit is not explicitly marked as non-structural, and (ii) the target
+reference does not contain \textit{Index} selector, i.e., it targets all elements of any involved
+lists. Also note that matching references in the document may be more specific than the edit
+target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
+a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. If the document
+contains a reference that would be invalidated by an edit, the edit is rejected.
 
-\subsection{Properties}
+\subsubsection*{Conflict Detection}
+Two edits are conflicting if the order in which they are applied matters. This is the case if
+they target the same node, or if one targets a node that is nested inside the node targetted by
+the other. For edits with dependencies (\textit{AppendFrom} and \textit{Copy}), a conflict also
+occurs if the other edit modifies the dependency.
 
-PROPERTIES
-- can change any document to any other - sure, via remove add - but also more semantically
-(if it contained all values, can we do it without removing and adding them?)
+Given two edit histories, one way to resolve conflicts is by removing (or marking as disabled) all
+edits from one of the histories that conflict with edits done by the other history. To do this,
+we first collect all targets of edits in the other history. We then iterate over edits from the first
+history to see if they depend on or target any of the affected targets. If an edit is removed,
+its target reference is added to the set of affected targets, so that other subsequent edits that depend
+on its original result are also removed.
+
+\subsubsection*{Merging Edit Histories}
+Removing all conflicting edits is sometimes useful, but there are many cases where conflicting
+edits, in the above sense, can be reconcilliated. Say we have edit histories with
+the same common shared part $E, E_1$ and $E, E_2$. If the edits in $E_1$ and $E_2$ conflict,
+the results of applying $E, E_1, E_2$ and $E, E_2, E_1$ would be different. We can, however,
+construct edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
+$E, E_2, E_1'$ are the same.
+
+The key operation that enables such reconcilliation takes two individual edits that occurred
+indpendently, $e_1$ and $e_2$, and produces $e_1'$ that has the same logical effect as
+$e_1$, but can be applied after $e_2$. There are two aspects of such reconcilliation:
+
+\begin{enumerate}
+\item \emph{Apply Edit to Newly Added.} If $e_1$ is adding a new list item, but~$e_2$ is
+  changing elements of the target list, we apply the edit $e_2$ to the node added by $e_1$ so
+  that, when it is added after the transformation, the added node has the new structure.
+
+\item \emph{Transform Matching References.} If $e_1$ targets a node that is inside a node
+  whose structure is changed by $e_2$, the target reference in $e_1$ is updated in a way that
+  corresponds to the new structure. That is, using the same rules, shown in Fig.~\ref{fig:updates},
+  that apply when transforming references inside a document.
+\end{enumerate}
+
+Applying an edit to a newly added node involves a number of cases. When the new node is added
+using \textit{Add} or \textit{Copy}, but the target location is modified, this is an unresolvable
+conflict (because those two operations would overwrite existing nodes). When the new node is
+added as a new list item using \textit{Append}, it can be transformed (by applying the other
+edit direclty to the new node).
+
+Finally, when the new node is added using \textit{AppendFrom},
+it is copied from another document location. We cannot transform the node in the soource location
+(this would have unintended effects) or after adding it (we do not know its new index because of
+aforementioned \emph{non-conditionality} of edits). Denicek reconcilliates such edits by first
+copying the source node to a new temporary location, transforming it there and then using
+\textit{AppendFrom} from the temporary location.
+
+% ==================================================================================================
+
+\begin{figure}
+\newcommand{\tttablecol}[5]{
+\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
+\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
+}
+\begin{tabular}{|l|}
+\hline
+\\[-1em]
+\tttablecol{UpdateField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
+  {Replace Field for matching references.}
+\tttablecol{DeleteItem}{target, index}{\Verb|/target/n/nested|}{\Verb|/target/(n-1)/nested|}
+  {Decrement Index greater than $n$ in matching refs.}
+\tttablecol{Reorder}{target, permutation}{\Verb|/target/n/nested|}{\Verb|/target/permutation(n)/nested|}
+  {Update Index using permutation in matching refs.}
+\tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
+  {Insert extra Field selector after matching prefix.}
+\tttablecol{WrapList}{target, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
+  {Insert extra All selector after matching prefix.}
+\hline
+\end{tabular}
+\vspace{-0.5em}
+\caption{How document edits transform references}
+\label{fig:updates}
+\vspace{-1em}
+\end{figure}
+
+% ==================================================================================================
 
@@
 
-3) Interaction (add button)
+\subsection{Interaction (add button + remove done)}
+\label{sec:impl-interaction}
 
@@
 
+AppendFrom - hard to avoid! because recorded edits cannot know length
+(but having -1 in index would break too)
+
 \section{Case study}
@@
 * show that they can transform document from any to any without removing/readding values
+
+PROPERTIES
+- can change any document to any other - sure, via remove add - but also more semantically
+(if it contained all values, can we do it without removing and adding them?)
 

### 2024-11-02 paper.tex
Intro to implementation

@@
+% TODO: Figure out how to call those "programming experiences" (also is this a good name?)
+%   the text uses "concrete, collaborative, interactive" - but this is a silly list
+%   really, they are "cool" but how to say that.
+%
+% TODO: Also, what exactly goes into the list?
+%
+%
 \documentclass[sigconf]{acmart}
@@
 \begin{document}
-\title[Denicek: Programming Substrate for Concrete, Collaborative, Interactive
-  Programming]{{\scshape Denicek}: Programming Substrate for Concrete, \\Collaborative, Interactive Programming}
+\title[Denicek: Computational Substrate for Concrete, Collaborative, Interactive
+  Programming]{{\scshape Denicek}: Computational Substrate for Concrete, \\Collaborative, Interactive Programming}
 
@@
 Research on interactive programming systems gave rise to a range of programming experiences,
-including programming by demonstration, local-first collaborative editing, structure
-editing, schema and code co-evolution, provenance tracking and output invalidation. Those
+including programming by demonstration, local-first collaborative editing, %structure editing,
+schema and code co-evo\-lution, provenance tracking and incremental recomputation. Those
 experiences are compelling, but they are hard to implement on the basis of existing programming
@@
 Computational documents?
+
+explain computational substrate
 
@@
 \label{sec:impl}
-(how each of the features is achieved)
 
-1) PBD (create textbox but not button)
+The key claim of this paper is that the computational substrate described in the previous
+section makes it easy to support a range of experiences that make programming more concrete,
+collaborative and interactive. In this section, we describe how to use the substrate to support
+local-first collaborative editing (\S\ref{sec:impl-collab}), programming by
+demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
+schema and code co-evolution (\S\ref{sec:impl-schema}), incremental recomputation
+(\S\ref{sec:impl-eval}), provenance tracking (\S\ref{sec:impl-provenance})
+and concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
+We describe the programming experience in isolation in this section. Next section provides a
+more comprehensive evaluation through a case study that combines multiple of them together.
 
-2) Merging (refactoring)
+\subsection{Merging (refactoring)}
+\label{sec:impl-collab}
+
+\subsection{PBD (create textbox but not button)}
+\label{sec:impl-pbd}
 
@@
 
-\subsection{Evaluation and invalidation}
+\subsection{Schema code co-evolution}
+\label{sec:impl-schema}
+
+\subsection{Incremental recomputation}
 \label{sec:impl-eval}
 
-5) Schema evolution
+\subsection{Provenance tracking}
+\label{sec:impl-provenance}
+
+\subsection{Managed copy \& paste}
+\label{sec:impl-copy}
 

### 2024-11-04 paper.tex
Write walkthrough

@@
 \usepackage{fancyvrb}
+\usepackage{tikz}
+\usepackage[breakable,theorems,skins]{tcolorbox}
+\usepackage{lettrine}
+
+\newcommand*\circled[1]{\textnormal{\footnotesize\sffamily\bfseries\protect\tikz[baseline=(char.base)]{
+  \node[shape=circle,fill=black,text=white,draw,inner sep=1pt] (char) {#1};}}}
+
+\DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
+  boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
+  colback=gray!20,colframe=gray!20,
+  enlarge bottom by=0pt,enlarge top by=0pt,
+  arc=0pt,outer arc=0pt]
+\lettrine[lraise=0.3]{\LARGE #1}{~}
+\small \textbf{#2.} #3
+\end{tcolorbox}
+}
+
 \setlist{leftmargin=1.5em}
@@
 
-\section{Walkthrough}
-conference organizer
+% ==================================================================================================
 
-1) PBD and interaction
-
-2) Merging and schema evolution
-
-3) Evaluation and maybe provenance
+\begin{figure*}[t]
+\vspace{-0.5em}
+\includegraphics[width=0.95\textwidth,clip,trim=1cm 1cm 1cm 0.5cm]{fig/walkthrough.pdf}
+\vspace{-1em}
+\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
+  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
+  formulas with schema and code co-evolution (F, G).}
+\label{fig:walkthrough}
+\end{figure*}
 
 \newpage
+
+\section{Walkthrough}
+To providea an overview of the programming experiences supported by the Denicek
+substrate, we use a web-based implementation of Denicek with a structure
+editor that supports navigation in document and issuing of edit commands. The editor
+implements support for a range of programming experiences, discussed in \S\ref{sec:impl}.
+
+\subsubsection*{\circled{A} Adding a Speaker}
+The user starts with an empty document, which is represented as a record. They add a field for
+each heading and a list \Verb_<ul>_ as a field named \Verb_speakers_. They use the command
+toolbox to add the first speaker.
+
+\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
+further speakers, the user creates a textbox and a button. They enter new speaker
+details, construct a temporary \Verb_<li>_ element outside of the main list, copy the value
+from the textbox into the new element using source view, and then copy the new element into the
+\Verb_speakers_ list.
+
+\subsubsection*{\circled{C} Saving the Interaction} After adding the speaker from the textbox,
+the user opens history view, selects edits that added the speaker and saves those as the
+\Verb_add-speaker_ interaction. They attach this as a handler for the \Verb_click_ event of the button.
+
+\keyideabox{\faLightbulbO}{Programming by Demonstration}{Denicek implements programming by
+demonstration (\S\ref{sec:impl-pbd}) by letting the user to save and replay past interactions,
+either directly or by attaching them to a UI event.}
+
+\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
+version of the document and turns the list into a table. This is done by invoking a series of
+commands that change tags, wrap elements, copy and transform values.
+
+\subsubsection*{\circled{E} Merging Edits} The two document versions can be automatically merged.
+The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
+button are also automatically turned into the new format.
+
+\keyideabox{\faLightbulbO}{Collaborative Editing and Interactivity}{Denicek's merging
+reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
+Merging is asymmetic, but the order does not matter here. The same merging operation is used
+when handling user interaction (\S\ref{sec:impl-interaction}).}
+
+\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds budget calculation to
+the initial document. Formulas are represented as special \Verb_x-formula_ nodes whose arguments
+are other nodes or references to other nodes or formulas (highlighted on mouse hover).
+
+\keyideabox{\faLightbulbO}{Incremental Recomputation}{Evaluating a formula yields additional
+edits that augment the document with the result. Those edits are kept
+at the top of the document history and are removed in case of a conflict (\S\ref{sec:impl-eval}),
+providing incremental recomputation.}
+
+\subsubsection*{\circled{G} Merging Formulas} When the budget calculation is merged with the
+other edits, references in formulas are automatically updated to point to the new list.
+Adding new speaker via the ``Add speaker!'' button invalidates the evaluated result.
+
+\keyideabox{\faLightbulbO}{Schema Code Co-Evolution}{References are understood
+by the substrate and so they are updated by structural edits (\S\ref{sec:impl-schema}). The
+evaluation mechanism can replace formulas with values, but also augment them to keep an evaluation
+trace for provenance analysis (\S\ref{sec:impl-provenance}).}
 
@@
 
-\subsection{Merging (refactoring)}
+\subsection{Local-First Collaborative Editing}
 \label{sec:impl-collab}
+
+like git rebase
 
@@
 
+\subsection{Incremental recomputation}
+\label{sec:impl-eval}
+
 \subsection{Schema code co-evolution}
 \label{sec:impl-schema}
-
-\subsection{Incremental recomputation}
-\label{sec:impl-eval}
 

### 2024-11-09 paper.tex
Implementation - merge and pbd sections

@@
 \usepackage{lettrine}
+
+\newcommand{\ident}[1]{{\sffamily #1}}
 
@@
 
+Medium is the message
+
+~
+
+~
+
 Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
@@
 JUSTIFICATION - this is a technical paper but belongs here!
+
+DESIGN PROCESS = formative examples + evaluation case study
 
@@
 \section{Walkthrough}
+\label{sec:walk}
+
 To providea an overview of the programming experiences supported by the Denicek
@@
 The user starts with an empty document, which is represented as a record. They add a field for
-each heading and a list \Verb_<ul>_ as a field named \Verb_speakers_. They use the command
+each heading and a list {\small\Verb_<ul>_} as a field named {\small\Verb_speakers_}. They use the command
 toolbox to add the first speaker.
@@
 further speakers, the user creates a textbox and a button. They enter new speaker
-details, construct a temporary \Verb_<li>_ element outside of the main list, copy the value
+details, construct a temporary {\small\Verb_<li>_} element outside of the main list, copy the value
 from the textbox into the new element using source view, and then copy the new element into the
-\Verb_speakers_ list.
+{\small\Verb_speakers_} list.
 
@@
 the user opens history view, selects edits that added the speaker and saves those as the
-\Verb_add-speaker_ interaction. They attach this as a handler for the \Verb_click_ event of the button.
+{\small\Verb_add-speaker_} interaction. They attach this as a handler for the {\small\Verb_click_} event of the button.
 
@@
 \subsubsection*{\circled{F} Adding Budget Calculation} A third user adds budget calculation to
-the initial document. Formulas are represented as special \Verb_x-formula_ nodes whose arguments
+the initial document. Formulas are represented as special {\small\Verb_x-formula_} nodes whose arguments
 are other nodes or references to other nodes or formulas (highlighted on mouse hover).
@@
 The document model assumes that lists are homogeneous and records heterogeneous, and so the
-\textit{Any} selector makes it possible to refer to all children of a list, but there is no
+\ident{Any} selector makes it possible to refer to all children of a list, but there is no
 way to refer to all children of a record.
@@
 require \emph{target} to which they are applied. Target is a reference and can contain the
-\textit{Any} selector, in which case the edit is applied to multiple nodes simultaneously.
+\ident{Any} selector, in which case the edit is applied to multiple nodes simultaneously.
 Most edits can only be applied to target node(s) of a particular kind.
@@
 There are two situations in which automatic update of references is undesirable. If an edit is
-applied to a singular element of a list (reference contains the \textit{Index} selector),
+applied to a singular element of a list (reference contains the \ident{Index} selector),
 references that refer into any element of the list should be unchanged.  Such edits may turn
@@
 define merging solely in terms of edits, without reference to current document state.
-(The effect of \textit{Copy} depends only on the existing document structure, but not on its value;
-while \textit{AppendFrom} affects only single list element and is thus labelled as a
+(The effect of \ident{Copy} depends only on the existing document structure, but not on its value;
+while \ident{AppendFrom} affects only single list element and is thus labelled as a
 non-structural edit.)
@@
 \subsection{Primitive Operations}
+\label{sec:system-ops}
 Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
@@
 provided that (i) the edit is not explicitly marked as non-structural, and (ii) the target
-reference does not contain \textit{Index} selector, i.e., it targets all elements of any involved
+reference does not contain \ident{Index} selector, i.e., it targets all elements of any involved
 lists. Also note that matching references in the document may be more specific than the edit
@@
 they target the same node, or if one targets a node that is nested inside the node targetted by
-the other. For edits with dependencies (\textit{AppendFrom} and \textit{Copy}), a conflict also
+the other. For edits with dependencies (\ident{AppendFrom} and \ident{Copy}), a conflict also
 occurs if the other edit modifies the dependency.
@@
 Applying an edit to a newly added node involves a number of cases. When the new node is added
-using \textit{Add} or \textit{Copy}, but the target location is modified, this is an unresolvable
+using \ident{Add} or \ident{Copy}, but the target location is modified, this is an unresolvable
 conflict (because those two operations would overwrite existing nodes). When the new node is
-added as a new list item using \textit{Append}, it can be transformed (by applying the other
+added as a new list item using \ident{Append}, it can be transformed (by applying the other
 edit direclty to the new node).
 
-Finally, when the new node is added using \textit{AppendFrom},
+Finally, when the new node is added using \ident{AppendFrom},
 it is copied from another document location. We cannot transform the node in the soource location
@@
 copying the source node to a new temporary location, transforming it there and then using
-\textit{AppendFrom} from the temporary location.
+\ident{AppendFrom} from the temporary location.
 
@@
 
-\section{Implementation}
+\section{Programming Experiences Implementation}
 \label{sec:impl}
@@
 
-like git rebase
+The Denicek representation enables \emph{local-first} collaboration~\cite{kleppmann-2019-local},
+as illustrated in \S\ref{sec:walk}E. If a document is edited by multiple users, they can
+each make edits to their local copy and eventually merge document variants using the operation to
+merge edit histories.
 
-\subsection{PBD (create textbox but not button)}
+Merging of histories behaves akin to git rebase in that it keeps a linear history. Synchronization
+in a distributed system thus requires first reapplying local edits on top of the remote history,
+before updating the remote history. Denicek thus implements the \emph{convergence model} of
+document variants \cite{edwards-2025-schema}, i.e., the user cannot, for example, maintain their
+own local document structure and import new data from another variant (this would require an
+inverse of the edit reconcilliation operation).
+
+Merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
+can be obtained either by appending $C'$ (produced by the edit reconcilliation operation)
+on top of $A,B$ or by appending $B'$ on top of $A,C$. Although the two edits are conflicting,
+$C$ only affects data and $B$ only affects structure and so the resulting document is the same
+in both cases. However, the histories differ. In the first case, the new node added by the
+\ident{Append} edit is transformed (from primitive string to a record representing table row).
+In the second case, the structural transformations are automatically applied to all rows.
+As discussed in \S\ref{sec:system-ops}, conflicts during merging can be resolved either by
+removing conflicting edits or by letting the later edits overwrite the former ones.
+
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
+\vspace{-0.5em}
+\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
+\label{fig:merging}
+\vspace{-0.5em}
+\end{figure}
+
+\subsection{Programming By Demonstration}
 \label{sec:impl-pbd}
+
+In \emph{programming by demonstration} \cite{cypher-1993-pbd}, the user demonstrates a task to the
+system and the system then repeats it, directly or in a generalized way. To use direct
+repetition with Denicek (Fig.~\ref{fig:pbd}), the user can select edits from the edit history,
+name them and replay them. In case of general-purpose document editing (in our formative prototype),
+this requires certain forethought (e.g., constructing a new list item in a temporary document
+field before appending it), but as illustrated in \S\ref{sec:case}, the mechanism is effective in
+a restricted domain such as data wrangling \cite{kandel-2011-wrangler}.
+
+There are two notable aspects of our implementation. First, Deni\-cek saves edits in the document
+itself (by representing individual edits as nodes and storing them in list inside a
+{\small\Verb_<saved-interactions>_} field). This means that no other implementation mechanism
+outside of the system is needed and also that the stored edits can be modified by the user
+(or tools working with the document).
+
+Second, to replay edits, Denicek does not simply append them on top of the current history.
+When saving edits, it records the hash of the history at the time of saving. When replaying edits,
+it appends the saved edits to the top of the original history (at the time of saving) and merges this new
+sequence with the current history. This pushes the saved edits through all subsequent edits made
+by the user. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where a newly added speaker
+is transformed from a primitive string to a table row.
+
+The implementation also has to account for the case where edits saved in the document are themselves
+transformed (when they are reconcilliated with other edits during merging). In this case, Denicek
+updates the saved edits (as discussed in \S\ref{sec:discuss}, this would not be needed if the
+history was stored as a graph, akin to ordinary git merging rather than rebasing).
 
@@
 
+x
+
+\newpage
+
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 7cm 9cm 0cm]{fig/pbd.pdf}
+\vspace{-0.5em}
+\caption{Programming by demonstration is implemented by selecting edits from the document
+history (1), saving them in the document (2) and replaying them (3).}
+\label{fig:pbd}
+\vspace{-0.5em}
+\end{figure}
+
 \subsection{Incremental recomputation}
 \label{sec:impl-eval}
+
+x
 
@@
 \section{Design Discussion}
+\label{sec:discuss}
+
+save graph of edits rather than list
+(would make replaying of saved edits easier because the histhash never changes)
 
@@
 \section{Case study}
+\label{sec:case}
 (data science environment)
@@
 maybe
-
-DESIGN PROCESS = formative examples + evaluation case study
 

### 2024-11-10 paper.tex
Interaction, formula evaluation, part 1

@@
 edits that augment the document with the result. Those edits are kept
-at the top of the document history and are removed in case of a conflict (\S\ref{sec:impl-eval}),
+at the top of the document history and are removed in case of a conflict (\S\ref{sec:impl-incremental}),
 providing incremental recomputation.}
@@
 schema and code co-evolution (\S\ref{sec:impl-schema}), incremental recomputation
-(\S\ref{sec:impl-eval}), provenance tracking (\S\ref{sec:impl-provenance})
+(\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}), provenance tracking (\S\ref{sec:impl-provenance})
 and concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
@@
 
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 7cm 9cm 0cm]{fig/pbd.pdf}
+\vspace{-0.5em}
+\caption{Programming by demonstration is implemented by selecting edits from the document
+history (1), saving them in the document (2) and replaying them (3).}
+\label{fig:pbd}
+\vspace{-0.5em}
+\end{figure}
+
 \subsection{Programming By Demonstration}
@@
 
-\subsection{Interaction (add button + remove done)}
+\begin{figure}[t]
+\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
+\vspace{-1em}
+\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
+the ``Remove completed'' button (2) modifies target and specifies a condition. }
+\label{fig:interactive}
+\vspace{-0.5em}
+\end{figure}
+
+\subsection{Interactive User Interfaces}
 \label{sec:impl-interaction}
 
-x
+Programming by demonstration can be used to define interactive elements in the document. In the
+simple case (Fig.~\ref{fig:interactive} (1)), the {\small\Verb_click_} event handler is set to a
+reference to a sequence of edits saved in the document. Clicking the button executes the edits
+using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., by appending them to a history at the
+time of saving and merging them with the current history.
+
+The Denicek substrate can be used to generalize the interactions saved through programming by
+demonstration. Our prototype illustrates this option in a limited way. As shown in
+Fig.~\ref{fig:interactive} (2), a button to remove all completed TODO items can be generalized
+from the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
+In addition to the saved interaction, we manually specify (in the source view) that the edits
+should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
+the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
+applied to elements for which the formula (which tests if the checkbox is checked) specified by a
+relative selector {\small\Verb_./condition/comp/result_} evaluates to true.
+
+Specifying such generalization manually is cumbersome, but a programming by demonstration system
+that implements typical heuristic for generalization from examples \cite{myers-2000-intelligence}
+would be able to infer and suggest such generalization to the user (possibly also automatically
+adding the formula based on positive and negative examples of selected items).
+
+\begin{figure}[t]
+\includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
+\caption{Increment wraps the existing count in a formula that adds 1 to the
+previous value (1), (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).}
+\label{fig:counter}
+\vspace{-0.5em}
+\end{figure}
+
+\subsection{Formula Language and Evaluation}
+\label{sec:impl-eval}
+
+Denicek documents can contain formulas inspired by the spreadsheet
+paradigm \cite{nardi-1990-spreadsheets}. Formulas can specify richer computations than
+what can be expressed using document edits. Formulas do not transform the document and their
+results are transient, although their evaluation also leverages the substrate operations,
+namely merging of edit histories.
+
+As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
+special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
+rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
+edits and the Denicek substrate treates them as stnadard nodes.
+
+To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
+record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
+formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
+(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
+as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
+with the document is discussed in \S\ref{sec:impl-incremental}
+
+The counter example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
+and programming by demonstration. To implement a counter, the Increment and Decrement buttons
+wrap the current counter value in a formula that adds or subtracts 1.
+
+\subsection{Incremental Recomputation}
+\label{sec:impl-incremental}
+
+xx
 
@@
 \begin{figure}[t]
-\includegraphics[width=0.9\columnwidth,clip,trim=0cm 7cm 9cm 0cm]{fig/pbd.pdf}
-\vspace{-0.5em}
-\caption{Programming by demonstration is implemented by selecting edits from the document
-history (1), saving them in the document (2) and replaying them (3).}
-\label{fig:pbd}
+  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
+  \vspace{-1em}
+  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
+  is added (3), only the results of affected formulas are invalidated.}
+  \label{fig:incremental}
   \vspace{-0.5em}
@@
 
-\subsection{Incremental recomputation}
-\label{sec:impl-eval}
-
-x
+\newpage
 
@@
 \label{sec:impl-copy}
+
+% TODO: Start Implementation section with some overall description of the "formative prototype"
+% TODO: In operations - better conflict resolution in terms of effect analysis
+% (collect selectors with structure/data effects; merging structure x data effects is OK,
+% but data x data or structure x structure is not); - Implement this
+% TODO: Add a section after Local-First Editing on the new conflict resolution (with demos)
+% TODO: In implementation use effects to better replay saved edits
+% (push through new structural but not data edits)
 

### 2024-11-11 paper.tex
Evaluation, coevolution, provenance figure

@@
 \begin{figure}[t]
+\vspace{-0.5em}
 \includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
-\vspace{-1em}
+\vspace{-1.25em}
 \caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
@@
 \label{fig:interactive}
-\vspace{-0.5em}
+\vspace{-2em}
 \end{figure}
@@
 
-Specifying such generalization manually is cumbersome, but a programming by demonstration system
-that implements typical heuristic for generalization from examples \cite{myers-2000-intelligence}
-would be able to infer and suggest such generalization to the user (possibly also automatically
-adding the formula based on positive and negative examples of selected items).
+\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
+Programming by demonstration systems typically implement heuristic for generalization
+\cite{myers-2000-intelligence} infers and suggests such generalizations. When integrated into
+a system based on Denicek, such heuristic may also be able to automatically add a formula
+based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}
 
@@
 \label{fig:counter}
-\vspace{-0.5em}
+\vspace{-1em}
 \end{figure}
@@
 
+\begin{figure}[t]
+\centering
+\vspace{-0.5em}
+\begin{minipage}{0.6\columnwidth}
+  ~\\[1em]
+  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
+  \end{minipage}%
+  \begin{minipage}{0.4\columnwidth}
+    \caption{Evaluated edits (A, B) are kept at the top of the history. When they conflict with
+    ordinary edits (4), evaluated edits are removed (A).}
+    \label{fig:eval}
+  \end{minipage}
+  \vspace{-1em}
+\end{figure}
+
 \subsection{Incremental Recomputation}
@@
 
-xx
+As illustrated in Fig.~\ref{fig:incremental}, Denicek supports incremental re-computa\-tion.
+In the example, the cost of speaker travel depends on the number of speakers, but the cost
+of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
+former.
 
-\newpage
+The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
+Denicek generates \emph{evaluated edits} that are appended to the top of the history.
+When subsequent edits are made, they are appended after all non-evaluated edits and the
+evaluated edits are pushed through the newly added edits. If the edits conflict (according to
+Denicek's conflict detection), affected evaluated edits and edits that depend on them are dropped.
+
+\keyideabox{\faRefresh}{Live Programming}{In the exploratory prototype, Denicek does not
+automatically evaluate formulas. This makes it easier to see and understand the evaluation
+mechanism, but a realistic system based on the substrate would likely automatically evaluate
+formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}.}
 
@@
   \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
-  \vspace{-1em}
+  \vspace{-0.5em}
   \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
@@
   \label{fig:incremental}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
+\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
+\label{fig:coevolution}
+\end{figure}
+
+\subsection{Schema Code Co-evolution}
+\label{sec:impl-schema}
+
+The problem of schema evolution is that, when the schema used in a database or a program evolves,
+data and code that depend on the schema need to evolve correspondingly. The problem is well-known
+in database systems \cite{rahm-2006-schema} and has recently been studied in the context of
+programming systems \cite{edwards-2025-schema}.
+
+Although Denicek does not explicitly track document structure (schema or type), all documents
+have an implicit structure and some of the edits transform the document structure. As
+discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
+document in this case. This enables a form of schema code co-evolution \cite{edwards-2025-schema}.
+The formulas embedded in Denicek documents use reference nodes to refer to both data sources
+(in the document) and the results of other computations. Consequently, if the document structure
+changes, the formulas are automatically updated.
+
+Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\Verb_<ul>_} is
+turned into {\small\Verb_<tbody>_} using \ident{UpdateTag} and wrapped inside {\small\Verb_<table>_}
+with a field {\small\Verb_body_} using \ident{WrapRecord}. For the latter, Denicek updates the
+reference accordingly turning the original {\small\Verb_/speaker_} reference in the formula
+into {\small\Verb_/speaker/body_}.
+
+As noted earier, evaluated edits transform the formula structure (wrapping it in the
+{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
+references to formulas are not transformed when the formula is evaluated (otherwise, references
+to evaluated formulas would not point to the evaluated result, but to the original unevaluated
+definition).
+
+\begin{figure}[t]
+\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
+\vspace{-1em}
+\caption{Using provenance tracking to highlight document parts that contributed to the
+calculation of refreshments costs (1), travel costs (2) and all costs (3).}
+\label{fig:provenance}
 \vspace{-0.5em}
@@
 \newpage
-
-\subsection{Schema code co-evolution}
-\label{sec:impl-schema}
 

### 2024-12-08 paper.tex
Save notes, working on new intro

@@
 \begin{abstract}
-Research on interactive programming systems gave rise to a range of programming experiences,
-including programming by demonstration, local-first collaborative editing, %structure editing,
-schema and code co-evo\-lution, provenance tracking and incremental recomputation. Those
-experiences are compelling, but they are hard to implement on the basis of existing programming
-languages and systems.
-
-We contribute the Denicek computational substrate. Denicek represents a program as a series of edits
-that construct or transform a document consisting of data and formulas. Denicek provides two
+Research on interactive programming systems gave rise to a range of compelling programming
+experiences, including local-first collaborative editing, programming by demonstration,
+incremental recomputation, schema and code co-evolution, provenance tracking and managed copy and
+paste. Those experiences advance the state of the art of programming, but they are hard to
+implement on the basis of existing programming languages and systems.
+
+We contribute Denicek, a computational substrate that reduces the complexity of
+implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
+that construct and transform a document consisting of data and formulas. Denicek provides two
 primitive operations on series of edits, merging and conflict checking, that form the backbone
-of the implementation of the aforementioned programming experiences.
+of the implementation of a range of programming experiences.
 
@@
 architecture, we use Denicek as the basis of a simple innovative data exploration environment.
-The case study shows that the Denicek computational substrate provides a pathway to the design
-of richer and more accessible interactive programming systems.
+The case study shows that the Denicek computational substrate provides an appealign basis for the
+design of richer and more accessible interactive programming systems.
 \end{abstract}
 
-\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
-  Your, Paper}
+\keywords{Programming Systems, Computational Substrate, Programming by Demonstration,
+  Local-First Software}
 
@@
 
-Medium is the message
+A computational substrate defines the basic structures using which programs are constructed, how
+the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
+a substrate affects what programming experiences can be readily supported. For example,
+object-oriented programming has been historically linked to graphical user
+interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
+a ``language laboratory'' \cite{steele-1993-lisp}.
+
+In principle, any programming experience can be developed on top of any computational
+substrate. However, a program representation and operations for working with programs provided by
+a suitable substrate can eliminate much of the complexity of implementing an interesting
+programming experience. For example, the reflective capabilities of Smalltalk make it easy
+to provide rich debugging tools \cite{rauch-2019-babylonian} that would be difficult to provide
+for C/C++~\cite{kell-2024-debugging}.
+
+\subsubsection*{Programming Experiences}
+
+In this paper, we ask what would be the ideal programming substrate
+for making programming easier, collaborative and more transparent? We seek a programming
+substrate that makes it easy to develop programming experiences such as:
+
+\begin{itemize}
+\item \emph{Programming by demonstration} --
+  Allow non-programmers to construct simple programs by performing examples of the expected behaviour. \cite{leiva-2021-rapido,cypher-1993-pbd}.
+\item \emph{Local-first collaboration} --
+  Multiple users should be able to use and modify a single program, preferrably without requiring a central server. \cite{kleppmann-2019-local}
+\item \emph{Provenance tracking} --
+  The execution of the program should leave an understandable trace that lets the user understand why program resulted in a particular result.
+\item \emph{Schema evolution [extra-ish]} --
+  When the user evolves the structure of the program, data and code should co-evolve automatically to match the new structure.
+\item \emph{Notational freedom [extra-ish?]} --
+  Allow users to adapt the program using a notation that suits them and is appropriate for the programming task at hand. [Joel]
+\item \emph{Concrete programming [extra?]} --
+  It should be possible to reuse parts of program or program logic without constructing abstractions, for example by managed copy \& paste.\cite{edwards-2006-copypaste,edwards-2022-copypaste}
+\end{itemize}
+
+\subsubsection*{Methodology}
+
+citizen programmers?
+
+DESIGN PROCESS = formative examples + evaluation case study
 
@@
 
-Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.
+substrate as defined by \cite{jakubovic-2022-ladder}
+
+% pbd
+\cite{leiva-2021-rapido,cypher-1993-pbd}
+\cite{chen-2023-miwa}
+
+% lcoal first
+\cite{kleppmann-2019-local,klokmose-2024-mywebstrates}
+% provenance
+\cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
+\cite{ricciotti-2017-imperative,perera-2012-functional}
+\cite{perera-2022-linked}
+
+% visualizations of results or execution
+% https://www.dcs.warwick.ac.uk/pvw04/p01.pdf
+% https://dl.acm.org/doi/10.1145/3313831.3376494
+
+% previews - livelits
+
+% Schema evolution
+% our paper; see https://openproceedings.org/2023/conf/edbt/paper-160.pdf - good refs in 2.2
+
+% https://x.com/jonathoda/status/1185888711210389504
+
+% also cool: https://arxiv.org/pdf/2303.06777
+
+% Hazel family of things
+% structure editing - sandblocks
+
+~
+
+~
 
@@
 JUSTIFICATION - this is a technical paper but belongs here!
-
-DESIGN PROCESS = formative examples + evaluation case study
 
@@
 
+PRINCIPLES
+* merging of edits should not require looking at current document (non-computationality)
+* two ways of pushing edits should be equivalent, i.e.
+
+push e1 e2 e3 over b1 b2 b3 can be done via
+  push e1 over b1 = e1'; over b2 = e1''; over b3 = e1'''
+  then push e2, then push e3
+or
+  push e1, e2, e3 over b1, then over b2, then over b3
+(implies that no "extra" edits can be appended later)
+
+USE MRDTs for list order!
+
+DIFFICULTIRES
+* if you are not able to refer to newly added items, you cannot reasonably do PBD
+  => need non-numerical indices
+
+how to transform appends?
+* apply to added? (does not work for copy)
+* copy to doc and edit? (good because edits are isolated from other things, but messy code!)
+(if we edit in doc, it works and we also do not need indices for ListAppend)
+* keep nested edits for append op? also does not work, because of the tricky case
+
+Tricky case is:
+- add new list item
+- then set its fields
+  (this breaks things because if we push "add" through structural edits, they will not affect the new values added in the second step)
+
+TODO: Delete ListAppend
+
+// cf shadow dom v reactu
+// something like google forms!
+% https://gowthamk.github.io/docs/mrdt.pdf page 11
+
+TAG selectors - sound nice, but they break
+(you cannot tell if two selectors overlap e.g. when there is index >< tag)
+
 save graph of edits rather than list
@@
 
-\section{Introduction}
+\section{Introduction old}
 
@@
 running software.
+
+substrate as defined by \cite{jakubovic-2022-ladder}
 
@@
 but it will not work with the rest of the ordinary C/C++ ecosystem.)
+
+Medium is the message
 
@@
 
+append /speakers \#hamilton <li>
+addfld /speakers/\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
+
+>>> wraprec /speakers body <table>
+
+[update selectors - add /body]
+append /speakers/body \#hamilton <li>
+[dtto]
+addfld /speakers/body\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
+
+>>> retag /spekers/body <tbody>
+
+[na - out of scope]
+
+>>> retag /speakers/body/* <tr>
+
+[apply to added - update node (* => \#hamilton scoping)]
+append /speakers/body \#hamilton <li>
++ retag /speakers/body/\#hamilton <tr>
+[na - out of scope]
+addfld /speakers/body\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
+
+>>> wraprec /speakers/body/*/speaker value <td>
+
+append /speakers/body \#hamilton <li>
++ retag /speakers/body/\#hamilton <tr>
++ wraprec /speakers/body/\#hamilton/speaker value <td> [ no op - it is not there]
+addfld /speakers/body/\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
++ wraprec /speakers/body/\#hamilton/speaker value <td>
+
+>> updid /speakers/body/* speaker=>name
+
+append /speakers/body \#hamilton <li>
++ retag /speakers/body/\#hamilton <tr>
++ wraprec /speakers/body/\#hamilton/speaker value <td> [ no op - it is not there]
++ updid /speakers/body/\#hamilton speaker=>name [ no op - it is not there ]
+addfld /speakers/body \#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
++ wraprec /speakers/body/\#hamilton/speaker value <td>
++ updid /speakers/body/\#hamilton speaker=>name
+
+>>> addfld /speakers/body/* email ""
+
+append /speakers/body \#hamilton <li>
++ retag /speakers/body/\#hamilton <tr>
++ wraprec /speakers/body/\#hamilton/speaker value <td> [ no op - it is not there]
++ updid /speakers/body/\#hamilton speaker=>name [ no op - it is not there ]
++ addfld /speakers/body/\#hamilton email "" [ no op - it is not there ]
+addfld /speakers/body \#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
++ wraprec /speakers/body/\#hamilton/speaker value <td>
++ updid /speakers/body/\#hamilton speaker=>name
++ addfld /speakers/body/\#hamilton email ""
+
+...
+
+copy /speakers/body/*/name => /speakers/body/*/email
+edit /speakers/body/*/name (fn1)
+edit /speakers/body/*/email (fn2)
+
 \bibliographystyle{plain}

### 2024-12-08 paper.tex
Experiences, methodology

@@
 %
-\documentclass[sigconf]{acmart}
+\documentclass[sigconf,anonymous]{acmart}
 %\documentclass[sigconf,review,anonymous]{acmart}
@@
 
-\setcopyright{acmlicensed}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[Conference acronym 'XX]{Make sure to enter the correct
-  conference title from your rights confirmation emai}{June 03--05,
-  2018}{Woodstock, NY}
+\setcopyright{none}
+\copyrightyear{2025}
+\acmYear{2025}
+\acmDOI{XXXX.XXXX}
+
+\acmConference[Conference acronym 'XX]{}{June 03--05, 2018}{Woodstock, NY}
 \acmISBN{978-1-4503-XXXX-X/18/06}
@@
 \begin{document}
-\title[Denicek: Computational Substrate for Concrete, Collaborative, Interactive
-  Programming]{{\scshape Denicek}: Computational Substrate for Concrete, \\Collaborative, Interactive Programming}
+\title[Denicek: Computational Substrate for Document-Oriented End-User
+  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming}
 
@@
 \begin{abstract}
-Research on interactive programming systems gave rise to a range of compelling programming
-experiences, including local-first collaborative editing, programming by demonstration,
-incremental recomputation, schema and code co-evolution, provenance tracking and managed copy and
-paste. Those experiences advance the state of the art of programming, but they are hard to
-implement on the basis of existing programming languages and systems.
+Human-computer interaction research gave rise to a wide range of compelling programming experiences,
+including local-first collaborative editing, programming by demonstration, incremental
+recomputation, schema change control, end-user debugging and concrete programming.
+Those experiences advance the state of the art of end-user programming, but they are hard to
+implement on the basis of established programming languages and systems.
 
@@
 implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
-that construct and transform a document consisting of data and formulas. Denicek provides two
-primitive operations on series of edits, merging and conflict checking, that form the backbone
-of the implementation of a range of programming experiences.
+that construct and transform a document consisting of data and formulas. Denicek provides three
+primitive operations on series of edits, application, merging of histories and conflict checking.
+We show that the three operations form the backbone of an easy implementation of a range of
+programming experiences.
 
@@
 architecture, we use Denicek as the basis of a simple innovative data exploration environment.
-The case study shows that the Denicek computational substrate provides an appealign basis for the
-design of richer and more accessible interactive programming systems.
+The case study shows that the Denicek computational substrate provides an appealing basis for the
+design of rich, interactive end-user programming systems.
 \end{abstract}
 
-\keywords{Programming Systems, Computational Substrate, Programming by Demonstration,
-  Local-First Software}
+\keywords{Programming Systems, Computational Substrate, End-User Programming,
+  Programming by Demonstration, Local-First Software}
 
@@
 
-\newpage
-
 \section{Introduction}
@@
 the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
-a substrate affects what programming experiences can be readily supported. For example,
+a substrate affects what programming experiences can be readily supported by a system. For example,
 object-oriented programming has been historically linked to graphical user
 interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
-a ``language laboratory'' \cite{steele-1993-lisp}.
+a language laboratory \cite{steele-1993-lisp}.
 
 In principle, any programming experience can be developed on top of any computational
-substrate. However, a program representation and operations for working with programs provided by
-a suitable substrate can eliminate much of the complexity of implementing an interesting
-programming experience. For example, the reflective capabilities of Smalltalk make it easy
-to provide rich debugging tools \cite{rauch-2019-babylonian} that would be difficult to provide
+substrate. However, a suitable program representation and operations for working with programs
+provided by the substrate can eliminate much of the complexity of implementing interesting
+programming experiences. For example, the reflective capabilities of Smalltalk make it easy
+to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
 for C/C++~\cite{kell-2024-debugging}.
@@
 
-In this paper, we ask what would be the ideal programming substrate
-for making programming easier, collaborative and more transparent? We seek a programming
-substrate that makes it easy to develop programming experiences such as:
+What would be the ideal programming substrate for rich, interactive end-user programming systems?
+In this paper, we aim to design a programming substrate that makes it easy to develop a range of
+programming experiences \cite{myers-2006-eup}:
 
 \begin{itemize}
-\item \emph{Programming by demonstration} --
-  Allow non-programmers to construct simple programs by performing examples of the expected behaviour. \cite{leiva-2021-rapido,cypher-1993-pbd}.
-\item \emph{Local-first collaboration} --
-  Multiple users should be able to use and modify a single program, preferrably without requiring a central server. \cite{kleppmann-2019-local}
-\item \emph{Provenance tracking} --
-  The execution of the program should leave an understandable trace that lets the user understand why program resulted in a particular result.
-\item \emph{Schema evolution [extra-ish]} --
-  When the user evolves the structure of the program, data and code should co-evolve automatically to match the new structure.
-\item \emph{Notational freedom [extra-ish?]} --
-  Allow users to adapt the program using a notation that suits them and is appropriate for the programming task at hand. [Joel]
-\item \emph{Concrete programming [extra?]} --
-  It should be possible to reuse parts of program or program logic without constructing abstractions, for example by managed copy \& paste.\cite{edwards-2006-copypaste,edwards-2022-copypaste}
+\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
+  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
+\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
+  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
+\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
+  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
+\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
+  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
+\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
+  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
 \end{itemize}
@@
 \subsubsection*{Methodology}
-
-citizen programmers?
-
-DESIGN PROCESS = formative examples + evaluation case study
+We use a two-stage research methodology. In the first stage, we identify a number
+of \emph{formative examples} -- simple programming tasks, drawn from literature, that manifest one or
+more of the desired programming experiences. Using those examples, we co-design the Denicek substrate
+and Webnicek (Section~\ref{sec:walk}), a simple web-based end-user programming environment based on the substrate.
+Although Webnicek can be used to complete end-user programming tasks, it is optimized for
+testing the underlying substrate rather than for usability.
+
+In the second stage, we use the Denicek substrate as the basis of Datnicek, an end-user data
+exploration environment inspired by existing research systems\cite{kandel-2011-wrangler,drossos-2020-wrex}.
+We use the second stage to evaluate suitability of the Denicek substrate for the development of
+end-user programming systems and report the results of our evaluation (Section~\ref{sec:eval}).
+
+\subsubsection*{Substrate}
+how it works very roughly
+
+The coolest trick - almost everything (PBD, collaboration, evaluation) is done through merge!
+
+\subsubsection*{Contributions}
+one main thing - substrate - with other things
+
+\newpage
+
+% The problems include a counter app \cite{kiss-2014-7guis},
+% Todo list app and conference planning \cite{edwards-2025-schema}.
 
@@
 
+% debugging
+% https://dl.acm.org/doi/pdf/10.1145/1133265.1133293
+
 % visualizations of results or execution
@@
 
+citizen programmers?
+
 TODO: Maybe do not talk about ``programs'' because this is more documents with formulas - not that fancy programs.
@@
 
-\subsection{Programming Experiences}
-list
-
-\subsection{Substrate}
-how it works very roughly
-
-The coolest trick - almost everything (PBD, collaboration, evaluation) is done through merge!
-
-\subsection{Contributions}
-one main thing - substrate - with other things
+METHODOLOGY (add somewhere belo) - table of "formative examples" and experiences they require?
++ SOURCE (literature?)
 
@@
 all the references
+
+what is this about?
+end-user programming ? Cite % https://dl.acm.org/doi/pdf/10.1145/1125451.1125472
+but also Live, Interactive etc. - perhaps for a lack of better term use end-user
+(also notebooks)
 
@@
 
+\section{Evaluation}
+\label{sec:eval}
+how well did Denicek work for implementing Datnicek
+
 \section{Discussion}

### 2024-12-11 paper.tex
Contributions, tweaks, tree and refs

@@
 %
-\documentclass[sigconf,anonymous]{acmart}
+\documentclass[sigconf,anonymous,screen]{acmart}
 %\documentclass[sigconf,review,anonymous]{acmart}
@@
 \usepackage{lettrine}
-
+\usepackage{hyperref}
 \newcommand{\ident}[1]{{\sffamily #1}}
@@
 \begin{abstract}
-Human-computer interaction research gave rise to a wide range of compelling programming experiences,
+User-centric programming research gave rise to a wide range of compelling programming experiences,
 including local-first collaborative editing, programming by demonstration, incremental
@@
 Those experiences advance the state of the art of end-user programming, but they are hard to
-implement on the basis of established programming languages and systems.
+implement on the basis of established programming language and system paradigms.
 
@@
 that construct and transform a document consisting of data and formulas. Denicek provides three
-primitive operations on series of edits, application, merging of histories and conflict checking.
-We show that the three operations form the backbone of an easy implementation of a range of
+primitive operations on series of edits, application of edits, merging of histories and conflict checking.
+We show that the three operations form the backbone of a direct implementation of a range of
 programming experiences.
@@
 the implementation of the programming experiences listed above. To evaluate the proposed
-architecture, we use Denicek as the basis of a simple innovative data exploration environment.
+architecture, we use Denicek as the basis of a simple end-user data exploration environment.
 The case study shows that the Denicek computational substrate provides an appealing basis for the
@@
 interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
-a language laboratory \cite{steele-1993-lisp}.
+a ``language laboratory'' \cite{steele-1993-lisp}.
 
@@
 to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
-for C/C++~\cite{kell-2024-debugging}.
+for C/C++~\cite{kell-2018-unix,kell-2024-debugging}.
 
@@
 
-\subsubsection*{Methodology}
-We use a two-stage research methodology. In the first stage, we identify a number
-of \emph{formative examples} -- simple programming tasks, drawn from literature, that manifest one or
-more of the desired programming experiences. Using those examples, we co-design the Denicek substrate
-and Webnicek (Section~\ref{sec:walk}), a simple web-based end-user programming environment based on the substrate.
-Although Webnicek can be used to complete end-user programming tasks, it is optimized for
-testing the underlying substrate rather than for usability.
-
-In the second stage, we use the Denicek substrate as the basis of Datnicek, an end-user data
-exploration environment inspired by existing research systems\cite{kandel-2011-wrangler,drossos-2020-wrex}.
+\subsubsection*{Two-Stage Methodology}
+Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
+To \emph{design} Denicek, we identify a number of \emph{formative examples} -- simple programming
+tasks that manifest one or more of the desired programming experiences.
+Using those examples, we co-design the Denicek substrate and Webnicek (Section~\ref{sec:walk}),
+a simple web-based end-user programming environment based on the substrate. Although Webnicek can
+be used to complete end-user programming tasks, it is optimized for testing the underlying
+substrate rather than for usability.
+
+To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek, an end-user data
+exploration environment inspired by existing systems~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
 We use the second stage to evaluate suitability of the Denicek substrate for the development of
@@
 \subsubsection*{Substrate}
-how it works very roughly
-
-The coolest trick - almost everything (PBD, collaboration, evaluation) is done through merge!
+The Denicek substrate brings together two central design ideas. First, it represents programs as
+document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
+static content. Second, Denicek does not store the document tree itself, but instead, it maintains
+a sequence of edit operations through which the tree was constructed and transformed.
+
+The substrate then provides three primitive operations for working with sequences of edits.
+First, it can apply a series of edits to reconstruct a document. Second, it can merge two
+diverging edit histories. Finally, it can detect conflicts when merging histories and, for
+example, remove conflicting edits from one branch.
+
+The key insight that we present in this paper is that many of the programming experiences can
+be obtained by leveraging the uniform document representation alongside with a suitable combination
+of the three primitive operations.
+
+Changing data and formulas is done using the same primitive edit operations to
+manipulate the document structure. A user-interface may provide a specialized editor, but still
+trigger the pritive edits behind the scenes. However, interacting with elements in the document,
+such as a entering text in a textbox, also generates a document edit that can be merged or checked
+for conflicts (\S\ref{sec:impl-interaction}).
+
+As we will see, past edits that demonstrate an operation done to the
+document can be recorded, allowing programming by demonstration (\S\ref{sec:impl-pbd}).  Replaying
+such recorded edits is implemented using the merging operation, which means that recorded operations
+continue working even if the document structure later changes. Moreover, structural changes to the
+document can be done independently and then merged (\S\ref{sec:impl-collab}). Evaluation of formulas
+also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
+edits done later by the user, the evaluated edits are removed, implementing an incremental
+recomputation (\S\ref{sec:impl-incremental}).
 
 \subsubsection*{Contributions}
-one main thing - substrate - with other things
-
-\newpage
-
-% The problems include a counter app \cite{kiss-2014-7guis},
-% Todo list app and conference planning \cite{edwards-2025-schema}.
-
-~
-
-~
-
-substrate as defined by \cite{jakubovic-2022-ladder}
-
-% pbd
+The structure and contributions of this paper are:
+
+\begin{itemize}
+\item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
+  document representation, edit operations and the key three operations.
+\item We explain how a wide range of end-user programming experiences can be readily implemented using
+  the substrate (\S\ref{sec:impl}) in the context of a simple web-based prototype (Webnicek).
+\item We document important design decisions and alternatives (\S\ref{sec:discuss}). The analysis
+  shows that the desired functionality requires a careful choice among interconnected design options.
+\item We develop an end-user data exploration system (Datnicek) on top of Denicek (\S\ref{sec:case})
+  and evaluate the extent to which the Denicek substrate enables such development (\S\ref{sec:eval}).
+\end{itemize}
+
+\noindent
+To enable others build on top of Denicek, we share our compact and documented
+source code at: \url{https://github.com/d3sprog/denicek}
+
+\section{Background}
+What is this about? end-user programming \cite{myers-2006-eup}
+but also programming systems \cite{jakubovic-2023-techdims},
+Live \cite{rein-2019-live}, Interactive etc. - perhaps for a lack of better term use end-user
+(also notebooks)
+
+explain computational substrate
+Substrate defined by \cite{jakubovic-2022-ladder}
+
+citizen programmers? Maybe do not talk about ``programs'' because this is more documents with
+formulas - not that fancy programs. Computational documents?
+
+programming by demonstration
 \cite{leiva-2021-rapido,cypher-1993-pbd}
@@
 
-% lcoal first
+local first
 \cite{kleppmann-2019-local,klokmose-2024-mywebstrates}
-% provenance
+
+provenance/debugging
 \cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
@@
 
-% debugging
-% https://dl.acm.org/doi/pdf/10.1145/1133265.1133293
-
-% visualizations of results or execution
-% https://www.dcs.warwick.ac.uk/pvw04/p01.pdf
-% https://dl.acm.org/doi/10.1145/3313831.3376494
-
-% previews - livelits
-
-% Schema evolution
+end-user debugging
+\cite{kissinger-2006-debugging}
+
+visualizing results of execution
+projection boxes \cite{lerner-2020-boxes}
+
+livelits (provide editors)
+\cite{omar-2021-livelits}
+
+edit calculus (based on tree navigation though)
+\cite{omar-2017-hazelnut}
+
+structure editing see \cite{beckman-2023-sandblocks}
+
+Schema evolution
 % our paper; see https://openproceedings.org/2023/conf/edbt/paper-160.pdf - good refs in 2.2
 
+Also on substrate see
 % https://x.com/jonathoda/status/1185888711210389504
@@
 
-% Hazel family of things
-% structure editing - sandblocks
-
 ~
@@
 
-citizen programmers?
-
-TODO: Maybe do not talk about ``programs'' because this is more documents with formulas - not that fancy programs.
-Computational documents?
-
-explain computational substrate
-
-JUSTIFICATION - this is a technical paper but belongs here!
-
-METHODOLOGY (add somewhere belo) - table of "formative examples" and experiences they require?
-+ SOURCE (literature?)
-
-\section{Background}
-all the references
-
-what is this about?
-end-user programming ? Cite % https://dl.acm.org/doi/pdf/10.1145/1125451.1125472
-but also Live, Interactive etc. - perhaps for a lack of better term use end-user
-(also notebooks)
-
+\newpage
+~
 % ==================================================================================================
@@
 &&\\[-1em]
+\seltablecol{\Verb|..|}{Parent}{Refers to a parent of a node}
 \seltablecol{\Verb|field|}{Field}{Refers to record field of a given name}
-\seltablecol{\Verb|number|}{Index}{Refers to list element at a given index}
+\seltablecol{\Verb|\#index|}{Index}{Refers to list element at a given index}
 \seltablecol{\Verb|*|}{Any}{Refers to all children of a list node}
@@
 &\\[-1em]
-\ndtablecol{\faListUl}{List}{tag, child$_1$, $\ldots$, child$_n$}
-  {Ordered list of nodes, addressable by index. Renders as \Verb|<tag>| with children}
+\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
+  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
 \ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
-  {Record with children addressable by \textit{field} name. Renders as \Verb|<tag>|.}
+  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
 \ndtablecol{\faExternalLink}{Reference}{selectors}
-  {Reference to another document location. Displays the \Verb|/selectors| as a link.}
+  {Reference to another document location. Displays the \Verb|selectors| as a link.}
 \ndtablecol{\faFont}{Primitivie}{string \textnormal{or} number}
@@
 document. In this section, we describe the structure of documents and edits, as well as the
-operations that form the backbone of the system and are used to implement a wide range of
-programming experiences, discussed in \S\ref{sec:impl}.
+operations that form the backbone of the system and are used to implement a range of
+end-user programming experiences (\S\ref{sec:impl}).
 
@@
 A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
-Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
-document visible to the user, although parts can be collapsible or hidden using CSS.
-Records and lists are rendered as HTML elements of a specified tag with children becoming child
-elements. Field names are hidden in the rendered document.
-
-References to document location are used in both document itself (reference nodes) and in edits
-(target of the edit). They are represented as a sequence of selectors (Fig.~\ref{fig:doc}).
+References to document locations can be relative or absolute. Both kinds can appear in a
+document (reference nodes) and absolute references are used in edits (to denote a target node).
+A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
 The document model assumes that lists are homogeneous and records heterogeneous, and so the
@@
 way to refer to all children of a record.
+%
+Note that Denicek does not use implicit numerical indices for lists and indices of new items have
+to be supplied explicitly. The design choice simplifies merging and is discussed later
+(\S\ref{sec:discuss}).
 
@@
 
+Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
+document visible to the user, although parts can be collapsible or hidden using CSS.
+
+[webnicek?]
+
 The key claim of this paper is that the computational substrate described in the previous
@@
 
+LISTS vs RECORDS
+- are they really different? also MRDT idea - needed for both
+
 PRINCIPLES
@@
 \section{Discussion}
+
+table of "formative examples" and experiences they require?
++ SOURCE (literature?) => Appendix
+
+The problems include a counter app \cite{kiss-2014-7guis},
+Todo list app and conference planning \cite{edwards-2025-schema}.
+
 \subsection{Heuristic evaluation}

### 2024-12-13 paper.tex
Tweaks, updates in walkthrough and substrate

@@
 \begin{itemize}
+\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
+  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
 \item \emph{Programming by Demonstration.} Allow users to construct simple programs by
   demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
-\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
-  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
+\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
+  depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
+\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
+  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
 \item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
   why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
-\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
-  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
 \item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
@@
 To \emph{design} Denicek, we identify a number of \emph{formative examples} -- simple programming
-tasks that manifest one or more of the desired programming experiences.
-Using those examples, we co-design the Denicek substrate and Webnicek (Section~\ref{sec:walk}),
+tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
+Using those examples, we co-design the Denicek substrate and Webnicek (\S\ref{sec:walk}),
 a simple web-based end-user programming environment based on the substrate. Although Webnicek can
@@
 
-To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek, an end-user data
+To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek (\S\ref{sec:case}), an end-user data
 exploration environment inspired by existing systems~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
 We use the second stage to evaluate suitability of the Denicek substrate for the development of
-end-user programming systems and report the results of our evaluation (Section~\ref{sec:eval}).
+end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).
 
@@
 
+live/incremental recomputation
+\cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
+
 visualizing results of execution
@@
 
-\newpage
 ~
@@
 
-To providea an overview of the programming experiences supported by the Denicek
-substrate, we use a web-based implementation of Denicek with a structure
-editor that supports navigation in document and issuing of edit commands. The editor
-implements support for a range of programming experiences, discussed in \S\ref{sec:impl}.
+In this section, we use Webnicek, our simple web-based prototype, to illustrate the programming
+experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
+editor that supports navigation in the document and issuing of edit commands. We show two
+formative examples (\S\ref{app:examples}) related to conference organization.
 
@@
 The user starts with an empty document, which is represented as a record. They add a field for
-each heading and a list {\small\Verb_<ul>_} as a field named {\small\Verb_speakers_}. They use the command
+each heading and a list {\small\Verb_<ul>_}, represented by a field named {\small\Verb_speakers_}. They use the command
 toolbox to add the first speaker.
@@
 further speakers, the user creates a textbox and a button. They enter new speaker
-details, construct a temporary {\small\Verb_<li>_} element outside of the main list, copy the value
-from the textbox into the new element using source view, and then copy the new element into the
-{\small\Verb_speakers_} list.
+details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
+from the textbox into the new element using source view.
 
@@
 The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
-button are also automatically turned into the new format.
-
-\keyideabox{\faLightbulbO}{Collaborative Editing and Interactivity}{Denicek's merging
+button are also automatically added in the new format.
+
+\keyideabox{\faLightbulbO}{Local-First Collaboration}{Denicek's merging
 reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
-Merging is asymmetic, but the order does not matter here. The same merging operation is used
-when handling user interaction (\S\ref{sec:impl-interaction}).}
+Merging is asymmetic, but the order does not matter in the above scenario. The same merging
+operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}
 
@@
 
-\keyideabox{\faLightbulbO}{Schema Code Co-Evolution}{References are understood
-by the substrate and so they are updated by structural edits (\S\ref{sec:impl-schema}). The
-evaluation mechanism can replace formulas with values, but also augment them to keep an evaluation
-trace for provenance analysis (\S\ref{sec:impl-provenance}).}
+\keyideabox{\faLightbulbO}{Schema Change Control}{The substrate understands references in
+the document and updates them when applying structural edits (\S\ref{sec:impl-schema}).
+Evaluation can replace formulas with values, but also augment them to enable
+end-user debugging via provenance analysis (\S\ref{sec:impl-provenance}).}
 
@@
 document. In this section, we describe the structure of documents and edits, as well as the
-operations that form the backbone of the system and are used to implement a range of
-end-user programming experiences (\S\ref{sec:impl}).
+operations that form the backbone of the system and are used to implement the
+end-user programming experiences discussed in \S\ref{sec:impl}.
 
@@
 References to document locations can be relative or absolute. Both kinds can appear in a
-document (reference nodes) and absolute references are used in edits (to denote a target node).
+document (reference nodes) and absolute references are also used in edits (to denote a target~node).
+
 A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
@@
 \ident{Any} selector makes it possible to refer to all children of a list, but there is no
-way to refer to all children of a record.
-%
-Note that Denicek does not use implicit numerical indices for lists and indices of new items have
-to be supplied explicitly. The design choice simplifies merging and is discussed later
-(\S\ref{sec:discuss}).
+way to refer to all children of a record. Note that Denicek does not use implicit numerical
+indices for lists. The index of a new item has to be supplied explicitly. The design choice
+simplifies merging and is discussed in \S\ref{sec:discuss}.
 
@@
 The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
-require \emph{target} to which they are applied. Target is a reference and can contain the
-\ident{Any} selector, in which case the edit is applied to multiple nodes simultaneously.
-Most edits can only be applied to target node(s) of a particular kind.
+require \emph{target} to which they are applied. Target is an absolute reference not containing the
+\ident{Parent} selector. It can contain the \ident{Any} selector, in which case the edit is applied
+to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a particular
+kind.
 
@@
 between edits that keep existing references in a document unchanged (above) and edits that affect
-references (below). When a selector is \emph{invalidated}, e.g., when deleting a field or a
-list item to which there is a reference, Denicek rejects the edit. Copying also invalidates
-selectors because it is ambiguous whether selectors referring to the original location should
-refer to the source or the target of the copying after the edit (and references cannot refer to
-multiple structurally incompatible locations). Updating a field, wrapping, reordering or deleting
-a list item requires updating references in the document correspondingly.
+references (below).
+
+Renaming a field or wrapping a node require corresponding change to references in the document
+(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
+rejects the edit. A copy edit can also be rejected, because it is ambiguous whether references
+referring to the original location should continue referring to the source or should be changed to
+point to the target of the copying after the edit (a reference cannot refer to two structurally
+independent locations).
 
@@
 \begin{figure}
-\newcommand{\ektablecol}[6]{
-\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 & \sffamily\footnotesize #5 \\[-0.2em]
-&\multicolumn{3}{l|}{\sffamily\footnotesize #6}\\[0.3em]
+\newcommand{\ektablecol}[5]{
+\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 \\[-0.2em]
+&\multicolumn{2}{l|}{\sffamily\footnotesize #5}\\[0.3em]
 }
-\begin{tabular}{|cp{13em}rr|}
+\begin{tabular}{|cp{17em}r|}
 \hline
 \rowcolor{ekgray}
-&&&\\[-1em]
+&&\\[-1em]
 \rowcolor{ekgray}
- & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target & \sffamily\footnotesize\bfseries Selectors \\[0.2em]
+ & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target \\[0.2em]
 \hline
-&&&\\[-1em]
-\ektablecol{\faPlus}{Add}{field, node}{Record}{Unchanged}
-  {Add \textit{node} as a \textit{field} to the specified record.}
-\ektablecol{\faAt}{Append}{node}{List}{Unchanged}
-  {Append \textit{node} to the end of the specified list.}
-\ektablecol{\faPaperclip}{AppendFrom}{selectors}{List}{Unchanged}
-  {Append node from \emph{selectors} to the end of the specified list.}
-\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}{Unchanged}
+&&\\[-1em]
+\ektablecol{\faPlus}{Add}{field, after, node}{Record}
+  {Add \textit{node} as a \textit{field} to the specified record \textit{after} a given field.}
+\ektablecol{\faAt}{Append}{index, after, node}{List}
+  {Append \textit{node} to the end of the specified list \textit{after} a given field.}
+\ektablecol{\faSort}{Reorder}{permutation}{List}
+  {Reorder items of a specified list according to a \textit{permutation}.}
+\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}
+  {Delete the item at a given \textit{index} of a specified list.}
+\ektablecol{\faCode}{UpdateTag}{tag}{Tagged}
+  {Change the tag of a specified list or record from to a new \textit{tag}.}
+\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}
   {Apply primitive \textit{transform} to the specified primitive.}
-\ektablecol{\faCode}{UpdateTag}{old tag, new tag}{Tagged}{Unchanged}
-  {Change the tag of a specified list or record from \textit{old} to \textit{new}.}
 \hline
-&&&\\[-1em]
-\ektablecol{\faFont}{UpdateField}{old field, new field}{Record}{Update}
+&&\\[-1em]
+\ektablecol{\faFont}{RenameField}{old field, new field}{Record}
   {Rename the field of a specified record from \textit{old} to \textit{new}.}
-\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}{Invalidate}
+\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}
   {Delete the field \textit{field} of a specified record.}
-\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Update}
-  {Delete the item at a given \textit{index} of a specified list.}
-\ektablecol{\faSort}{Reorder}{permutation}{List}{Update}
-  {Reorder items of a specified list according to a \textit{permutation}.}
-\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}{Update}
+\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}
   {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
-\ektablecol{\faListUl}{WrapList}{tag}{Any}{Update}
+\ektablecol{\faListUl}{WrapList}{tag, index}{Any}
   {Wrap the specified node as a sole element of a new list with \textit{tag}.}
-\ektablecol{\faCopy}{Copy}{selectors}{Any}{Invalidate}
+\ektablecol{\faCopy}{Copy}{selectors}{Any}
   {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
@@
 applied to a singular element of a list (reference contains the \ident{Index} selector),
-references that refer into any element of the list should be unchanged.  Such edits may turn
-document into an inconsistent state, but they typically do so temporarily during document
-construction (checking such edits is discussed in \S\ref{sec:discuss-future}).
+references that refer into the list should be unchanged. Although such edits violate the invariant
+that collections are homogeneous, they typically do so only temporarily, for example during
+construction of a new list item (checking such edits is discussed in \S\ref{sec:discuss-future}).
 
@@
 a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
-involves wrapping, but this should not affect references to the formula.
-To support those cases, it is possible to annotate edits that normally affect selectors
-(Fig.~\ref{fig:edits}, below) as non-structural. This annotation is required when the
-target reference contains the \emph{Index} selector.
+involves wrapping the formula node, but this should not affect references to the formula.
+To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits}, below)
+have a \emph{reference behavior} field that can be set to disable automatic reference
+updating. This annotation is required when the target contains the \emph{Index} selector.
 
@@
 define merging solely in terms of edits, without reference to current document state.
-(The effect of \ident{Copy} depends only on the existing document structure, but not on its value;
-while \ident{AppendFrom} affects only single list element and is thus labelled as a
-non-structural edit.)
+(The effect \ident{Copy} has on the document structure depends on the existing document
+structure, but not on its value).
 
@@
 (e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
-has to be provided as an additional  mechanism on top of the underlying Denicek substrate.
+can be provided as an additional mechanism on top of the underlying Denicek substrate.
 
@@
 Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
-sequences of edits can be checked for conflicts or merged. Denicek identifies edit histories by
-a (git-like) hash, computed from the hash of the parent and the current edit, which is used to identify
-common shared part of the history during conflict checking and merging.
+sequences of edits can be merged and also checked for conflicts. Denicek identifies edit histories by
+a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
+identify common shared part of the history during conflict checking and merging.
 
@@
 updates matching references in the document according to the rules shown in Fig.~\ref{fig:updates}
-provided that (i) the edit is not explicitly marked as non-structural, and (ii) the target
-reference does not contain \ident{Index} selector, i.e., it targets all elements of any involved
-lists. Also note that matching references in the document may be more specific than the edit
+provided that the reference behavior is not set to disable the updating (which is required when
+the target reference contains the \ident{Index} selector). Although \ident{WrapList} has to specify
+the index to be used for the newly created list item, the references in the document are updated to
+use the \ident{All} selector. We assume wrapping introduces a homogeneous list and references
+pointing to individual items should thus be generalized to point to all list items.
+
+Also note that affected references in the document may be more specific than the edit
 target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
 a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. If the document
-contains a reference that would be invalidated by an edit, the edit is rejected.
+contains a reference that would be invalidated by the \ident{Copy} or \ident{DeleteField} edit,
+the edit is rejected.
+
+~
+
+~
+
+todo
+
+~
+
+~
 
@@
 \\[-1em]
-\tttablecol{UpdateField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
+\tttablecol{RenameField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
   {Replace Field for matching references.}
-\tttablecol{DeleteItem}{target, index}{\Verb|/target/n/nested|}{\Verb|/target/(n-1)/nested|}
-  {Decrement Index greater than $n$ in matching refs.}
-\tttablecol{Reorder}{target, permutation}{\Verb|/target/n/nested|}{\Verb|/target/permutation(n)/nested|}
-  {Update Index using permutation in matching refs.}
 \tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
   {Insert extra Field selector after matching prefix.}
-\tttablecol{WrapList}{target, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
+\tttablecol{WrapList}{target, index, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
   {Insert extra All selector after matching prefix.}
@@
 
-table of "formative examples" and experiences they require?
-+ SOURCE (literature?) => Appendix
-
-The problems include a counter app \cite{kiss-2014-7guis},
-Todo list app and conference planning \cite{edwards-2025-schema}.
-
 \subsection{Heuristic evaluation}
@@
 
-maybe
+\bibliographystyle{plain}
+\bibliography{paper}
+
+\begin{figure}
+\newcommand{\extablecol}[5]{
+\small{\bfseries #1} #2\,\; --\,\; #3\\[-0.1em]
+\quad \footnotesize #4 \\[-0.2em]
+\quad \footnotesize \emph{#5} \\[0.4em]
+}
+
+\begin{tabular}{|l|}
+\hline
+\\[-1em]
+\extablecol{Counter}{\cite{kiss-2014-7guis}}
+  {Counter with increment and decrement buttons.}
+  {The current count is represented by a formula that is modified by the buttons.\\[-0.2em]\quad\footnotesize
+  The user can inspect the evaluation trace to see how the count was modified.}
+  {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
+\extablecol{Todo}{\cite{osmani-2024-todomvc}}
+  {App with controls to add an item and remove all completed.}
+  {Adding an item must correctly merge with independently added functionality to \\[-0.2em]\quad\footnotesize
+   compute which items are completed and remove them based on a formula result.}
+  {Programming by Demonstration, Local-First Collaboration, Incremental Recomputation}
+\extablecol{Conference list}{\cite{edwards-2025-schema}}
+  {Manage a list of invited conference speakers.}
+  {Adding speakers to a list through an in-document user interface merges with \\[-0.2em]\quad\footnotesize
+   refactoring that turns the list into a table and separates name from an email.}
+  {Local-First Collaboration, Programming by Demonstration}
+\extablecol{Conference budget}{\cite{edwards-2025-schema}}
+  {Calculate budget based on a speaker list.}
+  {References are updated  when the list is refactored. Only affected formulas\\[-0.2em]\quad\footnotesize
+   are recomputed and the user can view elements on which the result depends.}
+  {Local-First Collaboration, Incremental Recomputation, End-User Debugging}
+\extablecol{Hello world}{\cite{miller-2001-simult}}
+  {Normalize the capitalization of two word messages.}
+  {An operation to normalize the text in a list item can be recorded and applied to\\[-0.2em]\quad\footnotesize
+   all list items or, alternatively, applied directly to all list items.}
+  {Programming by Demonstration}
+\hline
+\end{tabular}
+
+\caption{Structure of selectors and document nodes}
+\label{fig:examples}
+\vspace{-1em}
+\end{figure}
+
+\appendix
+\section{Formative Examples}
+\label{app:examples}
+
+hello world is only loosely
+
+table of "formative examples" and experiences they require?
++ SOURCE (literature?) => Appendix
+
+The problems include a counter app \cite{kiss-2014-7guis},
+Todo list app and conference planning \cite{edwards-2025-schema}.
+
+\cite{osmani-2024-todomvc}
 
@@
 
-\bibliographystyle{plain}
-\bibliography{paper}
-
 \end{document}

### 2024-12-14 paper.tex
A bit about merging of histories

@@
 a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
-identify common shared part of the history during conflict checking and merging.
+identify a common shared part of the history during conflict checking and merging.
 
 \subsubsection*{Applying Edits}
-When applying an edit, Denicek locates the target node and transform it according to the
-edit. If the edit may affect references in the document (Fig~\ref{fig:edits}, below), Denicek
-updates matching references in the document according to the rules shown in Fig.~\ref{fig:updates}
-provided that the reference behavior is not set to disable the updating (which is required when
-the target reference contains the \ident{Index} selector). Although \ident{WrapList} has to specify
-the index to be used for the newly created list item, the references in the document are updated to
-use the \ident{All} selector. We assume wrapping introduces a homogeneous list and references
-pointing to individual items should thus be generalized to point to all list items.
+When applying an edit, Denicek locates the target node and transform it according to the edit. If
+the edit can affect references in the document (Fig~\ref{fig:edits}, below), Denicek updates
+matching references in the document according to the rules shown in Fig.~\ref{fig:updates} provided
+that the reference behavior of the edit is not set to disable the updating (this is required when
+the target reference contains the \ident{Index} selector). Reference update behavior for
+\ident{WrapRecord} and \ident{WrapList} differs in that the updated references as a result of
+\ident{WrapList} use the \ident{All} selector. Although \ident{WrapList} specifies the index to
+be used for the newly created list item, we assume that the operation introduces a homogeneous
+list and the new reference should point to all eventual list items.
 
@@
 target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
-a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. If the document
-contains a reference that would be invalidated by the \ident{Copy} or \ident{DeleteField} edit,
-the edit is rejected.
+a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (A reference in the
+document cannot be more general, because such edit would have to contain the \ident{Index} selector
+and this would require setting reference behavior to not trigger reference update.) Finally,
+if the document contains a reference that would be invalidated by the \ident{Copy} or
+\ident{DeleteField} edit, the edit is rejected.
+
+\subsubsection*{Merging Edit Histories}
+
+Merging edit histories is used when two users edit document independently, but also
+when replaying edits in programming by demonstration. The
+operation works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix $E$.
+The merging operation is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that
+can be reapplied on top of the other edit history, resulting in a new history $E, E_1, E_2'$.
+Note that the operation is not symmetric. If there are conflicts among the edits in $E_1$ and
+$E_2$, the result of $E, E_1, E_2'$ will differ from the result of $E, E_2, E_1'$.
+We return to this problem in the next section when discussing conflict detection.
+
+The key operation that enables such reconcilliation takes two individual edits that occurred
+indpendently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
+can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
+There are two aspects of such reconcilliation:
+
+\begin{enumerate}
+\item \emph{Apply to Newly Added.} If $e_2$ is adding new nodes to the document, but
+  $e_1$ modified the document through a selector that would also affect the new nodes added by $e_2$,
+  we need to apply the transformation represented by $e_1$ to the nodes newly added by $e_2$.
+  This is done by generating an additional edit, to be applied after $e_2$, that is based on
+  $e_1$ but targets only the newly added nodes (more details can be found in \S\ref{app:merge-apply-to-added}).
+
+\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
+  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
+  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
+  that apply when transforming references inside a document, although we also support the case
+  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
+\end{enumerate}
+
+no longer
 
@@
 
+Applying an edit to a newly added node involves a number of cases. When the new node is added
+using \ident{Add} or \ident{Copy}, but the target location is modified, this is an unresolvable
+conflict (because those two operations would overwrite existing nodes). When the new node is
+added as a new list item using \ident{Append}, it can be transformed (by applying the other
+edit direclty to the new node).
+
+Finally, when the new node is added using \ident{AppendFrom},
+it is copied from another document location. We cannot transform the node in the soource location
+(this would have unintended effects) or after adding it (we do not know its new index because of
+aforementioned \emph{non-conditionality} of edits). Denicek reconcilliates such edits by first
+copying the source node to a new temporary location, transforming it there and then using
+\ident{AppendFrom} from the temporary location.
+
+~
+
+~
+
 todo
@@
 ~
+
+Removing all conflicting edits is sometimes useful, but there are many cases where conflicting
+edits, in the above sense, can be reconcilliated.
+
+If the edits in $E_1$ and $E_2$ conflict,
+the results of applying $E, E_1, E_2$ and $E, E_2, E_1$ would be different. We can, however,
+construct edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
+$E, E_2, E_1'$ are the same.
 
@@
 on its original result are also removed.
-
-\subsubsection*{Merging Edit Histories}
-Removing all conflicting edits is sometimes useful, but there are many cases where conflicting
-edits, in the above sense, can be reconcilliated. Say we have edit histories with
-the same common shared part $E, E_1$ and $E, E_2$. If the edits in $E_1$ and $E_2$ conflict,
-the results of applying $E, E_1, E_2$ and $E, E_2, E_1$ would be different. We can, however,
-construct edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
-$E, E_2, E_1'$ are the same.
-
-The key operation that enables such reconcilliation takes two individual edits that occurred
-indpendently, $e_1$ and $e_2$, and produces $e_1'$ that has the same logical effect as
-$e_1$, but can be applied after $e_2$. There are two aspects of such reconcilliation:
-
-\begin{enumerate}
-\item \emph{Apply Edit to Newly Added.} If $e_1$ is adding a new list item, but~$e_2$ is
-  changing elements of the target list, we apply the edit $e_2$ to the node added by $e_1$ so
-  that, when it is added after the transformation, the added node has the new structure.
-
-\item \emph{Transform Matching References.} If $e_1$ targets a node that is inside a node
-  whose structure is changed by $e_2$, the target reference in $e_1$ is updated in a way that
-  corresponds to the new structure. That is, using the same rules, shown in Fig.~\ref{fig:updates},
-  that apply when transforming references inside a document.
-\end{enumerate}
-
-Applying an edit to a newly added node involves a number of cases. When the new node is added
-using \ident{Add} or \ident{Copy}, but the target location is modified, this is an unresolvable
-conflict (because those two operations would overwrite existing nodes). When the new node is
-added as a new list item using \ident{Append}, it can be transformed (by applying the other
-edit direclty to the new node).
-
-Finally, when the new node is added using \ident{AppendFrom},
-it is copied from another document location. We cannot transform the node in the soource location
-(this would have unintended effects) or after adding it (we do not know its new index because of
-aforementioned \emph{non-conditionality} of edits). Denicek reconcilliates such edits by first
-copying the source node to a new temporary location, transforming it there and then using
-\ident{AppendFrom} from the temporary location.
 
@@
 
+\newpage
+~
+\newpage
+
+\appendix
+\section{Formative Examples}
+\label{app:examples}
+
+The design the Denicek substrate, we identified five formative examples shown in
+Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
+(Todo and Counter apps) to problems posed as schema change challenges \cite{edwards-2025-schema}.
+The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
+built (as directly as possible) on top of the substrate and was used to solve implement the
+formative examples.
+
+Many of the formative examples include a small programming challenge, such as adding
+user interface to add a new speaker, a new list item or modify the count. Our aim was for
+the substrate to enable solving  those through programming by demonstration.
+Programming by Demonstration is often used in data wrangling
+\cite{gulwani-2012-spreadsheet,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
+is only a minimalistic illustration of such use, loosely inspired by earlier work \cite{miller-2001-simult}.
+
+\section{Merging Edit Histories}
+Recall that merging takes two edit histories, $E, E_1$ and $E, E_2$,
+transforms edits $E_2$ into $E_2'$ that can be reapplied on top of the first history
+resulting in $E, E_1, E_2'$. The key operation takes two individual edits, $e_1$ and $e_2$
+and produces a sequence of edits $e_2', e_2'', \ldots$ that can be applied after $e_1$,
+and combine the two edits. This section provides details about the two aspects of this operation.
+
+\subsection{Apply to Newly Added}
+\label{app:merge-apply-to-added}
+
+Assume that edits $e_1$ and $e_2$ occurred independently. We want to modify $e_2$ so that
+it can be placed after $e_1$. If the edit $e_2$ added new nodes to the document that the
+edit $e_1$ would affect, we generate an additional edit that apply the transformation of $e_1$
+to the newly added nodes (and only to those).
+
+The only edits that add new document nodes are \ident{Add}, \ident{Append}, \ident{Copy} and
+so we consider this case if the edit $e_2$ is one of those. If so, we check
+whether the target of $e_1$ is within the target of $e_2$, i.e., the list of selectors that
+forms the target of $e_2$ is a prefix of the list of selectors that forms the target of $e_1$.
+
+Along the way, we compute a \emph{more specific prefix}. If the target of $e_1$ contains
+the \ident{All} selector, it can be matched against a specific \ident{Index} selector in the
+target of $e_2$ (if the selector of $e_1$ is more specific than that of $e_2$, the targets
+are not matched). We then replace the original prefix in $e_1$ with the \emph{more specific prefix}
+that contains \ident{Index} selector in places where the original edit contained \ident{All}.
+This way, we obtain $e_1'$ which is a focused version of $e_1$ that applies only to the
+nodes newly added by $e_2$. The edit $e_2$ thus becomes a pair of edits $e_2, e_1'$.
+The final document will contain edits $e_1, e_2, e_1'$ -- that is, it will first apply the edit
+$e_1$ to nodes already in the document, then add new nodes and then apply the transformation
+represented by $e_1$ to the newly added nodes.
+
+\subsection{Transform Matching References}
+\label{app:merge-transform-refs}
+
+As above, assume that edits $e_1$ and $e_2$ occurred independently. We want to
+modify $e_2$ so that it can be placed after $e_1$. If $e_1$ is any of the three edits
+listed in Fig.~\ref{fig:updates} (\ident{RenameField}, \ident{WrapRecord}, \ident{WrapList}),
+we collect all references that appear inside $e_2$ (the target, the source of \ident{Copy}
+and any references occurring in the nodes added by \ident{Add} or \ident{Append}).
+If the target of $e_1$ is a prefix of any of those references, we update the references
+accordingly and obtain a new edit $e_2'$. Note that it would be an error to match specific
+\ident{Index} in $e_1$ with more general \ident{All} in $e_2$, but this cannot happen --
+reference updating is not done when the target of $e_1$ contains \ident{Index}.
+
+Now consider the case when $e_1$ is \ident{Copy} and the edit $e_2$ targets a node that is
+the source node of the copy operation (or any of its children). In this case, it is reasonable
+to require that the edit $e_2$ is applied to both the source and the target of the copy.
+(This is required by the refactoring done in the Conference Budget example.)
+We handle the case by creating a copy of $e_2$ with transformed selectors (target and, if $e_2$
+is also \ident{Copy}, also its source). To transform the selectors, we replace the prefix
+formed by the source of the \ident{Copy} by a new prefix, formed by the target target of
+the \ident{Copy}. We then add the new operation as $e_2'$ if at least one of its selectors
+was transformed (typically target, but possibly also source).
+
 \begin{figure}
@@
   {Programming by Demonstration, Local-First Collaboration, Incremental Recomputation}
-\extablecol{Conference list}{\cite{edwards-2025-schema}}
+\extablecol{Conference List}{\cite{edwards-2025-schema}}
   {Manage a list of invited conference speakers.}
@@
   {Local-First Collaboration, Programming by Demonstration}
-\extablecol{Conference budget}{\cite{edwards-2025-schema}}
+\extablecol{Conference Budget}{\cite{edwards-2025-schema}}
   {Calculate budget based on a speaker list.}
@@
   {Local-First Collaboration, Incremental Recomputation, End-User Debugging}
-\extablecol{Hello world}{\cite{miller-2001-simult}}
+\extablecol{Hello World}{\cite{miller-2001-simult}}
   {Normalize the capitalization of two word messages.}
@@
 
-\caption{Structure of selectors and document nodes}
+\caption{Formative examples used in Denicek design}
 \label{fig:examples}
@@
 
-\appendix
-\section{Formative Examples}
-\label{app:examples}
-
-hello world is only loosely
-
-table of "formative examples" and experiences they require?
-+ SOURCE (literature?) => Appendix
-
-The problems include a counter app \cite{kiss-2014-7guis},
-Todo list app and conference planning \cite{edwards-2025-schema}.
-
-\cite{osmani-2024-todomvc}
-
-xx
 \newpage

### 2024-12-15 paper.tex
Conflict resolution, some other tweaks

@@
 \usepackage{hyperref}
+
 \newcommand{\ident}[1]{{\sffamily #1}}
+\newcommand{\srcid}[1]{\textnormal{\sffamily #1}}
+\newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}
 
@@
 that construct and transform a document consisting of data and formulas. Denicek provides three
-primitive operations on series of edits, application of edits, merging of histories and conflict checking.
+primitive operations on series of edits, application of edits, merging of histories and conflict resolution.
 We show that the three operations form the backbone of a direct implementation of a range of
@@
 \maketitle
+
+% ==================================================================================================
 
@@
 
+% ==================================================================================================
+
 \section{Background}
@@
 % also cool: https://arxiv.org/pdf/2303.06777
+
+Webnicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
+document visible to the user, although parts can be collapsible or hidden using CSS.
+
+Merging - c.f. mywebstrates - we need more than Automerge (cite)
+which has a bunch of pre-built structures
 
@@
 
+% ==================================================================================================
+
 \section{Walkthrough}
@@
 
+% ==================================================================================================
+
 \subsection{Selectors, Documents and Edits}
@@
 References to document locations can be relative or absolute. Both kinds can appear in a
-document (reference nodes) and absolute references are also used in edits (to denote a target~node).
+document (reference nodes); absolute references are used in edits (to denote a target~node).
 
@@
 way to refer to all children of a record. Note that Denicek does not use implicit numerical
-indices for lists. The index of a new item has to be supplied explicitly. The design choice
-simplifies merging and is discussed in \S\ref{sec:discuss}.
+indices for lists. The index of a new item has to be supplied explicitly. The reasoning behind
+this design choice is discussed in \S\ref{sec:discuss}.
 
@@
 \ident{Parent} selector. It can contain the \ident{Any} selector, in which case the edit is applied
-to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a particular
-kind.
-
-The edits are designed to allow any transformation of a document
-through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
-Denicek updates references when document structure changes. Fig.~\ref{fig:edits} distinguishes
-between edits that keep existing references in a document unchanged (above) and edits that affect
-references (below).
+to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
+kind. As discussed in \S\ref{sec:discuss}, both fields of a record and list elements are ordered
+and edits that add a new item take the index or field name of a previous node.
+
+The edits allow any transformation of a document through a series of steps whose effect can be
+tracked by the substrate. As we saw, Denicek updates references when document structure changes.
+Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
+(above) and edits that affect references (below).
 
@@
 referring to the original location should continue referring to the source or should be changed to
-point to the target of the copying after the edit (a reference cannot refer to two structurally
-independent locations).
+point to the target of the copying (a reference cannot refer to two locations).
 
@@
 }
-\begin{tabular}{|cp{17em}r|}
+\begin{tabular}{|cp{16em}r|}
 \hline
@@
   {Delete the item at a given \textit{index} of a specified list.}
-\ektablecol{\faCode}{UpdateTag}{tag}{Tagged}
+\ektablecol{\faCode}{UpdateTag}{tag}{List or Record}
   {Change the tag of a specified list or record from to a new \textit{tag}.}
@@
 
+% ==================================================================================================
+
 \subsection{Primitive Operations}
@@
 a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
-identify a common shared part of the history during conflict checking and merging.
+identify a common shared part of the history during conflict resolution and merging.
 
@@
 \subsubsection*{Merging Edit Histories}
-
 Merging edit histories is used when two users edit document independently, but also
@@
 
-no longer
-
-~
-
-~
-
-Applying an edit to a newly added node involves a number of cases. When the new node is added
-using \ident{Add} or \ident{Copy}, but the target location is modified, this is an unresolvable
-conflict (because those two operations would overwrite existing nodes). When the new node is
-added as a new list item using \ident{Append}, it can be transformed (by applying the other
-edit direclty to the new node).
-
-Finally, when the new node is added using \ident{AppendFrom},
-it is copied from another document location. We cannot transform the node in the soource location
-(this would have unintended effects) or after adding it (we do not know its new index because of
-aforementioned \emph{non-conditionality} of edits). Denicek reconcilliates such edits by first
-copying the source node to a new temporary location, transforming it there and then using
-\ident{AppendFrom} from the temporary location.
-
-~
-
-~
-
-todo
-
-~
-
-~
-
-Removing all conflicting edits is sometimes useful, but there are many cases where conflicting
-edits, in the above sense, can be reconcilliated.
-
-If the edits in $E_1$ and $E_2$ conflict,
-the results of applying $E, E_1, E_2$ and $E, E_2, E_1$ would be different. We can, however,
-construct edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
-$E, E_2, E_1'$ are the same.
-
-\subsubsection*{Conflict Detection}
-Two edits are conflicting if the order in which they are applied matters. This is the case if
-they target the same node, or if one targets a node that is nested inside the node targetted by
-the other. For edits with dependencies (\ident{AppendFrom} and \ident{Copy}), a conflict also
-occurs if the other edit modifies the dependency.
-
-Given two edit histories, one way to resolve conflicts is by removing (or marking as disabled) all
-edits from one of the histories that conflict with edits done by the other history. To do this,
-we first collect all targets of edits in the other history. We then iterate over edits from the first
-history to see if they depend on or target any of the affected targets. If an edit is removed,
-its target reference is added to the set of affected targets, so that other subsequent edits that depend
-on its original result are also removed.
-
 % ==================================================================================================
@@
 }
-\begin{tabular}{|l|}
+\begin{tabular}{|p{27em}|}
 \hline
@@
 
+\begin{figure}
+\newcommand{\eftablecol}[3]{
+\small{\bfseries #1}\;\footnotesize\textit{target}\,\; --\,\; {\footnotesize #2}\\[-0.1em]
+\quad \small #3\\[0.3em]
+}
+\begin{tabular}{|p{27em}|}
+\hline
+\\[-1em]
+\eftablecol{StructureEffect}{Affects fields or structure of the target node.}
+  {\ident{RenameField}, \ident{DeleteField}, \ident{WrapRecord}, \ident{WrapList}, \ident{Copy}}
+\eftablecol{ValueEffect}{Transforms value, modifies list or adds an additional field.}
+  {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
+\eftablecol{TagEffect}{Modifies the tage of the target node.}
+  {\ident{UpdateTag}}
+\hline
+\end{tabular}
+\vspace{-0.5em}
+\caption{Effects of individual edit operations}
+\label{fig:effects}
+\vspace{-1em}
+\end{figure}
+
+% ==================================================================================================
+
+To illustrate merging, consider a case where we created a list of work items {\small\Verb|/todo|}.
+In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
+list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
+%
+\begin{equation*}
+\arraycolsep=2pt
+\begin{array}{rcl}
+e_2 &=& \srckvd{Append}(\textnormal{\small\Verb|/todo|}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
+e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
+e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+\end{array}
+\end{equation*}
+%
+If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
+reflect the additional wrapping (2) and we need to create and additional \ident{Add} operation
+that will add the checkbox to the new item (1). The result is two edits $e_2', e_2''$:
+%
+\begin{equation*}
+\arraycolsep=2pt
+\begin{array}{rcl}
+e_2'  &=& \srckvd{Append}(\textnormal{\small\Verb|/todo/items|}, \#1, \#0, \srcid{<li>Do some work</li>})\\
+e_2'' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/\#1|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+\end{array}
+\end{equation*}
+%
+If we performed the merge operation in the other order, we would simply append $e_1, e_1'$ after
+$e_2$. Although this is not the case in general, the result will be the same regardless of the order here.
+
+\subsubsection*{Conflict Resolution}
+When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is sometimes possible to construct
+edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
+$E, E_2, E_1'$ are the same. However, this is not always the case. Two edits can conflict if
+they both modify the same value, if they transfrom the structure of a node in incompatible ways
+or if one modifies a node nested inside a node replaced or deleted by the other.
+Detected conflicts can be reported to the user, but we also use conflict detection to
+implement incremental evaluation (\S\ref{sec:impl-incremental}).
+
+The Denicek substrate implements a conflict detection mechanism inspired by effect
+systems~\cite{lucassen-1988-effects}. The mechanism is simple and tractable, but over-approximates
+conflicts, i.e.~it may report a conflict even if two edits can be merged successfully
+(see \S\ref{sec:discuss-future}). An effects describe how an edit affects the document structure
+and we distinguish between three types of effects as shown in Fig.~\ref{fig:effects}.
+
+We say that a set of effects $F_1$ conflicts with another set of effects $F_2$ if there
+are effects $f_1\in F_1$ and $f_2\in F_2$ such that they are of the same kind and the target
+of $f_1$ is a prefix of the target of $f_2$ or vice versa (allowing specific \ident{Index} to
+match against \ident{All} in any direction).
+
+Given two edit histories $E, E_1$ and $E, E_2$, Denicek can use conflict detection to remove
+all conflicting edits from $E_2$ and produce a sequence of remaining edits $e_2', e_2'', \ldots$
+that can be added after $E, E_1$ and do not conflict with edits in $E_1$.
+
+To do this, we iterate over edits $e_2$ from $E_2$ and check if the dependencies of $e_2$
+conflict with effects of (1) any of the effect $e_1$ from $E_1$ or (2) effects of any of the
+previously removed edits. Here, the dependencies of $e_2$ include its target, but also source
+of \ident{Copy} and additional dependencies recorded explicitly as discussed in
+\S\ref{sec:impl-incremental}. If a conflict is detected, the edit $e_2$ is removed and its effect
+is recorded, so that we remove any subsequent edits that would depend on the removed edit.
+
+% ==================================================================================================
+
 \section{Programming Experiences Implementation}
@@
 
-Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
-document visible to the user, although parts can be collapsible or hidden using CSS.
-
-[webnicek?]
-
-The key claim of this paper is that the computational substrate described in the previous
-section makes it easy to support a range of experiences that make programming more concrete,
-collaborative and interactive. In this section, we describe how to use the substrate to support
-local-first collaborative editing (\S\ref{sec:impl-collab}), programming by
-demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
-schema and code co-evolution (\S\ref{sec:impl-schema}), incremental recomputation
-(\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}), provenance tracking (\S\ref{sec:impl-provenance})
+The key claim of this paper is that the Denicek computational substrate
+makes it easy to support a range of experiences that make programming more concrete,
+collaborative and interactive.
+
+In this section, we describe how Webnicek uses the substrate
+to support local-first collaboration (\S\ref{sec:impl-collab}), programming by
+demonstration (\S\ref{sec:impl-pbd}, \S\ref{sec:impl-interaction}), incremental recomputation
+(\S\ref{sec:impl-eval}, \S\ref{sec:impl-incremental}), schema change control (\S\ref{sec:impl-schema}),
+end-user debugging via provenance tracking (\S\ref{sec:impl-provenance})
 and concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
-We describe the programming experience in isolation in this section. Next section provides a
+
+We describe the programming experience in isolation in this section. The next section provides a
 more comprehensive evaluation through a case study that combines multiple of them together.
 
-\subsection{Local-First Collaborative Editing}
+% ==================================================================================================
+
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
+\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
+\label{fig:merging}
+\end{figure}
+
+% ==================================================================================================
+
+\subsection{Local-First Collaboratation}
 \label{sec:impl-collab}
 
-The Denicek representation enables \emph{local-first} collaboration~\cite{kleppmann-2019-local},
-as illustrated in \S\ref{sec:walk}E. If a document is edited by multiple users, they can
-each make edits to their local copy and eventually merge document variants using the operation to
+The Denicek representation enables local-first collaboration~\cite{kleppmann-2019-local},
+as illustrated previously in \S\ref{sec:walk} (E). If a document is edited by multiple users, they can
+each make edits to their local copy and eventually merge the variants using the operation to
 merge edit histories.
@@
 document variants \cite{edwards-2025-schema}, i.e., the user cannot, for example, maintain their
-own local document structure and import new data from another variant (this would require an
-inverse of the edit reconcilliation operation).
-
-Merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
+own local document structure and import new data from another variant (an alternative discussed
+in \S\ref{sec:discuss}).
+
+Recall that merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
 can be obtained either by appending $C'$ (produced by the edit reconcilliation operation)
-on top of $A,B$ or by appending $B'$ on top of $A,C$. Although the two edits are conflicting,
-$C$ only affects data and $B$ only affects structure and so the resulting document is the same
-in both cases. However, the histories differ. In the first case, the new node added by the
-\ident{Append} edit is transformed (from primitive string to a record representing table row).
+on top of $A,B$ or by appending $B'$ on top of $A,C$. According to our effect analysis, the two
+operations are conflicting. Although $B$ primarily affects the document structure, it also adds
+a new field to the record (email), which is a \ident{ValueEffect}, conflicting with
+\ident{ValueEffect} of $C$. In this scenario, the conflict can be ignored and the resulting document
+is the same in both cases.
+
+However, the resulting two histories will differ. In the first case, the
+\ident{Append} edit that adds a new node is supplemented by further focused edits that transform
+the added node from a list item to a new table row.
 In the second case, the structural transformations are automatically applied to all rows.
@@
 \begin{figure}[t]
-\includegraphics[width=0.9\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
-\vspace{-0.5em}
-\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
-\label{fig:merging}
-\vspace{-0.5em}
-\end{figure}
-
-\begin{figure}[t]
 \includegraphics[width=0.9\columnwidth,clip,trim=0cm 7cm 9cm 0cm]{fig/pbd.pdf}
-\vspace{-0.5em}
 \caption{Programming by demonstration is implemented by selecting edits from the document
@@
 \label{fig:pbd}
-\vspace{-0.5em}
 \end{figure}
@@
 
-In \emph{programming by demonstration} \cite{cypher-1993-pbd}, the user demonstrates a task to the
+In programming by demonstration \cite{cypher-1993-pbd}, the user demonstrates a task to the
 system and the system then repeats it, directly or in a generalized way. To use direct
 repetition with Denicek (Fig.~\ref{fig:pbd}), the user can select edits from the edit history,
-name them and replay them. In case of general-purpose document editing (in our formative prototype),
-this requires certain forethought (e.g., constructing a new list item in a temporary document
-field before appending it), but as illustrated in \S\ref{sec:case}, the mechanism is effective in
-a restricted domain such as data wrangling \cite{kandel-2011-wrangler}.
-
-There are two notable aspects of our implementation. First, Deni\-cek saves edits in the document
-itself (by representing individual edits as nodes and storing them in list inside a
-{\small\Verb_<saved-interactions>_} field). This means that no other implementation mechanism
-outside of the system is needed and also that the stored edits can be modified by the user
-(or tools working with the document).
-
-Second, to replay edits, Denicek does not simply append them on top of the current history.
-When saving edits, it records the hash of the history at the time of saving. When replaying edits,
-it appends the saved edits to the top of the original history (at the time of saving) and merges this new
-sequence with the current history. This pushes the saved edits through all subsequent edits made
-by the user. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where a newly added speaker
-is transformed from a primitive string to a table row.
-
-The implementation also has to account for the case where edits saved in the document are themselves
+name them and replay them. In case of general-purpose document editing in Webnicek, this requires
+certain forethought, but as illustrated in \S\ref{sec:case}, the mechanism is very effective in
+a domain such as data wrangling \cite{kandel-2011-wrangler}.
+
+There are two notable aspects of our implementation of programming by demonstration in Webnicek.
+First, Webnicek records edits in the document itself (by representing individual edits as nodes and
+storing them in list inside a {\small\Verb_/saved-interactions_} field). This means that no other
+implementation mechanism outside of the system is needed and also that the stored edits can be
+modified by the user or tools working with the document.
+
+Second, to replay edits, Webniceks does not append the recorded edits on top of the current history.
+When saving edits, it stores the hash of the history at the time of saving. When replaying,
+it appends the recorded edits to the top of the original history (at the time of saving) and merges
+this new sequence of edits with the current history. This pushes the recorded edits through all
+subsequent edits made by the user. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
+a newly added speaker is transformed from a list item to a table row.
+
+Webnicek also accounts for the case where edits recorded in the document are themselves
 transformed (when they are reconcilliated with other edits during merging). In this case, Denicek
-updates the saved edits (as discussed in \S\ref{sec:discuss}, this would not be needed if the
-history was stored as a graph, akin to ordinary git merging rather than rebasing).
+updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).
 
@@
 simple case (Fig.~\ref{fig:interactive} (1)), the {\small\Verb_click_} event handler is set to a
-reference to a sequence of edits saved in the document. Clicking the button executes the edits
+reference to a sequence of edits recorded in the document. Clicking the button executes the edits
 using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., by appending them to a history at the
@@
 
-The Denicek substrate can be used to generalize the interactions saved through programming by
+The Denicek substrate can be used to generalize the interactions recorded through programming by
 demonstration. Our prototype illustrates this option in a limited way. As shown in
@@
 from the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
-In addition to the saved interaction, we manually specify (in the source view) that the edits
+In addition to the recorded interaction, we manually specify (in the source view) that the edits
 should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
@@
 based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}
+
+\subsection{Formula Language and Evaluation}
+\label{sec:impl-eval}
+
+Denicek documents can contain formulas inspired by the spreadsheet
+paradigm \cite{nardi-1990-spreadsheets}. Formulas can specify richer computations than
+what can be expressed using document edits. Formulas do not transform the document and their
+results are transient, although their evaluation also leverages the substrate operations,
+namely merging of edit histories.
+
+As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
+special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
+rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
+edits and the Denicek substrate treates them as stnadard nodes.
+
+To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
+record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
+formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
+(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
+as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
+with the document is discussed in \S\ref{sec:impl-incremental}
+
+The counter example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
+and programming by demonstration. To implement a counter, the Increment and Decrement buttons
+wrap the current counter value in a formula that adds or subtracts 1.
 
@@
 \end{figure}
-
-\subsection{Formula Language and Evaluation}
-\label{sec:impl-eval}
-
-Denicek documents can contain formulas inspired by the spreadsheet
-paradigm \cite{nardi-1990-spreadsheets}. Formulas can specify richer computations than
-what can be expressed using document edits. Formulas do not transform the document and their
-results are transient, although their evaluation also leverages the substrate operations,
-namely merging of edit histories.
-
-As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
-special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
-rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
-edits and the Denicek substrate treates them as stnadard nodes.
-
-To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
-record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
-formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
-(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
-as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
-with the document is discussed in \S\ref{sec:impl-incremental}
-
-The counter example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
-and programming by demonstration. To implement a counter, the Increment and Decrement buttons
-wrap the current counter value in a formula that adds or subtracts 1.
 
@@
 % TODO: Add a section after Local-First Editing on the new conflict resolution (with demos)
-% TODO: In implementation use effects to better replay saved edits
+% TODO: In implementation use effects to better replay recorded edits
 % (push through new structural but not data edits)
@@
 
+divergence model of challenges paper - requires inverse operation for moving edits backwards
+
+alternative approache is discussed in \S\ref{sec:discuss}
+=> PBD - updates the recorded edits this would not be needed if the
+history was stored as a graph, akin to ordinary git merging rather than rebasing).
+
 \section{Case study}
@@
 use type system to check temporarily invalid state
+
+better effect system
 

### 2024-12-19 paper.tex
Debugging, concrete programming, design sections

@@
 Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
-To \emph{design} Denicek, we identify a number of \emph{formative examples} -- simple programming
+To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
 tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
@@
 ~
-% ==================================================================================================
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-In this section, we use Webnicek, our simple web-based prototype, to illustrate the programming
+In this section, we use the web-based prototype Webnicek to illustrate programming
 experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
-editor that supports navigation in the document and issuing of edit commands. We show two
-formative examples (\S\ref{app:examples}) related to conference organization.
+editor that supports navigation in the document and issuing of edit commands. We follow two
+formative examples (\S\ref{app:examples}) related to conference planning.
 
@@
 
-% ==================================================================================================
+% --------------------------------------------------------------------------------------------------
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}[t]
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Programming By Demonstration}
@@
 Webnicek also accounts for the case where edits recorded in the document are themselves
-transformed (when they are reconcilliated with other edits during merging). In this case, Denicek
+transformed (when they are reconcilliated with other edits during merging). In this case, Webnicek
 updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Interactive User Interfaces}
@@
 
-Programming by demonstration can be used to define interactive elements in the document. In the
-simple case (Fig.~\ref{fig:interactive} (1)), the {\small\Verb_click_} event handler is set to a
-reference to a sequence of edits recorded in the document. Clicking the button executes the edits
-using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., by appending them to a history at the
-time of saving and merging them with the current history.
+Programming by demonstration can be used to define interactive elements in the document. In a
+simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\Verb_click_} event
+handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
+executes the edits using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., Webnicek appends
+the edits to a history at the time when the edits were recorded and merges the edits with the
+current history.
+
+The use of merging when replaying recorded edits is crucial in both the Todo App example and the
+Conference List example (see \S\ref{app:examples}). In both cases, it makes it possible to define
+a user interface for adding items (new Todo items, new speakers) and later change the document
+structure (refactor speaker list to a table) or add functionality (formula to evaluate whether a
+Todo item has been completed) without having to recreate the user interface for adding items.
+The use of merging ensures that new items are added in a correct format or with the additional
+functionality.
 
@@
 demonstration. Our prototype illustrates this option in a limited way. As shown in
-Fig.~\ref{fig:interactive} (2), a button to remove all completed TODO items can be generalized
-from the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
+Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
+generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
 In addition to the recorded interaction, we manually specify (in the source view) that the edits
@@
 Programming by demonstration systems typically implement heuristic for generalization
-\cite{myers-2000-intelligence} infers and suggests such generalizations. When integrated into
-a system based on Denicek, such heuristic may also be able to automatically add a formula
+\cite{myers-2000-intelligence} infers and suggests such generalizations. If integrated into
+a system based on Webnicek, such heuristic could automatically construct a formula
 based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
-with the document is discussed in \S\ref{sec:impl-incremental}
-
-The counter example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
+with the document is discussed in the next section.
+
+The Counter App example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
 and programming by demonstration. To implement a counter, the Increment and Decrement buttons
@@
 
+\keyideabox{\faSuperscript}{Formula Language}{
+Webnicek exposes the underlying representation of formulas to the user, but the same representation
+can be edited through a user-friendly mechanism such as a block-based editor
+\cite{jansen-2019-xlblocks} or a calculation view \cite{sarkar-2018-calc}. The key point
+is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
+way, edit them and merge them with other changes.}
+
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}[t]
+\vspace{-0.5em}
 \includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
+\vspace{-0.5em}
 \caption{Increment wraps the existing count in a formula that adds 1 to the
@@
 \vspace{-0.5em}
-\begin{minipage}{0.6\columnwidth}
+\begin{minipage}{0.55\columnwidth}
   ~\\[1em]
@@
   \end{minipage}%
-  \begin{minipage}{0.4\columnwidth}
+  \begin{minipage}{0.45\columnwidth}
     \caption{Evaluated edits (A, B) are kept at the top of the history. When they conflict with
-    ordinary edits (4), evaluated edits are removed (A).}
+    ordinary edits (4), they are removed (A).}
     \label{fig:eval}
@@
 
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Incremental Recomputation}
@@
 
-As illustrated in Fig.~\ref{fig:incremental}, Denicek supports incremental re-computa\-tion.
+As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental re-computa\-tion.
 In the example, the cost of speaker travel depends on the number of speakers, but the cost
@@
 The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
-Denicek generates \emph{evaluated edits} that are appended to the top of the history.
+Webnicek generates \emph{evaluated edits} that are appended to the top of the history.
 When subsequent edits are made, they are appended after all non-evaluated edits and the
 evaluated edits are pushed through the newly added edits. If the edits conflict (according to
-Denicek's conflict detection), affected evaluated edits and edits that depend on them are dropped.
-
-\keyideabox{\faRefresh}{Live Programming}{In the exploratory prototype, Denicek does not
-automatically evaluate formulas. This makes it easier to see and understand the evaluation
-mechanism, but a realistic system based on the substrate would likely automatically evaluate
-formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}.}
+conflict detection discussed in \S\ref{sec:system-ops}), affected evaluated edits and edits that
+depend on them are dropped.
+
+\keyideabox{\faRefresh}{Live Programming}{The Webnicek prototype does not automatically
+evaluate formulas. This makes it easier to understand the evaluation
+mechanism, but a realistic system based on the substrate could automatically evaluate
+formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}
+and use incremental recomputation for performance~reasons.}
+
+% --------------------------------------------------------------------------------------------------
 
@@
   \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
-  \vspace{-0.5em}
+  \vspace{-1em}
   \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
@@
   \label{fig:incremental}
+  \vspace{-1em}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Schema Change Control}
+\label{sec:impl-schema}
+
+Document structure often needs to evolve \cite{burnett-2014-silos}, as illustrated by our
+Conference List example where a list is transformed into a table. When this happens,
+data and code that depend on the structure of the document need to evolve correspondingly.
+The problem is well-known in database systems \cite{rahm-2006-schema} and has recently been
+studied in the context of programming systems \cite{edwards-2025-schema}.
+
+Although Denicek does not explicitly track document structure (schema or type), all documents
+have an implicit structure and some edits transform this structure. As
+discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
+document when edits modify the document structure. This enables a form of schema code
+co-evolution \cite{edwards-2025-schema}. Formulas embedded in Denicek documents use reference
+nodes to refer to both data sources (in the document) and the results of other computations.
+Consequently, if the document structure changes, the formulas are automatically updated.
+
+Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\Verb_<ul>_} is
+turned into {\small\Verb_<tbody>_} using \ident{UpdateTag} and wrapped inside {\small\Verb_<table>_}
+with a field {\small\Verb_body_} using \ident{WrapRecord}. For the latter, Denicek updates the
+reference accordingly turning the original {\small\Verb_/speaker_} reference in the formula
+into {\small\Verb_/speaker/body_}.
+
+As noted earier, evaluated edits transform the formula structure (wrapping it in the
+{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
+references to formulas are not transformed when the formula is evaluated (otherwise, references
+would be updated to point to the original unevaluated formulas).
+
+% --------------------------------------------------------------------------------------------------
 
@@
 
-\subsection{Schema Code Co-evolution}
-\label{sec:impl-schema}
-
-The problem of schema evolution is that, when the schema used in a database or a program evolves,
-data and code that depend on the schema need to evolve correspondingly. The problem is well-known
-in database systems \cite{rahm-2006-schema} and has recently been studied in the context of
-programming systems \cite{edwards-2025-schema}.
-
-Although Denicek does not explicitly track document structure (schema or type), all documents
-have an implicit structure and some of the edits transform the document structure. As
-discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
-document in this case. This enables a form of schema code co-evolution \cite{edwards-2025-schema}.
-The formulas embedded in Denicek documents use reference nodes to refer to both data sources
-(in the document) and the results of other computations. Consequently, if the document structure
-changes, the formulas are automatically updated.
-
-Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\Verb_<ul>_} is
-turned into {\small\Verb_<tbody>_} using \ident{UpdateTag} and wrapped inside {\small\Verb_<table>_}
-with a field {\small\Verb_body_} using \ident{WrapRecord}. For the latter, Denicek updates the
-reference accordingly turning the original {\small\Verb_/speaker_} reference in the formula
-into {\small\Verb_/speaker/body_}.
-
-As noted earier, evaluated edits transform the formula structure (wrapping it in the
-{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
-references to formulas are not transformed when the formula is evaluated (otherwise, references
-to evaluated formulas would not point to the evaluated result, but to the original unevaluated
-definition).
+% --------------------------------------------------------------------------------------------------
 
@@
 
-\newpage
-
-\subsection{Provenance tracking}
+% --------------------------------------------------------------------------------------------------
+
+\subsection{End-User Debugging}
 \label{sec:impl-provenance}
 
-\subsection{Managed copy \& paste}
+The most common kind of end-user programming question is determining whether a value they observe
+is right or wrong \cite{kissinger-2006-debugging}. One way to help users answer the question is
+to provide an explanation how a value was obtained \cite{ko-2009-whyline}. Webnicek provides a basic
+mechanism that highlights document nodes that contributed to a specific computed result,
+illustrated in Figure~\ref{fig:provenance}.
+
+The implementation leverages the fact that evaluation keeps the original formula, as discussed
+in \S\ref{sec:impl-eval}. When a formula is evaluated, the final {\small\Verb_<x-evaluated>_}
+document node contains the result, but also sub-tree that represents a full evaluation trace
+\cite{perera-2012-functional}. We analyse the trace, collect all reference nodes in the trace and
+highlight all nodes referred to in the computation. Denicek makes this easy as we
+only need to collect reference nodes nested in a formula node.
+
+\keyideabox{\faBarChart}{Explanations and Linked Visualizations}{In Webnicek, we implement
+provenance analysis to show inputs involved in computation, but information from the execution
+trace collected by the Denicek substrate can also be used to provide a detailed
+explanation~\cite{perera-2012-functional} or to automatically generate linked visualizations~\cite{perera-2022-linked}.}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Concrete Programming}
 \label{sec:impl-copy}
 
-% TODO: Start Implementation section with some overall description of the "formative prototype"
-% TODO: In operations - better conflict resolution in terms of effect analysis
-% (collect selectors with structure/data effects; merging structure x data effects is OK,
-% but data x data or structure x structure is not); - Implement this
-% TODO: Add a section after Local-First Editing on the new conflict resolution (with demos)
-% TODO: In implementation use effects to better replay recorded edits
-% (push through new structural but not data edits)
+Abstraction is an essential feature of programming, but it has a high cognitive
+cost~\cite{blackwell-2002-attention}. Programming by demonstration (\S\ref{sec:impl-pbd})
+offers one way of reducing the cost. Another way of making programming more
+concrete~\cite{edwards-2004-example,smith-1975-pygmalion} is to support copying of functionality
+as in prototype-based object-oriented programming~\cite{randall-1995-self}.
+
+Webnicek supports a functionality akin to managed copy \&
+paste~\cite{edwards-2006-copypaste,edwards-2022-copypaste} for formulas. Rather than introducing
+abstractions (functions), users can copy and modify formulas to reuse them. However, when the user
+discovers an error in the original formula, Webnicek lets them use the merging mechanism to correct
+the error in the original formula and all its copies.
+
+The mechanism is shown in Figure~\ref{fig:copypaste}. The user copies an incorrect formula using
+the \ident{Copy} edit and then modifies it to use a different data source. They then navigate back
+in history to the point before the copying and create a temporary fork of the document. In the
+fork, they use \ident{RenameField} to switch the argument of the division. When they merge the
+temporary fork into the original document, the \emph{Apply to Newly Added} logic of the merge
+operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and applies them also
+to the copied formula. In other words, the \ident{Copy} edit of the Denicek substrate, alongside
+with the fact that formulas are ordinary document nodes, provide keys component for a
+straightforward implementation of the managed copy \& paste functionality.
+
+\keyideabox{\faClipboard}{Linked Editing}{Webnicek currently requires users to explicitly
+manipulate history to correct error across multiple code clones. Research on managing duplicated
+code resulted in multiple tools \cite{toomim-2004-linked,duala-ekoko-2008-clone} with dedicated
+user interface to manage clones. The Denicek substrate provides the underlying mechanism that
+could be used to implement a more user-friendly interface inspired by those systems.}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
+\vspace{-1em}
+\caption{Correcting an incorrect formula. The user uses Copy to create the second
+formula (1). They notice an error (2), go back in history to switch left and right argument (3),
+merge the change and re-evaluate both formulas (4).}
+\label{fig:copypaste}
+\vspace{-0.5em}
+\end{figure}
+
+% ==================================================================================================
 
@@
 
+\subsection{Design principles}
+\label{sec:discuss-principles}
+
+PRINCIPLES
+* merging of edits should not require looking at current document (non-computationality)
+
+\subsection{Design considerations}
+\label{sec:discuss-considerations}
+
+\subsubsection*{explicit indices}
+DIFFICULTIRES
+* if you are not able to refer to newly added items, you cannot reasonably do PBD
+  => need non-numerical indices
+
+unifying lists and records?
+
+\subsubsection*{mrdts and ordering}
 LISTS vs RECORDS
@@
 
-PRINCIPLES
-* merging of edits should not require looking at current document (non-computationality)
+% https://gowthamk.github.io/docs/mrdt.pdf page 11
+
+USE MRDTs for list order!
+
+\subsubsection*{two-way merging}
+divergence model of challenges paper - requires inverse operation for moving edits backwards
+
+\subsubsection*{keeping git tree}
+
+save graph of edits rather than list
+(would make replaying of saved edits easier because the histhash never changes)
+
+=> PBD - updates the recorded edits this would not be needed if the
+history was stored as a graph, akin to ordinary git merging rather than rebasing).
+
+\subsubsection*{Better apply to added}
+
+not so easy because of the non-computationality principle
+(we cannot know what will need to happen)
+
+how to transform appends?
+* apply to added? (does not work for copy)
+* copy to doc and edit? (good because edits are isolated from other things, but messy code!)
+(if we edit in doc, it works and we also do not need indices for ListAppend)
+* keep nested edits for append op? also does not work, because of the tricky case
+
+Tricky case is:
+- add new list item
+- then set its fields
+  (this breaks things because if we push "add" through structural edits, they will not affect the new values added in the second step)
+
+TODO: Delete ListAppend
+
+AppendFrom - hard to avoid! because recorded edits cannot know length
+(but having -1 in index would break too)
+
+\subsubsection*{Schema vs data}
+make schema explicit?
+
+\subsection{Future work}
+\label{sec:discuss-future}
+Although Denicek does not explicitly track document structure (or schema, or type), all documents
+have an implicit structure.
+=> use type system to check temporarily invalid state
+
+better effect system
+
+~
+
+~
+
+~
+
+% memory mapped graphics etc.
+
+\section{Case study}
+\label{sec:case}
+(data science environment)
+
+\section{Evaluation}
+\label{sec:eval}
+how well did Denicek work for implementing Datnicek
+
+\section{Discussion}
+
+\subsection{Heuristic evaluation}
+\subsection{Limitations}
+
+RANDOM FORMAL STUFF
+
+ohter formal things
+* non-conditionality of edits
+* show that they can transform document from any to any without removing/readding values
+
+PROPERTIES
+- can change any document to any other - sure, via remove add - but also more semantically
+(if it contained all values, can we do it without removing and adding them?)
+
+DEMOS
+
+// cf shadow dom v reactu
+// something like google forms!
+
 * two ways of pushing edits should be equivalent, i.e.
@@
 
-USE MRDTs for list order!
-
-DIFFICULTIRES
-* if you are not able to refer to newly added items, you cannot reasonably do PBD
-  => need non-numerical indices
-
-how to transform appends?
-* apply to added? (does not work for copy)
-* copy to doc and edit? (good because edits are isolated from other things, but messy code!)
-(if we edit in doc, it works and we also do not need indices for ListAppend)
-* keep nested edits for append op? also does not work, because of the tricky case
-
-Tricky case is:
-- add new list item
-- then set its fields
-  (this breaks things because if we push "add" through structural edits, they will not affect the new values added in the second step)
-
-TODO: Delete ListAppend
-
-// cf shadow dom v reactu
-// something like google forms!
-% https://gowthamk.github.io/docs/mrdt.pdf page 11
-
-TAG selectors - sound nice, but they break
-(you cannot tell if two selectors overlap e.g. when there is index >< tag)
-
-save graph of edits rather than list
-(would make replaying of saved edits easier because the histhash never changes)
-
-Although Denicek does not explicitly track document structure (or schema, or type), all documents
-have an implicit structure.
-
-c.f. notes
-
-memory mapped graphics etc.
-
-unifying lists and records?
-
-AppendFrom - hard to avoid! because recorded edits cannot know length
-(but having -1 in index would break too)
-
-divergence model of challenges paper - requires inverse operation for moving edits backwards
-
-alternative approache is discussed in \S\ref{sec:discuss}
-=> PBD - updates the recorded edits this would not be needed if the
-history was stored as a graph, akin to ordinary git merging rather than rebasing).
-
-\section{Case study}
-\label{sec:case}
-(data science environment)
-
-\section{Evaluation}
-\label{sec:eval}
-how well did Denicek work for implementing Datnicek
-
-\section{Discussion}
-
-\subsection{Heuristic evaluation}
-\subsection{Limitations}
-\subsection{Future work}
-\label{sec:discuss-future}
-use type system to check temporarily invalid state
-
-better effect system
-
-ohter formal things
-* non-conditionality of edits
-* show that they can transform document from any to any without removing/readding values
-
-PROPERTIES
-- can change any document to any other - sure, via remove add - but also more semantically
-(if it contained all values, can we do it without removing and adding them?)
-
 \bibliographystyle{plain}
@@
 
-The design the Denicek substrate, we identified five formative examples shown in
+The design the Denicek substrate, we identified six formative examples shown in
 Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
-(Todo and Counter apps) to problems posed as schema change challenges \cite{edwards-2025-schema}.
+(Todo and Counter apps) to cases from literature~\cite{edwards-2022-copypaste} and
+problems posed as schema change challenges \cite{edwards-2025-schema}.
 The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
@@
 \\[-1em]
-\extablecol{Counter}{\cite{kiss-2014-7guis}}
+\extablecol{Counter App}{\cite{kiss-2014-7guis}}
   {Counter with increment and decrement buttons.}
@@
   {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
-\extablecol{Todo}{\cite{osmani-2024-todomvc}}
-  {App with controls to add an item and remove all completed.}
+\extablecol{Todo App}{\cite{osmani-2024-todomvc}}
+  {Buttons to add an item and remove all completed.}
   {Adding an item must correctly merge with independently added functionality to \\[-0.2em]\quad\footnotesize
@@
   {Programming by Demonstration}
+\extablecol{Traffic Accidents}{\cite{edwards-2022-copypaste}}
+  {Compute statistics using two data sources.}
+  {Formula to compute statistics can be reused with a different data source;\\[-0.2em]\quad\footnotesize
+   error correction is propagated automtically to the copied version of the formula.}
+  {Concrete Programming, Incremental Recomputation}
 \hline

### 2024-12-21 paper.tex
Work on design considerations section

@@
 \section{Background}
+\label{sec:background}
+
 What is this about? end-user programming \cite{myers-2006-eup}
@@
 ~
+
+\cite{gobert-2023-lorgnette}
+
+~
+
+The Denicek substrate owes much to systems discussed in \S\ref{sec:background},
+especially Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}.
 
@@
 \label{fig:interactive}
-\vspace{-2em}
+\vspace{-1em}
 \end{figure}
@@
 
-The Denicek substrate can be used to generalize the interactions recorded through programming by
+Denicek can be used to generalize the interactions recorded through programming by
 demonstration. Our prototype illustrates this option in a limited way. As shown in
@@
 \includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
-\vspace{-1em}
+\vspace{-1.5em}
 \caption{Correcting an incorrect formula. The user uses Copy to create the second
@@
 
-\section{Design Discussion}
+\section{Design Considerations}
 \label{sec:discuss}
 
-\subsection{Design principles}
-\label{sec:discuss-principles}
-
-PRINCIPLES
-* merging of edits should not require looking at current document (non-computationality)
-
-\subsection{Design considerations}
-\label{sec:discuss-considerations}
-
-\subsubsection*{explicit indices}
-DIFFICULTIRES
-* if you are not able to refer to newly added items, you cannot reasonably do PBD
-  => need non-numerical indices
-
-unifying lists and records?
-
-\subsubsection*{mrdts and ordering}
-LISTS vs RECORDS
-- are they really different? also MRDT idea - needed for both
-
-% https://gowthamk.github.io/docs/mrdt.pdf page 11
-
-USE MRDTs for list order!
+The design of the Denicek substrate is the result of an iterative process in which we repeatedly
+adapted the Denicek design until we obtained a satisfactory solution for the six formative examples
+(\S\ref{app:examples}) in the Webnicek prototype. In this section, we document the design challenges,
+many of which have until now been personal knowledge of researchers working on related
+systems \cite{jakubovic-2022-ladder,edwards-2005-subtext,hall-2017-infra,omar-2021-livelits}.
+
+Two guiding principles for Denicek have been \emph{composability} and
+\emph{uniformity}~\cite{jakubovic-2023-techdims}. The design should cover a large number of
+cases using a small number of concepts. This is apparent in the design of the document
+structure (a node can be data, code or rich text) as well as in the role of edits
+(an edit can be a value or structure change, the result of user interaction or the
+result of evaluation).
+
+The inconvenience associated with composability and uniformity~\cite{jakubovic-2023-techdims} is not
+a problem for Denicek. We view Denicek as an underlying structure on top of which more
+convenient end-user programming systems can be built. The uniform design offers a degree of open-endedness,
+making it possible to use the basic structures of Denicek in new ways, not anticipated in this paper.
+
+\subsubsection*{Implicit vs. Explicit Structure}
+A key design choice for a substrate like Denicek is whether to track the document structure
+explicitly or make it implicit. In Denicek, we assume that elements of a list node have the same
+structure (i.e.~records should have the same fields, so that they can be targetted using the \ident{All}
+selector), but we do not enforce this. Often, the property is temporarily violated when adding a
+list item, but then restored once the item is created.
+
+Keeping the structure (type) explicit has theoretical appeal and it simplifies some
+aspects of the implementation. In particular, edits to document structure can be clearly separated
+from edits of document values. The same separation has disadvantages in that it complicates the
+edit language and it requires novel user interface. (In Webnicek, we can select all list items
+and edit them at once using the same mechanism as when editing individual list items. With explicit
+structure, changing a list structure has to be either a different kind of edit or possibly an edit
+of a virtual ``prototype'' element.)
+
+\subsubsection*{Merging and Document State}
+One specific design choice in Denicek is that the merge operation operates solely on two
+sequences of edits, but it does not need access to the current document state (or full edit
+histories). The reconcilliation of edits (\S\ref{sec:system-ops}) is defined for two individual
+edits $e_1$ and $e_2$.
+
+This limits the language of edits in a number of ways. The first consequence is that we cannot, in
+general, support conditional edits (edits that are applied only when a certain condition holds for
+the current document state). This is the case because the merging operation would not be able to
+determine whether the edit is applied (and whether its effects should take place). One consequence
+of this choice is that generalization of recorded interactions (\S\ref{sec:impl-interaction}) has to
+be done on a meta-level (by specifying how to invoke edits, rather than in the edits themselves).
+The second consequence of this design choice is that our langauge of selectors has to be abstract.
+We can target all list elements using the \ident{All} selector, but cannot convert between this
+representation and a representation that explicitly lists indices of all items currently in the list.
+
+\subsubsection*{Explicit List Indices}
+Denicek does not use implicit numerical indices for lists. When adding an item, the
+user (or the system built on top of Denicek) has to explicitly provide an index. This design
+is motivated by a typical programming by demonstration scenario where the user adds a
+new list item and then modifies it (adding a speaker or a Todo item). When using
+explicit indices, the edits following \ident{Append} can easily refer to the newly added
+item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
+indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}).
+
+Using implicit numerical indices is possible, but it requires treating all list-related edits
+as operations that transform document selectors. This means adding multiple new rules to
+Figure~\ref{fig:updates}, which specifies how edits transform references. Merging then has to
+modify indices when edits add, remove or reorder list items. Somewhat, counterintuitively, list
+order then becomes a part of document structure rather than a property of the list value.
+
+The problem could also be avoided by not letting the user modify a list item after adding it.
+(We can require new items to be constructed outside of the list and then added through a
+single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easier to
+implement, but it requires unintuitive user interaction and it also does not address all
+use cases for modifying list item via index.
+
+\subsubsection*{Ordering List Items and Record Fields}
+When list indices are not numerical and supplied explicitly, we need another mechanism for
+tracking order of list items. Denicek uses a data structure inspired by Mergeable Replicated
+Data Types \cite{kaki-2019-mrdts} and requires specifying the index of a preceding list item
+when appending to a list. (If two edits append to the same list independently with the same
+preceding item, the order is non-deterministic.)
+
+Note that Denicek uses the same mechanism for record fields. Keeping record fields ordered is
+desirable because Denicek documents map directly to HTML documents and so the order is visible
+to a user. Moreover, different ways of merging can result in different order of adding fields
+to a record. Making order explicit ensures those result in equivalent documents.
+
+Denicek still makes a distinction between records and lists, even if the two are technically
+extremely similar. The difference is conceptual. A list is expected to contain items of the
+same structure (that can be addressed using the \ident{All} selector), whereas record fields
+are expected to be of different types. If we unified lists and records, \ident{All} would have
+to be applicable to records too and inferring and checking document structure would be
+challenging.
 
@@
 
-\subsubsection*{Better apply to added}
-
-not so easy because of the non-computationality principle
-(we cannot know what will need to happen)
-
-how to transform appends?
-* apply to added? (does not work for copy)
-* copy to doc and edit? (good because edits are isolated from other things, but messy code!)
-(if we edit in doc, it works and we also do not need indices for ListAppend)
-* keep nested edits for append op? also does not work, because of the tricky case
-
-Tricky case is:
-- add new list item
-- then set its fields
-  (this breaks things because if we push "add" through structural edits, they will not affect the new values added in the second step)
-
-TODO: Delete ListAppend
-
-AppendFrom - hard to avoid! because recorded edits cannot know length
-(but having -1 in index would break too)
-
-\subsubsection*{Schema vs data}
-make schema explicit?
+ISSUES
+* if we merge a thing with saved-interactions with something, hashes will change!
+
+\subsubsection*{Fancy selectors}
+
+* many things with <tag> selectors currently do not work
+  (e.g. `matches` for highlighting) because if we collect path of a current node,
+  we collect indices and get /some/2/another - and cannot tell if this matches
+  /some/<li>/another - we'd have to collect more detailed path info!
+
+  * selectors - all for structure / index for data
+    (but it is useful to allow others...)
+    (multiselect also bad for checks!)
+
+\subsubsection*{Dependency tracking}
+
+NOTES
+type Edit =
+  { Kind : EditKind
+    Dependencies : Selectors list }  -- only needed for evaluated edits
+
+% ---------------------------------
+
+Maybe have 'enabled' for edits afterall?
+(we can merge with conflicts and disable some edits, but keep them in history for info)
+
+TODO - things to work on
+* "represent" edits somewhere in document as "library of functions"
+  and then call those from buttons (rather than embedding them directly)
+  allow some kind of abstraction (as in Histogram) to make them reusable
+* figure out how to do evaluation better
+  (based on the stored abstractions? but need to store provenance...)
+
+SEMANTIC CONDITIONS
+\url{https://www.youtube.com/watch?v=NBnc2ToS_j0}
+(has a section on this in background)
+
+INTERACTION
+* replay stored event handlers against the old version?
+  (this way, adding an item to a speakers list gets migrated \& adds a new table row!)
+* simlarly!! we need merge in order to apply edits to multiple targets
+  (when you remove all items in a list, the indices change)
+  (but I guess we should do this against version at the time of saving too....)
+
+[this \& evaluation = the unreasonable effectiveness of merging]
+
+Notes on storing and reusing edits
+* references need to be represented as references so that they get updated
+  (NO! not if we reply them against old version, which seems better - but there are 2 design choices)
+* how to apply them to multiple targets? use Move to update the selectors instead of
+  replacing the prefix manually
+
+CONDITIONALS
+https://toby.li/files/p311-radensky.pdf
 
@@
 
+IDEA: Type check edit groups to ensure they preserve structure but not individual edits eg when adding list item
+
 ~
@@
 \subsection{Heuristic evaluation}
+
+techdims
+
 \subsection{Limitations}
@@
 
-Maybe have 'enabled' for edits afterall?
-(we can merge with conflicts and disable some edits, but keep them in history for info)
-
-NOTES
-type Edit =
-  { Kind : EditKind
-    Dependencies : Selectors list }  -- only needed for evaluated edits
-
-VALUE vs STRUCTURE distinction
-* good in theory, nice for implementation
-* tricky to use! needs some assistance tools
-
-TODO - things to work on
-* "represent" edits somewhere in document as "library of functions"
-  and then call those from buttons (rather than embedding them directly)
-  allow some kind of abstraction (as in Histogram) to make them reusable
-* figure out how to do evaluation better
-  (based on the stored abstractions? but need to store provenance...)
-
-SEMANTIC CONDITIONS
-\url{https://www.youtube.com/watch?v=NBnc2ToS_j0}
-(has a section on this in background)
-
-SUBSTRATE DESIGN PROBLEMS
-* selectors - all for structure / index for data
-  (but it is useful to allow others...)
-  (multiselect also bad for checks!)
-* groups/conditions/preconditions
-  (c.f. email to jonathan)
-  tried conditions on edits; trying groups with check edits
-* what to do with "disabled edits"? for example when we remove all checked
-  (before, this created edit groups with "check" but if the check was false,
-  the group was ignored and this messed up merging - because we wouldn't know if the
-  edit had any effect or not)
-
-Evaluation
-* evaluated edits have to be migrated to the end
-  (if there are conflicts, they are dropped)
-  Think of this as maintaining a tree:
-
-  e3
-  |
-  e2   evaluated
-  |  /
-  e1
-  |
-  e0
-
-  this has to be serialized as e0 -> e1 -> e2 -> e3 -> evaluated
-
-  evaluated edits do not became part of the main history
-  but hang on the side
-
-ISSUES
-* if we merge a thing with saved-interactions with something, hashes will change!
-
-NOTES
-* ListAppendFrom - we need this, because we cannot encode this.
-* for records, we can RecordAdd(sel, fld, ..) @ Copy(sel @ [Field fld], src) but
-  this does not work for lists - because we do not know the index!
-  (and we cannot look into current document, because it will differ for saved-interactions)
-
-TODO
-* many things with <tag> selectors currently do not work
-  (e.g. `matches` for highlighting) because if we collect path of a current node,
-  we collect indices and get /some/2/another - and cannot tell if this matches
-  /some/<li>/another - we'd have to collect more detailed path info!
-
-INTERACTION
-* replay stored event handlers against the old version?
-  (this way, adding an item to a speakers list gets migrated \& adds a new table row!)
-* simlarly!! we need merge in order to apply edits to multiple targets
-  (when you remove all items in a list, the indices change)
-  (but I guess we should do this against version at the time of saving too....)
-
-[this \& evaluation = the unreasonable effectiveness of merging]
-
-Notes on storing and reusing edits
-* references need to be represented as references so that they get updated
-  (NO! not if we reply them against old version, which seems better - but there are 2 design choices)
-* how to apply them to multiple targets? use Move to update the selectors instead of
-  replacing the prefix manually
-
-IDEA: Type check edit groups to ensure they preserve structure but not individual edits eg when adding list item
-
-CONDITIONALS
-https://toby.li/files/p311-radensky.pdf
-
 REMAINING IMPLEMENTATION TODOs:

### 2024-12-24 paper.tex
more design considerations

@@
 cases using a small number of concepts. This is apparent in the design of the document
-structure (a node can be data, code or rich text) as well as in the role of edits
+structure (a node can represent data, code or rich text) as well as in the role of edits
 (an edit can be a value or structure change, the result of user interaction or the
@@
 
-Keeping the structure (type) explicit has theoretical appeal and it simplifies some
+Keeping the structure (type information) explicit has theoretical appeal and it simplifies some
 aspects of the implementation. In particular, edits to document structure can be clearly separated
 from edits of document values. The same separation has disadvantages in that it complicates the
-edit language and it requires novel user interface. (In Webnicek, we can select all list items
-and edit them at once using the same mechanism as when editing individual list items. With explicit
-structure, changing a list structure has to be either a different kind of edit or possibly an edit
-of a virtual ``prototype'' element.)
+edit language and it requires user interface that departs further from document editor.
+(In Webnicek, users can select all list items and edit them at once using the same mechanism as
+when editing individual list items. With explicit structure, changing a list structure has to be
+either a different kind of edit or possibly an edit of a virtual ``prototype'' element.)
 
@@
 
-This limits the language of edits in a number of ways. The first consequence is that we cannot, in
-general, support conditional edits (edits that are applied only when a certain condition holds for
-the current document state). This is the case because the merging operation would not be able to
-determine whether the edit is applied (and whether its effects should take place). One consequence
-of this choice is that generalization of recorded interactions (\S\ref{sec:impl-interaction}) has to
-be done on a meta-level (by specifying how to invoke edits, rather than in the edits themselves).
-The second consequence of this design choice is that our langauge of selectors has to be abstract.
-We can target all list elements using the \ident{All} selector, but cannot convert between this
-representation and a representation that explicitly lists indices of all items currently in the list.
+This design limits the possible language of edits. We cannot, in general, support
+conditional edits (that are applied only when a certain condition holds for
+the current document state). This is because the merging operation would not be able to
+determine whether the edit is applied (and whether its effects should~take~place).
+
+One consequence of the design choice is that the generalization of recorded interactions (\S\ref{sec:impl-interaction}) has to
+be done on a meta-level (by specifying how to invoke edits, rather than through the edits themselves).
+Another consequence is that our langauge of selectors has to be abstract.
+We can target all list elements using the \ident{All} selector, but cannot combine this
+representation with a representation that explicitly lists indices of all items currently in the list.
 
@@
 is motivated by a typical programming by demonstration scenario where the user adds a
-new list item and then modifies it (adding a speaker or a Todo item). When using
-explicit indices, the edits following \ident{Append} can easily refer to the newly added
+new list item and then modifies it (adding a speaker or a Todo item). When using explicit indices,
+the edits following \ident{Append} can use the index to refer to and modify the newly added
 item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
-indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}).
+indices of newly added items when replaying recorded interactions in \S\ref{sec:impl-pbd}).
 
@@
 Figure~\ref{fig:updates}, which specifies how edits transform references. Merging then has to
-modify indices when edits add, remove or reorder list items. Somewhat, counterintuitively, list
-order then becomes a part of document structure rather than a property of the list value.
-
-The problem could also be avoided by not letting the user modify a list item after adding it.
+modify indices when edits add, remove or reorder list items. List order then, somewhat confusingly,
+becomes a part of document structure rather than a property of the list value.
+
+The problem could be avoided by not letting the user modify a list item after adding it.
 (We can require new items to be constructed outside of the list and then added through a
-single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easier to
+single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easy to
 implement, but it requires unintuitive user interaction and it also does not address all
@@
 Data Types \cite{kaki-2019-mrdts} and requires specifying the index of a preceding list item
-when appending to a list. (If two edits append to the same list independently with the same
-preceding item, the order is non-deterministic.)
+when appending to a list. However, Webnicek supplies the index of preceding item automatically.
+(If two edits append to the same list independently with the same preceding item, the order
+is non-deterministic.)
 
@@
 desirable because Denicek documents map directly to HTML documents and so the order is visible
-to a user. Moreover, different ways of merging can result in different order of adding fields
+to a user. Moreover, different ways of merging edits can result in different order of adding fields
 to a record. Making order explicit ensures those result in equivalent documents.
 
-Denicek still makes a distinction between records and lists, even if the two are technically
-extremely similar. The difference is conceptual. A list is expected to contain items of the
+Denicek still makes a distinction between records and lists, even if the two are similar
+technically. The difference is conceptual. A list is expected to contain items of the
 same structure (that can be addressed using the \ident{All} selector), whereas record fields
@@
 
-\subsubsection*{two-way merging}
-divergence model of challenges paper - requires inverse operation for moving edits backwards
-
-\subsubsection*{keeping git tree}
-
-save graph of edits rather than list
-(would make replaying of saved edits easier because the histhash never changes)
-
-=> PBD - updates the recorded edits this would not be needed if the
-history was stored as a graph, akin to ordinary git merging rather than rebasing).
-
-ISSUES
-* if we merge a thing with saved-interactions with something, hashes will change!
-
-\subsubsection*{Fancy selectors}
-
-* many things with <tag> selectors currently do not work
-  (e.g. `matches` for highlighting) because if we collect path of a current node,
-  we collect indices and get /some/2/another - and cannot tell if this matches
-  /some/<li>/another - we'd have to collect more detailed path info!
-
-  * selectors - all for structure / index for data
-    (but it is useful to allow others...)
-    (multiselect also bad for checks!)
-
-\subsubsection*{Dependency tracking}
-
-NOTES
-type Edit =
-  { Kind : EditKind
-    Dependencies : Selectors list }  -- only needed for evaluated edits
+\subsubsection*{Convergence vs. Divergence of Document Variants}
+There are two ways of managing document variants in collaborative editing~\cite{edwards-2025-schema}.
+In the \emph{divergence} model, users can maintain their own variant indefinitely (share data,
+but use a different schema locally). In the \emph{convergence} model, users have to adopt all
+earlier changes in order to adopt a later change (accept new schema before new data).
+
+Denicek supports only the simpler \emph{convergence} model. To support the \emph{divergence}
+model, we would need an operation dual to our edit reconcilliation (given two subsequent edits,
+$e_1, e_2$, we want to generate $e_2'$ that has the same effect as $e_2$ but can occur before
+$e_1$). The pair of operations has been called \emph{project} and \emph{retract}
+\cite{edwards-2021-typed}. We choose not to support this as it would complicate the basic
+Denicek substrate structure.
+
+\subsubsection*{Linear History vs. Graph of Edits}
+Denicek keeps a linear history of edits. When merging edits, new edits have to be transformed and
+added to the top of the history. This model is akin to git rebase. An alternative representation
+would be to maintain a graph of edits akin to when using git merge. In this representation,
+we would need to keep parents of edits and special merge edits would have multiple parents.
+
+Maintaining a graph of edits would make recording of edits in programming by demonstration
+easier (\S\ref{sec:impl-pbd}) as we would not need to update recorded edits if they are
+transformed during merging. (Their hashes would remain the same.) However, supporting special
+merge edits and non-linear history would make the basic substrate more complex.
+
+\subsubsection*{Supporting Rich Selectors}
+Denicek supports only a limited set of selectors. An absolute reference can select a record
+field (\ident{Field}), specific list item (\ident{Index}) or all list items (\ident{All}).
+Once can imagine a range of other useful selectors, for example to select list items with a
+given tag or select items satisfying other conditions.
+
+We could support such selectors for those edits that do not affect the document
+structure. However, supporting such conditional selectors in general is incompatible with our
+design decision that merging should not depend on document state. (If an edit wraps the body of all
+{\small\Verb_<h3>_} elements, we need to know the document state in order to determine whether
+we need to modify the selector pointing to an item at a specific index.)
+
+\subsubsection*{Dependency Tracking}
+Recall that the incremental recomputation (\S\ref{sec:impl-incremental}) is based on the
+conflict detection mechanism of Denicek (\S\ref{sec:system-ops}). When pushing edit $e_2$ through
+$e_1$, a conflict is detected if they affect overlapping targets or if $e_1$ modifies a
+document node that is a source of $e_2$, typically, the source of a \ident{Copy} edit.
+
+For edits generated during formula evaluation (\S\ref{sec:impl-eval}), we need to track an
+additional kind of dependency. Evaluating a formula that adds two numbers from two document
+location results in an edit that sets the result of the formula to the sum of the two numbers.
+The edit contains the resulting number, but we also need to record that the edit has the two
+source locations as dependencies. For this reason, Denicek makes it possible to annotate edits
+with additional dependencies. One alternative would be to introduce a special \ident{Evaluate}
+edit that would identify an operation to compute and a list of arguments, specified as a list
+of references. We choose the former to keep the set of edits smaller.
 
@@
 
+\begin{comment}
 Maybe have 'enabled' for edits afterall?
@@
 
+\end{comment}
+
 \subsection{Future work}
@@
 \section{Discussion}
+
+\begin{comment}
 
@@
 (implies that no "extra" edits can be appended later)
+\end{comment}
 
@@
 
+\begin{comment}
 \newpage
@@
 \newpage
-
 \section{Introduction old}
@@
 
+\end{comment}
+
 \end{document}

### 2025-01-13 paper.tex
Notes and suggestions

--- 2024-12-21 paper.tex
@@
 \newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}
+\newcommand{\note}[1]{\textcolor{red}{#1}}
 
@@
 
-We contribute Denicek, a computational substrate that reduces the complexity of
+We contribute Denicek, a system that reduces the complexity of
 implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
 that construct and transform a document consisting of data and formulas. Denicek provides three
-primitive operations on series of edits, application of edits, merging of histories and conflict resolution.
-We show that the three operations form the backbone of a direct implementation of a range of
+primitive operations on series of edits: application of edits, merging of histories, and conflict resolution.
+We show how these three operations can be composed to implement a range of
 programming experiences.
+We call Denicek a \emph{computational substrate} because it serves as a shared platform for such varied programming experiences.
 
@@
 design of rich, interactive end-user programming systems.
+
+\note{This is a very promising paper.
+It's so much simpler and more complete than my stuff.
+I need to consider some of the different choices you've made.
+I think I am finally getting what your code execution model is.
+I hope the comments below are helpful.}
 \end{abstract}
@@
 
-A computational substrate defines the basic structures using which programs are constructed, how
-the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
+A computational substrate defines the basic structures with which programs are constructed, how
+the progam state is represented, and how it is changed \cite{jakubovic-2022-ladder}. The choice of
 a substrate affects what programming experiences can be readily supported by a system. For example,
 object-oriented programming has been historically linked to graphical user
-interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
+interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
 a ``language laboratory'' \cite{steele-1993-lisp}.
@@
 substrate. However, a suitable program representation and operations for working with programs
-provided by the substrate can eliminate much of the complexity of implementing interesting
+provided by the substrate can reduce the complexity of implementing interesting
 programming experiences. For example, the reflective capabilities of Smalltalk make it easy
@@
 
-What would be the ideal programming substrate for rich, interactive end-user programming systems?
-In this paper, we aim to design a programming substrate that makes it easy to develop a range of
-programming experiences \cite{myers-2006-eup}:
+In this paper, we aim to design a programming substrate that makes it easier to develop a range of
+rich, interactive programming experiences \cite{myers-2006-eup}:
 
@@
 \item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
-  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
+  modify a single document and automatically merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
 \item \emph{Programming by Demonstration.} Allow users to construct simple programs by
@@
   depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
+  \note{cite classic incremental computation literature}
 \item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
@@
 \item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
-  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
+  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}. \note{find more references}
 \end{itemize}
@@
 manipulate the document structure. A user-interface may provide a specialized editor, but still
-trigger the pritive edits behind the scenes. However, interacting with elements in the document,
+trigger the primitive edits behind the scenes. However, interacting with elements in the document,
 such as a entering text in a textbox, also generates a document edit that can be merged or checked
@@
   document representation, edit operations and the key three operations.
-\item We explain how a wide range of end-user programming experiences can be readily implemented using
+\item We explain how a wide range of end-user programming experiences can be implemented using
   the substrate (\S\ref{sec:impl}) in the context of a simple web-based prototype (Webnicek).
@@
   shows that the desired functionality requires a careful choice among interconnected design options.
+  \note{We also identify problems requiring future work.}
 \item We develop an end-user data exploration system (Datnicek) on top of Denicek (\S\ref{sec:case})
@@
 
-\subsubsection*{\circled{C} Saving the Interaction} After adding the speaker from the textbox,
+\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker from the textbox,
 the user opens history view, selects edits that added the speaker and saves those as the
@@
 reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
-Merging is asymmetic, but the order does not matter in the above scenario. The same merging
+Merging is asymmetric, but the order does not matter in the above scenario. The same merging
 operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}
@@
 A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
-References to document locations can be relative or absolute. Both kinds can appear in a
+Non-leaf nodes are records or lists.
+Both of these are named with a \emph{tag} as in HTML.
+Record fields have \note{unique?} names.
+Lists contain elements of the same type and assign a unique index to each element \note{It seems that these indexes are meant to be unique IDs, in which case it would be clearer to call them that.}
+Leaf nodes can be primitive values (strings and numbers) or references to a location in the document tree.
+References can be relative or absolute. Both kinds can appear in a
 document (reference nodes); absolute references are used in edits (to denote a target~node).
+\note{Why are relative references needed? I only use absolute references.}
 
@@
 
-Renaming a field or wrapping a node require corresponding change to references in the document
+Renaming a field or wrapping a node converts any references into that location
 (\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
@@
 point to the target of the copying (a reference cannot refer to two locations).
+\note{why not have distinct Copy and Move operations to resolve this?}
 
@@
 construction of a new list item (checking such edits is discussed in \S\ref{sec:discuss-future}).
+\note{Wouldn't this leave dangling references?
+Wouldn't it be better to evolve the references as these edits occur?} \note{Insertion into a list can default the contents based on the type to avoid such a transient type error.}
 
 Documents can also contain values that can be of multiple different kinds (i.e., a union type).
-In such case, references should not be updated when the kind of the value changes. For example,
+\note{This is surprsing since there has been no prior mention of union types, and lists are homogeneous.}
+In such cases, references should not be updated when the kind of the value changes. For example,
 a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
 involves wrapping the formula node, but this should not affect references to the formula.
-To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits}, below)
+To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits})
 have a \emph{reference behavior} field that can be set to disable automatic reference
 updating. This annotation is required when the target contains the \emph{Index} selector.
-
-\subsubsection*{No Conditional Edits.}
-An important aspect of the design is that the effect an edit has on references inside
-the document does not depend on the current value of the document. This makes it possible to
+\note{I think it would be desireable to avoid all these complications. References should be automatically maintained so they ``just work''.}
+\note{I believe that evaluation should not change the type/shape of the document.
+The space for evaluation results should be be preallocated in some sense, either as extra structure in a formula, or as metadata attached to the tree.
+That said, I've tried various alternative approaches to this in Subtext without ever settling on a preferred solution.}
+
+\subsubsection*{Edits are ``static''.}
+An important principle is that the effect an edit has on references inside
+the document does not depend on the current value of the document, only its type. This makes it possible to
 define merging solely in terms of edits, without reference to current document state.
@@
 identify a common shared part of the history during conflict resolution and merging.
+\note{what exactly does parent mean here? History prefix?}
 
 \subsubsection*{Applying Edits}
-When applying an edit, Denicek locates the target node and transform it according to the edit. If
+When applying an edit, Denicek locates the target node and transforms it according to the edit. If
 the edit can affect references in the document (Fig~\ref{fig:edits}, below), Denicek updates
@@
 
-The key operation that enables such reconcilliation takes two individual edits that occurred
-indpendently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
+\note{It might help to define a function signature for merging to make this clearer: $\mathcal{M}(P, L, R) = M$ or $\mathcal{M}_P(L, R) = M$.}
+\note{Merging is often thought of as symmetric, perhaps choose a name that is clearly asymmetric, like rebase, translate, or project.}
+
+The key operation that enables such reconciliation takes two individual edits that occurred
+independently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
 can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
-There are two aspects of such reconcilliation:
+There are two aspects of such reconciliation:
 
@@
 \end{enumerate}
+
+\note{I believe that in general a ``backwards'' form of merge is also needed, as in my \emph{retract} and the OT \emph{Exclusion Transform}.
+A regular merge takes forking edits and composes them.
+The backwards form converts composed edits into forking edits (and can be combined with a forward merge to recompose in the opposite order).
+This is needed for cherry picking and reverts, however you don't seem to need that for this paper.}
+
+\note{Related question: how do you merge an Append(target, index, after, node) with DeleteItem(target, after)?
+A famous counterexample in OT was that two insertions adjacent to a deletion could flip their location if the edits were reordered.
+I'll dig up the details of this counterexample for you.
+Some systems handle this by leaving behind a ``tombstone'' on delete to remember the position of the deleted item.
+I have a novel solution but I'm probably going to go back to tombstones anyway because they are simpler and handy in the diff UI.}
 
@@
   {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
-\eftablecol{TagEffect}{Modifies the tage of the target node.}
+\eftablecol{TagEffect}{Modifies the tag of the target node.}
   {\ident{UpdateTag}}
@@
 implement incremental evaluation (\S\ref{sec:impl-incremental}).
+Our treatment of conflicts necessarily differs in this respect from replication algorithms like Operational Transformation~\cite{OT} and Convergent Replicated DataTypes~\cite{CRDT} where conflicts are arbitrarily but deterministically resolved.
 
@@
 
+\note{The effect-style conflict detection is a nice idea! Removing conflicts before merging is also an interesting idea I hadn't considered. That might allow me to simplify my transformation rules. }
+
 % ==================================================================================================
@@
 The key claim of this paper is that the Denicek computational substrate
-makes it easy to support a range of experiences that make programming more concrete,
-collaborative and interactive.
-
-In this section, we describe how Webnicek uses the substrate
-to support local-first collaboration (\S\ref{sec:impl-collab}), programming by
-demonstration (\S\ref{sec:impl-pbd}, \S\ref{sec:impl-interaction}), incremental recomputation
-(\S\ref{sec:impl-eval}, \S\ref{sec:impl-incremental}), schema change control (\S\ref{sec:impl-schema}),
-end-user debugging via provenance tracking (\S\ref{sec:impl-provenance})
-and concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
+suppports a range of experiences that make programming more concrete,
+collaborative, and interactive.
+In particular, Denicek's ability to merge edits is a unifying mechanism that provides essential features across the range.
+We demonstrate this claim with the Webnicek web-based programming system, which uses Denicek to support:
+\begin{enumerate}
+\item Local-first collaboration (\S\ref{sec:impl-collab})
+\item Programming By Demonstration (\S\ref{sec:impl-pbd}\S\ref{sec:impl-interaction})
+\item Incremental recomputation (\S\ref{sec:impl-eval}, \S\ref{sec:impl-incremental})
+\item Schema change control (\S\ref{sec:impl-schema})
+\item End-user debugging via provenance tracking (\S\ref{sec:impl-provenance})
+\item Concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
+\end{enumerate}
 
@@
 Recall that merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
-can be obtained either by appending $C'$ (produced by the edit reconcilliation operation)
+can be obtained either by appending $C'$ (produced by the edit reconciliation operation)
 on top of $A,B$ or by appending $B'$ on top of $A,C$. According to our effect analysis, the two
@@
 
+\note{To guarantee that replicas converge it is necessary to have a global order on merges.
+OT and CRDT algorithms use timestamps combined with a unique ID for each client to establish this global order without a centralized coordinator.
+We can do the same.
+This means you never know if your changes will dominate or whether they will be eventually overriden but people seem to be willing to live with it.
+Personally, I'm a skeptical: I expect that like with NoSQL people will eventually realize this sucks and go back to centralized systems.}
+
 % --------------------------------------------------------------------------------------------------
@@
 
-Second, to replay edits, Webniceks does not append the recorded edits on top of the current history.
+Second, to replay edits, Webnicek does not append the recorded edits on top of the current history.
 When saving edits, it stores the hash of the history at the time of saving. When replaying,
@@
 Webnicek also accounts for the case where edits recorded in the document are themselves
-transformed (when they are reconcilliated with other edits during merging). In this case, Webnicek
+transformed (when they are reconciliated with other edits during merging). In this case, Webnicek
 updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).
+
+Simply replaying recorded edits provides limited programming capabilities.
+We discuss in the next section how recorded edits can themselves be edited by the user to generalize and conditionalize their effects.
+More general and conventional programming constructs are discussed in \S\ref{sec:discuss}.
 
@@
 
+\note{The above paragraph was surprising to me and I'm not sure I fully understand it even after staring at the figure for a while.
+Edits are recorded as an AST using x-tags to hide them in the UI?
+That became clear only after reading the next section.
+Our audience might be a lot happier if we used a textual grammar.
+Earlier it was said that edits can't be conditional but then a conditional predicate appears here.
+This is a special feature of button nodes?
+Or event handler nodes?
+This starts to resemble Angular and similar template languages (which I find hard to read).}
+
 \keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
@@
 what can be expressed using document edits. Formulas do not transform the document and their
-results are transient, although their evaluation also leverages the substrate operations,
-namely merging of edit histories.
-
+results are transient.
+
+Formula evaluation leverages Denicek's ability to merge edit histories.
 As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
@@
 rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
-edits and the Denicek substrate treates them as stnadard nodes.
-
-To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
+edits and the Denicek substrate treats them as standard nodes.
+
+To evaluate formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
 record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
@@
 
+\note{As I've said before, I'd prefer to store the result of the formula inside the <x-formula> record instead of wrapping it on evaluation.
+This avoids changing the shape of the document on evaluation, which I recall added some complexity earlier.}
+
 The Counter App example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
@@
 wrap the current counter value in a formula that adds or subtracts 1.
+\note{I'd like to see the code for the increment button.}
+
+\note{OK maybe I'm finally getting it.
+You aren't mutating the value of the counter, you're wrapping the value in an increment formula.
+I never got that in our previous discussions.
+Presumably that evaluation is now frozen so it serves as an execution trace.
+This is a lot different from my approach where a formula stays unchanged except for the results.
+One advantage is that fields contain a record of their computed mutations whereas in my approach that is recorded in the past state of the doc whence the mutation occured.}
 
@@
 is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
-way, edit them and merge them with other changes.}
+way, edit them and merge them with other changes. \note{Some readers may disagree that it is easy: AST editing is controversial. }}
 
@@
 
-As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental re-computa\-tion.
+As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental recomputation.
 In the example, the cost of speaker travel depends on the number of speakers, but the cost
-of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
-former.
-
-The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
+of refreshments depends only on two constants. The edit that adds a speaker invalidates only the former computation.
+
+The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. \note{This figure is out of order.} When the formulas are evaluated,
 Webnicek generates \emph{evaluated edits} that are appended to the top of the history.
+\note{What exactly are these evaluated edits?}
 When subsequent edits are made, they are appended after all non-evaluated edits and the
@@
 and use incremental recomputation for performance~reasons.}
+
+\note{I'm confused.
+Changing the value of a location conflicts with an edit that evaluates a formula containing a reference to that location?
+And these evaluation edits are second-class edits that get retroactively dropped to ``unevaluate'' the formula?
+I'd prefer what you suggest above: formulas are seemingly globally re-evaluated after every change, with incremental computation a hidden perf optimization.
+}
+
+\note{An important property of incremental computation systems is avoiding redundant execution.
+Usually this involves a topological sort of the computation dependency graph.
+We may need to have a story about that to avoid hate from reviewer 2.
+Personally, I think incremental recomputation is about performance, not the programming experience, so it isn't that important for this paper.
+}
 
@@
 
+\note{Nice and simple! My approach tried to be a lot fancier but ended up being too heavywieght. }
+
 \keyideabox{\faClipboard}{Linked Editing}{Webnicek currently requires users to explicitly
@@
 sequences of edits, but it does not need access to the current document state (or full edit
-histories). The reconcilliation of edits (\S\ref{sec:system-ops}) is defined for two individual
+histories). The reconciliation of edits (\S\ref{sec:system-ops}) is defined for two individual
 edits $e_1$ and $e_2$.
@@
 item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
-indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}).
+indices of newly added items in recorded interactions in \S\ref{sec:impl-pbd}). \note{I don't think that was explained back there.}
 
@@
 
+\note{I'm still unclear whether your list indices are ordinal numbers like in an array or are unique IDs.
+Your usage of implicit/explicit isn't obvious to me.
+The above paragraph seems to imply they are IDs.
+Note that automatically generating unique IDs in code execution can lead to nondeterminism.
+If you are using IDs, do they affect equality of lists?}
+
 The problem could also be avoided by not letting the user modify a list item after adding it.
@@
 implement, but it requires unintuitive user interaction and it also does not address all
-use cases for modifying list item via index.
+use cases for modifying list item via index. \note{I'm still confused, but this sounds like it really isn't a solution afterall.}
 
@@
 when appending to a list. (If two edits append to the same list independently with the same
-preceding item, the order is non-deterministic.)
+preceding item, the order is non-deterministic.) \note{Why is this needed? I let the order of operations determine the list order. }
 
@@
   (this way, adding an item to a speakers list gets migrated \& adds a new table row!)
-* simlarly!! we need merge in order to apply edits to multiple targets
+* similarly!! we need merge in order to apply edits to multiple targets
   (when you remove all items in a list, the indices change)
@@
 
-ohter formal things
+other formal things
 * non-conditionality of edits
@@
 
-// cf shadow dom v reactu
+// cf shadow dom v react
 // something like google forms!
@@
   {Formula to compute statistics can be reused with a different data source;\\[-0.2em]\quad\footnotesize
-   error correction is propagated automtically to the copied version of the formula.}
+   error correction is propagated automatically to the copied version of the formula.}
   {Concrete Programming, Incremental Recomputation}
@@
 
-% lcoal first
+% local first
 \cite{kleppmann-2019-local,klokmose-2024-mywebstrates}
@@
     forms (allow partial edits only / def by selector?)
-* scehma change - change data \& code accordingly
+* schema change - change data \& code accordingly
 * everything is an edit
@@
 * augmenters
-  - cf. bonnie nardi (calls them something else - Jonathan says)
+  - cf. bonnie Nardi (calls them something else - Jonathan says)
+  \note{data detectors https://dl.acm.org/doi/10.1145/272287.272331}
+  \note{also see Potluck https://www.inkandswitch.com/potluck/}
   - add programming by demonstration data wrangling gui to table (trigger interactions)
@@
 append /speakers/body \#hamilton <li>
-[dtto]
+[ditto]
 addfld /speakers/body\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
 
->>> retag /spekers/body <tbody>
+>>> retag /speakers/body <tbody>
 

### 2025-03-23 paper.tex
Start small tweaks

--- 2024-12-24 paper.tex
@@
 \begin{abstract}
-User-centric programming research gave rise to a wide range of compelling programming experiences,
+User-centric programming research gave rise to a variety of compelling programming experiences,
 including local-first collaborative editing, programming by demonstration, incremental
@@
 Those experiences advance the state of the art of end-user programming, but they are hard to
-implement on the basis of established programming language and system paradigms.
-
-We contribute Denicek, a computational substrate that reduces the complexity of
-implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
-that construct and transform a document consisting of data and formulas. Denicek provides three
-primitive operations on series of edits, application of edits, merging of histories and conflict resolution.
-We show that the three operations form the backbone of a direct implementation of a range of
-programming experiences.
+implement on the basis of established programming language and system.
+
+We contribute Denicek, a computational substrate that makes it easier to
+implement the aforementioned programming experiences. Denicek represents a program as a series of edits
+that construct and transform a document consisting of data and formulas. It provides three
+core operations on edit histories: (i) edit application, (ii) merging of histories and (iii) conflict resolution.
+We show that these operations serve as fundamental building blocks for a direct implementation
+of a range of programming experiences.
 
@@
 the implementation of the programming experiences listed above. To evaluate the proposed
-architecture, we use Denicek as the basis of a simple end-user data exploration environment.
-The case study shows that the Denicek computational substrate provides an appealing basis for the
+architecture, we use Denicek as the basis of an interactive data science notebook.
+The case study shows that the Denicek computational substrate provides a robust basis for the
 design of rich, interactive end-user programming systems.
@@
 
-A computational substrate defines the basic structures using which programs are constructed, how
-the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
+A computational substrate defines the structures from which programs are constructed, how
+the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
 a substrate affects what programming experiences can be readily supported by a system. For example,
@@
 In principle, any programming experience can be developed on top of any computational
-substrate. However, a suitable program representation and operations for working with programs
-provided by the substrate can eliminate much of the complexity of implementing interesting
+substrate. However, a suitable program representation can eliminate much of the complexity of implementing interesting
 programming experiences. For example, the reflective capabilities of Smalltalk make it easy
-to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
+to build rich debugging tools \cite{rauch-2019-babylonian} that are difficult to implement
 for C/C++~\cite{kell-2018-unix,kell-2024-debugging}.
@@
 What would be the ideal programming substrate for rich, interactive end-user programming systems?
-In this paper, we aim to design a programming substrate that makes it easy to develop a range of
-programming experiences \cite{myers-2006-eup}:
+In this paper, we aim to design a programming substrate that makes it easy to develop diverse
+programming experiences \cite{myers-2006-eup} including:
 
@@
 
+NICE THINGS IN DATNICEK
+* merging of code edits (e.g. change parameter merges with adding a call to chain)
+* remove cell to which there are refs - does the susbtrate catch this?
+
+RELATED WORK
+
+TIJS
+https://homepages.cwi.nl/~storm/publications/icsews13live-p17-p-16201-preprint.pdf
+
+via Pierre
+https://pijul.org/manual/why_pijul.html
+
+Geoffrey Litt
+https://dl.acm.org/doi/pdf/10.1145/3426428.3426914
+https://dl.acm.org/doi/pdf/10.1145/3397537.3397541
+
+Programming By Example: Intelligence in Demonstrational Interfaces
+https://cacm.acm.org/research/programming-by-example-intelligence-in-demonstrational-interfaces/
+
+Rapido: Prototyping Interactive AR Experiences through
+Programming by Demonstration
+https://dl.acm.org/doi/pdf/10.1145/3472749.3474774
+
+openrefine
+- facet (group by) - then change the key to normalize
+  (https://www.youtube.com/watch?v=yTJ6x6zEQmI 51:40)
+
+Don't Step on My Toes: Resolving Editing Conflicts in Real-Time
+Collaboration in Computational Notebooks
+https://arxiv.org/pdf/2404.04695
+
+AI assistants in notebook interfaces
+https://dl.acm.org/doi/pdf/10.1145/3544548.3580940
+
+Unravel - understand pipelines
+https://dl.acm.org/doi/pdf/10.1145/3472749.3474744
+
+WREX - nicest tool
+https://dl.acm.org/doi/pdf/10.1145/3313831.3376442
+
 \begin{comment}

### 2025-03-28 paper.tex
Edit design considerations

@@
 Those experiences advance the state of the art of end-user programming, but they are hard to
-implement on the basis of established programming language and system.
-
-We contribute Denicek, a computational substrate that makes it easier to
-implement the aforementioned programming experiences. Denicek represents a program as a series of edits
+implement on the basis of established programming languages and system.
+
+We contribute Denicek, a computational substrate that simplifies the implementation of the
+above programming experiences. Denicek represents a program as a series of edits
 that construct and transform a document consisting of data and formulas. It provides three
-core operations on edit histories: (i) edit application, (ii) merging of histories and (iii) conflict resolution.
-We show that these operations serve as fundamental building blocks for a direct implementation
-of a range of programming experiences.
-
-We discuss the architecture of Denicek, document key design considerations and elaborate
-the implementation of the programming experiences listed above. To evaluate the proposed
-architecture, we use Denicek as the basis of an interactive data science notebook.
-The case study shows that the Denicek computational substrate provides a robust basis for the
+operations on edit histories: edit application, merging of histories and conflict resolution.
+By composing these operations, we can  easily implement a range of programming experiences.
+
+We present the architecture of Denicek, discuss key design considerations and elaborate
+the implementation of a variety of programming experiences. To evaluate the proposed
+substrate, we use Denicek to develop an innovative interactive data science notebook system.
+The case study shows that the Denicek computational substrate provides a suitable basis for the
 design of rich, interactive end-user programming systems.
@@
 
-A computational substrate defines the structures from which programs are constructed, how
+A computational substrate defines the structures with which programs are constructed, how
 the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
@@
 object-oriented programming has been historically linked to graphical user
-interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
-a ``language laboratory'' \cite{steele-1993-lisp}.
+interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
+a language laboratory \cite{steele-1993-lisp}.
 
@@
 
-What would be the ideal programming substrate for rich, interactive end-user programming systems?
-In this paper, we aim to design a programming substrate that makes it easy to develop diverse
-programming experiences \cite{myers-2006-eup} including:
+In this paper, we describe a programming substrate that makes it easy to develop diverse
+programming experiences \cite{myers-2006-eup}. Those include:
 
@@
 \item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
-  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
+  modify a single document and automatically merge their changes, preferrably without a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
 \item \emph{Programming by Demonstration.} Allow users to construct simple programs by
-  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
+  enacting the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
 \item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
-  depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
+  depends on it are invalidated and, possibly, automatically recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
 \item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
@@
 \subsubsection*{Two-Stage Methodology}
-Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
+The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
 To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
-tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
-Using those examples, we co-design the Denicek substrate and Webnicek (\S\ref{sec:walk}),
-a simple web-based end-user programming environment based on the substrate. Although Webnicek can
-be used to complete end-user programming tasks, it is optimized for testing the underlying
+tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
+Using those examples, we co-design the Denicek substrate and Webnicek~(\S\ref{sec:walk}),
+a simple web-based end-user programming environment built directly on top of the substrate. Although Webnicek can
+be used to complete end-user programming tasks, it is optimized for developing the underlying
 substrate rather than for usability.
 
-To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek (\S\ref{sec:case}), an end-user data
-exploration environment inspired by existing systems~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
+To \emph{evaluate} Denicek, we use the substrate to develop Datnicek (\S\ref{sec:case}), an interactive data
+science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
 We use the second stage to evaluate suitability of the Denicek substrate for the development of
@@
 document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
-static content. Second, Denicek does not store the document tree itself, but instead, it maintains
+static content. Second, Denicek does not store the document tree itself, but instead, maintains
 a sequence of edit operations through which the tree was constructed and transformed.
@@
 The substrate then provides three primitive operations for working with sequences of edits.
-First, it can apply a series of edits to reconstruct a document. Second, it can merge two
+First, it can apply a series of edits to reconstruct the document. Second, it can merge two
 diverging edit histories. Finally, it can detect conflicts when merging histories and, for
@@
 
-The key insight that we present in this paper is that many of the programming experiences can
-be obtained by leveraging the uniform document representation alongside with a suitable combination
-of the three primitive operations.
-
-Changing data and formulas is done using the same primitive edit operations to
+The key insight that we present in this paper is that many compelling programming experiences can
+be implemented by leveraging the uniform document representation alongside with a suitable
+composition of the three primitive operations.
+
+Changing data and formulas is done using the same primitive edit operations that
 manipulate the document structure. A user-interface may provide a specialized editor, but still
-trigger the pritive edits behind the scenes. However, interacting with elements in the document,
-such as a entering text in a textbox, also generates a document edit that can be merged or checked
+trigger the primitive edits behind the scenes. However, interacting with elements in the document,
+such as a entering text in a textbox can also generate a document edit that can be merged or checked
 for conflicts (\S\ref{sec:impl-interaction}).
@@
 such recorded edits is implemented using the merging operation, which means that recorded operations
-continue working even if the document structure later changes. Moreover, structural changes to the
-document can be done independently and then merged (\S\ref{sec:impl-collab}). Evaluation of formulas
+continue working even if the document structure later evolves. Moreover, structural changes to the
+document can be merged with concurrent data edits  (\S\ref{sec:impl-collab}). Evaluation of formulas
 also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
@@
 \item We present the Denicek substrate (\S\ref{sec:system}) and provide a detailed description of its
-  document representation, edit operations and the key three operations.
-\item We explain how a wide range of end-user programming experiences can be readily implemented using
-  the substrate (\S\ref{sec:impl}) in the context of a simple web-based prototype (Webnicek).
-\item We document important design decisions and alternatives (\S\ref{sec:discuss}). The analysis
-  shows that the desired functionality requires a careful choice among interconnected design options.
-\item We develop an end-user data exploration system (Datnicek) on top of Denicek (\S\ref{sec:case})
-  and evaluate the extent to which the Denicek substrate enables such development (\S\ref{sec:eval}).
+  document representation, edit operations and the key three operations for working with edit histories.
+\item We illustrate a range of end-user programming experiences supported in Webnicek, a simple
+  web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
+  implemented using the Denicek substrate (\S\ref{sec:impl}).
+\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
+  The analysis shows that the desired functionality requires a careful choice among interconnected
+  design options.
+\item To evaluate how Denicek simplifies the development of program\-ming systems, we
+  build an innovative data science notebook Datnicek (\S\ref{sec:case})
+  and assess its impementation complexity~(\S\ref{sec:eval}).
 \end{itemize}
@@
 To enable others build on top of Denicek, we share our compact and documented
-source code at: \url{https://github.com/d3sprog/denicek}
+source code at: \url{https://github.com/removed/for/review} %d3sprog/denicek
 
@@
 \label{sec:background}
-
-What is this about? end-user programming \cite{myers-2006-eup}
-but also programming systems \cite{jakubovic-2023-techdims},
-Live \cite{rein-2019-live}, Interactive etc. - perhaps for a lack of better term use end-user
-(also notebooks)
-
-explain computational substrate
-Substrate defined by \cite{jakubovic-2022-ladder}
-
-citizen programmers? Maybe do not talk about ``programs'' because this is more documents with
-formulas - not that fancy programs. Computational documents?
-
-programming by demonstration
-\cite{leiva-2021-rapido,cypher-1993-pbd}
-\cite{chen-2023-miwa}
-
-local first
-\cite{kleppmann-2019-local,klokmose-2024-mywebstrates}
-
-provenance/debugging
-\cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
-\cite{ricciotti-2017-imperative,perera-2012-functional}
-\cite{perera-2022-linked}
-
-end-user debugging
-\cite{kissinger-2006-debugging}
-
-live/incremental recomputation
-\cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
-
-visualizing results of execution
-projection boxes \cite{lerner-2020-boxes}
-
-livelits (provide editors)
-\cite{omar-2021-livelits}
-
-edit calculus (based on tree navigation though)
-\cite{omar-2017-hazelnut}
-
-structure editing see \cite{beckman-2023-sandblocks}
-
-Schema evolution
-% our paper; see https://openproceedings.org/2023/conf/edbt/paper-160.pdf - good refs in 2.2
-
-Also on substrate see
-% https://x.com/jonathoda/status/1185888711210389504
-
-% also cool: https://arxiv.org/pdf/2303.06777
-
-Webnicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
-document visible to the user, although parts can be collapsible or hidden using CSS.
-
-Merging - c.f. mywebstrates - we need more than Automerge (cite)
-which has a bunch of pre-built structures
-
-~
-
-~
-
-\cite{gobert-2023-lorgnette}
-
-~
-
-The Denicek substrate owes much to systems discussed in \S\ref{sec:background},
-especially Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}.
-
-~
+Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
+notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness, interactivity
+and richness \cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
+or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
+We focus on systems that allow gradual progression from a user to a developer, as envisioned
+in Smalltalk~\cite{reenskaug-1981-byte} and use the term end-user programming loosely to refer to a
+part of this spectrum, also including spreadsheet systems and notebooks for data science.
+We first review related systems and their technical design (\S\ref{sec:background-sys})
+before discussing programming experiences (\S\ref{sec:background-exp}).
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Programming Systems and Substrates}
+\label{sec:background-sys}
+
+\subsubsection*{Programming Systems and Substrates}
+A number of systems illustrate the qualities Denicek aims to support and have a related internal
+structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
+use structured document-based program representation and provide some of the desired programming
+experiences on top of this representation. Many of those design ideas can be traced back to
+Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
+sees is all there is) that we follow in Webnicek.
+
+To indicate that Denicek is intended as an underlying infrastructure on top of which programming
+systems can be built, we use the term \emph{computational substrate}, which also dates back to
+Boxer~\cite{disessa-1995-epistemology} and is closely related to the notion of dynamic media
+introduced by Kay and Goldberg \cite{kay-1977-media}. Webstrates \cite{klokmose-2015-webstrates}
+is a substrate based on synchronization of documents (but without edit histories) that has been
+use as the basis for multiple programming systems \cite{radle-2017-codestrates,borowski-2022-varv}.
+
+\subsubsection*{Edit Histories and Merging}
+Manipulating programs through semantically meaningful edits is a technique used by structure
+editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The langauge of
+edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
+edits have also been used as the basis of live programming environments \cite{storm-2013-deltas}.
+
+Merging of edits is most frequently done in version control systems such as git. Systems based
+on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
+by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
+programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
+structure similar to Pijul, as the basis for a collaborative structure editor.
+
+More generally, merging of edits can be based on the operational transform (OT) approach
+\cite{davis-2002-otsgml}, where edit conflicts are reconciled,
+or conflict-free representation (CRDTs) \cite{shapiro-2011-crdts,kleppmann-2019-local,litt-2022-peritext}.
+In both approaches, supporting complex edits on tree structures
+remains a challenge \cite{jungnickel-2016-otjson,da-2024-jsoncrdt}. Mergeable replicated data
+types (MRDTs) \cite{kaki-2019-mrdts} can merge updates provided there is a suitable relational
+representation of the data. Recent work on MySubstrates and Grove \cite{adams-2025-grove,klokmose-2024-mywebstrates}
+has been based on CRDTs, whereas Edwards et al. \cite{edwards-2021-typed} use OT.
 
@@
 \begin{figure*}[t]
-\vspace{-0.5em}
-\includegraphics[width=0.95\textwidth,clip,trim=1cm 1cm 1cm 0.5cm]{fig/walkthrough.pdf}
-\vspace{-1em}
+\includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
 \caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
@@
 
-\newpage
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Programming Experiences}
+\label{sec:background-exp}
+
+We review a range of compelling programming experiences explored in recent research,
+focusing on those supported by Denicek.
+
+\subsubsection*{Local-First Collaboration}
+Since the early collaborative programming environments such as Collabode \cite{goldman-2011-collaborative},
+real-time collaboration has became widely used, if not always without challenges \cite{tan-2024-vslive}.
+Merging concurrent edits is one such challenge.
+In addition to CRDT-based approaches \cite{klokmose-2024-mywebstrates}, conflicts arising during
+collaboration have been solved through fine-grained locking \cite{wang-2024-nbconflicts}.
+
+\subsubsection*{Programming by Demonstration}
+Earliest PbD systems used the paradigm for tasks ranging from graphics
+and user interfaces to general-purpose programming \cite{smith-1975-pygmalion,cypher-1993-pbd}.
+Wrangler \cite{kandel-2011-wrangler} showed the effectivity of PbD for data cleaning,
+whereas more recent uses range from augmented reality prototyping \cite{leiva-2021-rapido}
+and web automation \cite{chen-2023-miwa} to end-user software customizations
+\cite{litt-2020-customization,litt-2020-wildcard}. Nevertheless, expressing conditions in PbD remains an
+active research topic~\cite{pu-2022-semanticon,radensky-2018-conditionals}. A more general class of demonstrational
+interfaces \cite{myers-2000-demonstrational} also includes programming by example, used
+for example to infer data transformations in spreadsheets \cite{gulwani-2012-spreadsheets}.
+Demonstrational systems can be used to directly perform actions, but also to generate code
+that captures the interaction, as done e.g. in WREX \cite{drossos-2020-wrex}.
+
+\subsubsection*{Incremental Recomputation}
+Interactive programming systems with live previews \cite{petricek-2020-live,mcdirmid-2013-usable}
+have been attempting to update the previews without full recomputation at least since the pioneering
+work on the Cornell Program Synthesizer \cite{teitelbaum-1981-cps}. Outside live programming,
+more work has been focused on updating computation when data change, although this can also
+be source code of a program passed to an adaptive interpreter \cite{acar-2006-adaptive}.
+Incremental recomputation is also a concern in notebook systems where cells can be run out of
+order \cite{singer-2020-jollity}. This can be addressed by maintaining a dependency graph
+\cite{petricek-2018-wrattler,koop-2017-dataflow} that also allows incremental recomputation
+on code change.
+
+\subsubsection*{Schema Change Control}
+Schema change control is concerned with adapting data and code to reflect changes in schema.
+The problem is well-studied in the context of databases \cite{brahmia-2024-evolution},
+although only few systems also automatically adapt database queries \cite{wang-2019-schema}.
+However, the problem is starting to be recognized in programming systems research
+\cite{litt-2020-cambria,edwards-2025-schema} as well as live programming \cite{barenz-2020-live}
+where state needs to be preserved during program editing.
+
+\subsubsection*{End-User Debugging}
+Non-programmers also need to be able understand and debug their programs \cite{kissinger-2006-debugging}.
+Systems such as Whyline and Probe Log \cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
+record information about program execution to let users analyse why they see a particular result,
+whereas displaying intermediate steps can aid understanding of data science pipelines \cite{shrestha-2021-unravel}.
+More generally, such functionality can leverage provenance tracking \cite{cheney-2009-provenance}
+and program slicing \cite{ricciotti-2017-imperative,perera-2012-functional}. The same infrastructure
+can also be used to build linked visualizations \cite{perera-2022-linked}.
+
+\subsubsection*{Concrete Programming}
+Programming can be simplified by working with concrete values instead of abstractions, an idea
+pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. Self uses
+prototypes at the object level. At the expression level, similar functionality can be provided
+by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
+mechanism as central, whereas other systems view tracking of copy \& paste as an additional
+editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
+also tracked in spreadsheets \cite{hermans-2015-copypaste} and Gridlets \cite{joharizadeh-2020-gridlets}
+offer an appealing alternative design.
+
+\subsubsection*{Other Programming Experiences}
+Several experiences not directly addressed in this paper are worth further investigation.
+Projectional editors such as Lorgnette \cite{gobert-2023-lorgnette},
+live literals \cite{omar-2021-livelits}, projection boxes \cite{lerner-2020-boxes} and data detectors
+\cite{nardi-1998-agents} allow visualization or editing of aspects of programs through a user interface.
+The problem of making live and rich editors compositional is addressed by
+Engraft \cite{horowitz-2023-engraft}. Programming environments may also integrate
+live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.
 
@@
 In this section, we use the web-based prototype Webnicek to illustrate programming
-experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
-editor that supports navigation in the document and issuing of edit commands. We follow two
+experiences enabled by the Denicek substrate. Webnicek is based on a structure
+editor that supports navigation in the document and issuing of edit commands. In this section, we follow two
 formative examples (\S\ref{app:examples}) related to conference planning.
@@
 \subsubsection*{\circled{A} Adding a Speaker}
-The user starts with an empty document, which is represented as a record. They add a field for
-each heading and a list {\small\Verb_<ul>_}, represented by a field named {\small\Verb_speakers_}. They use the command
-toolbox to add the first speaker.
+The first user starts with an empty document, which is represented as a record. They add a field for
+each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. They
+use the command toolbox to add the first speaker.
 
@@
 details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
-from the textbox into the new element using source view.
-
-\subsubsection*{\circled{C} Saving the Interaction} After adding the speaker from the textbox,
+from the textbox into the new element using a source view.
+
+\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker from the textbox,
 the user opens history view, selects edits that added the speaker and saves those as the
-{\small\Verb_add-speaker_} interaction. They attach this as a handler for the {\small\Verb_click_} event of the button.
+{\small\Verb_add-speaker_} interaction. They attach this interaction as a event handler for the
+{\small\Verb_click_} event of the button.
 
 \keyideabox{\faLightbulbO}{Programming by Demonstration}{Denicek implements programming by
-demonstration (\S\ref{sec:impl-pbd}) by letting the user to save and replay past interactions,
-either directly or by attaching them to a UI event.}
+demonstration (\S\ref{sec:impl-pbd}) by enabling the user to save and replay past interactions.
+Those can be replayed directly, or used as event handlers in order to construct interactive user
+interfaces (\S\ref{sec:impl-interaction}).}
 
 \subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
-version of the document and turns the list into a table. This is done by invoking a series of
+version of the document (A) and turns the list into a table. This is done by invoking a series of
 commands that change tags, wrap elements, copy and transform values.
 
-\subsubsection*{\circled{E} Merging Edits} The two document versions can be automatically merged.
+\subsubsection*{\circled{E} Merging Edits} The two document versions (C and D) can be automatically merged.
 The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
@@
 reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
-Merging is asymmetic, but the order does not matter in the above scenario. The same merging
+Merging is asymmetric, but the order does not matter in the above scenario. The same merging
 operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}
 
-\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds budget calculation to
+\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
 the initial document. Formulas are represented as special {\small\Verb_x-formula_} nodes whose arguments
@@
 
-\keyideabox{\faLightbulbO}{Incremental Recomputation}{Evaluating a formula yields additional
-edits that augment the document with the result. Those edits are kept
-at the top of the document history and are removed in case of a conflict (\S\ref{sec:impl-incremental}),
-providing incremental recomputation.}
+\keyideabox{\faLightbulbO}{Incremental Recomputation}{Formulas are document nodes and
+evaluating a formula yields additional edits that augment the document with the result
+(\S\ref{sec:impl-eval}). Those evaluated edits are kept at the top of the document history and are
+removed in case of a conflict (\S\ref{sec:impl-incremental}), resulting in the incremental
+recomputation behavior.}
 
@@
 end-user debugging via provenance analysis (\S\ref{sec:impl-provenance}).}
+
+% ==================================================================================================
+
+\section{The Denicek Substrate}
+\label{sec:system}
+Denicek represents programs as sequences of edits that construct and transform a computational
+document. In this section, we describe the structure of documents and edits, as well as the
+operations that form the backbone of the system and are used to implement the
+end-user programming experiences as discussed in \S\ref{sec:impl}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Selectors, Documents and Edits}
+\label{sec:system-defs}
+A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
+Records and lists are labeled with a \emph{tag} as in HTML. Record fields have unique names.
+Lists are expected to contain elements of the same structure, identified by a unique index (serving
+as an ID). Primitive nodes can be strings, numbers or references to another location in the
+document tree. References can be relative or absolute. Edit operations only use absolute
+references (to denote a target node), but relative references can appear in the document (e.g., to
+refer to another node within a list element).
+
+A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
+The document model assumes that lists are homogeneous and records heterogeneous, and so the
+\ident{Any} selector makes it possible to refer to all elements of a list, but there is no
+way to refer to all fields of a record. Note that Denicek does not use implicit numerical
+indices for lists. The index of a new list item has to be supplied explicitly. The reasoning behind
+this design choice is discussed in \S\ref{sec:discuss}.
+
+\subsubsection*{Document Edits}
+The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
+require a \emph{target} to which they are applied. Targets are absolute references not containing the
+\ident{Parent} selector. They can contain the \ident{Any} selector, in which case the edit is applied
+to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
+kind. As discussed in \S\ref{sec:discuss}, list elements as well as fields of a record are ordered
+and edits that add a new item take the index or field name of a previous node.
+
+The edits allow an arbitrary transformation of a document via a series of steps whose effect can be
+tracked by the substrate. As we saw, Denicek updates references when document structure changes.
+Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
+(above) and edits that affect references (below).
+
+Renaming a field or wrapping a node updates any references to within the target location
+(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
+rejects the edit. A copy edit of a node to which there is a reference is also rejected, because it
+is ambiguous whether references referring to the original location should be left unchanged,
+or modified to point to the target of the copying (a reference cannot refer to two unrelated locations).
+
+% \note{why not have distinct Copy and Move operations to resolve this?}
+% TP: Good point - this is a nice motivation for Move! I should probably add that :-)
 
@@
 \end{figure}
-
-% ==================================================================================================
-
-\section{The Denicek Substrate}
-\label{sec:system}
-Denicek represents programs as sequences of edits that construct and transform a computational
-document. In this section, we describe the structure of documents and edits, as well as the
-operations that form the backbone of the system and are used to implement the
-end-user programming experiences discussed in \S\ref{sec:impl}.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Selectors, Documents and Edits}
-A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
-References to document locations can be relative or absolute. Both kinds can appear in a
-document (reference nodes); absolute references are used in edits (to denote a target~node).
-
-A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
-The document model assumes that lists are homogeneous and records heterogeneous, and so the
-\ident{Any} selector makes it possible to refer to all children of a list, but there is no
-way to refer to all children of a record. Note that Denicek does not use implicit numerical
-indices for lists. The index of a new item has to be supplied explicitly. The reasoning behind
-this design choice is discussed in \S\ref{sec:discuss}.
-
-\subsubsection*{Document Edits}
-The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
-require \emph{target} to which they are applied. Target is an absolute reference not containing the
-\ident{Parent} selector. It can contain the \ident{Any} selector, in which case the edit is applied
-to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
-kind. As discussed in \S\ref{sec:discuss}, both fields of a record and list elements are ordered
-and edits that add a new item take the index or field name of a previous node.
-
-The edits allow any transformation of a document through a series of steps whose effect can be
-tracked by the substrate. As we saw, Denicek updates references when document structure changes.
-Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
-(above) and edits that affect references (below).
-
-Renaming a field or wrapping a node require corresponding change to references in the document
-(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
-rejects the edit. A copy edit can also be rejected, because it is ambiguous whether references
-referring to the original location should continue referring to the source or should be changed to
-point to the target of the copying (a reference cannot refer to two locations).
 
@@
 \end{tabular}
-\vspace{-0.5em}
+\vspace{-0.75em}
 \caption{Summary of document edit types in Denicek}
@@
 \subsubsection*{Automatic Reference Update}
-There are two situations in which automatic update of references is undesirable. If an edit is
-applied to a singular element of a list (reference contains the \ident{Index} selector),
-references that refer into the list should be unchanged. Although such edits violate the invariant
-that collections are homogeneous, they typically do so only temporarily, for example during
-construction of a new list item (checking such edits is discussed in \S\ref{sec:discuss-future}).
-
-Documents can also contain values that can be of multiple different kinds (i.e., a union type).
-In such case, references should not be updated when the kind of the value changes. For example,
-a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
-involves wrapping the formula node, but this should not affect references to the formula.
-To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits}, below)
-have a \emph{reference behavior} field that can be set to disable automatic reference
-updating. This annotation is required when the target contains the \emph{Index} selector.
-
-\subsubsection*{No Conditional Edits.}
-An important aspect of the design is that the effect an edit has on references inside
-the document does not depend on the current value of the document. This makes it possible to
+As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
+statically enforced (the substrate can be seen as dynamically typed). As a consequence, edits
+that transform the structure can be used as structural edits (affecting the structure)
+or as value edits (affecting the value of specific nodes). The latter can be used, for example,
+when constructing an additional list item. An item can be first constructed and then populated
+with values. This temporarily violates the invariant that collections are homogeneous
+(an issue we might address by using transactions as discussed in \S\ref{sec:discuss}).
+Another use of value edits is when evaluating a formula, which involves wrapping the formula
+node (\S\ref{sec:impl-eval}), but the operation should not affect references to the formula
+itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
+have a \emph{reference behavior} option that can be set to disable automatic reference
+updating. This annotation is required when the target contains the \emph{Index} selector,
+which is necessarilly a value edit affecting a specific node.
+
+% \note{I think it would be desireable to avoid all these complications. References
+% should be automatically maintained so they ``just work''.}
+% \note{I believe that evaluation should not change the type/shape of the document. The space
+% for evaluation results should be be preallocated in some sense, either as extra structure in
+% a formula, or as metadata attached to the tree. That said, I've tried various alternative
+% approaches to this in Subtext without ever settling on a preferred solution.}
+%
+% TP: I think this is true - but that would make the system more "statically typed"
+% and Denicek is dynamically typed - my only reason for this is to make the editing experience
+% more "document-like" - but it is perhaps not necessary.
+
+An important principle of the Denicek design is that the effect an edit has on references inside
+the document does not depend on the current value of the document, only its type. This makes it possible to
 define merging solely in terms of edits, without reference to current document state.
-(The effect \ident{Copy} has on the document structure depends on the existing document
-structure, but not on its value).
-
+% (The effect \ident{Copy} has on the document structure depends on the existing document
+% structure, but not on its value).
 This design choice makes it impossible to encode computational logic directly in the edits
@@
 \label{sec:system-ops}
-Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
-sequences of edits can be merged and also checked for conflicts. Denicek identifies edit histories by
-a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
-identify a common shared part of the history during conflict resolution and merging.
+Denicek provides three primitive operations. A sequence of edits can be applied to a
+document, two sequences of edits can be merged and they can be checked for conflicts. Denicek
+identifies edit histories by a (git-like) hash, computed from the current edit and the hash of the
+preceding edit. The hash is used to identify a common shared prefix of the
+history during conflict resolution and merging.
 
 \subsubsection*{Applying Edits}
-When applying an edit, Denicek locates the target node and transform it according to the edit. If
-the edit can affect references in the document (Fig~\ref{fig:edits}, below), Denicek updates
-matching references in the document according to the rules shown in Fig.~\ref{fig:updates} provided
-that the reference behavior of the edit is not set to disable the updating (this is required when
-the target reference contains the \ident{Index} selector). Reference update behavior for
-\ident{WrapRecord} and \ident{WrapList} differs in that the updated references as a result of
-\ident{WrapList} use the \ident{All} selector. Although \ident{WrapList} specifies the index to
-be used for the newly created list item, we assume that the operation introduces a homogeneous
-list and the new reference should point to all eventual list items.
-
-Also note that affected references in the document may be more specific than the edit
-target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
-a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (A reference in the
-document cannot be more general, because such edit would have to contain the \ident{Index} selector
+When applying an edit, Denicek locates the target node and transforms it according to the edit. If
+the edit affects references in the document (Fig~\ref{fig:edits}, below), Denicek updates
+the relevant references according to the rules shown in Fig.~\ref{fig:updates}, provided
+that the reference behavior of the edit is not set to disable this.
+%(this is required when %the target reference contains the \ident{Index} selector).
+
+Reference update behavior for \ident{WrapRecord} and \ident{WrapList} differs in that the updated
+references as a result of \ident{WrapList} use the \ident{All} selector. Although \ident{WrapList}
+specifies the index to be used for the newly created list item, we assume that the operation
+introduces a homogeneous list, initially containing a single element, and the updated references
+should point to all eventual list items.
+
+Note that the affected references in the document may be more specific than the edit target. For
+example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|}, a reference
+{\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (Contrariwise, a reference in the
+document cannot be more general. A more specific edit would have to contain the \ident{Index} selector
 and this would require setting reference behavior to not trigger reference update.) Finally,
@@
 \ident{DeleteField} edit, the edit is rejected.
-
-\subsubsection*{Merging Edit Histories}
-Merging edit histories is used when two users edit document independently, but also
-when replaying edits in programming by demonstration. The
-operation works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix $E$.
-The merging operation is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that
-can be reapplied on top of the other edit history, resulting in a new history $E, E_1, E_2'$.
-Note that the operation is not symmetric. If there are conflicts among the edits in $E_1$ and
-$E_2$, the result of $E, E_1, E_2'$ will differ from the result of $E, E_2, E_1'$.
-We return to this problem in the next section when discussing conflict detection.
-
-The key operation that enables such reconcilliation takes two individual edits that occurred
-indpendently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
-can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
-There are two aspects of such reconcilliation:
-
-\begin{enumerate}
-\item \emph{Apply to Newly Added.} If $e_2$ is adding new nodes to the document, but
-  $e_1$ modified the document through a selector that would also affect the new nodes added by $e_2$,
-  we need to apply the transformation represented by $e_1$ to the nodes newly added by $e_2$.
-  This is done by generating an additional edit, to be applied after $e_2$, that is based on
-  $e_1$ but targets only the newly added nodes (more details can be found in \S\ref{app:merge-apply-to-added}).
-
-\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
-  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
-  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
-  that apply when transforming references inside a document, although we also support the case
-  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
-\end{enumerate}
 
@@
 \end{tabular}
-\vspace{-0.5em}
+\vspace{-0.75em}
 \caption{How document edits transform references}
 \label{fig:updates}
-\vspace{-1em}
+\vspace{-0.5em}
 \end{figure}
@@
 \end{tabular}
-\vspace{-0.5em}
+\vspace{-0.75em}
 \caption{Effects of individual edit operations}
@@
 % --------------------------------------------------------------------------------------------------
+
+\subsubsection*{Merging Edit Histories}
+Merging edit histories is used when two users edit document independently, but also
+when replaying edits in programming by demonstration. The operation $\mathcal{M}_E(E_1, E_2)$
+works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix $E$.
+The operation is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that can be reapplied
+on top of the other edit history. Formally, $\mathcal{M}_E(E_1, E_2) = E, E_1, E_2'$.
+Note that the operation is not symmetric. $\mathcal{M}_E(E_2, E_1) = E, E_2, E_1'$ and the
+result of applying the two histories to the same node will differ if there are conflicts
+among the edits in $E_1$ and $E_2$. We return to conflict detection in the next section.
+
+% \note{It might help to define a function signature for merging to make this clearer: $\mathcal{M}(P, L, R) = M$ or $\mathcal{M}_P(L, R) = M$.}
+% \note{Merging is often thought of as symmetric, perhaps choose a name that is clearly asymmetric, like rebase, translate, or project.}
+
+The key operation that enables such reconciliation takes two individual edits that occurred
+independently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
+can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
+There are two aspects of such reconciliation:
+
+\begin{enumerate}
+\item \emph{Apply to Newly Added.} If $e_2$ is adding new nodes to the document, but
+  $e_1$ modified the document through a selector that would also affect the new nodes added by $e_2$,
+  we need to apply the transformation represented by $e_1$ to the nodes newly added by $e_2$.
+  This is done by generating an additional edit, to be applied after $e_2$, that is based on
+  $e_1$ but targets only the newly added nodes (more details can be found in \S\ref{app:merge-apply-to-added}).
+
+\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
+  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
+  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
+  that apply when transforming references inside a document, although we now also support the case
+  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
+\end{enumerate}
+
+% \note{I believe that in general a ``backwards'' form of merge is also needed, as in my
+% \emph{retract} and the OT \emph{Exclusion Transform}. A regular merge takes forking edits
+% and composes them. The backwards form converts composed edits into forking edits (and can
+% be combined with a forward merge to recompose in the opposite order). This is needed for
+% cherry picking and reverts, however you don't seem to need that for this paper.}
+%
+% TP: Correct, I do not support those - but cherry-picking is a nice example of a thing
+% that needs then and would be very nice to have!
+
+% \note{Related question: how do you merge an Append(target, index, after, node) with
+% DeleteItem(target, after)? A famous counterexample in OT was that two insertions adjacent
+% to a deletion could flip their location if the edits were reordered. I'll dig up the
+% details of this counterexample for you. Some systems handle this by leaving behind a
+% ``tombstone'' on delete to remember the position of the deleted item. I have a novel
+% solution but I'm probably going to go back to tombstones anyway because they are
+% simpler and handy in the diff UI.}
+%
+% TP: This is a good question. My ordering of records is a somewhat last minute addition
+% so it is not super well debugged. But in theory, it should be fine - because the ordering relation
+% could keep the IDs of removed elements (only for ordering purposes). I think this is
+% compatible with the MRDT data structure I'm trying to use.
 
@@
 If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
-reflect the additional wrapping (2) and we need to create and additional \ident{Add} operation
-that will add the checkbox to the new item (1). The result is two edits $e_2', e_2''$:
+reflect the wrapping (2) and we need to create and additional \ident{Add} operation
+that will add the checkbox to the new item (1). The result is a sequence with two addtional edits $e_2', e_2''$:
 %
@@
 \begin{array}{rcl}
+  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
+  e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
 e_2'  &=& \srckvd{Append}(\textnormal{\small\Verb|/todo/items|}, \#1, \#0, \srcid{<li>Do some work</li>})\\
@@
 %
-If we performed the merge operation in the other order, we would simply append $e_1, e_1'$ after
-$e_2$. Although this is not the case in general, the result will be the same regardless of the order here.
+If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
+$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
+same.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\includegraphics[width=0.95\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
+\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
+\label{fig:merging}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
 \subsubsection*{Conflict Resolution}
-When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is sometimes possible to construct
-edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
-$E, E_2, E_1'$ are the same. However, this is not always the case. Two edits can conflict if
-they both modify the same value, if they transfrom the structure of a node in incompatible ways
-or if one modifies a node nested inside a node replaced or deleted by the other.
-Detected conflicts can be reported to the user, but we also use conflict detection to
-implement incremental evaluation (\S\ref{sec:impl-incremental}).
+When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is desirable that
+applying $\mathcal{M}_E(E_1, E_2)$ and $\mathcal{M}_E(E_2, E_1)$ results in the same document.
+This is not always the case. If the two edits modify the same value, transform the structure of
+a node in incompatible ways or one edit modifies a node deleted by the other, the merge operation
+cannot reconcile. Such conflicting edits can be detected, reported to the user and optionally
+removed. Conflict detection is also used to implement incremental evaluation~(\S\ref{sec:impl-incremental}).
+
+% Our treatment of conflicts necessarily differs in this respect from replication algorithms like
+% Operational Transformation~\cite{OT} and Convergent Replicated DataTypes~\cite{CRDT} where
+% conflicts are arbitrarily but deterministically resolved.
+%
+% TP: The merge operation in Denicek does also resolve them deterministically and arbitrarily
+% though this does not always result in "sensible" resolution. (The dynamic nature again, I guess...)
+% It is an interesting question how CRDTs could do this though - because we do want to have
+% detectable conflicts (for the purpose of the evaluation)
 
@@
 conflicts, i.e.~it may report a conflict even if two edits can be merged successfully
-(see \S\ref{sec:discuss-future}). An effects describe how an edit affects the document structure
+(see \S\ref{sec:discuss}). Effects describe how an edit affects the document structure
 and we distinguish between three types of effects as shown in Fig.~\ref{fig:effects}.
@@
 We say that a set of effects $F_1$ conflicts with another set of effects $F_2$ if there
-are effects $f_1\in F_1$ and $f_2\in F_2$ such that they are of the same kind and the target
-of $f_1$ is a prefix of the target of $f_2$ or vice versa (allowing specific \ident{Index} to
-match against \ident{All} in any direction).
+are effects $f_1\in F_1$ and $f_2\in F_2$ that are of the same kind and the target
+of one is a prefix of the target of the other (allowing a specific \ident{Index} to
+match against \ident{All} in both directions).
 
@@
 conflict with effects of (1) any of the effect $e_1$ from $E_1$ or (2) effects of any of the
-previously removed edits. Here, the dependencies of $e_2$ include its target, but also source
+previously removed edits. Here, the dependencies of $e_2$ include its target, but also the source
 of \ident{Copy} and additional dependencies recorded explicitly as discussed in
@@
 
+% \note{The effect-style conflict detection is a nice idea! Removing conflicts before merging is
+%  also an interesting idea I hadn't considered. That might allow me to simplify my transformation rules. }
+
 % ==================================================================================================
 
-\section{Programming Experiences Implementation}
+\section{Programming Experiences}
 \label{sec:impl}
 
-The key claim of this paper is that the Denicek computational substrate
-makes it easy to support a range of experiences that make programming more concrete,
-collaborative and interactive.
-
-In this section, we describe how Webnicek uses the substrate
-to support local-first collaboration (\S\ref{sec:impl-collab}), programming by
-demonstration (\S\ref{sec:impl-pbd}, \S\ref{sec:impl-interaction}), incremental recomputation
-(\S\ref{sec:impl-eval}, \S\ref{sec:impl-incremental}), schema change control (\S\ref{sec:impl-schema}),
-end-user debugging via provenance tracking (\S\ref{sec:impl-provenance})
-and concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
-
-We describe the programming experience in isolation in this section. The next section provides a
-more comprehensive evaluation through a case study that combines multiple of them together.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\includegraphics[width=0.9\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
-\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
-\label{fig:merging}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
+The key claim of this paper is that the Denicek computational substrate suppports a range of
+compelling user experiences, many of which crucially rely on Denicek's ability to merge edits.
+We demonstrate this claim with the Webnicek web-based programming system, which uses Denicek to support:
+%
+\begin{enumerate}[itemsep=1.5pt,label=(\roman*),topsep=6pt,leftmargin=20pt]
+\item local-first collaboration (\S\ref{sec:impl-collab}),
+\item programming by demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
+\item incremental recomputation (\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}),
+\item schema change control (\S\ref{sec:impl-schema}),
+\item end-user debugging via provenance tracking (\S\ref{sec:impl-provenance}),
+\item concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
+\end{enumerate}
+%
+We describe the programming experience in isolation in the context of Webnicek in this section.
+The data science notebook system Datnicek, discussed in \S\ref{sec:case}, provides
+a more comprehensive evaluation by combining multiple user experiences together.
 
@@
 
-The Denicek representation enables local-first collaboration~\cite{kleppmann-2019-local},
-as illustrated previously in \S\ref{sec:walk} (E). If a document is edited by multiple users, they can
+The Denicek substrate enables local-first collaboration~\cite{kleppmann-2019-local},
+as illustrated earlier (\S\ref{sec:walk}, E). If a document is edited by multiple users, they can
 each make edits to their local copy and eventually merge the variants using the operation to
@@
 
-Recall that merging of edit histories is not symmetric. Document $D$ in Fig.~\ref{fig:merging}
-can be obtained either by appending $C'$ (produced by the edit reconcilliation operation)
-on top of $A,B$ or by appending $B'$ on top of $A,C$. According to our effect analysis, the two
-operations are conflicting. Although $B$ primarily affects the document structure, it also adds
-a new field to the record (email), which is a \ident{ValueEffect}, conflicting with
-\ident{ValueEffect} of $C$. In this scenario, the conflict can be ignored and the resulting document
-is the same in both cases.
-
-However, the resulting two histories will differ. In the first case, the
-\ident{Append} edit that adds a new node is supplemented by further focused edits that transform
-the added node from a list item to a new table row.
-In the second case, the structural transformations are automatically applied to all rows.
-As discussed in \S\ref{sec:system-ops}, conflicts during merging can be resolved either by
-removing conflicting edits or by letting the later edits overwrite the former ones.
+As an example, consider the scenario in Fig.~\ref{fig:merging}. Here, document $D$ can be obtained
+either by appending $C'$ (produced by the edit reconciliation operation) on top of $A,B$ or by
+appending $B'$ on top of $A,C$. The edit histories resulting from the two ways of merging will
+differ. In the first case ($B$ then $C'$), the \ident{Append} edit that adds a new node is followed
+by further focused edits that transform the newly added node from a list item to a table row.
+In the second case ($C$ then $B'$), the structural transformations are applied to all rows.
+
+According to our effect analysis, the two operations are conflicting.
+Although $B$ primarily affects the document structure, it also adds a new field to the record (email),
+which is a \ident{ValueEffect}, conflicting with the \ident{ValueEffect} of $C$. In this scenario, the
+conflict can be ignored and the resulting document will be the same.  However, as discussed
+in \S\ref{sec:system-ops}, merging of edit histories is not symmetric and arising conflicts
+can also be resolved by removing conflicting edits.
+
+~
+
+% \note{To guarantee that replicas converge it is necessary to have a global order on merges.
+% OT and CRDT algorithms use timestamps combined with a unique ID for each client to establish
+% this global order without a centralized coordinator. We can do the same. This means you never
+% know if your changes will dominate or whether they will be eventually overriden but people
+% seem to be willing to live with it. Personally, I'm a skeptical: I expect that like with NoSQL
+% people will eventually realize this sucks and go back to centralized systems.}
 
@@
 system and the system then repeats it, directly or in a generalized way. To use direct
-repetition with Denicek (Fig.~\ref{fig:pbd}), the user can select edits from the edit history,
-name them and replay them. In case of general-purpose document editing in Webnicek, this requires
-certain forethought, but as illustrated in \S\ref{sec:case}, the mechanism is very effective in
-a domain such as data wrangling \cite{kandel-2011-wrangler}.
+repetition with Denicek (Fig.~\ref{fig:pbd}), the user selects edits from the edit history,
+name them and replay them. For general-purpose document editing in Webnicek, this requires
+certain forethought, but as illustrated in \S\ref{sec:case}, the mechanism is very effective
+in a domain such as data wrangling \cite{kandel-2011-wrangler}.
 
 There are two notable aspects of our implementation of programming by demonstration in Webnicek.
-First, Webnicek records edits in the document itself (by representing individual edits as nodes and
-storing them in list inside a {\small\Verb_/saved-interactions_} field). This means that no other
-implementation mechanism outside of the system is needed and also that the stored edits can be
-modified by the user or tools working with the document.
-
-Second, to replay edits, Webniceks does not append the recorded edits on top of the current history.
+First, Webnicek records the saved edits in the document itself (by representing individual edits
+as nodes and storing them in a list inside the {\small\Verb_/saved-interactions_} field).
+This means that no other implementation mechanism outside of the system is needed and also
+that the stored edits can be modified by the user or tools working with the document.
+
+Second, to replay edits, Webnicek does not append the recorded edits on top of the current history.
 When saving edits, it stores the hash of the history at the time of saving. When replaying,
@@
 Webnicek also accounts for the case where edits recorded in the document are themselves
-transformed (when they are reconcilliated with other edits during merging). In this case, Webnicek
+transformed (when they are reconciliated with other edits during merging). In this case, Webnicek
 updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).
+
+% Simply replaying recorded edits provides limited programming capabilities. We discuss in the
+% next section how recorded edits can themselves be edited by the user to generalize and
+% conditionalize their effects. More general and conventional programming constructs are discussed in \S\ref{sec:discuss}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Interactive User Interfaces}
+\label{sec:impl-interaction}
+
+Programming by demonstration can be used to define interactive elements in the document. In a
+simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\Verb_click_} event
+handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
+executes the edits using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., Webnicek appends
+the edits to a history at the time when the edits were recorded and merges the edits with the
+current history.
+
+The use of merging when replaying recorded edits is crucial in both the Todo App and the Conference
+List formative examples (see \S\ref{app:examples}). In both cases, the merging makes it possible to define
+a user interface for adding items (new Todo items, new speakers) and later change the document
+structure (refactor the speaker list into a table) or add functionality (formula to evaluate whether a
+Todo item has been completed) without having to recreate the user interface for adding items.
+The use of merging ensures that new items are added in a correct format or with the additional
+functionality.
+
+In addition to the core functionality provided by Denicek, the Webnicek system also makes it possible
+to generalize the interactions recorded through programming by demonstration in a limited way. As shown
+in Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
+generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
+In addition to the recorded interaction, we manually specify (in the source view) that the edits
+should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
+the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
+applied to elements for which the formula (which tests if the checkbox is checked) specified by a
+relative selector {\small\Verb_./condition/comp/result_} evaluates to true.
+
+Note that the generalization mechanism does not violate the Denicek principle discussed in \S\ref{sec:system-defs}
+that edits cannot depend on values. Webnicek finds all nodes for which the condition
+holds and generates one specific (non-conditional) edit for each of the nodes.
+
+% \note{The above paragraph was surprising to me and I'm not sure I fully understand it even after
+% staring at the figure for a while. Edits are recorded as an AST using x-tags to hide them
+% in the UI? That became clear only after reading the next section. Our audience might be a lot
+% happier if we used a textual grammar. Earlier it was said that edits can't be conditional but
+% then a conditional predicate appears here. This is a special feature of button nodes? Or event
+% handler nodes? This starts to resemble Angular and similar template languages (which I find hard to read).}
+%
+% TP: Haha, Angualr - oh no! But I don't have a better solution right now sadly...
+
+\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
+Programming by demonstration systems typically implement heuristic for generalization
+\cite{myers-2000-intelligence} infers and suggests such generalizations. If integrated into
+Webnicek, such heuristic could for example automatically construct a formula
+based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}
 
@@
 
-\subsection{Interactive User Interfaces}
-\label{sec:impl-interaction}
-
-Programming by demonstration can be used to define interactive elements in the document. In a
-simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\Verb_click_} event
-handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
-executes the edits using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., Webnicek appends
-the edits to a history at the time when the edits were recorded and merges the edits with the
-current history.
-
-The use of merging when replaying recorded edits is crucial in both the Todo App example and the
-Conference List example (see \S\ref{app:examples}). In both cases, it makes it possible to define
-a user interface for adding items (new Todo items, new speakers) and later change the document
-structure (refactor speaker list to a table) or add functionality (formula to evaluate whether a
-Todo item has been completed) without having to recreate the user interface for adding items.
-The use of merging ensures that new items are added in a correct format or with the additional
-functionality.
-
-Denicek can be used to generalize the interactions recorded through programming by
-demonstration. Our prototype illustrates this option in a limited way. As shown in
-Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
-generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
-In addition to the recorded interaction, we manually specify (in the source view) that the edits
-should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
-the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
-applied to elements for which the formula (which tests if the checkbox is checked) specified by a
-relative selector {\small\Verb_./condition/comp/result_} evaluates to true.
-
-\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
-Programming by demonstration systems typically implement heuristic for generalization
-\cite{myers-2000-intelligence} infers and suggests such generalizations. If integrated into
-a system based on Webnicek, such heuristic could automatically construct a formula
-based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}
-
-% --------------------------------------------------------------------------------------------------
-
 \subsection{Formula Language and Evaluation}
@@
 
-Denicek documents can contain formulas inspired by the spreadsheet
-paradigm \cite{nardi-1990-spreadsheets}. Formulas can specify richer computations than
-what can be expressed using document edits. Formulas do not transform the document and their
-results are transient, although their evaluation also leverages the substrate operations,
-namely merging of edit histories.
-
+As illustrated earlier (\S\ref{sec:walk}, F), Denicek documents can contain formulas inspired by the
+spreadsheet paradigm \cite{nardi-1990-spreadsheets}. Formulas in Webnicek do not transform the
+document and their results are transient, but they can describe richer computations than what can
+be expressed using document edits.
+
+Formula evaluation leverages Denicek's ability to merge edit histories.
 As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
 special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
-rendered in a special way (\circled{1} and \circled{4}), but they are created using ordinary
-edits and the Denicek substrate treates them as stnadard nodes.
-
-To evalaute formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_}
-record into {\small\Verb_<x-evaluated>_} (\circled{2} and \circled{3}), which keeps the previous
+rendered in a special way (4), but they are created using ordinary
+edits and the Denicek substrate treats them as standard nodes.
+
+To evaluate formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_} (2)
+record into {\small\Verb_<x-evaluated>_} (3), which keeps the previous
 formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
-(Keeping the previous state of the formula is not necessary, but it enables provenance analysis
-as discussed in \S\ref{sec:impl-provenance}.) The way edits generated by evaluation are merged
+Keeping the previous state of the formula is not necessary, but it enables provenance analysis
+as discussed in \S\ref{sec:impl-provenance}. The way edits generated by evaluation are merged
 with the document is discussed in the next section.
 
+% \note{As I've said before, I'd prefer to store the result of the formula inside the <x-formula>
+% record instead of wrapping it on evaluation. This avoids changing the shape of the document on
+% evaluation, which I recall added some complexity earlier.}
+%
+% TP: I think I tried that, but it still did not solve my problem - because I needed to add
+% child nodes to the result. So this can probably only be done for a statically typed version.
+
 The Counter App example shown in Fig.~\ref{fig:counter} illustrates the interaction between formulas
-and programming by demonstration. To implement a counter, the Increment and Decrement buttons
-wrap the current counter value in a formula that adds or subtracts 1.
+and programming by demonstration. To implement a counter, we first add the initial value 0 to the
+document. We then perform three edits that wrap the current counter value in a formula that adds
+1 to the value. The three edits are used as an event handler for the button and so clicking the
+button creates an increasingly long sequence of increments. One advantage of this approach that
+we will leverage later is that the system also records a trace of how the final computed value
+was obtained.
+
+% \note{OK maybe I'm finally getting it. You aren't mutating the value of the counter, you're wrapping
+% the value in an increment formula. I never got that in our previous discussions. Presumably that
+% evaluation is now frozen so it serves as an execution trace. This is a lot different from my
+% approach where a formula stays unchanged except for the results. One advantage is that fields
+% contain a record of their computed mutations whereas in my approach that is recorded in the
+% past state of the doc whence the mutation occured.}
 
@@
 Webnicek exposes the underlying representation of formulas to the user, but the same representation
-can be edited through a user-friendly mechanism such as a block-based editor
-\cite{jansen-2019-xlblocks} or a calculation view \cite{sarkar-2018-calc}. The key point
+can be edited through a user-friendly mechanism such as a textual calculation view \cite{sarkar-2018-calc}
+a block-based editor \cite{jansen-2019-xlblocks}. The key point
 is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
-way, edit them and merge them with other changes.}
+way, edit them as any other node and merge them with other changes.}
 
@@
 \vspace{-0.5em}
-\includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
+\includegraphics[width=0.95\columnwidth,clip,trim=0cm 5cm 11.5cm 0cm]{fig/counter.pdf}
 \vspace{-0.5em}
-\caption{Increment wraps the existing count in a formula that adds 1 to the
-previous value (1), (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).}
+\caption{We wrap the initial count (1) in a formula that adds 1 to the
+value, (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).
+The button replays the saved edits that wrap the current count in a formula~(5).}
 \label{fig:counter}
+\vspace{-1em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Incremental Recomputation}
+\label{sec:impl-incremental}
+
+As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental recomputation.
+The cost of speaker travel depends on the number of speakers, but the cost of refreshments
+depends only on two constants. The edit that adds a speaker invalidates only the former computation.
+
+The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
+Denicek generates edits that wrap the formula in {\small\Verb_<x-evaluated>_} and set the result
+to the computed value. In Webnicek, those evaluated edits are kept in a separate list that is always
+appended to the top of the current edit history. When the user makes subsequent edits, the evaluated
+edits are pushed through the newly added edits. If the edits conflict (according to the conflict
+detection discussed in \S\ref{sec:system-ops}), the affected evaluated edits and edits that
+depend on them are dropped.
+
+The dependencies between edits alongside with the conflict detection mechanism provide functionality
+that other systems implement using a dependency graph using topological sort.
+
+\keyideabox{\faRefresh}{Live Programming}{The Webnicek prototype does not automatically
+evaluate formulas. This makes it easier to understand the evaluation
+mechanism, but a realistic system based on the substrate could automatically evaluate
+formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}
+and use incremental recomputation for performance~reasons.}
+
+% \note{I'm confused. Changing the value of a location conflicts with an edit that evaluates a
+% formula containing a reference to that location? And these evaluation edits are second-class
+% edits that get retroactively dropped to ``unevaluate'' the formula? I'd prefer what you suggest
+% above: formulas are seemingly globally re-evaluated after every change, with incremental
+% computation a hidden perf optimization. }
+
+% \note{An important property of incremental computation systems is avoiding redundant execution.
+% Usually this involves a topological sort of the computation dependency graph. We may need to
+% have a story about that to avoid hate from reviewer 2. Personally, I think incremental
+% recomputation is about performance, not the programming experience, so it isn't that important for this paper. }
+%
+% TP: The cool thing is we get this for free! Clarified.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
+  \vspace{-1em}
+  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
+  is added (3), only the results of affected formulas are invalidated.}
+  \label{fig:incremental}
   \vspace{-1em}
@@
 
-\subsection{Incremental Recomputation}
-\label{sec:impl-incremental}
-
-As illustrated in Fig.~\ref{fig:incremental}, Webnicek supports incremental re-computa\-tion.
-In the example, the cost of speaker travel depends on the number of speakers, but the cost
-of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
-former.
-
-The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
-Webnicek generates \emph{evaluated edits} that are appended to the top of the history.
-When subsequent edits are made, they are appended after all non-evaluated edits and the
-evaluated edits are pushed through the newly added edits. If the edits conflict (according to
-conflict detection discussed in \S\ref{sec:system-ops}), affected evaluated edits and edits that
-depend on them are dropped.
-
-\keyideabox{\faRefresh}{Live Programming}{The Webnicek prototype does not automatically
-evaluate formulas. This makes it easier to understand the evaluation
-mechanism, but a realistic system based on the substrate could automatically evaluate
-formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}
-and use incremental recomputation for performance~reasons.}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
-  \vspace{-1em}
-  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
-  is added (3), only the results of affected formulas are invalidated.}
-  \label{fig:incremental}
-  \vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \subsection{Schema Change Control}
@@
 
-Document structure often needs to evolve \cite{burnett-2014-silos}, as illustrated by our
-Conference List example where a list is transformed into a table. When this happens,
+Document structure often needs to evolve \cite{burnett-2014-silos}, as illustrated by the formative
+example Conference List where a list is transformed into a table. When this happens,
 data and code that depend on the structure of the document need to evolve correspondingly.
 The problem is well-known in database systems \cite{rahm-2006-schema} and has recently been
-studied in the context of programming systems \cite{edwards-2025-schema}.
+explored in the context of programming systems \cite{edwards-2025-schema}.
 
@@
 discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
-document when edits modify the document structure. This enables a form of schema code
-co-evolution \cite{edwards-2025-schema}. Formulas embedded in Denicek documents use reference
+document when edits modify the document structure. This enables a form of schema and code
+co-evolution \cite{edwards-2025-schema}. Formulas embedded in documents use reference
 nodes to refer to both data sources (in the document) and the results of other computations.
@@
 
-As noted earier, evaluated edits transform the formula structure (wrapping it in the
-{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
-references to formulas are not transformed when the formula is evaluated (otherwise, references
+Edits produced by document evaluation transform the formula structure (wrapping it in the
+{\small\Verb_<x-evaluated>_} record). However, the reference behaviour of those edits is set
+to keep references unchanged (as discussed in \S\ref{sec:system-defs}) and so references to
+formulas are not transformed when the formula is evaluated (otherwise, references
 would be updated to point to the original unevaluated formulas).
@@
 \begin{figure}[t]
-\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
-\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
-\label{fig:coevolution}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
+\vspace{-0.25em}
 \includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
-\vspace{-1em}
+\vspace{-2em}
 \caption{Using provenance tracking to highlight document parts that contributed to the
@@
 \label{fig:provenance}
-\vspace{-0.5em}
+\vspace{-0.75em}
 \end{figure}
@@
 in \S\ref{sec:impl-eval}. When a formula is evaluated, the final {\small\Verb_<x-evaluated>_}
-document node contains the result, but also sub-tree that represents a full evaluation trace
+document node contains the result, but also sub-tree that represents the full evaluation trace
 \cite{perera-2012-functional}. We analyse the trace, collect all reference nodes in the trace and
 highlight all nodes referred to in the computation. Denicek makes this easy as we
-only need to collect reference nodes nested in a formula node.
+need to collect references directly nested in the formula node.
 
@@
 trace collected by the Denicek substrate can also be used to provide a detailed
-explanation~\cite{perera-2012-functional} or to automatically generate linked visualizations~\cite{perera-2022-linked}.}
+explanation~\cite{perera-2012-functional} or, for example, to automatically construct
+linked visualizations~\cite{perera-2022-linked}.}
 
@@
 concrete~\cite{edwards-2004-example,smith-1975-pygmalion} is to support copying of functionality
-as in prototype-based object-oriented programming~\cite{randall-1995-self}.
+as in prototype-based object-oriented programming~\cite{ungar-1987-self,randall-1995-self}.
 
@@
 paste~\cite{edwards-2006-copypaste,edwards-2022-copypaste} for formulas. Rather than introducing
-abstractions (functions), users can copy and modify formulas to reuse them. However, when the user
+abstractions (functions), users can copy and modify formulas to reuse them. When the user
 discovers an error in the original formula, Webnicek lets them use the merging mechanism to correct
@@
 
-The mechanism is shown in Figure~\ref{fig:copypaste}. The user copies an incorrect formula using
+In Figure~\ref{fig:copypaste}, the user copies an incorrect formula using
 the \ident{Copy} edit and then modifies it to use a different data source. They then navigate back
@@
 temporary fork into the original document, the \emph{Apply to Newly Added} logic of the merge
-operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and applies them also
-to the copied formula. In other words, the \ident{Copy} edit of the Denicek substrate, alongside
-with the fact that formulas are ordinary document nodes, provide keys component for a
-straightforward implementation of the managed copy \& paste functionality.
+operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and also applies them
+to the copied formula. The merging operation with the \ident{Copy} edit and the fact that formulas
+are ordinary document nodes, provide keys component for a straightforward implementation of the
+managed copy \& paste functionality.
+
+%\note{Nice and simple! My approach tried to be a lot fancier but ended up being too heavywieght. }
 
@@
 \begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
+\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
+\label{fig:coevolution}
+\vspace{-1em}
+\end{figure}
+
+% ==================================================================================================
+
+\section{Design Considerations}
+\label{sec:discuss}
+
+The design of the Denicek substrate is the result of an iterative process in which we repeatedly
+adapted the Denicek design and revisited the implementation of the Webnicek system until we obtained
+a satisfactory solution for the six formative examples (\S\ref{app:examples}). In this section, we
+document the design challenges, many of which are shared with related systems
+\cite{jakubovic-2022-ladder,edwards-2005-subtext,hall-2017-infra,omar-2021-livelits}.
+
+\keyideabox{\faCubes}{Uniformity and Composability}{
+Two guiding principles for the design of Denicek have been \emph{composability} and
+\emph{uniformity}~\cite{jakubovic-2023-techdims}. The substrate should cover a large number of
+scenarios using a small number of concepts. This is apparent in the design of the document
+structure---a node can represent data, code or rich text---as well as in the role of
+edits---an edit can be a value or structure change, the result of user interaction or the
+result of evaluation.
+}
+
+% The inconvenience associated with composability and uniformity~\cite{jakubovic-2023-techdims} is not
+% a problem for Denicek. We view Denicek as an underlying structure on top of which more
+% convenient end-user programming systems can be built. The uniform design offers a degree of open-endedness,
+% making it possible to use the basic structures of Denicek in new ways, not anticipated in this paper.
+
+\subsubsection*{Dynamic and Static Typing}
+Denicek does not explicitly track the document structure (it is dynamically typed). We assume that lists
+are homogeneous, but do not enforce this property. A statically typed system could distinguish
+structural edits from value edits. This has a theoretical appeal and it simplifies aspects
+of the implementation, e.g. by removing the need to control reference updating, but it makes
+working with the system less akin to document editing.
+
+We violate structural invariants when constructing values gradually (e.g., when adding a new
+speaker to a table), as well as in evaluation (wrapping {\small\Verb_<x-formula>_}
+in {\small\Verb_<x-evaluated>_}). A system with static types could use a form of edit
+transactions, where structural invariants are reestablished only after a sequence of edits.
+
+% A key design choice for a substrate like Denicek is whether to track the document structure
+% explicitly or make it implicit. In Denicek, we assume that elements of a list node have the same
+% structure (i.e.~records should have the same fields, so that they can be targetted using the \ident{All}
+% selector), but we do not enforce this. Often, the property is temporarily violated when adding a
+% list item, but then restored once the item is created.
+%
+% Keeping the structure (type information) explicit has theoretical appeal and it simplifies some
+% aspects of the implementation. In particular, edits to document structure can be clearly separated
+% from edits of document values. The same separation has disadvantages in that it complicates the
+% edit language and it requires user interface that departs further from document editor.
+% (In Webnicek, users can select all list items and edit them at once using the same mechanism as
+% when editing individual list items. With explicit structure, changing a list structure has to be
+% either a different kind of edit or possibly an edit of a virtual ``prototype'' element.)
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
 \includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
-\vspace{-1.5em}
+\vspace{-1em}
 \caption{Correcting an incorrect formula. The user uses Copy to create the second
@@
 
-% ==================================================================================================
-
-\section{Design Considerations}
-\label{sec:discuss}
-
-The design of the Denicek substrate is the result of an iterative process in which we repeatedly
-adapted the Denicek design until we obtained a satisfactory solution for the six formative examples
-(\S\ref{app:examples}) in the Webnicek prototype. In this section, we document the design challenges,
-many of which have until now been personal knowledge of researchers working on related
-systems \cite{jakubovic-2022-ladder,edwards-2005-subtext,hall-2017-infra,omar-2021-livelits}.
-
-Two guiding principles for Denicek have been \emph{composability} and
-\emph{uniformity}~\cite{jakubovic-2023-techdims}. The design should cover a large number of
-cases using a small number of concepts. This is apparent in the design of the document
-structure (a node can represent data, code or rich text) as well as in the role of edits
-(an edit can be a value or structure change, the result of user interaction or the
-result of evaluation).
-
-The inconvenience associated with composability and uniformity~\cite{jakubovic-2023-techdims} is not
-a problem for Denicek. We view Denicek as an underlying structure on top of which more
-convenient end-user programming systems can be built. The uniform design offers a degree of open-endedness,
-making it possible to use the basic structures of Denicek in new ways, not anticipated in this paper.
-
-\subsubsection*{Implicit vs. Explicit Structure}
-A key design choice for a substrate like Denicek is whether to track the document structure
-explicitly or make it implicit. In Denicek, we assume that elements of a list node have the same
-structure (i.e.~records should have the same fields, so that they can be targetted using the \ident{All}
-selector), but we do not enforce this. Often, the property is temporarily violated when adding a
-list item, but then restored once the item is created.
-
-Keeping the structure (type information) explicit has theoretical appeal and it simplifies some
-aspects of the implementation. In particular, edits to document structure can be clearly separated
-from edits of document values. The same separation has disadvantages in that it complicates the
-edit language and it requires user interface that departs further from document editor.
-(In Webnicek, users can select all list items and edit them at once using the same mechanism as
-when editing individual list items. With explicit structure, changing a list structure has to be
-either a different kind of edit or possibly an edit of a virtual ``prototype'' element.)
-
-\subsubsection*{Merging and Document State}
-One specific design choice in Denicek is that the merge operation operates solely on two
-sequences of edits, but it does not need access to the current document state (or full edit
-histories). The reconcilliation of edits (\S\ref{sec:system-ops}) is defined for two individual
-edits $e_1$ and $e_2$.
-
-This design limits the possible language of edits. We cannot, in general, support
-conditional edits (that are applied only when a certain condition holds for
-the current document state). This is because the merging operation would not be able to
-determine whether the edit is applied (and whether its effects should~take~place).
-
-One consequence of the design choice is that the generalization of recorded interactions (\S\ref{sec:impl-interaction}) has to
-be done on a meta-level (by specifying how to invoke edits, rather than through the edits themselves).
-Another consequence is that our langauge of selectors has to be abstract.
-We can target all list elements using the \ident{All} selector, but cannot combine this
-representation with a representation that explicitly lists indices of all items currently in the list.
-
-\subsubsection*{Explicit List Indices}
-Denicek does not use implicit numerical indices for lists. When adding an item, the
-user (or the system built on top of Denicek) has to explicitly provide an index. This design
-is motivated by a typical programming by demonstration scenario where the user adds a
-new list item and then modifies it (adding a speaker or a Todo item). When using explicit indices,
-the edits following \ident{Append} can use the index to refer to and modify the newly added
-item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
-indices of newly added items when replaying recorded interactions in \S\ref{sec:impl-pbd}).
-
-Using implicit numerical indices is possible, but it requires treating all list-related edits
-as operations that transform document selectors. This means adding multiple new rules to
-Figure~\ref{fig:updates}, which specifies how edits transform references. Merging then has to
-modify indices when edits add, remove or reorder list items. List order then, somewhat confusingly,
-becomes a part of document structure rather than a property of the list value.
-
-The problem could be avoided by not letting the user modify a list item after adding it.
-(We can require new items to be constructed outside of the list and then added through a
-single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easy to
-implement, but it requires unintuitive user interaction and it also does not address all
-use cases for modifying list item via index.
-
-\subsubsection*{Ordering List Items and Record Fields}
-When list indices are not numerical and supplied explicitly, we need another mechanism for
-tracking order of list items. Denicek uses a data structure inspired by Mergeable Replicated
-Data Types \cite{kaki-2019-mrdts} and requires specifying the index of a preceding list item
-when appending to a list. However, Webnicek supplies the index of preceding item automatically.
-(If two edits append to the same list independently with the same preceding item, the order
-is non-deterministic.)
-
-Note that Denicek uses the same mechanism for record fields. Keeping record fields ordered is
-desirable because Denicek documents map directly to HTML documents and so the order is visible
-to a user. Moreover, different ways of merging edits can result in different order of adding fields
-to a record. Making order explicit ensures those result in equivalent documents.
-
-Denicek still makes a distinction between records and lists, even if the two are similar
-technically. The difference is conceptual. A list is expected to contain items of the
-same structure (that can be addressed using the \ident{All} selector), whereas record fields
-are expected to be of different types. If we unified lists and records, \ident{All} would have
-to be applicable to records too and inferring and checking document structure would be
-challenging.
-
-\subsubsection*{Convergence vs. Divergence of Document Variants}
-There are two ways of managing document variants in collaborative editing~\cite{edwards-2025-schema}.
-In the \emph{divergence} model, users can maintain their own variant indefinitely (share data,
-but use a different schema locally). In the \emph{convergence} model, users have to adopt all
-earlier changes in order to adopt a later change (accept new schema before new data).
-
-Denicek supports only the simpler \emph{convergence} model. To support the \emph{divergence}
-model, we would need an operation dual to our edit reconcilliation (given two subsequent edits,
-$e_1, e_2$, we want to generate $e_2'$ that has the same effect as $e_2$ but can occur before
-$e_1$). The pair of operations has been called \emph{project} and \emph{retract}
-\cite{edwards-2021-typed}. We choose not to support this as it would complicate the basic
-Denicek substrate structure.
-
-\subsubsection*{Linear History vs. Graph of Edits}
-Denicek keeps a linear history of edits. When merging edits, new edits have to be transformed and
-added to the top of the history. This model is akin to git rebase. An alternative representation
-would be to maintain a graph of edits akin to when using git merge. In this representation,
-we would need to keep parents of edits and special merge edits would have multiple parents.
-
-Maintaining a graph of edits would make recording of edits in programming by demonstration
-easier (\S\ref{sec:impl-pbd}) as we would not need to update recorded edits if they are
-transformed during merging. (Their hashes would remain the same.) However, supporting special
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection*{Expressive Selectors and Edits}
+Our merge operation operates on two sequences of edits, but it does not need access
+to the current document state or history. This limits the expressiveness of edits. Denicek does not
+support conditional edits (applied only when a specific condition holds), because merging would then
+not be able to determine if two edits are conflicting. As a result, generalization of recorded
+interactions (\S\ref{sec:impl-interaction}) has to be done at a meta level and generates a series
+of individual edits. This design also limits the expressiveness of selectors. For example,
+an edit targeting nodes with a specific tag (e.g., all {\small\Verb_<h3>_} elements) would also be
+conditional.
+
+% One specific design choice in Denicek is that the merge operation operates solely on two
+% sequences of edits, but it does not need access to the current document state (or full edit
+% histories). The reconciliation of edits (\S\ref{sec:system-ops}) is defined for two individual
+% edits $e_1$ and $e_2$.
+%
+% This design limits the possible language of edits. We cannot, in general, support
+% conditional edits (that are applied only when a certain condition holds for
+% the current document state). This is because the merging operation would not be able to
+% determine whether the edit is applied (and whether its effects should~take~place).
+%
+% One consequence of the design choice is that the generalization of recorded interactions (\S\ref{sec:impl-interaction}) has to
+% be done on a meta-level (by specifying how to invoke edits, rather than through the edits themselves).
+% Another consequence is that our langauge of selectors has to be abstract.
+% We can target all list elements using the \ident{All} selector, but cannot combine this
+% representation with a representation that explicitly lists indices of all items currently in the list.
+
+% Denicek supports only a limited set of selectors. An absolute reference can select a record
+% field (\ident{Field}), specific list item (\ident{Index}) or all list items (\ident{All}).
+% Once can imagine a range of other useful selectors, for example to select list items with a
+% given tag or select items satisfying other conditions.
+%
+% We could support such selectors for those edits that do not affect the document
+% structure. However, supporting such conditional selectors in general is incompatible with our
+% design decision that merging should not depend on document state. (If an edit wraps the body of all
+% {\small\Verb_<h3>_} elements, we need to know the document state in order to determine whether
+% we need to modify the selector pointing to an item at a specific index.)
+
+\subsubsection*{List Indices and Ordering}
+Denicek does not use numerical indices for lists. Indices are unique identifiers that have to
+be provided explicitly. (Although Webnicek generates indices automatically when editing lists.)
+The design supports a programming by demonstration scenario where the user adds a new list item
+and then modifies it (adding a speaker or a Todo item).
+
+With numerical indices, the edits following \ident{Append} would not have access to the index of
+the newly added item. Using a numerical index computed from the list length would require
+complex logic to update indices when merging edit operations that affect lists. Another alternative
+is to disallow modification of a newly added item, but this limits programming by demonstration.
+
+% Denicek does not use implicit numerical indices for lists. When adding an item, the
+% user (or the system built on top of Denicek) has to explicitly provide an index. This design
+% is motivated by a typical programming by demonstration scenario where the user adds a
+% new list item and then modifies it (adding a speaker or a Todo item). When using explicit indices,
+% the edits following \ident{Append} can use the index to refer to and modify the newly added
+% item. (To avoid overwriting existing items when replaying recorded edits, Webnicek replaces
+% indices of newly added items when replaying recorded interactions in \S\ref{sec:impl-pbd}). \note{I don't think that was explained back there.}
+%
+% Using implicit numerical indices is possible, but it requires treating all list-related edits
+% as operations that transform document selectors. This means adding multiple new rules to
+% Figure~\ref{fig:updates}, which specifies how edits transform references. Merging then has to
+% modify indices when edits add, remove or reorder list items. List order then, somewhat confusingly,
+% becomes a part of document structure rather than a property of the list value.
+%
+% \note{I'm still unclear whether your list indices are ordinal numbers like in an array or are unique IDs.
+Your usage of implicit/explicit isn't obvious to me.
+The above paragraph seems to imply they are IDs.
+Note that automatically generating unique IDs in code execution can lead to nondeterminism.
+If you are using IDs, do they affect equality of lists?}
+% The problem could also be avoided by not letting the user modify a list item after adding it.
+% (We can require new items to be constructed outside of the list and then added through a
+% single edit that combines the logic of \ident{Append} and \ident{Copy}.) This is easy to
+% implement, but it requires unintuitive user interaction and it also does not address all
+% use cases for modifying list item via index. \note{I'm still confused, but this sounds like it really isn't a solution afterall.}
+
+To maintain order of list elements, Denicek uses a data structure inspired by the list MRDT
+\cite{kaki-2019-mrdts} and requires specifying the index of a preceding item when appending
+or inserting into a list. Denicek uses the same mechanism for record fields as the order of fields
+may be visible to the user, for example in Webnicek where fields represent children of a HTML node.
+Making the order of fields explicit ensures that it is maintained during merging.
+
+% \subsubsection*{Ordering List Items and Record Fields}
+% When list indices are not numerical and supplied explicitly, we need another mechanism for
+% tracking order of list items. Denicek uses a data structure inspired by Mergeable Replicated
+% Data Types \cite{kaki-2019-mrdts} and requires specifying the index of a preceding list item
+% when appending to a list. However, Webnicek supplies the index of preceding item automatically.
+% (If two edits append to the same list independently with the same preceding item, the order
+% is non-deterministic.) \note{Why is this needed? I let the order of operations determine the list order. }
+%
+% Note that Denicek uses the same mechanism for record fields. Keeping record fields ordered is
+% desirable because Denicek documents map directly to HTML documents and so the order is visible
+% to a user. Moreover, different ways of merging edits can result in different order of adding fields
+% to a record. Making order explicit ensures those result in equivalent documents.
+%
+% Denicek still makes a distinction between records and lists, even if the two are similar
+% technically. The difference is conceptual. A list is expected to contain items of the
+% same structure (that can be addressed using the \ident{All} selector), whereas record fields
+% are expected to be of different types. If we unified lists and records, \ident{All} would have
+% to be applicable to records too and inferring and checking document structure would be
+% challenging.
+
+% TODO: This is potentially interesting too, but it is maybe way too complex and weird
+%
+% \subsubsection*{Dependency Tracking}
+% Recall that the incremental recomputation (\S\ref{sec:impl-incremental}) is based on the
+% conflict detection mechanism of Denicek (\S\ref{sec:system-ops}). When pushing edit $e_2$ through
+% $e_1$, a conflict is detected if they affect overlapping targets or if $e_1$ modifies a
+% document node that is a source of $e_2$, typically, the source of a \ident{Copy} edit.
+%
+% For edits generated during formula evaluation (\S\ref{sec:impl-eval}), we need to track an
+% additional kind of dependency. Evaluating a formula that adds two numbers from two document
+% location results in an edit that sets the result of the formula to the sum of the two numbers.
+% The edit contains the resulting number, but we also need to record that the edit has the two
+% source locations as dependencies. For this reason, Denicek makes it possible to annotate edits
+% with additional dependencies. One alternative would be to introduce a special \ident{Evaluate}
+% edit that would identify an operation to compute and a list of arguments, specified as a list
+% of references. We choose the former to keep the set of edits smaller.
+
+\subsubsection*{Merging Structure and Capabilities}
+Denicek keeps a linear history and merging appends edits to the top of the history.
+This model is akin to git rebase. One alternative is to maintain a graph of edits akin
+to git merge. This would simplify recording of edits in programming by demonstration
+(\S\ref{sec:impl-pbd}) as we would not need to update saved edits transformed
+during merging. (Their hashes would remain the same.) However, supporting special
 merge edits and non-linear history would make the basic substrate more complex.
 
-\subsubsection*{Supporting Rich Selectors}
-Denicek supports only a limited set of selectors. An absolute reference can select a record
-field (\ident{Field}), specific list item (\ident{Index}) or all list items (\ident{All}).
-Once can imagine a range of other useful selectors, for example to select list items with a
-given tag or select items satisfying other conditions.
-
-We could support such selectors for those edits that do not affect the document
-structure. However, supporting such conditional selectors in general is incompatible with our
-design decision that merging should not depend on document state. (If an edit wraps the body of all
-{\small\Verb_<h3>_} elements, we need to know the document state in order to determine whether
-we need to modify the selector pointing to an item at a specific index.)
-
-\subsubsection*{Dependency Tracking}
-Recall that the incremental recomputation (\S\ref{sec:impl-incremental}) is based on the
-conflict detection mechanism of Denicek (\S\ref{sec:system-ops}). When pushing edit $e_2$ through
-$e_1$, a conflict is detected if they affect overlapping targets or if $e_1$ modifies a
-document node that is a source of $e_2$, typically, the source of a \ident{Copy} edit.
-
-For edits generated during formula evaluation (\S\ref{sec:impl-eval}), we need to track an
-additional kind of dependency. Evaluating a formula that adds two numbers from two document
-location results in an edit that sets the result of the formula to the sum of the two numbers.
-The edit contains the resulting number, but we also need to record that the edit has the two
-source locations as dependencies. For this reason, Denicek makes it possible to annotate edits
-with additional dependencies. One alternative would be to introduce a special \ident{Evaluate}
-edit that would identify an operation to compute and a list of arguments, specified as a list
-of references. We choose the former to keep the set of edits smaller.
-
-% ---------------------------------
-
-\begin{comment}
-Maybe have 'enabled' for edits afterall?
-(we can merge with conflicts and disable some edits, but keep them in history for info)
-
-TODO - things to work on
-* "represent" edits somewhere in document as "library of functions"
-  and then call those from buttons (rather than embedding them directly)
-  allow some kind of abstraction (as in Histogram) to make them reusable
-* figure out how to do evaluation better
-  (based on the stored abstractions? but need to store provenance...)
-
-SEMANTIC CONDITIONS
-\url{https://www.youtube.com/watch?v=NBnc2ToS_j0}
-(has a section on this in background)
-
-INTERACTION
-* replay stored event handlers against the old version?
-  (this way, adding an item to a speakers list gets migrated \& adds a new table row!)
-* simlarly!! we need merge in order to apply edits to multiple targets
-  (when you remove all items in a list, the indices change)
-  (but I guess we should do this against version at the time of saving too....)
-
-[this \& evaluation = the unreasonable effectiveness of merging]
-
-Notes on storing and reusing edits
-* references need to be represented as references so that they get updated
-  (NO! not if we reply them against old version, which seems better - but there are 2 design choices)
-* how to apply them to multiple targets? use Move to update the selectors instead of
-  replacing the prefix manually
-
-CONDITIONALS
-https://toby.li/files/p311-radensky.pdf
-
-\end{comment}
-
-\subsection{Future work}
-\label{sec:discuss-future}
-Although Denicek does not explicitly track document structure (or schema, or type), all documents
-have an implicit structure.
-=> use type system to check temporarily invalid state
-
-better effect system
-
-IDEA: Type check edit groups to ensure they preserve structure but not individual edits eg when adding list item
-
-~
-
-~
-
-~
-
-% memory mapped graphics etc.
+Denicek also supports only the simpler \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
+where users have to merge all changes in order and cannot merge only selected edits (cherry
+picking in git). The \emph{divergence} model lets users, for example, keep their own schema but
+import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
+that is dual to our edit reconcilliation (given two subsequent edits, $e_1, e_2$, we want
+to generate $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).
 
@@
 
+what are the limitations of Datnicek
+
+x
+
+x
+
+x
+
+x
+
 \section{Evaluation}
@@
 
+NICE THINGS IN DATNICEK
+* merging of code edits (e.g. change parameter merges with adding a call to chain)
+* remove cell to which there are refs - does the susbtrate catch this?
+
 \section{Discussion}
-
-\begin{comment}
-
-\subsection{Heuristic evaluation}
-
-techdims
-
-\subsection{Limitations}
-
-RANDOM FORMAL STUFF
-
-ohter formal things
-* non-conditionality of edits
-* show that they can transform document from any to any without removing/readding values
-
-PROPERTIES
-- can change any document to any other - sure, via remove add - but also more semantically
-(if it contained all values, can we do it without removing and adding them?)
-
-DEMOS
-
-// cf shadow dom v reactu
-// something like google forms!
-
-* two ways of pushing edits should be equivalent, i.e.
-
-push e1 e2 e3 over b1 b2 b3 can be done via
-  push e1 over b1 = e1'; over b2 = e1''; over b3 = e1'''
-  then push e2, then push e3
-or
-  push e1, e2, e3 over b1, then over b2, then over b3
-(implies that no "extra" edits can be appended later)
-\end{comment}
 
@@
 
-NICE THINGS IN DATNICEK
-* merging of code edits (e.g. change parameter merges with adding a call to chain)
-* remove cell to which there are refs - does the susbtrate catch this?
-
-RELATED WORK
-
-TIJS
-https://homepages.cwi.nl/~storm/publications/icsews13live-p17-p-16201-preprint.pdf
-
-via Pierre
-https://pijul.org/manual/why_pijul.html
-
-Geoffrey Litt
-https://dl.acm.org/doi/pdf/10.1145/3426428.3426914
-https://dl.acm.org/doi/pdf/10.1145/3397537.3397541
-
-Programming By Example: Intelligence in Demonstrational Interfaces
-https://cacm.acm.org/research/programming-by-example-intelligence-in-demonstrational-interfaces/
-
-Rapido: Prototyping Interactive AR Experiences through
-Programming by Demonstration
-https://dl.acm.org/doi/pdf/10.1145/3472749.3474774
-
-openrefine
-- facet (group by) - then change the key to normalize
-  (https://www.youtube.com/watch?v=yTJ6x6zEQmI 51:40)
-
-Don't Step on My Toes: Resolving Editing Conflicts in Real-Time
-Collaboration in Computational Notebooks
-https://arxiv.org/pdf/2404.04695
-
-AI assistants in notebook interfaces
-https://dl.acm.org/doi/pdf/10.1145/3544548.3580940
-
-Unravel - understand pipelines
-https://dl.acm.org/doi/pdf/10.1145/3472749.3474744
-
-WREX - nicest tool
-https://dl.acm.org/doi/pdf/10.1145/3313831.3376442
-
-\begin{comment}
-\newpage
-xx
-
-~\\
-introduction \\
-background \\
-related work \\
-design process / goals \\
-case study \\
-formative research \\
-formative study \\
-analysis \\
-system \\
-implementation \\
-evaluation / heuristic evaluation \\
-discussion and limitations \\
-
-\newpage
-\section{Introduction old}
-
-The computational substrate using which software is built determines the capabilities that the
-software can provide. An imperative substrate that views programs as instructions modifying
-bytes in memory makes it almost impossible to allow end-user inspection or reprogramming of
-running software.
-
-substrate as defined by \cite{jakubovic-2022-ladder}
-
-A computational substrate defines what software is built from. This may be objects as in
-Smalltalk, lists as in Lisp, or memory with data and code as in UNIX/C.
-The different substrates enable different kinds of programming experiences.
-For example, object-oriented programming has historically been linked to the development
-of graphical user interfaces (where objects can correspond to elements on the screen).
-It has also enabled the development of visual programming environments such as the Alternate
-Reality Kit, based on message sending between objects.
-
-In principle, any computational substrate can be used to develop any programming experience,
-but the greater the impedance mismatch between the substrate and the desired experience,
-the more difficult it will be to provide the experience and combine it with the rest of the
-system and other programming experiences developed for the system. (One can implement support
-for programming-by-demonstration using C/C++, for example as part of a game scripting engine,
-but it will not work with the rest of the ordinary C/C++ ecosystem.)
-
-Medium is the message
-
-\subsection{Substrate}
-
-The question asked in this paper is, what would be the ideal programming substrate
-for supporting a range of programming experiences that make programs more
-collaborative, transparent and allows for a gradual transition from non-programmer
-to a programmer. We want a programming substrate that makes it easy to develop
-programming experiences such as:
-
-\begin{itemize}
-\item \emph{Programming by demonstration} --
-  Allow non-programmers to construct simple programs by performing examples of the expected behaviour. \cite{leiva-2021-rapido}.
-\item \emph{Local-first collaboration} --
-  Multiple users should be able to use and modify a single program, preferrably without requiring a central server. \cite{kleppmann-2019-local}
-\item \emph{Provenance tracking} --
-  The execution of the program should leave an understandable trace that lets the user understand why program resulted in a particular result.
-\item \emph{Schema evolution [extra-ish]} --
-  When the user evolves the structure of the program, data and code should co-evolve automatically to match the new structure.
-\item \emph{Notational freedom [extra-ish?]} --
-  Allow users to adapt the program using a notation that suits them and is appropriate for the programming task at hand. [Joel]
-\item \emph{Concrete programming [extra?]} --
-  It should be possible to reuse parts of program or program logic without constructing abstractions, for example by managed copy \& paste.\cite{edwards-2006-copypaste,edwards-2022-copypaste}
-\end{itemize}
-
-substrate as defined by \cite{jakubovic-2022-ladder}
-
-% pbd
-\cite{leiva-2021-rapido,cypher-1993-pbd}
-\cite{chen-2023-miwa}
-
-% lcoal first
-\cite{kleppmann-2019-local,klokmose-2024-mywebstrates}
-% provenance
-\cite{ko-2004-whyline,ko-2009-whyline,krebs-2023-probelog}
-\cite{ricciotti-2017-imperative,perera-2012-functional}
-\cite{perera-2022-linked}
-
-% visualizations of results or execution
-% https://www.dcs.warwick.ac.uk/pvw04/p01.pdf
-% https://dl.acm.org/doi/10.1145/3313831.3376494
-
-% previews - livelits
-
-% Schema evolution
-% our paper; see https://openproceedings.org/2023/conf/edbt/paper-160.pdf - good refs in 2.2
-
-% https://x.com/jonathoda/status/1185888711210389504
-
-% also cool: https://arxiv.org/pdf/2303.06777
-
-% Hazel family of things
-% structure editing - sandblocks
-
-Joel's definition of substrate in Onward!
-Bret Victor talk
-\url{https://www.youtube.com/watch?v=ef2jpjTEB5U}
-
-In what ways is a substrate "natural"?
-
-thinglab - create line by cloning, it sticks to mouse pointer, clicking sticks it to something else
-squeak - has all the browsers (method search...)
-
-computational substrate
-how it differs from computational media?
-more low-level - media suggests that there it comes
-
-\section{The whatever system}
-\subsection{Document + Edits}
-defines
-\begin{itemize}
-  \item selectors
-  \item nodes
-  \item edits
-\end{itemize}
-
-\subsection{Walkthrough}
-* todo list? (or counter, but that is a bit boring)
-
-\section{Themes}
-* programming by demonstration
-  - binding interactions to gui elements (event handlers)
-* provenance tracking
-  - Amy Ko's whyline, Probe Log by HPI, enables linked visualizations
-* merging of edit histories / collaborative editing
-  - bonus - can share restricted link to allow users fill out
-    forms (allow partial edits only / def by selector?)
-* scehma change - change data \& code accordingly
-* everything is an edit
-  - interaction with the GUI
-  - evaluation? tbd
-* copy \& paste abstraction
-    (requires finishing new approach to formulas!)
-  - edit before copy to propagate edit to other places
-    (or edit after copy to make it specific to a case)
-  - higher order copying from https://tomasp.net/academic/papers/copy-paste/paint22.pdf
-* augmenters
-  - cf. bonnie nardi (calls them something else - Jonathan says)
-  - add programming by demonstration data wrangling gui to table (trigger interactions)
-    cf. lorgnette
-
-\section{Applications}
-* todo list / counter / maybe too simple
-* (if used in the walkthrough, maybe something else? board game as in varv - tic tac toe? or 7guis?)
-* conference organizer
-* data exploration (ala histogram)
-* linked charts
-
-\section{Extras}
-* metablocks?
-* self-sustainability
-* some non-browser implementation of this (as in Varv?)
-
-explicit structure
-self-sustainability
-notational freedom
-
-\newpage
-~
-
-REMAINING IMPLEMENTATION TODOs:
-
-* Some kind of provenance visualization
-* Some kind of matchers/transformers mechanism (ideally to add interactive buttons to tables)
-* Apply to all (remove completed in TODO)
-
-\newpage
-~
-
-append /speakers \#hamilton <li>
-addfld /speakers/\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
-
->>> wraprec /speakers body <table>
-
-[update selectors - add /body]
-append /speakers/body \#hamilton <li>
-[dtto]
-addfld /speakers/body\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
-
->>> retag /spekers/body <tbody>
-
-[na - out of scope]
-
->>> retag /speakers/body/* <tr>
-
-[apply to added - update node (* => \#hamilton scoping)]
-append /speakers/body \#hamilton <li>
-+ retag /speakers/body/\#hamilton <tr>
-[na - out of scope]
-addfld /speakers/body\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
-
->>> wraprec /speakers/body/*/speaker value <td>
-
-append /speakers/body \#hamilton <li>
-+ retag /speakers/body/\#hamilton <tr>
-+ wraprec /speakers/body/\#hamilton/speaker value <td> [ no op - it is not there]
-addfld /speakers/body/\#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
-+ wraprec /speakers/body/\#hamilton/speaker value <td>
-
->> updid /speakers/body/* speaker=>name
-
-append /speakers/body \#hamilton <li>
-+ retag /speakers/body/\#hamilton <tr>
-+ wraprec /speakers/body/\#hamilton/speaker value <td> [ no op - it is not there]
-+ updid /speakers/body/\#hamilton speaker=>name [ no op - it is not there ]
-addfld /speakers/body \#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
-+ wraprec /speakers/body/\#hamilton/speaker value <td>
-+ updid /speakers/body/\#hamilton speaker=>name
-
->>> addfld /speakers/body/* email ""
-
-append /speakers/body \#hamilton <li>
-+ retag /speakers/body/\#hamilton <tr>
-+ wraprec /speakers/body/\#hamilton/speaker value <td> [ no op - it is not there]
-+ updid /speakers/body/\#hamilton speaker=>name [ no op - it is not there ]
-+ addfld /speakers/body/\#hamilton email "" [ no op - it is not there ]
-addfld /speakers/body \#hamilton speaker "Margaret Hamilton, hamilton@mit.edu"
-+ wraprec /speakers/body/\#hamilton/speaker value <td>
-+ updid /speakers/body/\#hamilton speaker=>name
-+ addfld /speakers/body/\#hamilton email ""
-
-...
-
-copy /speakers/body/*/name => /speakers/body/*/email
-edit /speakers/body/*/name (fn1)
-edit /speakers/body/*/email (fn2)
-
-\end{comment}
-
 \end{document}

### 2025-03-28 paper.tex
Case study part 1

@@
 \begin{figure}[t]
+\vspace{-0.5em}
 \includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
-\vspace{-1em}
-\caption{Correcting an incorrect formula. The user uses Copy to create the second
-formula (1). They notice an error (2), go back in history to switch left and right argument (3),
-merge the change and re-evaluate both formulas (4).}
+\vspace{-1.5em}
+\caption{The user uses creates a copy of a formula (1). They notice an error (2), go back in
+history to switch the arguments (3), merge the change and re-evaluate both formulas (4).}
 \label{fig:copypaste}
@@
 
-% TODO: This is potentially interesting too, but it is maybe way too complex and weird
-%
-% \subsubsection*{Dependency Tracking}
+\subsubsection*{Merging Structure and Capabilities}
+Denicek keeps a linear history and merging appends edits to the top of the history.
+This model is akin to git rebase. An alternative is to maintain a graph of edits akin
+to git merge. This would simplify recording of edits in programming by demonstration
+(\S\ref{sec:impl-pbd}) as we would not need to update saved edits transformed
+during merging. (Their hashes would remain the same.) However, supporting special
+merge edits and non-linear history would make the basic substrate more complex.
+
+Denicek also only supports the \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
+where users have to merge all changes in order and cannot adopt selected edits (cherry
+picking in git). In the \emph{divergence} model, users could keep their own schema but
+import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
+that is dual to our edit reconcilliation (given subsequent edits $e_1, e_2$, we want
+to generate $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).
+
+\subsubsection*{Dependency Tracking}
+Conflict detection in Denicek is used when merging edits, but also to implement
+incremental recomputation (\S\ref{sec:impl-incremental}). Edits generated during formula evaluation
+(\S\ref{sec:impl-eval}) may be \ident{Copy} edits (when evaluating a reference), but also
+\ident{Add} edits (when setting the evaluation result to a computed value). For those edits, we
+record additional dependencies identifying the source of the operation arguments. Those additional
+dependencies also need to be considered in conflict detection.
+
 % Recall that the incremental recomputation (\S\ref{sec:impl-incremental}) is based on the
@@
 
-\subsubsection*{Merging Structure and Capabilities}
-Denicek keeps a linear history and merging appends edits to the top of the history.
-This model is akin to git rebase. One alternative is to maintain a graph of edits akin
-to git merge. This would simplify recording of edits in programming by demonstration
-(\S\ref{sec:impl-pbd}) as we would not need to update saved edits transformed
-during merging. (Their hashes would remain the same.) However, supporting special
-merge edits and non-linear history would make the basic substrate more complex.
-
-Denicek also supports only the simpler \emph{convergence} model of collaborative editing \cite{edwards-2025-schema}
-where users have to merge all changes in order and cannot merge only selected edits (cherry
-picking in git). The \emph{divergence} model lets users, for example, keep their own schema but
-import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
-that is dual to our edit reconcilliation (given two subsequent edits, $e_1, e_2$, we want
-to generate $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).
-
-\section{Case study}
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\vspace{-0.25em}
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 0.7cm 18.5cm 0.2cm]{fig/source.pdf}
+\vspace{-0.5em}
+\caption{Underlying document representation of a code cell and a grid cell in the Datnicek
+notebook system.}
+\label{fig:source}
+\vspace{-0.25em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}[t]
+\includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
+\caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
+a code cell (1), edits it in a grid that infers edit operations via programming by demonstration
+(2), turned to a code cell (3) and then visualized.}
+\label{fig:datnicek}
+\end{figure*}
+
+% ==================================================================================================
+
+\section{Case study: Datnicek Notebooks}
 \label{sec:case}
+
+Denicek is intended as a low-level computational substrate that can be used as the basis for
+interactive programming systems that view programs as documents. The most prominent example
+of such systems today are notebook environments for data science. We developed Datnicek, a
+simple notebook system that shows the ability of Denicek to support rich interactive user
+experiences. In this section, we present Datnicek and reflect on its development. A simple data
+exploration conducted in Datnicek is shown in Fig.~\ref{fig:datnicek}.
+
+\subsection{Requirements}
+The design of Datnicek brings together a range of recent research ideas on interactive programming
+environments for data science. Datnicek notebooks consist of code cells and markdown cells, but
+they also support interactive grid cells where the user can use programming by demonstration to
+construct data cleaning scripts. We aim to support the following features:
+
+\begin{itemize}
+\item \emph{Structure Editing.} Code in code cells should be edited through structure editor
+  as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
+  structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
+  We do not aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.
+
+\item \emph{Collaborative Editing.} It should be possible to merge independently done code changes
+  as in Grove~\cite{adams-2025-grove}, addressing the known pitfall of versioning Jupyter
+  notebooks~\cite{singer-2020-jollity}.
+
+\item \emph{Code Completion.} During editing, code completion should offer available data
+  transformations and operations as when using type providers \cite{syme-2013-inforich} or
+  iterative prompting in The Gamma \cite{petricek-2022-thegamma}.
+
+\item \emph{Output Invalidation.} When code is edited, the previously evaluated results that
+  depend on it, directly or transitively, should be invalidated as in Wrattler \cite{petricek-2018-wrattler,petricek-2020-live},
+  addressing another well-known limitation of Jupyter notebooks \cite{koop-2017-dataflow}.
+
+\item \emph{Interactive Data Cleaning.} It should be possible to edit data tables in an interactive
+  grid and use programming by demonstration for common data cleaning tasks as in
+  Wrangler or Vizier~\cite{kandel-2011-wrangler,kennedy-2022-vizier}.
+
+\item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
+  turned into code that can be checked and further edited as in Wrex \cite{drossos-2020-wrex}.
+\end{itemize}
+
+\noindent
+The Datnicek notebook system implements the above requirements on top of Denicek.
+Although Datnicek is a proof of concept, it shows that Denicek provides a suitable basis
+for the implementation.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Implementation}
+
+data representation
+
+(maybe show fig with source)
+
+operations
+
+\newpage
+
+\subsection{Reflections}
+
+foof
+
 (data science environment)

### 2025-03-30 paper.tex
Case study implementation

@@
 \begin{figure*}[t]
+\vspace{-0.5em}
 \includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
+\vspace{-0.5em}
 \caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
@@
 \label{fig:datnicek}
+\vspace{-0.5em}
 \end{figure*}
@@
 
-Denicek is intended as a low-level computational substrate that can be used as the basis for
-interactive programming systems that view programs as documents. The most prominent example
-of such systems today are notebook environments for data science. We developed Datnicek, a
-simple notebook system that shows the ability of Denicek to support rich interactive user
+Denicek is a low-level computational substrate. It serves best as the basis for interactive
+programming systems that view programs as documents. The most prominent example of such systems
+today are notebook environments for data science. To explore this use case, we have developed
+Datnicek, a simple notebook system that shows the ability of Denicek to support rich interactive user
 experiences. In this section, we present Datnicek and reflect on its development. A simple data
@@
   structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
-  We do not aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.
+  We do not yet aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.
 
@@
 \item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
-  turned into code that can be checked and further edited as in Wrex \cite{drossos-2020-wrex}.
+  turned into code that can be checked by the data scientist and further edited as in Wrex \cite{drossos-2020-wrex}.
 \end{itemize}
@@
 \subsection{Implementation}
-
-data representation
-
-(maybe show fig with source)
-
-operations
+Datnicek is implemented using the Elm architecture \cite{fowler-2020-mvu},
+where a system maintains a state that is updated through events. The state consists
+of the list of Denicek edits, transient state of the user interface, current computed document
+and other cached information. All interactions that affect the notebook
+use an event that appends edits to the list of Denicek edits. Remaining events
+update the transient user interface state. The support for collaborative editing has been
+implemented through an ad-hoc mechanism.
+
+\subsubsection*{User Interface}
+Datnicek notebooks consists of a series of cells. In the code cells, the user can
+use the ``+'' button to add a new variable, add an operation to the end of an existing chain
+of operations, but also to specify an argument of an operation. The button opens
+a menu showing all valid options based on the current context.
+
+In the interactive grid, the user can edit column headers and table cells. When they edit
+a value, Datnicek suggests generalised edits to be applied to the entire
+column or table. For example, when the user changes the ``0 p'' value to ``0'', Datnicek
+suggests to replace `` p'' with the empty string in all cells. Other suggested edits include renaming
+or deleting a column, splitting a column using a delimiter and filtering rows based on the selected
+value. Datnicek can also turn edits performed interactively into a new code cell.
+
+\subsubsection*{Programmatic Code Cells}
+A Denicek document representing a Datnicek notebook is shown in Fig.~\ref{fig:source}. A
+notebook is a record storing individual cells as fields. The tag determines the type of the cell.
+
+The programming language used in code cells is inspired by the data exploration calculus \cite{petricek-2020-live}.
+A cell consists of a sequence of bindings of the form $\textnormal{\ident{let}}~v=e$ that assign
+an expression $e$ to a variable $v$. An expression can be a reference to a variable,
+global value (such as \ident{data} and \ident{vis}) or a method invocation $e.m(e_1, \ldots, e_n)$.
+
+Bindings are represented as fields of a record and expressions use the formula representation
+discussed in \S\ref{sec:impl-eval}. Operations and global values are identified by an absolute
+reference pointing to the special \ident{/\$datnicek} namespace. Chains of method calls are
+represented using nested formulas. Evaluation of formulas proceeds as in Webnicek
+and wraps {\small\Verb_<x-formula>_} in {\small\Verb_<x-evaluated>_}.
+The user interface displays the original formula alongside with the evaluated result, supporting
+tables and Compost data visualizations \cite{petricek-2021-compost}. Edits to the formula
+invalidate dependent results as discussed in \S\ref{sec:impl-incremental}.
+
+\subsubsection*{Interactive Grid Cells}
+Grid cells consist of a reference to the data source they edit and a collection of
+{\small\Verb_<transform>_} nodes that represent individual data transformations constructed via
+programming by demonstration. A transformation consists of metadata,
+a sequence of edits and a formula representing the transformation (with {\small\Verb_<x-hole>_}
+standing for the data table to be transformed).
+
+A transformation may correspond to multiple underlying Denicek edits. For example, splitting
+``foo/bar'' using ``/'' as the delimiter adds a new column ``bar'', copies values from
+``foo/bar'', renames ``foo/bar'' to ``foo'' and then transforms the values in the new columns
+by dropping the part before and after the delimiter, respectively.
+
+Interactive grid also supports conditional transformations that affect rows satisfying a given
+condition (e.g., deleting rows with a specific value in a given column). As discussed
+in \S\ref{sec:system-defs}, Denicek does not allow specifying conditional selectors.
+To support conditional transformations, we add a special kind of (virtual) edit
+{\small\Verb_<x-expandable-edit>_} that stores the underlying edit and a condition.
+As in \S\ref{sec:impl-interaction}, when applying the edit, Datnicek finds all rows for which the
+condition holds and generates a single non-conditional edit for each such row.
 
@@
 \subsection{Reflections}
+
+edits and formula in interactive should be unified
 
@@
 * remove cell to which there are refs - does the susbtrate catch this?
+
+\newpage
+~
+\newpage
 

### 2025-03-30 paper.tex
Datnicek reflections

@@
 
-\subsubsection*{Two-Stage Methodology}
+\subsubsection*{Two-Phase Methodology}
 The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
@@
 science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
-We use the second stage to evaluate suitability of the Denicek substrate for the development of
+We use the second phase to evaluate suitability of the Denicek substrate for the development of
 end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).
@@
 editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
-also tracked in spreadsheets \cite{hermans-2015-copypaste} and Gridlets \cite{joharizadeh-2020-gridlets}
-offer an appealing alternative design.
+also tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
+offer a spreadsheet abstraction based on reusing concrete computation, similar to the one proposed
+for The Gamma \cite{petricek-2020-live}.
 
@@
 \subsection{Requirements}
+\label{sec:case-req}
+
 The design of Datnicek brings together a range of recent research ideas on interactive programming
@@
 
+\subsection{Reflections}
+The development of Datnicek only started after the design and implementation of Denicek and
+Webnicek was completed. Thanks to this two-phase methodology, the experience reported here
+provides a qualitative evaluation of the Denicek substrate. It also highlights limitations and
+suggests directions for future work.
+
+\subsubsection*{Effectiveness and Unifromity.}
+Many of the Datnicek requirements are addressed by Denicek with minimal implementation
+effort. The implementation represents all non-transient actions on a notebook as Denicek edits.
+Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
+changing a primitive value is an edit, and recommendations in the interactive grid are also edits
+(appending a {\small\Verb_<transform>_} node).
+
+The uniform and dynamic nature makes the development experience using Denicek more akin to dynamic
+object-oriented languages than to statically-typed functional languages. The system can quickly evolve,
+but it needs suitable debugging tools, such as the source view (Fig.~\ref{fig:source}) or an
+ability to step through the history.
+
+\subsubsection*{Notebook Editing.}
+Collaborative editing based on merging of edits makes conflicts less common when compared to
+textual merging. For example, changing a method parameter can be merged with adding a call
+to a chain. However, adding two independent calls still results in a conflict. A formula
+representation that avoids nesting, as in Histogram \cite{petricek-2019-histogram}, may be more flexible.
+
+Variables are represented as references and automatic reference updating %(\S\ref{sec:system-defs})
+prevents a number of errors. References are automatically updated when the field name changes.
+Denicek also prevents the deletion of cells that contain variables to which there are references.
+
+The proof of concept nature of Datnicek means that it currently lacks suitable user interface for
+resolving conflicts. Similarly, the structure editor for code is pointer-based and would benefit from
+a better support for keyboard based input \cite{moon-2022-tylr,beckman-2023-sandblocks}.
+However, Denicek provides a suitable infrastructure for implementing this.
+
 \newpage
 
-\subsection{Reflections}
-
-edits and formula in interactive should be unified
-
-foof
-
-(data science environment)
-
-what are the limitations of Datnicek
-
-x
-
-x
-
-x
-
-x
+\subsubsection*{Evaluation Model}
+The evaluation model of Denicek implements dependency tracking and incremental recomputation,
+which aid notebook reproducibility \cite{petricek-2018-wrattler,koop-2017-dataflow}. As in
+Webnicek (\S\ref{sec:impl-incremental}), Datnicek keeps evaluated edits on top of the current history
+and use conflict detection to remove invalidated edits. We also uses the same logic for edits
+performed in interactive grid cells, suggesting that the approach is a useful general
+implementation pattern.
+
+Because invalidation of evaluated edits is not built-in into Denicek, it is worth exploring
+if Denicek could support the standard evaluation model of Jupyter notebooks based on mutable
+state.
+
+The development of Datnicek uncovered some limitations of the underlying conflict detection
+in Denicek. In particular, edits generated by evaluation (such as \ident{Add} edits that set the
+evaluated result) need to be treated as conflicting with edits that have both value and
+structure effects. This is because they are mere results of a more complex evaluation process.
+
+Data loaded in Datnicek is represented as Denicek document nodes (a table is a list of records).
+Such uniform representation makes Denicek more general, but places a high demand on the system
+performance. We implemented a number of Denicek optimisations, but Datnicek is still cumbersome
+to use with data consisting of multiple thousands of rows. Improving system performance is
+thus important future work.
+
+\subsubsection*{Future Work}
+One of the design choices discussed in \S\ref{sec:discuss} is whether to track the document
+structure explicitly. The structure of documents created in Webnicek can be irregular,
+which justified the dynamic nature of Denicek. In contrast, Datnicek notebooks are more regular.
+Replicating Datnicek using a system akin to Denicek, but based on explicit structure would
+yield an interesting comparison.
+
+Our experience with conflict detection shows the importance of getting the details of the
+implementation of programming susbstrates like Denicek right. This could be aided by the
+development of a formally tractable model of the system and proving a correctness property
+akin to that of The Gamma live previews \cite{petricek-2020-foundations}.
+
+Finally, the representation of interactive grid transformations (Fig.~\ref{fig:source})
+highlights a duplication -- a transformation is represented as a sequence of edits,
+but also as a formula. An interesting future direction is unifying or integrating those two
+kinds of computation in the Denicek substrate.
 
@@
 
-NICE THINGS IN DATNICEK
-* merging of code edits (e.g. change parameter merges with adding a call to chain)
-* remove cell to which there are refs - does the susbtrate catch this?
+\subsection{Olsen}
+Situations, Tasks, Users + Importance
+
+Generality - webnicek, denicek, map space via TDPSs
+Flexibility - dynamic nature via edits - experienced in Datnicek
+Expressive Match - possibly - for document based things
+Can it Scale up - remains a question
+
+\subsection{Technical Dimensions}
+TDPS determine a space of designs
+how much of this space can be covered by systems based on Denicek?
+
+INTERACTION
+Feedback Loops
+Modes of Interaction
+Abstraction Construction
+
+NOTATION
+Notational Structure
+
+Surface and Internal Notation
+Uniformity of Notations
+
+CONCEPTUAL STRUCTRUE
+not much to say
+Conceptual integrity vs openness - integrity
+Composability Convenience
+
+CUSTOMIZABILITY
+self-sustainability - future work
+
+COMPLEXITY
+Factoring of Complexity - no reuse for now, remains to be explored
+
+\section{Conclusion}
+
+knowledge exists only in the brain of system designers!!
 
@@
 \newpage
-
-\section{Discussion}
 

### 2025-04-04 paper.tex
Evaluation, conclusions, editing

--- 2025-03-30 paper.tex
@@
 User-centric programming research gave rise to a variety of compelling programming experiences,
-including local-first collaborative editing, programming by demonstration, incremental
+including collaborative source code editing, programming by demonstration, incremental
 recomputation, schema change control, end-user debugging and concrete programming.
@@
 above programming experiences. Denicek represents a program as a series of edits
-that construct and transform a document consisting of data and formulas. It provides three
+that construct and transform a document consisting of data and formulas. Denicek provides three
 operations on edit histories: edit application, merging of histories and conflict resolution.
-By composing these operations, we can  easily implement a range of programming experiences.
+Many programming experiences can be easily implemented by composing these three operations.
 
@@
 substrate, we use Denicek to develop an innovative interactive data science notebook system.
-The case study shows that the Denicek computational substrate provides a suitable basis for the
-design of rich, interactive end-user programming systems.
+The case study shows that the Denicek computational substrate provides a suitable basis for
+the design of rich, interactive end-user programming systems.
 \end{abstract}
@@
 A computational substrate defines the structures with which programs are constructed, how
-the progam state is represented and how it evolves \cite{jakubovic-2022-ladder}. The choice of
-a substrate affects what programming experiences can be readily supported by a system. For example,
-object-oriented programming has been historically linked to graphical user
+the progam state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
+The choice of a substrate affects what programming experiences can be readily supported.
+For example, object-oriented programming has been historically linked to graphical user
 interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
@@
 In this paper, we describe a programming substrate that makes it easy to develop diverse
-programming experiences \cite{myers-2006-eup}. Those include:
+programming experiences \cite{myers-2006-eup}. The experiences we focus on include:
 
 \begin{itemize}
-\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
-  modify a single document and automatically merge their changes, preferrably without a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
+\item \emph{Collaborative editing.} Users should be able to locally modify a shared document
+  and merge their changes, preferrably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
 \item \emph{Programming by Demonstration.} Allow users to construct simple programs by
-  enacting the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
+  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
 \item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
-  depends on it are invalidated and, possibly, automatically recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
+  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
 \item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
   data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
-\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
+\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
   why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
@@
 tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
-Using those examples, we co-design the Denicek substrate and Webnicek~(\S\ref{sec:walk}),
-a simple web-based end-user programming environment built directly on top of the substrate. Although Webnicek can
-be used to complete end-user programming tasks, it is optimized for developing the underlying
-substrate rather than for usability.
-
-To \emph{evaluate} Denicek, we use the substrate to develop Datnicek (\S\ref{sec:case}), an interactive data
-science notebook system inspired by existing environments~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
-We use the second stage to evaluate suitability of the Denicek substrate for the development of
-end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).
+Using those examples, we co-design the Denicek substrate and a simple web-based end-user programming
+environment Webnicek~(\S\ref{sec:walk}), which is built directly on top of the substrate. Although
+Webnicek can be used to complete end-user programming tasks, it is optimized for developing the
+underlying substrate rather than for usability.
+
+To \emph{evaluate} the usability of Denicek for the development of end-user programming systems,
+we use it to build Datnicek (\S\ref{sec:case}), an innovative interactive data science notebook
+that brings together a range of recent research work~\cite{kandel-2011-wrangler,drossos-2020-wrex,petricek-2022-thegamma,adams-2025-grove}.
+We reflect on the development experience~(\S\ref{sec:case-reflection}) and provide a heuristic
+evaluation of Denicek characteristics including importance and generality (\S\ref{sec:eval}).
 
@@
 The key insight that we present in this paper is that many compelling programming experiences can
-be implemented by leveraging the uniform document representation alongside with a suitable
+be implemented by leveraging this uniform document representation alongside with a suitable
 composition of the three primitive operations.
 
-Changing data and formulas is done using the same primitive edit operations that
-manipulate the document structure. A user-interface may provide a specialized editor, but still
-trigger the primitive edits behind the scenes. However, interacting with elements in the document,
+Editing of data and formulas is done using a single set of primitive edit operations that
+manipulate the document. A user-interface may provide a specialized editor, but still
+trigger the primitive edits behind the scenes. Interacting with elements in the document,
 such as a entering text in a textbox can also generate a document edit that can be merged or checked
@@
 
-As we will see, past edits that demonstrate an operation done to the
+As we will see, past edits that demonstrate an operation done with the
 document can be recorded, allowing programming by demonstration (\S\ref{sec:impl-pbd}).  Replaying
@@
 also generates document edits (\S\ref{sec:impl-eval}). If the evaluated edits conflict with manual
-edits done later by the user, the evaluated edits are removed, implementing an incremental
-recomputation (\S\ref{sec:impl-incremental}).
+edits done later by the user, the evaluated edits are removed, yielding an incremental
+recomputation behavior (\S\ref{sec:impl-incremental}).
 
@@
   web-based prototype programming system (\S\ref{sec:walk}), and discuss how the experiences are
-  implemented using the Denicek substrate (\S\ref{sec:impl}).
+  implemented on top of the Denicek substrate (\S\ref{sec:impl}).
 \item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
@@
 \item To evaluate how Denicek simplifies the development of program\-ming systems, we
-  build an innovative data science notebook Datnicek (\S\ref{sec:case})
-  and assess its impementation complexity~(\S\ref{sec:eval}).
+  build an innovative data science notebook Datnicek and assess its impementation
+  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
 \end{itemize}
@@
 \noindent
-To enable others build on top of Denicek, we share our compact and documented
-source code at: \url{https://github.com/removed/for/review} %d3sprog/denicek
+To enable others build on top of Denicek, we share our compact open-source
+implementation at: \url{https://github/removed/for/review} %d3sprog/denicek
 
@@
 Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
-notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness, interactivity
-and richness \cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
+notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness and interactivity
+\cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
 or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
@@
 part of this spectrum, also including spreadsheet systems and notebooks for data science.
-We first review related systems and their technical design (\S\ref{sec:background-sys})
-before discussing programming experiences (\S\ref{sec:background-exp}).
+% We first review related systems and their technical design (\S\ref{sec:background-sys})
+% before discussing programming experiences (\S\ref{sec:background-exp}).
 
@@
 Boxer~\cite{disessa-1986-boxer}, which introduced the \emph{naive realism} principle (what the user
-sees is all there is) that we follow in Webnicek.
+sees is all there is) that we also follow in Webnicek.
 
 To indicate that Denicek is intended as an underlying infrastructure on top of which programming
-systems can be built, we use the term \emph{computational substrate}, which also dates back to
-Boxer~\cite{disessa-1995-epistemology} and is closely related to the notion of dynamic media
-introduced by Kay and Goldberg \cite{kay-1977-media}. Webstrates \cite{klokmose-2015-webstrates}
-is a substrate based on synchronization of documents (but without edit histories) that has been
-use as the basis for multiple programming systems \cite{radle-2017-codestrates,borowski-2022-varv}.
+systems can be built, we use the term \emph{computational substrate}. The term also dates back to
+Boxer~\cite{disessa-1995-epistemology} and is related to the notion of dynamic media of Kay and
+Goldberg \cite{kay-1977-media}. Webstrates \cite{klokmose-2015-webstrates}
+revisit the idea, providing a substrate based on synchronization of documents (without edit
+histories) that has been used as the basis for multiple programming systems \cite{radle-2017-codestrates,borowski-2022-varv}.
 
@@
 edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
-edits have also been used as the basis of live programming environments \cite{storm-2013-deltas}.
-
-Merging of edits is most frequently done in version control systems such as git. Systems based
-on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
+edit histories have also been recognized as a suitable basis for live programming environments \cite{storm-2013-deltas}.
+
+Merging of edits is most frequently done in version control systems such as git. Source control ystems
+based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
 by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
@@
 
-More generally, merging of edits can be based on the operational transform (OT) approach
-\cite{davis-2002-otsgml}, where edit conflicts are reconciled,
-or conflict-free representation (CRDTs) \cite{shapiro-2011-crdts,kleppmann-2019-local,litt-2022-peritext}.
+More generally, merging of edits can use the operational transform (OT) approach
+\cite{davis-2002-otsgml}, where edit conflicts are reconciled, or conflict-free
+representation (CRDTs) \cite{shapiro-2011-crdts,litt-2022-peritext}. The latter is commonly associated
+with local-first software \cite{kleppmann-2019-local} that operates without a central server.
 In both approaches, supporting complex edits on tree structures
 remains a challenge \cite{jungnickel-2016-otjson,da-2024-jsoncrdt}. Mergeable replicated data
-types (MRDTs) \cite{kaki-2019-mrdts} can merge updates provided there is a suitable relational
-representation of the data. Recent work on MySubstrates and Grove \cite{adams-2025-grove,klokmose-2024-mywebstrates}
-has been based on CRDTs, whereas Edwards et al. \cite{edwards-2021-typed} use OT.
+types (MRDTs) \cite{kaki-2019-mrdts} merge updates using a suitable relational
+representation. Recent work on MySubstrates and Grove \cite{adams-2025-grove,klokmose-2024-mywebstrates}
+has been based on CRDTs, whereas Edwards et al. \cite{edwards-2021-typed} prefer OT.
 
@@
 
-\subsubsection*{Local-First Collaboration}
+\subsubsection*{Collaborative Editing}
 Since the early collaborative programming environments such as Collabode \cite{goldman-2011-collaborative},
 real-time collaboration has became widely used, if not always without challenges \cite{tan-2024-vslive}.
-Merging concurrent edits is one such challenge.
-In addition to CRDT-based approaches \cite{klokmose-2024-mywebstrates}, conflicts arising during
-collaboration have been solved through fine-grained locking \cite{wang-2024-nbconflicts}.
+Merging concurrent edits is one such challenge. In addition to OT-based and CRDT-based
+approaches \cite{klokmose-2024-mywebstrates}, conflicts arising during
+collaboration have also been solved using fine-grained locking \cite{wang-2024-nbconflicts}.
 
@@
 whereas more recent uses range from augmented reality prototyping \cite{leiva-2021-rapido}
-and web automation \cite{chen-2023-miwa} to end-user software customizations
-\cite{litt-2020-customization,litt-2020-wildcard}. Nevertheless, expressing conditions in PbD remains an
+and web automation \cite{chen-2023-miwa} to end-user software customization
+\cite{litt-2020-customization,litt-2020-wildcard}. Expressing conditions in PbD remains an
 active research topic~\cite{pu-2022-semanticon,radensky-2018-conditionals}. A more general class of demonstrational
-interfaces \cite{myers-2000-demonstrational} also includes programming by example, used
-for example to infer data transformations in spreadsheets \cite{gulwani-2012-spreadsheets}.
-Demonstrational systems can be used to directly perform actions, but also to generate code
-that captures the interaction, as done e.g. in WREX \cite{drossos-2020-wrex}.
+interfaces \cite{myers-2000-intelligence} also includes programming by example, used
+for example for data transformation in spreadsheets \cite{gulwani-2012-spreadsheets}.
+Demonstrational interfaces can be used to directly perform actions, but also to generate code
+as in Wrex \cite{drossos-2020-wrex}.
 
@@
 work on the Cornell Program Synthesizer \cite{teitelbaum-1981-cps}. Outside live programming,
-more work has been focused on updating computation when data change, although this can also
-be source code of a program passed to an adaptive interpreter \cite{acar-2006-adaptive}.
+more work has been focused on updating computation when data change, although such data can also
+be source code passed to an adaptive interpreter \cite{acar-2006-adaptive}.
 Incremental recomputation is also a concern in notebook systems where cells can be run out of
-order \cite{singer-2020-jollity}. This can be addressed by maintaining a dependency graph
-\cite{petricek-2018-wrattler,koop-2017-dataflow} that also allows incremental recomputation
+order \cite{singer-2020-jollity}. The ordering problem can be addressed using a dependency graph
+\cite{petricek-2018-wrattler,koop-2017-dataflow}, allowing for incremental recomputation
 on code change.
@@
 although only few systems also automatically adapt database queries \cite{wang-2019-schema}.
-However, the problem is starting to be recognized in programming systems research
+The problem is starting to be recognized in programming systems research
 \cite{litt-2020-cambria,edwards-2025-schema} as well as live programming \cite{barenz-2020-live}
@@
 Programming can be simplified by working with concrete values instead of abstractions, an idea
-pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. Self uses
-prototypes at the object level. At the expression level, similar functionality can be provided
+pioneered by the prototype-based programming language Self \cite{ungar-1987-self}. In Self,
+prototypes are used at the object level. At the expression level, similar functionality can be provided
 by managed copy \& paste. Subtext \cite{edwards-2006-copypaste,edwards-2022-copypaste} treats this
-mechanism as central, whereas other systems view tracking of copy \& paste as an additional
-editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Notably, copy \& paste has been
-also tracked in spreadsheets \cite{hermans-2015-copypaste} and Gridlets \cite{joharizadeh-2020-gridlets}
-offer an appealing alternative design.
+mechanism as central, whereas other systems view tracking of copy \& paste as an extra
+editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Copy \& paste has also been
+tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
+offer a spreadsheet abstraction based on reusing concrete computation, similar to the one proposed
+for The Gamma~\cite{petricek-2020-live}.
 
@@
 The problem of making live and rich editors compositional is addressed by
-Engraft \cite{horowitz-2023-engraft}. Programming environments may also integrate
-live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.
+Engraft \cite{horowitz-2023-engraft}. Programming environments may be also improved by integrating
+live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce
+tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.
 
@@
 experiences enabled by the Denicek substrate. Webnicek is based on a structure
-editor that supports navigation in the document and issuing of edit commands. In this section, we follow two
-formative examples (\S\ref{app:examples}) related to conference planning.
+editor that supports navigation in the document and issuing of edit commands. We follow two
+formative examples (\S\ref{app:examples}) where users collaboratively plan a conference.
 
@@
 demonstration (\S\ref{sec:impl-pbd}) by enabling the user to save and replay past interactions.
-Those can be replayed directly, or used as event handlers in order to construct interactive user
+Past interactions can be replayed directly, or used as event handlers in order to construct interactive user
 interfaces (\S\ref{sec:impl-interaction}).}
@@
 version of the document (A) and turns the list into a table. This is done by invoking a series of
-commands that change tags, wrap elements, copy and transform values.
+commands that change tags, wrap elements, copy values and transform them.
 
@@
 The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
-button are also automatically added in the new format.
+button are also automatically transformed to the new format.
 
@@
 \subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
-the initial document. Formulas are represented as special {\small\Verb_x-formula_} nodes whose arguments
-are other nodes or references to other nodes or formulas (highlighted on mouse hover).
-
-\keyideabox{\faLightbulbO}{Incremental Recomputation}{Formulas are document nodes and
-evaluating a formula yields additional edits that augment the document with the result
+the initial document. Formulas are represented as nodes with the {\small\Verb_x-formula_} tag,
+whose arguments are other nodes, references to other nodes, or formulas (highlighted on mouse hover).
+
+\keyideabox{\faLightbulbO}{Incremental Recomputation}{Formulas are document nodes.
+Evaluating a formula yields edits that augment the document with the result
 (\S\ref{sec:impl-eval}). Those evaluated edits are kept at the top of the document history and are
-removed in case of a conflict (\S\ref{sec:impl-incremental}), resulting in the incremental
-recomputation behavior.}
-
-\subsubsection*{\circled{G} Merging Formulas} When the budget calculation is merged with the
-other edits, references in formulas are automatically updated to point to the new list.
+removed in case of a conflict (\S\ref{sec:impl-incremental}), a logic that implements
+incremental recomputation using core Denicek operations.}
+
+\subsubsection*{\circled{G} Merging Formulas} When the budget calculation (F) is merged with the
+earlier edits (E), references in formulas are automatically updated to point to the new list.
 Adding new speaker via the ``Add speaker!'' button invalidates the evaluated result.
@@
 references (to denote a target node), but relative references can appear in the document (e.g., to
-refer to another node within a list element).
+refer to another node within the same list element).
 
@@
 \ident{Any} selector makes it possible to refer to all elements of a list, but there is no
-way to refer to all fields of a record. Note that Denicek does not use implicit numerical
-indices for lists. The index of a new list item has to be supplied explicitly. The reasoning behind
+way to refer to all fields of a record. Because Denicek does not use implicit numerical
+indices for lists, the index of a new list item has to be supplied explicitly. The reasoning behind
 this design choice is discussed in \S\ref{sec:discuss}.
@@
 \subsubsection*{Document Edits}
-The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
+The supported document edits and their behaviors are listed in Fig.~\ref{fig:edits}. All edits
 require a \emph{target} to which they are applied. Targets are absolute references not containing the
@@
 
-The edits allow an arbitrary transformation of a document via a series of steps whose effect can be
-tracked by the substrate. As we saw, Denicek updates references when document structure changes.
+The edits can transform the document structure. Denicek tracks the effect of the edits on the
+structure and updates references when the document structure changes.
 Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
-(above) and edits that affect references (below).
+(above) and edits that can affect references (below).
 
@@
 As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
-statically enforced (the substrate can be seen as dynamically typed). As a consequence, edits
-that transform the structure can be used as structural edits (affecting the structure)
-or as value edits (affecting the value of specific nodes). The latter can be used, for example,
-when constructing an additional list item. An item can be first constructed and then populated
+statically enforced (the substrate can be seen as being dynamically typed). As a consequence, edits
+that transform the structure can serve both as structural edits (affecting the structure)
+and as value edits (affecting the value of specific nodes). The latter can be used, for example,
+when constructing an additional list item. An item is first constructed and then populated
 with values. This temporarily violates the invariant that collections are homogeneous
-(an issue we might address by using transactions as discussed in \S\ref{sec:discuss}).
+(an issue that could be address by using transactions as discussed in \S\ref{sec:discuss}).
+
 Another use of value edits is when evaluating a formula, which involves wrapping the formula
-node (\S\ref{sec:impl-eval}), but the operation should not affect references to the formula
+node (\S\ref{sec:impl-eval}), an operation that should not affect references to the formula
 itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
-have a \emph{reference behavior} option that can be set to disable automatic reference
+have a \emph{reference behavior} option that can disable automatic reference
 updating. This annotation is required when the target contains the \emph{Index} selector,
@@
 An important principle of the Denicek design is that the effect an edit has on references inside
-the document does not depend on the current value of the document, only its type. This makes it possible to
+the document does not depend on the current value of the document. This makes it possible to
 define merging solely in terms of edits, without reference to current document state.
@@
 the relevant references according to the rules shown in Fig.~\ref{fig:updates}, provided
-that the reference behavior of the edit is not set to disable this.
+that the reference behavior of the edit does not prevent updating.
 %(this is required when %the target reference contains the \ident{Index} selector).
@@
 example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|}, a reference
-{\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (Contrariwise, a reference in the
-document cannot be more general. A more specific edit would have to contain the \ident{Index} selector
+{\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (A reference in the
+document cannot be more general; a more specific edit would have to contain the \ident{Index} selector
 and this would require setting reference behavior to not trigger reference update.) Finally,
@@
 Merging edit histories is used when two users edit document independently, but also
-when replaying edits in programming by demonstration. The operation $\mathcal{M}_E(E_1, E_2)$
-works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix $E$.
-The operation is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that can be reapplied
-on top of the other edit history. Formally, $\mathcal{M}_E(E_1, E_2) = E, E_1, E_2'$.
-Note that the operation is not symmetric. $\mathcal{M}_E(E_2, E_1) = E, E_2, E_1'$ and the
+when replaying edits in programming by demonstration. The merge operation $\mathcal{M}_E(E_1, E_2)$
+works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix~$E$.
+Our merging is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that can be reapplied
+on top of the other edit history. That~is, $\mathcal{M}_E(E_1, E_2) = E, E_1, E_2'$.
+Note that the operation is not symmetric: $\mathcal{M}_E(E_2, E_1) = E, E_2, E_1'$. The
 result of applying the two histories to the same node will differ if there are conflicts
@@
 
-The key operation that enables such reconciliation takes two individual edits that occurred
+The key operation that enables edit history reconciliation takes two individual edits that occurred
 independently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
@@
   $e_1$ modified the document through a selector that would also affect the new nodes added by $e_2$,
-  we need to apply the transformation represented by $e_1$ to the nodes newly added by $e_2$.
+  we need to apply the transformation represented by $e_1$ to the nodes newly added by~$e_2$.
   This is done by generating an additional edit, to be applied after $e_2$, that is based on
@@
 reflect the wrapping (2) and we need to create and additional \ident{Add} operation
-that will add the checkbox to the new item (1). The result is a sequence with two addtional edits $e_2', e_2''$:
+that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
+that target the newly wrapped list and add the checkbox to the added item:
 %
@@
 \includegraphics[width=0.95\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
+\vspace{-0.5em}
 \caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
 \label{fig:merging}
+\vspace{-0.5em}
 \end{figure}
@@
 a node in incompatible ways or one edit modifies a node deleted by the other, the merge operation
-cannot reconcile. Such conflicting edits can be detected, reported to the user and optionally
+is not symmetric. Such conflicting edits can be detected, reported to the user and optionally
 removed. Conflict detection is also used to implement incremental evaluation~(\S\ref{sec:impl-incremental}).
@@
 systems~\cite{lucassen-1988-effects}. The mechanism is simple and tractable, but over-approximates
-conflicts, i.e.~it may report a conflict even if two edits can be merged successfully
-(see \S\ref{sec:discuss}). Effects describe how an edit affects the document structure
+conflicts, i.e.~it may report a conflict even if two edits can be merged successfully.
+Effects describe how an edit affects the document structure
 and we distinguish between three types of effects as shown in Fig.~\ref{fig:effects}.
@@
 \S\ref{sec:impl-incremental}. If a conflict is detected, the edit $e_2$ is removed and its effect
-is recorded, so that we remove any subsequent edits that would depend on the removed edit.
+is recorded, so that we remove any subsequent edits that depend on the removed edit.
 
@@
 
-The key claim of this paper is that the Denicek computational substrate suppports a range of
-compelling user experiences, many of which crucially rely on Denicek's ability to merge edits.
-We demonstrate this claim with the Webnicek web-based programming system, which uses Denicek to support:
-%
-\begin{enumerate}[itemsep=1.5pt,label=(\roman*),topsep=6pt,leftmargin=20pt]
-\item local-first collaboration (\S\ref{sec:impl-collab}),
+The key claim of this paper is that the Denicek computational substrate makes it easy to support
+a range of compelling user experiences. The experiences can be implemented by composing the
+primitive operations of the substrate, relying in particular on the operation for merging edits.
+We demonstrate this claim with the web-based Webnicek system, which uses Denicek to support:
+%
+\begin{enumerate}[itemsep=0pt,label=(\roman*),topsep=3pt,leftmargin=20pt]
+\item collaborative document editing (\S\ref{sec:impl-collab}),
 \item programming by demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
@@
 %
-We describe the programming experience in isolation in the context of Webnicek in this section.
+We describe the programming experiences in isolation in the context of Webnicek in this section.
 The data science notebook system Datnicek, discussed in \S\ref{sec:case}, provides
-a more comprehensive evaluation by combining multiple user experiences together.
-
-\subsection{Local-First Collaboratation}
+a more comprehensive case study by combining multiple user experiences together.
+
+\subsection{Collaboratative Editing}
 \label{sec:impl-collab}
 
-The Denicek substrate enables local-first collaboration~\cite{kleppmann-2019-local},
-as illustrated earlier (\S\ref{sec:walk}, E). If a document is edited by multiple users, they can
-each make edits to their local copy and eventually merge the variants using the operation to
-merge edit histories.
+Denicek enables collaborative editing as illustrated earlier (\S\ref{sec:walk}, E).
+If a document is edited by multiple users, they can each make edits to their local copy and
+eventually merge the variants using the operation to reconcile edit histories. Merging requires
+coordination among users, but not a central server as in local-first software~\cite{kleppmann-2019-local}.
 
@@
 
-As an example, consider the scenario in Fig.~\ref{fig:merging}. Here, document $D$ can be obtained
+Consider the scenario in Fig.~\ref{fig:merging}. Here, the document $D$ can be obtained
 either by appending $C'$ (produced by the edit reconciliation operation) on top of $A,B$ or by
@@
 
-According to our effect analysis, the two operations are conflicting.
+According to the current implementation of our effect analysis, the two example histories are conflicting.
 Although $B$ primarily affects the document structure, it also adds a new field to the record (email),
@@
 can also be resolved by removing conflicting edits.
-
-~
 
@@
 repetition with Denicek (Fig.~\ref{fig:pbd}), the user selects edits from the edit history,
-name them and replay them. For general-purpose document editing in Webnicek, this requires
-certain forethought, but as illustrated in \S\ref{sec:case}, the mechanism is very effective
+names them and replays them. For general-purpose document editing in Webnicek, this requires
+certain forethought, but as shown in \S\ref{sec:case}, the mechanism is very effective
 in a domain such as data wrangling \cite{kandel-2011-wrangler}.
@@
 as nodes and storing them in a list inside the {\small\Verb_/saved-interactions_} field).
-This means that no other implementation mechanism outside of the system is needed and also
+This means that no other implementation mechanism outside of Denicek is needed and also
 that the stored edits can be modified by the user or tools working with the document.
 
-Second, to replay edits, Webnicek does not append the recorded edits on top of the current history.
-When saving edits, it stores the hash of the history at the time of saving. When replaying,
-it appends the recorded edits to the top of the original history (at the time of saving) and merges
-this new sequence of edits with the current history. This pushes the recorded edits through all
-subsequent edits made by the user. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
-a newly added speaker is transformed from a list item to a table row.
+Second, when replaying edits, Webnicek does not append the recorded edits on top of the current
+history. Instead, it stores the hash of the history at the time of saving. To replay the edits,
+it then appends the edits to the top of the original history and merges this edit sequence
+with the current history. This pushes the recorded edits through later edits made by the user,
+reconciling them with potential structural edits. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
+a newly added list item becames a table row.
 
@@
 handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
-executes the edits using the mechanism discussed in \S\ref{sec:impl-pbd}, i.e., Webnicek appends
+executes the edits using the mechanism discussed in the previous section, i.e., Webnicek appends
 the edits to a history at the time when the edits were recorded and merges the edits with the
@@
 In addition to the core functionality provided by Denicek, the Webnicek system also makes it possible
-to generalize the interactions recorded through programming by demonstration in a limited way. As shown
+to generalize the interactions recorded through programming by demonstration. As shown
 in Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
@@
 
-Note that the generalization mechanism does not violate the Denicek principle discussed in \S\ref{sec:system-defs}
-that edits cannot depend on values. Webnicek finds all nodes for which the condition
+Note that the generalization mechanism does not violate the Denicek principle that edits cannot
+depend on values (\S\ref{sec:system-defs}). Webnicek finds all nodes for which the condition
 holds and generates one specific (non-conditional) edit for each of the nodes.
@@
 Programming by demonstration systems typically implement heuristic for generalization
-\cite{myers-2000-intelligence} infers and suggests such generalizations. If integrated into
+\cite{myers-2000-intelligence} that infers and suggests such generalizations. If integrated into
 Webnicek, such heuristic could for example automatically construct a formula
-based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}
+based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected Todo
+list items).}
 
@@
 
-Formula evaluation leverages Denicek's ability to merge edit histories.
+Formula evaluation also leverages Denicek's ability to merge edit histories.
 As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
@@
 document. We then perform three edits that wrap the current counter value in a formula that adds
-1 to the value. The three edits are used as an event handler for the button and so clicking the
+1 to the value. The three edits (5) are used as an event handler for the button and so clicking the
 button creates an increasingly long sequence of increments. One advantage of this approach that
-we will leverage later is that the system also records a trace of how the final computed value
+we will leverage later is that the evaluation also records a trace of how the final computed value
 was obtained.
@@
 Webnicek exposes the underlying representation of formulas to the user, but the same representation
-can be edited through a user-friendly mechanism such as a textual calculation view \cite{sarkar-2018-calc}
-a block-based editor \cite{jansen-2019-xlblocks}. The key point
+could be edited through a user-friendly mechanism such as a textual calculation view \cite{sarkar-2018-calc},
+or a block-based editor \cite{jansen-2019-xlblocks}. The key point
 is that Denicek's tree structure makes it easy to embed formulas in document in a uniform
-way, edit them as any other node and merge them with other changes.}
+way, edit them as other nodes and merge edits that change them.}
 
@@
 appended to the top of the current edit history. When the user makes subsequent edits, the evaluated
-edits are pushed through the newly added edits. If the edits conflict (according to the conflict
-detection discussed in \S\ref{sec:system-ops}), the affected evaluated edits and edits that
-depend on them are dropped.
+edits are pushed through the newly added edits using edit reconciliation. If the edits conflict
+(according to the conflict detection discussed in \S\ref{sec:system-ops}), the affected evaluated
+edits are dropped.
 
 The dependencies between edits alongside with the conflict detection mechanism provide functionality
-that other systems implement using a dependency graph using topological sort.
+that other systems implement using an explicit dependency graph or by topological sorting.
 
@@
 mechanism, but a realistic system based on the substrate could automatically evaluate
-formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}
+formulas to provide a live programming experience \cite{petricek-2020-live,rein-2019-live}
 and use incremental recomputation for performance~reasons.}
@@
   \begin{minipage}{0.45\columnwidth}
-    \caption{Evaluated edits (A, B) are kept at the top of the history. When they conflict with
-    ordinary edits (4), they are removed (A).}
+    \caption{Evaluated edits (A, B) are kept at the top of the history. Evaluated edits (A)
+      that conflict with ordinary edits (4) are dropped.}
     \label{fig:eval}
@@
 
-The implementation leverages the fact that evaluation keeps the original formula, as discussed
-in \S\ref{sec:impl-eval}. When a formula is evaluated, the final {\small\Verb_<x-evaluated>_}
-document node contains the result, but also sub-tree that represents the full evaluation trace
+The implementation leverages the fact that evaluation wraps the previous state of the formula
+(\S\ref{sec:impl-eval}). When a formula is fully evaluated, the {\small\Verb_<x-evaluated>_}
+document node contains the result, but also a sub-tree with the full evaluation trace
 \cite{perera-2012-functional}. We analyse the trace, collect all reference nodes in the trace and
 highlight all nodes referred to in the computation. Denicek makes this easy as we
-need to collect references directly nested in the formula node.
+can collect references directly nested in the formula node.
 
@@
 trace collected by the Denicek substrate can also be used to provide a detailed
-explanation~\cite{perera-2012-functional} or, for example, to automatically construct
-linked visualizations~\cite{perera-2022-linked}.}
+explanation~\cite{perera-2012-functional} or to automatically construct
+a linked visualization~\cite{perera-2022-linked}.}
 
@@
 cost~\cite{blackwell-2002-attention}. Programming by demonstration (\S\ref{sec:impl-pbd})
-offers one way of reducing the cost. Another way of making programming more
-concrete~\cite{edwards-2004-example,smith-1975-pygmalion} is to support copying of functionality
+offers one way of reducing the cost. Another way is making programming more
+concrete~\cite{edwards-2004-example,smith-1975-pygmalion}, i.e., to support copying of functionality
 as in prototype-based object-oriented programming~\cite{ungar-1987-self,randall-1995-self}.
@@
 
-In Figure~\ref{fig:copypaste}, the user copies an incorrect formula using
+In Fig.~\ref{fig:copypaste}, the user copies a formula with switched operands using
 the \ident{Copy} edit and then modifies it to use a different data source. They then navigate back
 in history to the point before the copying and create a temporary fork of the document. In the
-fork, they use \ident{RenameField} to switch the argument of the division. When they merge the
+fork, they use \ident{RenameField} to switch the operands of the division. When they merge the
 temporary fork into the original document, the \emph{Apply to Newly Added} logic of the merge
-operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and also applies them
-to the copied formula. The merging operation with the \ident{Copy} edit and the fact that formulas
-are ordinary document nodes, provide keys component for a straightforward implementation of the
+operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and applies them
+to the copied formula. Merging with the \ident{Copy} edit and the fact that formulas
+are ordinary document nodes, provide the key component for a straightforward implementation of the
 managed copy \& paste functionality.
@@
 \keyideabox{\faClipboard}{Linked Editing}{Webnicek currently requires users to explicitly
-manipulate history to correct error across multiple code clones. Research on managing duplicated
+manipulate history to correct errors across multiple code clones. Research on managing duplicated
 code resulted in multiple tools \cite{toomim-2004-linked,duala-ekoko-2008-clone} with dedicated
@@
 \begin{figure}[t]
+\vspace{-0.5em}
 \includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
+\vspace{-1em}
 \caption{When an edit changes the document structure, references in formulas are updated accordingly.}
 \label{fig:coevolution}
-\vspace{-1em}
+\vspace{-0.5em}
 \end{figure}
@@
 adapted the Denicek design and revisited the implementation of the Webnicek system until we obtained
-a satisfactory solution for the six formative examples (\S\ref{app:examples}). In this section, we
-document the design challenges, many of which are shared with related systems
+a satisfactory solution for the six formative examples detailed in \S\ref{app:examples}. In this
+section, we document the design challenges, many of which are shared with related systems
 \cite{jakubovic-2022-ladder,edwards-2005-subtext,hall-2017-infra,omar-2021-livelits}.
@@
 scenarios using a small number of concepts. This is apparent in the design of the document
-structure---a node can represent data, code or rich text---as well as in the role of
-edits---an edit can be a value or structure change, the result of user interaction or the
-result of evaluation.
+structure -- a node can represent data, code or rich text -- as well as in the role of
+edits -- an edit can be a value change, a structure change, the result of user interaction
+or the result of formula evaluation.
 }
@@
 \subsubsection*{Dynamic and Static Typing}
-Denicek does not explicitly track the document structure (it is dynamically typed). We assume that lists
-are homogeneous, but do not enforce this property. A statically typed system could distinguish
-structural edits from value edits. This has a theoretical appeal and it simplifies aspects
-of the implementation, e.g. by removing the need to control reference updating, but it makes
-working with the system less akin to document editing.
-
-We violate structural invariants when constructing values gradually (e.g., when adding a new
-speaker to a table), as well as in evaluation (wrapping {\small\Verb_<x-formula>_}
+Denicek does not explicitly track the document structure (it is dynamically typed). For example, we
+assume that lists are homogeneous, but do not enforce this property. A statically typed system could
+enforce such properties and explicitly distinguish structural edits from value edits. This has a
+theoretical appeal and it simplifies aspects of the implementation, e.g. by removing the need to
+control reference updating, but it makes working with the system less akin to document editing.
+
+The structural invariants are violated when constructing values gradually (e.g., when adding a new
+speaker to a table), as well as during evaluation (wrapping {\small\Verb_<x-formula>_}
 in {\small\Verb_<x-evaluated>_}). A system with static types could use a form of edit
-transactions, where structural invariants are reestablished only after a sequence of edits.
+transactions, where structural invariants are reestablished after a sequence of edits.
 
@@
 \subsubsection*{Expressive Selectors and Edits}
-Our merge operation operates on two sequences of edits, but it does not need access
+Our merge operation works on a pair of edit sequences, but it does not need access
 to the current document state or history. This limits the expressiveness of edits. Denicek does not
@@
 The design supports a programming by demonstration scenario where the user adds a new list item
-and then modifies it (adding a speaker or a Todo item).
-
-With numerical indices, the edits following \ident{Append} would not have access to the index of
-the newly added item. Using a numerical index computed from the list length would require
-complex logic to update indices when merging edit operations that affect lists. Another alternative
-is to disallow modification of a newly added item, but this limits programming by demonstration.
+and then modifies it (e.g.,~when adding a speaker or a Todo item).
+
+With numerical indices, the edits following \ident{Append} would not have a direct access to the
+index of the added item. Computing a numerical index from the length of the list would require
+a complex logic to update indices when merging edit operations that affect lists. An alternative
+is to disallow modification of the newly added item, but this makes programming by demonstration
+cumbersome.
 
@@
 
-\subsubsection*{Merging Structure and Capabilities}
+\subsubsection*{Structure and Capabilities of Merging}
 Denicek keeps a linear history and merging appends edits to the top of the history.
@@
 where users have to merge all changes in order and cannot adopt selected edits (cherry
-picking in git). In the \emph{divergence} model, users could keep their own schema but
+picking in git). The \emph{divergence} model would let users keep their own schema but
 import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
-that is dual to our edit reconcilliation (given subsequent edits $e_1, e_2$, we want
-to generate $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).
+that is dual to our edit reconcilliation (given subsequent edits $e_1, e_2$, generate
+$e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).
 
@@
 (\S\ref{sec:impl-eval}) may be \ident{Copy} edits (when evaluating a reference), but also
-\ident{Add} edits (when setting the evaluation result to a computed value). For those edits, we
-record additional dependencies identifying the source of the operation arguments. Those additional
-dependencies also need to be considered in conflict detection.
+\ident{Add} edits (when setting the evaluation result to a computed value). \ident{Add} edits
+need to record additional dependencies identifying the source of the operation arguments.
+Alternative evaluation models avoid this need, but lead to a less uniform system design.
 
@@
 \caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
-a code cell (1), edits it in a grid that infers edit operations via programming by demonstration
-(2), turned to a code cell (3) and then visualized.}
+a code cell (1) and edits it in a grid that infers edit operations via programming by demonstration
+(2). They turn the edits into a code cell (3) and add a chart.}
 \label{fig:datnicek}
@@
 
-Denicek is a low-level computational substrate. It serves best as the basis for interactive
+Denicek is a low-level computational substrate. It is intended as the basis for interactive
 programming systems that view programs as documents. The most prominent example of such systems
 today are notebook environments for data science. To explore this use case, we have developed
-Datnicek, a simple notebook system that shows the ability of Denicek to support rich interactive user
+Datnicek, a notebook system that shows the ability of Denicek to support rich interactive user
 experiences. In this section, we present Datnicek and reflect on its development. A simple data
@@
   structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
-  We do not yet aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.
+  We do not initially aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.
 
@@
 
-\item \emph{Interactive Data Cleaning.} It should be possible to edit data tables in an interactive
-  grid and use programming by demonstration for common data cleaning tasks as in
+\item \emph{Interactive Data Cleaning.} It should be possible to edit tabular data in an interactive
+  grid and use programming by demonstration for common cleaning tasks as in
   Wrangler or Vizier~\cite{kandel-2011-wrangler,kennedy-2022-vizier}.
@@
 \item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
-  turned into code that can be checked by the data scientist and further edited as in Wrex \cite{drossos-2020-wrex}.
+  convertible into code that can be checked by the data scientist and further edited as in Wrex \cite{drossos-2020-wrex}.
 \end{itemize}
@@
 of the list of Denicek edits, transient state of the user interface, current computed document
-and other cached information. All interactions that affect the notebook
-use an event that appends edits to the list of Denicek edits. Remaining events
-update the transient user interface state. The support for collaborative editing has been
+and other cached information. All interactions that affect the notebook trigger
+the same type of event that appends edits to the list of Denicek edits. Remaining events
+update the transient user interface state. The support for collaborative editing is
 implemented through an ad-hoc mechanism.
@@
 \subsubsection*{User Interface}
-Datnicek notebooks consists of a series of cells. In the code cells, the user can
+Datnicek notebooks consists of a series of cells. In code cells, the user can
 use the ``+'' button to add a new variable, add an operation to the end of an existing chain
@@
 
-A transformation may correspond to multiple underlying Denicek edits. For example, splitting
-``foo/bar'' using ``/'' as the delimiter adds a new column ``bar'', copies values from
+A single transformation may correspond to multiple underlying Denicek edits. For example, splitting a
+column named ``foo/bar'' with values of the form ``7/5'' using ``/'' as the delimiter adds a new column ``bar'', copies values from
 ``foo/bar'', renames ``foo/bar'' to ``foo'' and then transforms the values in the new columns
@@
 
-Interactive grid also supports conditional transformations that affect rows satisfying a given
+Interactive grid also supports transformations that affect rows satisfying a given
 condition (e.g., deleting rows with a specific value in a given column). As discussed
-in \S\ref{sec:system-defs}, Denicek does not allow specifying conditional selectors.
-To support conditional transformations, we add a special kind of (virtual) edit
+in \S\ref{sec:system-defs}, Denicek does not support conditional selectors.
+To implement conditional transformations, we add a special kind of (virtual) edit
 {\small\Verb_<x-expandable-edit>_} that stores the underlying edit and a condition.
@@
 
-\newpage
-
 \subsection{Reflections}
-
-edits and formula in interactive should be unified
-
-foof
-
-(data science environment)
-
-what are the limitations of Datnicek
-
-x
-
-x
-
-x
-
-x
+\label{sec:case-reflection}
+
+The development of Datnicek was only started after the design and implementation of Denicek and
+Webnicek was completed. Thanks to this two-phase methodology, the experience reported here
+provides a qualitative evaluation of Denicek's capabilities. It also highlights limitations and
+suggests directions for future work.
+
+\subsubsection*{Effectiveness and Unifromity.}
+Many of the Datnicek requirements are addressed by Denicek with minimal development
+effort. Denicek also provides a small and uniform set of primitives. In particular,
+the implementation represents all non-transient changes to a notebook as Denicek edits.
+Code completions offered by the code editor are edits (wrapping a formula or appending a parameter),
+changing a primitive value is an edit, and recommendations in the interactive grid are also edits
+(appending a {\small\Verb_<transform>_} node).
+
+The uniform and dynamic nature makes the development experience of using Denicek more akin to dynamic
+object-oriented languages than to statically-typed functional languages. The system can quickly evolve,
+but it needs suitable debugging tools, such as the source view (Fig.~\ref{fig:source}) or an
+ability to step through the history.
+
+\subsubsection*{Notebook Editing.}
+Collaborative editing based on merging of edits makes conflicts less common when compared to
+textual merging. For example, changing a method parameter can be merged with adding a call
+to a chain. However, adding two independent calls still results in a conflict. A formula
+representation that avoids nesting \cite{petricek-2019-histogram} may eliminate an even larger
+proportion of conflicts.
+%
+Variables are represented as references and automatic reference updating %(\S\ref{sec:system-defs})
+prevents a number of errors. References are automatically updated when the field (variable) name changes
+and Denicek prevents the deletion of cells that contain variables to which there are references.
+
+The proof of concept nature of Datnicek means that it currently lacks suitable user interface for
+resolving conflicts. Similarly, the structure editor for code is pointer-based and would benefit from
+a better support for keyboard-based input \cite{moon-2022-tylr,beckman-2023-sandblocks}.
+However, Denicek provides a suitable infrastructure for implementing this.
+
+\subsubsection*{Evaluation Model}
+The evaluation model of Denicek implements dependency tracking and incremental recomputation,
+which aid notebook reproducibility \cite{petricek-2018-wrattler,koop-2017-dataflow}. As in
+Webnicek (\S\ref{sec:impl-incremental}), Datnicek keeps evaluated edits on top of the current history
+and uses conflict detection to remove invalidated edits. We uses the same logic to evaluate
+edits performed in interactive grid cells, suggesting that the approach is a useful general
+implementation pattern.
+
+However, as invalidation of evaluated edits is not built into Deni\-cek, it is worth investigating
+if Denicek could support the standard evaluation model of Jupyter notebooks based on mutable state.
+
+The development of Datnicek uncovered some limitations of conflict detection in Denicek.
+In particular, edits generated by evaluation (such as \ident{Add} edits that set the
+result) need to be treated as conflicting with overlapping edits with both value and structure
+effects. This is because evaluated edits are the result of evaluation that considers both value
+and structure of source nodes.
+
+Finally, data loaded in Datnicek is represented as Denicek nodes (a table is a list of
+records). Such uniform representation makes De\-ni\-cek more versatile, but places a high demand on the
+system performance. We implemented a few basic Denicek optimisations, but Datnicek is still cumbersome
+to use with data consisting of multiple thousands of rows. Improving system performance is
+thus an important direction for future work.
+
+\subsubsection*{Future Work}
+One of the design choices discussed in \S\ref{sec:discuss} is whether to track the document
+structure explicitly. The structure of documents created in Webnicek can be irregular,
+which justified the dynamic nature of Denicek. In contrast, Datnicek notebooks are more regular.
+Replicating Datnicek using a system akin to Denicek, but based on explicit structure would
+yield an interesting comparison.
+
+Our experience with conflict detection shows the importance of getting the details of the
+implementation of programming susbstrates like Denicek right. This could be aided by the
+development of a formally tractable model of the system and proving a correctness property
+akin to that of The Gamma live previews \cite{petricek-2020-live}.
+
+Finally, the representation of transformations in interactive grid cells (Fig.~\ref{fig:source})
+highlights an unnecessary duplication -- a transformation stores a sequence of edits, as well as a formula.
+An interesting future direction is unifying or integrating those two notions of computation
+provided by the Denicek substrate.
 
@@
 \label{sec:eval}
-how well did Denicek work for implementing Datnicek
-
-NICE THINGS IN DATNICEK
-* merging of code edits (e.g. change parameter merges with adding a call to chain)
-* remove cell to which there are refs - does the susbtrate catch this?
-
-\newpage
-~
-\newpage
-
-\section{Discussion}
+
+The Datnicek case study provides a primary qualitative evaluation of Denicek. We complement this
+with heuristic evaluation according to the criteria proposed by Olsen \cite{olsen-2007-evaluation}
+and through the technical dimensions of programming systems framework (TDPS) developed by Jakubovic
+et al. \cite{jakubovic-2023-techdims}. Criteria and dimension names are in bold.
+
+\subsection{Complex System Evaluation}
+First and foremost, Olsen \cite{olsen-2007-evaluation} argues that a system must illustrate
+its \textbf{importance} for a particular class of users performing certain tasks in a given situation.
+Denicek is designed to support programmers and researchers developing novel interactive programming systems,
+an active research area, arguably constrained by the dependence on existing programming languages and systems.
+
+To support a wide range of novel systems, Denicek must satisfy Olsen's criterion of \textbf{generality}.
+We show that Denicek is suitable for building two very different kinds of programming systems, Webnicek
+and Datnicek. We further delineate the space of supported programming systems in the next section using TDPS.
+
+Denicek makes it easy to try different solutions, satisfying the \textbf{flexibility} condition.
+This was experienced during the development of Datnicek, where a uniform representation (everything
+is an edit) made it possible to quickly prototype different designs.
+
+For an emerging class of programming systems built around documents, such as Potluck \cite{litt-2020-potluck}
+and notebooks for data science, Denicek provides \textbf{expressive match} by being nearer to the
+problem being solved, i.e., by using document representation for a programming system built around
+documents.
+
+Denicek also satisfies the \textbf{inductive combination}
+criterion by offering a small set of primitives from which different designs can be built.
+Specifically, we used Denicek to build a document-based programming system, interactive
+grid for data cleaning and a structure editor with contextual code completion.
+
+Several Olsen's criteria point to limitations of our work. First, it is unclear if Denicek
+can empower new participants to be involved in programming system design. Second, the Datnicek
+case study did not address scalability, although there is arguably a number of domains for which
+the capabilities of the system already suffice.
+
+\subsection{Technical Dimensions}
+The technical dimensions (TDPS) framework \cite{jakubovic-2023-techdims} maps the design space of
+interactive programming systems. We use the framework to characterize the generality of the
+substrate.  For some dimensions, Denicek can cover the full range of options, while for others,
+it is limited to one fixed design or a subset of options.
+
+\subsubsection*{Interaction}
+In both Datnicek and Webnicek, there is one primary \textbf{mode of interaction}. Both
+systems offer immediate feedback during editing, meaning there is one main \textbf{feedback loop}.
+Denicek could be used to build systems with multiple modes of interaction and feedback loops
+(e.g., by separating editing, checking of edits and evaluation), but it is unsuitable for live
+systems that preserve evaluation state during source code editing \cite{burckhardt-2013-live}.
+
+Denicek covers the full range of \textbf{abstraction construction} approaches.
+Abstraction \emph{from concrete cases} can be implemented via programming by demonstration as
+shown in \S\ref{sec:impl-pbd} and \S\ref{sec:case}. Although Datnicek code cells do not currently
+support function declarations, these could be added to provide abstraction \emph{from the
+first principles}.
+
+\subsubsection*{Notation}
+Denicek supports both \textbf{notational structures}. \emph{Complementing notations} use
+distinct representations for different aspects of a program. For example, Datnicek combines code cells with
+interactively constructed transformations in grid cells. \emph{Overlapping notations} allow editing
+of the same underlying structure in multiple ways. Although not integrated in a single system,
+formula editor in Webnicek and code editor in Datnicek illustrate this option.
+
+Even though Denicek's underlying conceptual structure is based on composability with a small number of
+primitives, this does not determine \textbf{uniformity of notations} of systems built on top
+of Denicek. These can support representations formed by a small set of primitives, as well as
+by a large set of domain-specific constructs.
+
+In Denicek, the \textbf{surface and internal notation} are typically closely related. The
+internal notation has an explicit structure (document nodes), which can directly encode the surface
+notation. Using an unstructured representation (such as text) is possible, but would be
+incompatible with other Denicek capabilities such as merging.
+
+\subsubsection*{Conceptual Structure and Customizability.}
+On the scale of \textbf{conceptual integrity vs. openness}, the Denicek substrate favors integrity.
+Systems based on Denicek need to use its representation of documents and document edits. This reduces
+complexity, but leads to incompatibility with existing software stacks.
+
+Systems built using Denicek currently do not exhibit the \textbf{self-sustainability} property,
+i.e., the ability to be modified from within themselves. Supporting this is an appealing
+possibility. A further research on the Denicek computation model should attempt to make it possible
+to implement formula evaluation as, for example, conditional edits recorded in the document itself
+through programming by demonstration.
+
+% ==================================================================================================
+
+\section{Conclusion}
+Many recently developed programming experiences make the task easier, less error-prone,
+more direct and more collaborative. We believe that many more research advances are possible.
+Alas, novel programming experiences are difficult to implement on top of
+existing programming languages. A suitable computational substrate can
+significantly reduce the implementation effort needed to implement novel programming experiences.
+
+To this end, we present Denicek, a computational substrate that represents programs as sequences
+of document edits. We describe the design of Denicek and show how it supports a range of
+programming experiences. A remarkable property of the design is that many programming experiences are
+implemented through a straightforward combination of three basic operations provided by Denicek, namely edit
+application, merging of histories and conflict~detection.
+
+To evaluate the usability of Denicek for the development of interactive programming systems,
+we use it as the basis for Datnicek, a data science notebook system that supports a range of
+programming experiences pioneered in recent research systems. We also assess the substrate and
+its generality through heuristic evaluation.
+
+The design of Denicek resolves a number of challenges that have been faced by authors of related
+systems, but have never been explicitly documented. While we hope researchers will adopt
+Denicek as a foundation for future innovative programming systems, an equally important
+contribution of our work is that it documents this existing but unwritten tacit knowledge.
+
+% ==================================================================================================
 
@@
 Programming by Demonstration is often used in data wrangling
-\cite{gulwani-2012-spreadsheet,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
+\cite{gulwani-2012-spreadsheets,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
 is only a minimalistic illustration of such use, loosely inspired by earlier work \cite{miller-2001-simult}.

### 2025-04-04 paper.tex
Spellcheck

@@
 A computational substrate defines the structures with which programs are constructed, how
-the progam state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
+the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
 The choice of a substrate affects what programming experiences can be readily supported.
@@
 \item \emph{Collaborative editing.} Users should be able to locally modify a shared document
-  and merge their changes, preferrably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
+  and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
 \item \emph{Programming by Demonstration.} Allow users to construct simple programs by
@@
 \item To evaluate how Denicek simplifies the development of program\-ming systems, we
-  build an innovative data science notebook Datnicek and assess its impementation
+  build an innovative data science notebook Datnicek and assess its implementation
   complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
@@
 Manipulating programs through semantically meaningful edits is a technique used by structure
-editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The langauge of
+editors \cite{teitelbaum-1981-cps,hempel-2018-deuce,beckman-2023-sandblocks}. The language of
 edits has been captured formally as an \emph{edit calculus} \cite{omar-2017-hazelnut} and
@@
 
-Merging of edits is most frequently done in version control systems such as git. Source control ystems
+Merging of edits is most frequently done in version control systems such as git. Source control systems
 based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
@@
 mechanism as central, whereas other systems view tracking of copy \& paste as an extra
-editor featrue \cite{jablonski-2007-cren,toomim-2004-linked}. Copy \& paste has also been
+editor feature \cite{jablonski-2007-cren,toomim-2004-linked}. Copy \& paste has also been
 tracked in spreadsheets \cite{hermans-2015-copypaste}. Gridlets \cite{joharizadeh-2020-gridlets}
@@
 Engraft \cite{horowitz-2023-engraft}. Programming environments may be also improved by integrating
-live examples \cite{rauch-2019-babylonian} and a variety of AI assistantce
+live examples \cite{rauch-2019-babylonian} and a variety of AI assistance
 tools \cite{petricek-2023-aias,blinn-2024-llms,mcnutt-2023-nbai}.
@@
   {Reference to another document location. Displays the \Verb|selectors| as a link.}
-\ndtablecol{\faFont}{Primitivie}{string \textnormal{or} number}
+\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
   {Numerical or textual primitive value. Renders as an HTML text node.}
@@
 updating. This annotation is required when the target contains the \emph{Index} selector,
-which is necessarilly a value edit affecting a specific node.
+which is necessarily a value edit affecting a specific node.
 
@@
 
-\subsection{Collaboratative Editing}
+\subsection{Collaborative Editing}
 \label{sec:impl-collab}
@@
 reconciling them with potential structural edits. The result can be seen in Fig.~\ref{fig:walkthrough} (E), where
-a newly added list item becames a table row.
+a newly added list item becomes a table row.
 
@@
 import all data edits. Supporting the model requires a \emph{retract} operation \cite{edwards-2021-typed}
-that is dual to our edit reconcilliation (given subsequent edits $e_1, e_2$, generate
+that is dual to our edit reconciliation (given subsequent edits $e_1, e_2$, generate
 $e_2'$ that has the same effect as $e_2$ but can occur before $e_1$).
@@
 
-\subsubsection*{Effectiveness and Unifromity.}
+\subsubsection*{Effectiveness and Uniformity.}
 Many of the Datnicek requirements are addressed by Denicek with minimal development
@@
 Our experience with conflict detection shows the importance of getting the details of the
-implementation of programming susbstrates like Denicek right. This could be aided by the
+implementation of programming substrates like Denicek right. This could be aided by the
 development of a formally tractable model of the system and proving a correctness property

### 2025-07-10 paper.tex
Add teaser, edit page 1

@@
 %
-\documentclass[sigconf,anonymous,screen]{acmart}
+\documentclass[sigconf]{acmart}
 %\documentclass[sigconf,review,anonymous]{acmart}
@@
 
+\setlength{\marginparwidth}{1.4cm}
+
+\newcommand{\diff}[1]{{\textcolor{magenta}{#1}}}
+\newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}
+
 \newcommand{\ident}[1]{{\sffamily #1}}
@@
 \newcommand{\srckvd}[1]{\textnormal{\sffamily\bfseries #1}}
-\newcommand{\note}[1]{\textcolor{red}{#1}}
 
@@
 
-\setcopyright{none}
 \copyrightyear{2025}
 \acmYear{2025}
-\acmDOI{XXXX.XXXX}
-
-\acmConference[Conference acronym 'XX]{}{June 03--05, 2018}{Woodstock, NY}
-\acmISBN{978-1-4503-XXXX-X/18/06}
+\setcopyright{cc}
+\setcctype{by}
+\acmConference[UIST '25]{The 38th Annual ACM Symposium on User Interface Software and Technology}{September 28-October 1, 2025}{Busan, Republic of Korea}
+\acmBooktitle{The 38th Annual ACM Symposium on User Interface Software and Technology (UIST '25), September 28-October 1, 2025, Busan, Republic of Korea}\acmDOI{10.1145/3746059.3747646}
+\acmISBN{979-8-4007-2037-6/2025/09}
 
@@
 \title[Denicek: Computational Substrate for Document-Oriented End-User
-  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming}
+  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming\\[0.25em]}
 
@@
 
+\author{Jonathan Edwards}
+\email{jonathanmedwards@gmail.com}
+\orcid{0000-0003-1958-7967}
+\affiliation{%
+  \institution{Independent}
+  \city{Boston}
+  \country{USA}
+}
+
 \begin{abstract}
+\note{Added overview figure with examples showing relationships between Denicek \& co.}
 User-centric programming research gave rise to a variety of compelling programming experiences,
@@
 
-%\begin{teaserfigure}
-%  \includegraphics[width=\textwidth]{sampleteaser}
-%  \caption{Seattle Mariners at Spring Training, 2010.}
-%  \Description{Enjoying the baseball game from the third-base
-%  seats. Ichiro Suzuki preparing to bat.}
-%  \label{fig:teaser}
-%\end{teaserfigure}
+\begin{teaserfigure}
+  \vspace{-0.2em}
+  \includegraphics[width=1\textwidth,clip,trim=0cm 8.5cm 7.7cm 0cm]{fig/teaser.pdf}
+  \vspace{-1.5em}
+  \caption{\diff{Denicek is a computational substrate for document-oriented programming
+    based on document edit histories~(A). We co-design Denicek with a web-based end-user
+    programming environment Denicek (B). Here, Denicek is used to build a Todo list app via
+    programming-by-demonstration, by copying a value from input to a new list item.
+    We then evaluate the generality of Denicek by using it to build a data science notebook
+    system Datnicek (C). Here, Datnicek is used to interactively clean the data table, by
+    removing the `` p'' (provisional) marker from numerical columns.
+  }}
+  \Description{The Denicek diagram in the middle illustrates a document representation as an
+   edit history. Two arrows point from the diagram to two screenshots. Webnicek on the left
+   contains a Todo list app with a command palette showing add item operation. Datnicek on the
+   right contains two cells, one with code to load data and another with an interactive editor
+   for data cleaning.}
+  \label{fig:teaser}
+  \vspace{0.75em}
+\end{teaserfigure}
 \maketitle
@@
 % ==================================================================================================
-
 \section{Introduction}
@@
 interfaces~\cite{kay-1993-smalltalk}, while representing programs as lists enabled Lisp to become
-a language laboratory \cite{steele-1993-lisp}.
+a language laboratory~\cite{steele-1993-lisp}.
 
@@
 
-In this paper, we describe a programming substrate that makes it easy to develop diverse
-programming experiences \cite{myers-2006-eup}. The experiences we focus on include:
+\note{Revised to say Denicek is a substrate used to build programming systems.}
+\diff{We describe Denicek, a computational substrate that makes it easy to build programming systems
+supporting diverse compelling programming experiences \cite{myers-2006-eup}:}
 

### 2025-07-12 paper.tex
More editing and revising

@@
 \newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}
+
+%\newcommand{\diff}[1]{{#1}}
+%\newcommand{\note}[1]{}
 
@@
 The technical focus of this paper fits within the interior mode of design science research \cite{adam-2021-dsr}.
-To \emph{design} Denicek, we identify six \emph{formative examples} -- simple programming
+To \emph{design} Denicek (Fig.~\ref{fig:teaser} (A)), we identify six \emph{formative examples} -- simple programming
 tasks that manifest one or more of the desired programming experiences~(\S\ref{app:examples}).
 Using those examples, we co-design the Denicek substrate and a simple web-based end-user programming
-environment Webnicek~(\S\ref{sec:walk}), which is built directly on top of the substrate. Although
+environment Webnicek~(Fig.~\ref{fig:teaser} (B)), which is built directly on top of the substrate. Although
 Webnicek can be used to complete end-user programming tasks, it is optimized for developing the
@@
 To \emph{evaluate} the usability of Denicek for the development of end-user programming systems,
-we use it to build Datnicek (\S\ref{sec:case}), an innovative interactive data science notebook
+we use it to build Datnicek (Fig.~\ref{fig:teaser}~(C)), an innovative interactive data science notebook
 that brings together a range of recent research work~\cite{kandel-2011-wrangler,drossos-2020-wrex,petricek-2022-thegamma,adams-2025-grove}.
-We reflect on the development experience~(\S\ref{sec:case-reflection}) and provide a heuristic
-evaluation of Denicek characteristics including importance and generality (\S\ref{sec:eval}).
-
-\subsubsection*{Substrate}
-The Denicek substrate brings together two central design ideas. First, it represents programs as
+\note{Clarify that Datnicek was designed after Denicek was completed.}
+\diff{We reflect on the degree to which Denicek simplifies the implementation of a programming
+system that is based on current research advances and was conceived after the design of Denicek
+has been fully finalized~(\S\ref{sec:case-reflection}).} We also provide a heuristic evaluation
+of Denicek characteristics including importance and generality (\S\ref{sec:eval}).
+
+\subsubsection*{Computational Substrates}
+Denicek brings together two central design ideas. First, it represents programs as
 document trees consisting of nodes that can represent data, formulas, evaluated results, as well as
@@
 
-The key insight that we present in this paper is that many compelling programming experiences can
-be implemented by leveraging this uniform document representation alongside with a suitable
-composition of the three primitive operations.
+\subsubsection*{Key Takeaways}
+\note{Add more general takeaway (below) and clarify specific technical takeaway (here).}
+\diff{The key insights of this paper are twofold. A specific technical takeaway is that many
+compelling programming experiences can be implemented on top a uniform document representation,
+by using a suitable composition of three primitive operations that operate on sequences of
+edit operations.}
 
@@
 recomputation behavior (\S\ref{sec:impl-incremental}).
+
+\note{Add a more general takeaway suggested by Reviewer 3 (Thank you!)}
+\diff{A more basic takeaway, illustrated by the development of Denicek, is that it is worth looking
+for novel computational substrates that better support compelling programming experiences developed
+in user-centric programming research.}
 
@@
 To enable others build on top of Denicek, we share our compact open-source
-implementation at: \url{https://github/removed/for/review} %d3sprog/denicek
+implementation at: \url{https://github.com/d3sprog/denicek}
 
@@
 \label{sec:background}
-Denicek aims to support a class of systems associated with end-user programming \cite{myers-2006-eup},
-notational freedom and self-sustainability~\cite{jakubovic-2023-techdims}, liveness and interactivity
-\cite{horowitz-2023-qualities,rein-2019-live}. We see programming more as interacting with a medium
-or a substrate \cite{kay-1977-media,klokmose-2015-webstrates} than writing text \cite{rpg-2012-revolution}.
-We focus on systems that allow gradual progression from a user to a developer, as envisioned
-in Smalltalk~\cite{reenskaug-1981-byte} and use the term end-user programming loosely to refer to a
-part of this spectrum, also including spreadsheet systems and notebooks for data science.
-% We first review related systems and their technical design (\S\ref{sec:background-sys})
-% before discussing programming experiences (\S\ref{sec:background-exp}).
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Programming Systems and Substrates}
+\note{Clarify focus to make paper more self-contained}
+\diff{The premise of Denicek is that suitable structures for constructing programs can make it easier
+to support a range of compelling user experiences. A prime example is
+Smalltalk whose object-oriented structures
+enable malleable programming experience~\cite{girba-2017-moldable,tcher-2019-malleable}.}
+
+Denicek aims to support a class of systems associated with end-user programming, liveness and
+interactivity \cite{myers-2006-eup,horowitz-2023-qualities,rein-2019-live}, notational freedom and
+self-sustainability~\cite{jakubovic-2023-techdims}. We see programming as interacting with a medium
+or a substrate \cite{kay-1977-media,klokmose-2015-webstrates,rpg-2012-revolution} and
+use the term end-user programming loosely to refer to a part of this spectrum, also including
+spreadsheet systems and notebooks for data science.
+
+\note{Highlight related systems (discussed later) for better framing}
+\diff{We follow systems such as BootstrapLab~\cite{jakubovic-2022-ladder}, which aims to support
+gradual progression from a user to a developer envisioned in Smalltalk~\cite{reenskaug-1981-byte}
+in a primarily graphical environment and Subtext~\cite{edwards-2005-subtext},
+which develops a user-friendly programming environment based on object copying.}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Programming Systems}
 \label{sec:background-sys}
@@
 \subsubsection*{Programming Systems and Substrates}
-A number of systems illustrate the qualities Denicek aims to support and have a related internal
+A number of systems illustrate the qualities Denicek aims to support and have a related underlying
 structure. Subtext, BootstrapLab and Infra~\cite{edwards-2005-subtext,jakubovic-2022-ladder,hall-2017-infra}
@@
 
-Merging of edits is most frequently done in version control systems such as git. Source control systems
-based on a more rigorous design such as Pijul \cite{meunier-2024-pijul} delay merging to a later point
+Merging of edits is most frequently done in version control systems. The
+Pijul system \cite{meunier-2024-pijul} delays merging to a later point
 by using a graph or a lattice \cite{schurmann-2022-merging}. In the context of
-programming environments, Grove \cite{adams-2025-grove} uses a commutative patches, with a graph
-structure similar to Pijul, as the basis for a collaborative structure editor.
+programming environments, Grove \cite{adams-2025-grove} uses a commutative patches with a similar graph
+structure as the basis for a collaborative structure editor. \diff{Grove patches
+(construct, delete, relocate) are sufficient for a structure editor, but lack some structures and
+operations that our formative examples rely on (collections, copying). However, the core model
+of Grove provides a possible alternative basis for Denicek.}
+\note{Added why Grove cannot be used directly by Denicek, but is interesting direction.}
 
@@
 
+\note{Added report of our experience with what makes EUP systems hard to build.}
+\subsubsection*{Building Programming Experiences}
+\diff{
+The challenge addressed by Denicek is that end-user programming experiences are hard to~build.
+Our experience developing such systems \cite{petricek-2019-histogram,petricek-2022-thegamma,
+jakubovic-2022-ladder,edwards-2005-subtext,edwards-2004-example} suggests two difficulties.
+The first is the need to move between the concrete and the abstract \cite{jakubovic-2023-techdims}.
+A sequence edits strikes a balance between those opposing representations.
+The second is the need to combine multiple experiences, which requires
+a major engineering effort when using a conventional program representation.}
+
+% Wrangler~\cite{kandel-2011-wrangler}
+
+% * Reviewers also ask about the concrete problem addressed (R2) and what makes end-user programming
+% tools hard to build (R1). The latter question (aptly phrased by R1) deserves a deeper investigation.
+% We believe the difficulty emerges when combining multiple programming experiences in a single system
+% and from the need to move between the concrete and the abstract. This belief is based on our personal
+% experience developing multiple systems (to be cited in the revision) and on a technical dimensions
+% analysis (Sec. 8.2). A substrate based on edit history provides a middle ground between the
+% abstract and the concrete and we show that it can be carefully designed to serve as the basis for a range of experiences.
+
 % --------------------------------------------------------------------------------------------------
@@
 
-We review a range of compelling programming experiences explored in recent research,
-focusing on those supported by Denicek.
+%We review a range of compelling programming experiences explored in recent research,
+%focusing on those supported by Denicek.
 

### 2025-07-13 paper.tex
Camera ready

@@
 
-\newcommand{\diff}[1]{{\textcolor{magenta}{#1}}}
-\newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}
-
-%\newcommand{\diff}[1]{{#1}}
-%\newcommand{\note}[1]{}
+%\newcommand{\diff}[1]{{\textcolor{magenta}{#1}}}
+%\newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}
+
+\newcommand{\diff}[1]{{#1}}
+\newcommand{\note}[1]{}
 
@@
    edit history. Two arrows point from the diagram to two screenshots. Webnicek on the left
-   contains a Todo list app with a command palette showing add item operation. Datnicek on the
+   contains a Todo list app with a command toolbox showing add item operation. Datnicek on the
    right contains two cells, one with code to load data and another with an interactive editor
@@
 \begin{itemize}
-\item \emph{Collaborative editing.} Users should be able to locally modify a shared document
+\item \emph{Collaborative Editing.} Users should be able to locally modify a shared document
   and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
@@
 \includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
+\vspace{-0.5em}
 \caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
@@
 
-In this section, we use the web-based prototype Webnicek to illustrate programming
-experiences enabled by the Denicek substrate. Webnicek is based on a structure
-editor that supports navigation in the document and issuing of edit commands. We follow two
-formative examples (\S\ref{app:examples}) where users collaboratively plan a conference.
+We use the web-based prototype Webnicek to illustrate programming experiences enabled by Denicek.
+Webnicek is based on a structure editor that supports navigation in the document and issuing of
+edit commands. \diff{We follow a formative example (\S\ref{app:examples}) where Evelyn, Juliana and Alfred
+collaboratively plan a conference.}
+\note{Turn this into a concrete user story consistent with the video. Also add more details about
+the edit operations and UI interactions.}
 
 \subsubsection*{\circled{A} Adding a Speaker}
-The first user starts with an empty document, which is represented as a record. They add a field for
-each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. They
-use the command toolbox to add the first speaker.
+Evelyn starts with an empty document, which is represented as a record. She adds a field for
+each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. \diff{She
+uses the command toolbox to issue edit commands that add the first speaker.}
 
 \subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
-further speakers, the user creates a textbox and a button. They enter new speaker
-details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
-from the textbox into the new element using a source view.
-
-\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker from the textbox,
-the user opens history view, selects edits that added the speaker and saves those as the
-{\small\Verb_add-speaker_} interaction. They attach this interaction as a event handler for the
+further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
+the details of another speaker \diff{into the textbox}, adds a new {\small\Verb_<li>_} element
+and copies the speaker details from the textbox into the new element in the source view,
+\diff{using a copy command}.
+
+\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker,
+Evelyn opens the history view, selects edits that added the speaker and saves them as the
+{\small\Verb_add-speaker_} interaction. She attaches this interaction as an event handler for the
 {\small\Verb_click_} event of the button.
@@
 
-\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
-version of the document (A) and turns the list into a table. This is done by invoking a series of
-commands that change tags, wrap elements, copy values and transform them.
-
-\subsubsection*{\circled{E} Merging Edits} The two document versions (C and D) can be automatically merged.
-The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
+\subsubsection*{\circled{D} Refactoring Document Structure} Alfred starts with the initial
+version of the document (A) and turns the list into a table. He invokes a series of commands that
+\diff{rename tags, wrap elements, copy values and split strings using the comma as a separator}.
+\note{More concrete description of edits}
+
+\subsubsection*{\circled{E} Merging Edits}
+\diff{Alfred then merges the later edits (B, C) done by Evelyn into his version of the document.}
+The refactoring is applied to all speakers. New speakers added using the ``Add speaker!''
 button are also automatically transformed to the new format.
@@
 
-\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
-the initial document. Formulas are represented as nodes with the {\small\Verb_x-formula_} tag,
-whose arguments are other nodes, references to other nodes, or formulas (highlighted on mouse hover).
+\subsubsection*{\circled{F} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
+the initial document. \diff{She uses the command toolbox to create document nodes representing
+formulas. Formulas are regular nodes with a special} {\small\Verb_x-formula_} \diff{tag and
+arguments as child nodes. To count the speakers, she uses the} {\small\Verb_count_} \diff{builtin
+with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
+representation of formulas.}
 
@@
 
-\subsubsection*{\circled{G} Merging Formulas} When the budget calculation (F) is merged with the
-earlier edits (E), references in formulas are automatically updated to point to the new list.
-Adding new speaker via the ``Add speaker!'' button invalidates the evaluated result.
+\subsubsection*{\circled{G} Merging Formulas} When Juliana merges the budget calculation (F) with
+the earlier edits (E), references in formulas are automatically updated to point to the \diff{rows of the table}.
+Adding new speaker via the ``Add speaker!'' button \diff{invalidates results of only those formulas
+that depend on the number of speakers.}
 
@@
 this design choice is discussed in \S\ref{sec:discuss}.
-
-\subsubsection*{Document Edits}
-The supported document edits and their behaviors are listed in Fig.~\ref{fig:edits}. All edits
-require a \emph{target} to which they are applied. Targets are absolute references not containing the
-\ident{Parent} selector. They can contain the \ident{Any} selector, in which case the edit is applied
-to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
-kind. As discussed in \S\ref{sec:discuss}, list elements as well as fields of a record are ordered
-and edits that add a new item take the index or field name of a previous node.
-
-The edits can transform the document structure. Denicek tracks the effect of the edits on the
-structure and updates references when the document structure changes.
-Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
-(above) and edits that can affect references (below).
-
-Renaming a field or wrapping a node updates any references to within the target location
-(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
-rejects the edit. A copy edit of a node to which there is a reference is also rejected, because it
-is ambiguous whether references referring to the original location should be left unchanged,
-or modified to point to the target of the copying (a reference cannot refer to two unrelated locations).
-
-% \note{why not have distinct Copy and Move operations to resolve this?}
-% TP: Good point - this is a nice motivation for Move! I should probably add that :-)
-
-% --------------------------------------------------------------------------------------------------
 
@@
 \end{tabular}
-\vspace{-0.5em}
 \caption{Structure of selectors and document nodes}
 \label{fig:doc}
-\vspace{-1em}
 \end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection*{Document Edits}
+The supported document edits and their behaviors are listed in Fig.~\ref{fig:edits}. All edits
+require a \emph{target} to which they are applied. Targets are absolute references not containing the
+\ident{Parent} selector. They can contain the \ident{Any} selector, in which case the edit is applied
+to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a specified
+kind. As discussed in \S\ref{sec:discuss}, list elements as well as fields of a record are ordered
+and edits that add a new item take the index or field name of a previous node.
+
+The edits can transform the document structure. Denicek tracks the effect of the edits on the
+structure and updates references when the document structure changes.
+Fig.~\ref{fig:edits} distinguishes between edits that keep references in a document unchanged
+(above) and edits that can affect references (below).
+
+Renaming a field or wrapping a node updates any references to within the target location
+(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
+rejects the edit. A copy edit of a node to which there is a reference is also rejected, because it
+is ambiguous whether references referring to the original location should be left unchanged,
+or modified to point to the target of the copying (a reference cannot refer to two unrelated locations).
+
+% \note{why not have distinct Copy and Move operations to resolve this?}
+% TP: Good point - this is a nice motivation for Move! I should probably add that :-)
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection*{Automatic Reference Update}
+As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
+statically enforced (the substrate can be seen as being dynamically typed). As a consequence, edits
+that transform the structure can serve both as structural edits (affecting the structure)
+and as value edits (affecting the value of specific nodes). The latter can be used, for example,
+when constructing an additional list item. An item is first constructed and then populated
+with values. This temporarily violates the invariant that collections are homogeneous
+(an issue that could be address by using transactions as discussed in \S\ref{sec:discuss}).
+
+Another use of value edits is when evaluating a formula, which involves wrapping the formula
+node (\S\ref{sec:impl-eval}), an operation that should not affect references to the formula
+itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
+have a \emph{reference behavior} option that can disable automatic reference
+updating. This annotation is required when the target contains the \emph{Index} selector,
+which is necessarily a value edit affecting a specific node.
+
+% \note{I think it would be desireable to avoid all these complications. References
+% should be automatically maintained so they ``just work''.}
+% \note{I believe that evaluation should not change the type/shape of the document. The space
+% for evaluation results should be be preallocated in some sense, either as extra structure in
+% a formula, or as metadata attached to the tree. That said, I've tried various alternative
+% approaches to this in Subtext without ever settling on a preferred solution.}
+%
+% TP: I think this is true - but that would make the system more "statically typed"
+% and Denicek is dynamically typed - my only reason for this is to make the editing experience
+% more "document-like" - but it is perhaps not necessary.
+
+An important principle of the Denicek design is that the effect an edit has on references inside
+the document does not depend on the current value of the document. This makes it possible to
+define merging solely in terms of edits, without reference to current document state.
+% (The effect \ident{Copy} has on the document structure depends on the existing document
+% structure, but not on its value).
+This design choice makes it impossible to encode computational logic directly in the edits
+(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
+can be provided as an additional mechanism on top of the underlying Denicek substrate.
 
@@
 \end{tabular}
-\vspace{-0.75em}
 \caption{Summary of document edit types in Denicek}
 \label{fig:edits}
-\vspace{-1em}
 \end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsubsection*{Automatic Reference Update}
-As discussed in \S\ref{sec:discuss}, the structure of Denicek document is implicit and not
-statically enforced (the substrate can be seen as being dynamically typed). As a consequence, edits
-that transform the structure can serve both as structural edits (affecting the structure)
-and as value edits (affecting the value of specific nodes). The latter can be used, for example,
-when constructing an additional list item. An item is first constructed and then populated
-with values. This temporarily violates the invariant that collections are homogeneous
-(an issue that could be address by using transactions as discussed in \S\ref{sec:discuss}).
-
-Another use of value edits is when evaluating a formula, which involves wrapping the formula
-node (\S\ref{sec:impl-eval}), an operation that should not affect references to the formula
-itself. To support value edits, edits that normally transform references (Fig.~\ref{fig:edits} below)
-have a \emph{reference behavior} option that can disable automatic reference
-updating. This annotation is required when the target contains the \emph{Index} selector,
-which is necessarily a value edit affecting a specific node.
-
-% \note{I think it would be desireable to avoid all these complications. References
-% should be automatically maintained so they ``just work''.}
-% \note{I believe that evaluation should not change the type/shape of the document. The space
-% for evaluation results should be be preallocated in some sense, either as extra structure in
-% a formula, or as metadata attached to the tree. That said, I've tried various alternative
-% approaches to this in Subtext without ever settling on a preferred solution.}
-%
-% TP: I think this is true - but that would make the system more "statically typed"
-% and Denicek is dynamically typed - my only reason for this is to make the editing experience
-% more "document-like" - but it is perhaps not necessary.
-
-An important principle of the Denicek design is that the effect an edit has on references inside
-the document does not depend on the current value of the document. This makes it possible to
-define merging solely in terms of edits, without reference to current document state.
-% (The effect \ident{Copy} has on the document structure depends on the existing document
-% structure, but not on its value).
-This design choice makes it impossible to encode computational logic directly in the edits
-(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
-can be provided as an additional mechanism on top of the underlying Denicek substrate.
 
@@
 \ident{DeleteField} edit, the edit is rejected.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsubsection*{Merging Edit Histories}
+Merging edit histories is used when two users edit document independently, but also
+when replaying edits in programming by demonstration. The merge operation $\mathcal{M}_E(E_1, E_2)$
+works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix~$E$.
+Our merging is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that can be reapplied
+on top of the other edit history. That~is, $\mathcal{M}_E(E_1, E_2) = E, E_1, E_2'$.
+Note that the operation is not symmetric: $\mathcal{M}_E(E_2, E_1) = E, E_2, E_1'$. The
+result of applying the two histories to the same node will differ if there are conflicts
+among the edits in $E_1$ and $E_2$. We return to conflict detection in the next section.
+
+% \note{It might help to define a function signature for merging to make this clearer: $\mathcal{M}(P, L, R) = M$ or $\mathcal{M}_P(L, R) = M$.}
+% \note{Merging is often thought of as symmetric, perhaps choose a name that is clearly asymmetric, like rebase, translate, or project.}
+
+The key operation that enables edit history reconciliation takes two individual edits that occurred
+independently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
+can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
+There are two aspects of such reconciliation:
+
+\begin{enumerate}
+\item \emph{Apply to Newly Added.} If $e_2$ is adding new nodes to the document, but
+  $e_1$ modified the document through a selector that would also affect the new nodes added by $e_2$,
+  we need to apply the transformation represented by $e_1$ to the nodes newly added by~$e_2$.
+  This is done by generating an additional edit, to be applied after $e_2$, that is based on
+  $e_1$ but targets only the newly added nodes (more details can be found in \S\ref{app:merge-apply-to-added}).
+
+\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
+  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
+  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
+  that apply when transforming references inside a document, although we now also support the case
+  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
+\end{enumerate}
+
+% \note{I believe that in general a ``backwards'' form of merge is also needed, as in my
+% \emph{retract} and the OT \emph{Exclusion Transform}. A regular merge takes forking edits
+% and composes them. The backwards form converts composed edits into forking edits (and can
+% be combined with a forward merge to recompose in the opposite order). This is needed for
+% cherry picking and reverts, however you don't seem to need that for this paper.}
+%
+% TP: Correct, I do not support those - but cherry-picking is a nice example of a thing
+% that needs then and would be very nice to have!
+
+% \note{Related question: how do you merge an Append(target, index, after, node) with
+% DeleteItem(target, after)? A famous counterexample in OT was that two insertions adjacent
+% to a deletion could flip their location if the edits were reordered. I'll dig up the
+% details of this counterexample for you. Some systems handle this by leaving behind a
+% ``tombstone'' on delete to remember the position of the deleted item. I have a novel
+% solution but I'm probably going to go back to tombstones anyway because they are
+% simpler and handy in the diff UI.}
+%
+% TP: This is a good question. My ordering of records is a somewhat last minute addition
+% so it is not super well debugged. But in theory, it should be fine - because the ordering relation
+% could keep the IDs of removed elements (only for ordering purposes). I think this is
+% compatible with the MRDT data structure I'm trying to use.
+
+To illustrate merging, consider a case where we created a list of work items {\small\Verb|/todo|}.
+In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
+list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
+%
+\begin{equation*}
+\arraycolsep=2pt
+\begin{array}{rcl}
+e_2 &=& \srckvd{Append}(\textnormal{\small\Verb|/todo|}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
+e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
+e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+\end{array}
+\end{equation*}
+%
+If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
+reflect the wrapping (2) and we need to create and additional \ident{Add} operation
+that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
+that target the newly wrapped list and add the checkbox to the added item:
+%
+\begin{equation*}
+\arraycolsep=2pt
+\begin{array}{rcl}
+  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
+  e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+e_2'  &=& \srckvd{Append}(\textnormal{\small\Verb|/todo/items|}, \#1, \#0, \srcid{<li>Do some work</li>})\\
+e_2'' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/\#1|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+\end{array}
+\end{equation*}
+%
+If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
+$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
+same.
 
@@
 
-\subsubsection*{Merging Edit Histories}
-Merging edit histories is used when two users edit document independently, but also
-when replaying edits in programming by demonstration. The merge operation $\mathcal{M}_E(E_1, E_2)$
-works on two edit histories, $E, E_1$ and $E, E_2$, that have a shared prefix~$E$.
-Our merging is akin to git rebase. It turns edits $E_2$ into edits $E_2'$ that can be reapplied
-on top of the other edit history. That~is, $\mathcal{M}_E(E_1, E_2) = E, E_1, E_2'$.
-Note that the operation is not symmetric: $\mathcal{M}_E(E_2, E_1) = E, E_2, E_1'$. The
-result of applying the two histories to the same node will differ if there are conflicts
-among the edits in $E_1$ and $E_2$. We return to conflict detection in the next section.
-
-% \note{It might help to define a function signature for merging to make this clearer: $\mathcal{M}(P, L, R) = M$ or $\mathcal{M}_P(L, R) = M$.}
-% \note{Merging is often thought of as symmetric, perhaps choose a name that is clearly asymmetric, like rebase, translate, or project.}
-
-The key operation that enables edit history reconciliation takes two individual edits that occurred
-independently, $e_1$ and $e_2$, and produces a sequence of edits $e_2', e_2'', \ldots$ that
-can be applied after $e_1$ and have the effect of $e_2$, modified to respect the effects of $e_1$.
-There are two aspects of such reconciliation:
-
-\begin{enumerate}
-\item \emph{Apply to Newly Added.} If $e_2$ is adding new nodes to the document, but
-  $e_1$ modified the document through a selector that would also affect the new nodes added by $e_2$,
-  we need to apply the transformation represented by $e_1$ to the nodes newly added by~$e_2$.
-  This is done by generating an additional edit, to be applied after $e_2$, that is based on
-  $e_1$ but targets only the newly added nodes (more details can be found in \S\ref{app:merge-apply-to-added}).
-
-\item \emph{Transform Matching References.} If $e_2$ targets a node that is inside a node
-  whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
-  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
-  that apply when transforming references inside a document, although we now also support the case
-  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
-\end{enumerate}
-
-% \note{I believe that in general a ``backwards'' form of merge is also needed, as in my
-% \emph{retract} and the OT \emph{Exclusion Transform}. A regular merge takes forking edits
-% and composes them. The backwards form converts composed edits into forking edits (and can
-% be combined with a forward merge to recompose in the opposite order). This is needed for
-% cherry picking and reverts, however you don't seem to need that for this paper.}
-%
-% TP: Correct, I do not support those - but cherry-picking is a nice example of a thing
-% that needs then and would be very nice to have!
-
-% \note{Related question: how do you merge an Append(target, index, after, node) with
-% DeleteItem(target, after)? A famous counterexample in OT was that two insertions adjacent
-% to a deletion could flip their location if the edits were reordered. I'll dig up the
-% details of this counterexample for you. Some systems handle this by leaving behind a
-% ``tombstone'' on delete to remember the position of the deleted item. I have a novel
-% solution but I'm probably going to go back to tombstones anyway because they are
-% simpler and handy in the diff UI.}
-%
-% TP: This is a good question. My ordering of records is a somewhat last minute addition
-% so it is not super well debugged. But in theory, it should be fine - because the ordering relation
-% could keep the IDs of removed elements (only for ordering purposes). I think this is
-% compatible with the MRDT data structure I'm trying to use.
-
-To illustrate merging, consider a case where we created a list of work items {\small\Verb|/todo|}.
-In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
-list in an extra \ident{<div>} element ($e_1$) and add a checkbox to each work item ($e_1'$):
-%
-\begin{equation*}
-\arraycolsep=2pt
-\begin{array}{rcl}
-e_2 &=& \srckvd{Append}(\textnormal{\small\Verb|/todo|}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
-e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
-e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
-\end{array}
-\end{equation*}
-%
-If we want to append the edit $e_2$ after edits $e_1, e_1'$, we need to update its target to
-reflect the wrapping (2) and we need to create and additional \ident{Add} operation
-that will add the checkbox to the new item (1). The result is a sequence with two edits $e_2', e_2''$
-that target the newly wrapped list and add the checkbox to the added item:
-%
-\begin{equation*}
-\arraycolsep=2pt
-\begin{array}{rcl}
-  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
-  e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
-e_2'  &=& \srckvd{Append}(\textnormal{\small\Verb|/todo/items|}, \#1, \#0, \srcid{<li>Do some work</li>})\\
-e_2'' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/\#1|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
-\end{array}
-\end{equation*}
-%
-If we performed the merge operation in the other order, we would append $e_1, e_1'$ after
-$e_2$ without a change. In this case, the result of applying the two edit sequences would be the
-same.
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\includegraphics[width=0.95\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
-\vspace{-0.5em}
-\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
-\label{fig:merging}
-\vspace{-0.5em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \subsubsection*{Conflict Resolution}
 When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is desirable that
-applying $\mathcal{M}_E(E_1, E_2)$ and $\mathcal{M}_E(E_2, E_1)$ results in the same document.
-This is not always the case. If the two edits modify the same value, transform the structure of
-a node in incompatible ways or one edit modifies a node deleted by the other, the merge operation
-is not symmetric. Such conflicting edits can be detected, reported to the user and optionally
-removed. Conflict detection is also used to implement incremental evaluation~(\S\ref{sec:impl-incremental}).
+$\mathcal{M}_E(E_1, E_2)$ and $\mathcal{M}_E(E_2, E_1)$ result in the same document.
+This is not always the case. If the two edits modify the same value, transform the structure of a node
+in incompatible ways or one edit modifies a node deleted by the other, the merge operation
+is not symmetric. Such conflicting edits can be detected and reported to the user.
+\diff{Conflict detection is also used to implement incremental recomputation~(\S\ref{sec:impl-incremental}),
+in which case conflicting edits produced by evaluation are removed and formulas have to be
+recomputed.}\note{Clarify conflicting edits are only removed in incremental evaluation}
 
@@
 % ==================================================================================================
+
+\begin{figure}[t]
+\includegraphics[width=0.95\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
+\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
+\label{fig:merging}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
 
@@
 Webnicek also accounts for the case where edits recorded in the document are themselves
-transformed (when they are reconciliated with other edits during merging). In this case, Webnicek
+transformed (when they are reconciled with other edits during merging). In this case, Webnicek
 updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).
@@
 % conditionalize their effects. More general and conventional programming constructs are discussed in \S\ref{sec:discuss}.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\includegraphics[width=0.95\columnwidth,clip,trim=0cm 5cm 11.5cm 0cm]{fig/counter.pdf}
+\caption{We wrap the initial count (1) in a formula that adds 1 to the
+value, (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).
+The button replays the saved edits that wrap the current count in a formula~(5).}
+\label{fig:counter}
+\end{figure}
 
@@
 \keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
-Programming by demonstration systems typically implement heuristic for generalization
+Programming by demonstration systems typically implement heuristics for generalization
 \cite{myers-2000-intelligence} that infers and suggests such generalizations. If integrated into
@@
 \begin{figure}[t]
-\vspace{-0.5em}
 \includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
-\vspace{-1.25em}
 \caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
@@
 \label{fig:interactive}
-\vspace{-1em}
 \end{figure}
@@
 
-\begin{figure}[t]
-\vspace{-0.5em}
-\includegraphics[width=0.95\columnwidth,clip,trim=0cm 5cm 11.5cm 0cm]{fig/counter.pdf}
-\vspace{-0.5em}
-\caption{We wrap the initial count (1) in a formula that adds 1 to the
-value, (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).
-The button replays the saved edits that wrap the current count in a formula~(5).}
-\label{fig:counter}
-\vspace{-1em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \subsection{Incremental Recomputation}
@@
   \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
-  \vspace{-1em}
   \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
@@
 \centering
-\vspace{-0.5em}
 \begin{minipage}{0.55\columnwidth}
@@
   \end{minipage}
-  \vspace{-1em}
 \end{figure}
@@
 
+\diff{Note that Denicek can only reflect schema changes explicitly represented by the document
+structure. If a formula depended on the structure of the original string representing speakers
+(with a comma), Denicek would not be able to amend the logic of the formula. Systems supporting
+the divergence model \cite{edwards-2025-schema} may be able to lift this restriction by
+transforming the new value back before applying the original formula.}
+\note{Added response to R3's question about schema change}
+
 Edits produced by document evaluation transform the formula structure (wrapping it in the
@@
 \begin{figure}[t]
-\vspace{-0.25em}
 \includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
-\vspace{-2em}
+\vspace{-1em}
 \caption{Using provenance tracking to highlight document parts that contributed to the
@@
 \label{fig:provenance}
-\vspace{-0.75em}
 \end{figure}
@@
 
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
+\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
+\label{fig:coevolution}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Concrete Programming}
@@
 
-In Fig.~\ref{fig:copypaste}, the user copies a formula with switched operands using
-the \ident{Copy} edit and then modifies it to use a different data source. They then navigate back
+\note{Modified figure to add call to read-csv}
+\diff{The functionality is illustrated in Fig.~\ref{fig:copypaste}, which uses the builtin operation}
+{\small\Verb_read-csv_} \diff{to load a table from a file. The operation imports the data as a
+document table with rows that can be addressed by subsequent formulas. The then user writes two
+formulas to sum the rows and compute an average.}
+\note{Add explanation of a more complex builtin operation}
+
+They copy a formula with switched operands using
+the \ident{Copy} edit and then modify it to use a different data source. They then navigate back
 in history to the point before the copying and create a temporary fork of the document. In the
@@
 \begin{figure}[t]
-\vspace{-0.5em}
-\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
-\vspace{-1em}
-\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
-\label{fig:coevolution}
-\vspace{-0.5em}
+\includegraphics[width=1\columnwidth,clip,trim=0cm 1.7cm 1cm 0cm]{fig/copypaste.pdf}
+\caption{The user uses creates a copy of a formula (1). They notice an error (2), go back in
+history to switch the arguments (3), merge the change and re-evaluate both formulas (4).}
+\label{fig:copypaste}
 \end{figure}
@@
 % either a different kind of edit or possibly an edit of a virtual ``prototype'' element.)
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure}[t]
-\vspace{-0.5em}
-\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
-\vspace{-1.5em}
-\caption{The user uses creates a copy of a formula (1). They notice an error (2), go back in
-history to switch the arguments (3), merge the change and re-evaluate both formulas (4).}
-\label{fig:copypaste}
-\vspace{-0.5em}
-\end{figure}
 
@@
 conditional.
+
+\note{Respond to R3's question about avoiding edits depending on values and R1's concern about
+usefulness for PBD tools.}
+\diff{This limitation means that edits generated by generalized interactions (\S\ref{sec:impl-interaction})
+can be only merged in limited ways (an item added to a list through merging after
+invoking an action will not be affected by the action). We expect that the restriction on
+expressiveness of selectors can be lifted for edits that do not modify the document structure,
+but have not yet lifted this restriction in Denicek.}
 
@@
 
-\begin{figure}[t]
-\vspace{-0.25em}
-\includegraphics[width=0.9\columnwidth,clip,trim=0cm 0.7cm 18.5cm 0.2cm]{fig/source.pdf}
-\vspace{-0.5em}
-\caption{Underlying document representation of a code cell and a grid cell in the Datnicek
-notebook system.}
-\label{fig:source}
-\vspace{-0.25em}
-\end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
 \begin{figure*}[t]
@@
 \label{fig:datnicek}
-\vspace{-0.5em}
 \end{figure*}
@@
 
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}[t]
+\includegraphics[width=0.9\columnwidth,clip,trim=0cm 0.7cm 18.5cm 0.2cm]{fig/source.pdf}
+\caption{Underlying document representation of a code cell and a grid cell in the Datnicek
+notebook system.}
+\label{fig:source}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
 \subsection{Reflections}
@@
 
+\subsubsection*{External Validity.}
 The development of Datnicek was only started after the design and implementation of Denicek and
-Webnicek was completed. Thanks to this two-phase methodology, the experience reported here
-provides a qualitative evaluation of Denicek's capabilities. It also highlights limitations and
-suggests directions for future work.
+Webnicek was completed. Thanks to this two-phase methodology, the case study provides
+a qualitative evaluation of Denicek's capabilities. \diff{The design of Denicek is based
+on third-party research (\S\ref{sec:case-req}) and so the case study provides a limited external
+validation. However, several design choices of Datnicek align it with the capabilities of Denicek.
+This includes the use of a structure editor and
+a user interface for programming by demonstration that suggests edit operations. We revisit the
+generality of Denicek in \S\ref{sec:eval-tdps}, but establishing external validity ultimately
+requires adoption of the open-source Denicek package by other programming system researchers.}
+\note{Added discussion on external validity - although this is necessarilly limited.}
 
@@
 system performance. We implemented a few basic Denicek optimisations, but Datnicek is still cumbersome
-to use with data consisting of multiple thousands of rows. Improving system performance is
-thus an important direction for future work.
+to use with data consisting of multiple thousands of rows.
 
@@
 highlights an unnecessary duplication -- a transformation stores a sequence of edits, as well as a formula.
-An interesting future direction is unifying or integrating those two notions of computation
-provided by the Denicek substrate.
+Unifying these two notions is another interesting future direction.
 
@@
 \subsection{Technical Dimensions}
+\label{sec:eval-tdps}
+
 The technical dimensions (TDPS) framework \cite{jakubovic-2023-techdims} maps the design space of
@@
 
+\begin{acks}
+The authors thank Clemens Nylandsted Klokmose for research guidance and
+attendees of the IFIP WG 2.16 meeting in Serpiano, members of the Software Architecture Group at HPI
+and attendees of LIVE 2023 for feedback on the project.
+Anonymous reviewers provided invaluable actionable guidance for improving the final version
+of the paper. The work has been supported by the
+\grantsponsor{cuni}{Charles University}{https://d3s.mff.cuni.cz/projects/primus24/}
+grant \grantnum{cuni}{PRIMUS/24/SCI/021} and by the
+\grantsponsor{erccz}{Czech Ministry of Education, Youth and Sports}{https://starfos.tacr.cz/en/projekty/LL2325}
+grant ERC-CZ \grantnum{erccz}{LL2325}.
+\end{acks}
+
 \bibliographystyle{plain}
@@
 
-\newpage
-~
-\newpage
-
 \appendix
-\section{Formative Examples}
-\label{app:examples}
-
-The design the Denicek substrate, we identified six formative examples shown in
-Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
-(Todo and Counter apps) to cases from literature~\cite{edwards-2022-copypaste} and
-problems posed as schema change challenges \cite{edwards-2025-schema}.
-The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
-built (as directly as possible) on top of the substrate and was used to solve implement the
-formative examples.
-
-Many of the formative examples include a small programming challenge, such as adding
-user interface to add a new speaker, a new list item or modify the count. Our aim was for
-the substrate to enable solving  those through programming by demonstration.
-Programming by Demonstration is often used in data wrangling
-\cite{gulwani-2012-spreadsheets,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
-is only a minimalistic illustration of such use, loosely inspired by earlier work \cite{miller-2001-simult}.
-
-\section{Merging Edit Histories}
-Recall that merging takes two edit histories, $E, E_1$ and $E, E_2$,
-transforms edits $E_2$ into $E_2'$ that can be reapplied on top of the first history
-resulting in $E, E_1, E_2'$. The key operation takes two individual edits, $e_1$ and $e_2$
-and produces a sequence of edits $e_2', e_2'', \ldots$ that can be applied after $e_1$,
-and combine the two edits. This section provides details about the two aspects of this operation.
-
-\subsection{Apply to Newly Added}
-\label{app:merge-apply-to-added}
-
-Assume that edits $e_1$ and $e_2$ occurred independently. We want to modify $e_2$ so that
-it can be placed after $e_1$. If the edit $e_2$ added new nodes to the document that the
-edit $e_1$ would affect, we generate an additional edit that apply the transformation of $e_1$
-to the newly added nodes (and only to those).
-
-The only edits that add new document nodes are \ident{Add}, \ident{Append}, \ident{Copy} and
-so we consider this case if the edit $e_2$ is one of those. If so, we check
-whether the target of $e_1$ is within the target of $e_2$, i.e., the list of selectors that
-forms the target of $e_2$ is a prefix of the list of selectors that forms the target of $e_1$.
-
-Along the way, we compute a \emph{more specific prefix}. If the target of $e_1$ contains
-the \ident{All} selector, it can be matched against a specific \ident{Index} selector in the
-target of $e_2$ (if the selector of $e_1$ is more specific than that of $e_2$, the targets
-are not matched). We then replace the original prefix in $e_1$ with the \emph{more specific prefix}
-that contains \ident{Index} selector in places where the original edit contained \ident{All}.
-This way, we obtain $e_1'$ which is a focused version of $e_1$ that applies only to the
-nodes newly added by $e_2$. The edit $e_2$ thus becomes a pair of edits $e_2, e_1'$.
-The final document will contain edits $e_1, e_2, e_1'$ -- that is, it will first apply the edit
-$e_1$ to nodes already in the document, then add new nodes and then apply the transformation
-represented by $e_1$ to the newly added nodes.
-
-\subsection{Transform Matching References}
-\label{app:merge-transform-refs}
-
-As above, assume that edits $e_1$ and $e_2$ occurred independently. We want to
-modify $e_2$ so that it can be placed after $e_1$. If $e_1$ is any of the three edits
-listed in Fig.~\ref{fig:updates} (\ident{RenameField}, \ident{WrapRecord}, \ident{WrapList}),
-we collect all references that appear inside $e_2$ (the target, the source of \ident{Copy}
-and any references occurring in the nodes added by \ident{Add} or \ident{Append}).
-If the target of $e_1$ is a prefix of any of those references, we update the references
-accordingly and obtain a new edit $e_2'$. Note that it would be an error to match specific
-\ident{Index} in $e_1$ with more general \ident{All} in $e_2$, but this cannot happen --
-reference updating is not done when the target of $e_1$ contains \ident{Index}.
-
-Now consider the case when $e_1$ is \ident{Copy} and the edit $e_2$ targets a node that is
-the source node of the copy operation (or any of its children). In this case, it is reasonable
-to require that the edit $e_2$ is applied to both the source and the target of the copy.
-(This is required by the refactoring done in the Conference Budget example.)
-We handle the case by creating a copy of $e_2$ with transformed selectors (target and, if $e_2$
-is also \ident{Copy}, also its source). To transform the selectors, we replace the prefix
-formed by the source of the \ident{Copy} by a new prefix, formed by the target target of
-the \ident{Copy}. We then add the new operation as $e_2'$ if at least one of its selectors
-was transformed (typically target, but possibly also source).
 
@@
 \newcommand{\extablecol}[5]{
-\small{\bfseries #1} #2\,\; --\,\; #3\\[-0.1em]
+\small{\bfseries #1} #2\; --\; #3\\
 \quad \footnotesize #4 \\[-0.2em]
-\quad \footnotesize \emph{#5} \\[0.4em]
+\quad \footnotesize \emph{#5} \\[0.6em]
 }
-
-\begin{tabular}{|l|}
+\hspace*{-0.32cm}\begin{tabular}{|l|}
 \hline
@@
    compute which items are completed and remove them based on a formula result.}
-  {Programming by Demonstration, Local-First Collaboration, Incremental Recomputation}
+  {Programming by Demonstration, Collaborative Editing, Incremental Recomputation}
 \extablecol{Conference List}{\cite{edwards-2025-schema}}
@@
 \label{fig:examples}
-\vspace{-1em}
 \end{figure}
 
+\section{Formative Examples}
+\label{app:examples}
+
+The design the Denicek substrate, we identified six formative examples shown in
+Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
+(Todo and Counter apps) to cases from literature~\cite{edwards-2022-copypaste} and
+problems posed as schema change challenges \cite{edwards-2025-schema}.
+The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
+built (as directly as possible) on top of the substrate and was used to solve implement the
+formative examples.
+
+Many of the formative examples include a small programming challenge, such as adding
+user interface to add a new speaker, a new list item or modify the count. Our aim was for
+the substrate to enable solving  those through programming by demonstration.
+Programming by Demonstration is often used in data wrangling
+\cite{gulwani-2012-spreadsheets,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
+is only a minimalistic illustration of such use, loosely inspired by earlier work \cite{miller-2001-simult}.
+
+\section{Merging Edit Histories}
+Recall that merging takes two edit histories, $E, E_1$ and $E, E_2$,
+transforms edits $E_2$ into $E_2'$ that can be reapplied on top of the first history
+resulting in $E, E_1, E_2'$. The key operation takes two individual edits, $e_1$ and $e_2$
+and produces a sequence of edits $e_2', e_2'', \ldots$ that can be applied after $e_1$,
+and combine the two edits. This section provides details about the two aspects of this operation.
+
+\subsection{Apply to Newly Added}
+\label{app:merge-apply-to-added}
+
+Assume that edits $e_1$ and $e_2$ occurred independently. We want to modify $e_2$ so that
+it can be placed after $e_1$. If the edit $e_2$ added new nodes to the document that the
+edit $e_1$ would affect, we generate an additional edit that apply the transformation of $e_1$
+to the newly added nodes (and only to those).
+
+The only edits that add new document nodes are \ident{Add}, \ident{Append}, \ident{Copy} and
+so we consider this case if the edit $e_2$ is one of those. If so, we check
+whether the target of $e_1$ is within the target of $e_2$, i.e., the list of selectors that
+forms the target of $e_2$ is a prefix of the list of selectors that forms the target of $e_1$.
+
+Along the way, we compute a \emph{more specific prefix}. If the target of $e_1$ contains
+the \ident{All} selector, it can be matched against a specific \ident{Index} selector in the
+target of $e_2$ (if the selector of $e_1$ is more specific than that of $e_2$, the targets
+are not matched). We then replace the original prefix in $e_1$ with the \emph{more specific prefix}
+that contains \ident{Index} selector in places where the original edit contained \ident{All}.
+This way, we obtain $e_1'$ which is a focused version of $e_1$ that applies only to the
+nodes newly added by $e_2$. The edit $e_2$ thus becomes a pair of edits $e_2, e_1'$.
+The final document will contain edits $e_1, e_2, e_1'$ -- that is, it will first apply the edit
+$e_1$ to nodes already in the document, then add new nodes and then apply the transformation
+represented by $e_1$ to the newly added nodes.
+
+\subsection{Transform Matching References}
+\label{app:merge-transform-refs}
+
+As above, assume that edits $e_1$ and $e_2$ occurred independently. We want to
+modify $e_2$ so that it can be placed after $e_1$. If $e_1$ is any of the three edits
+listed in Fig.~\ref{fig:updates} (\ident{RenameField}, \ident{WrapRecord}, \ident{WrapList}),
+we collect all references that appear inside $e_2$ (the target, the source of \ident{Copy}
+and any references occurring in the nodes added by \ident{Add} or \ident{Append}).
+If the target of $e_1$ is a prefix of any of those references, we update the references
+accordingly and obtain a new edit $e_2'$. Note that it would be an error to match specific
+\ident{Index} in $e_1$ with more general \ident{All} in $e_2$, but this cannot happen --
+reference updating is not done when the target of $e_1$ contains \ident{Index}.
+
+Now consider the case when $e_1$ is \ident{Copy} and the edit $e_2$ targets a node that is
+the source node of the copy operation (or any of its children). In this case, it is reasonable
+to require that the edit $e_2$ is applied to both the source and the target of the copy.
+(This is required by the refactoring done in the Conference Budget example.)
+We handle the case by creating a copy of $e_2$ with transformed selectors (target and, if $e_2$
+is also \ident{Copy}, also its source). To transform the selectors, we replace the prefix
+formed by the source of the \ident{Copy} by a new prefix, formed by the target target of
+the \ident{Copy}. We then add the new operation as $e_2'$ if at least one of its selectors
+was transformed (typically target, but possibly also source).
+
 \end{document}

### 2025-07-19 paper.tex
Add ACM submission

@@
 \title[Denicek: Computational Substrate for Document-Oriented End-User
-  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming\\[0.25em]}
+  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming}
 
@@
 
-\keywords{Programming Systems, Computational Substrate, End-User Programming,
- Programming by Demonstration, Local-First Software}
+\begin{CCSXML}
+ <ccs2012>
+    <concept>
+        <concept_id>10011007.10011006.10011066</concept_id>
+        <concept_desc>Software and its engineering~Development frameworks and environments</concept_desc>
+        <concept_significance>500</concept_significance>
+        </concept>
+    <concept>
+        <concept_id>10003120.10003121.10003129.10011756</concept_id>
+        <concept_desc>Human-centered computing~User interface programming</concept_desc>
+        <concept_significance>300</concept_significance>
+        </concept>
+    <concept>
+        <concept_id>10003120.10003121.10003124.10010868</concept_id>
+        <concept_desc>Human-centered computing~Web-based interaction</concept_desc>
+        <concept_significance>300</concept_significance>
+        </concept>
+  </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Development frameworks and environments}
+\ccsdesc[300]{Human-centered computing~User interface programming}
+\ccsdesc[300]{Human-centered computing~Web-based interaction}
+
+\keywords{Programming Systems, End-User Programming}
+%Computational Substrate,  Programming by Demonstration, Local-First Software}
 
 \begin{teaserfigure}
-  \vspace{-0.2em}
+  \vspace{0.25em}
   \includegraphics[width=1\textwidth,clip,trim=0cm 8.5cm 7.7cm 0cm]{fig/teaser.pdf}
@@
 
-\begin{figure}
+\begin{figure*}
 \newcommand{\extablecol}[5]{
@@
 \quad \footnotesize #4 \\[-0.2em]
-\quad \footnotesize \emph{#5} \\[0.6em]
+\quad \footnotesize Programming experiences: \emph{#5} \\[0.6em]
 }
-\hspace*{-0.32cm}\begin{tabular}{|l|}
+\begin{tabular}{|l|}
 \hline
@@
   {Counter with increment and decrement buttons.}
-  {The current count is represented by a formula that is modified by the buttons.\\[-0.2em]\quad\footnotesize
+  {The current count is represented by a formula that is modified by the buttons.
   The user can inspect the evaluation trace to see how the count was modified.}
@@
 \extablecol{Todo App}{\cite{osmani-2024-todomvc}}
-  {Buttons to add an item and remove all completed.}
-  {Adding an item must correctly merge with independently added functionality to \\[-0.2em]\quad\footnotesize
+  {Todo with buttons to add an item and remove all completed items.}
+  {Adding an item must correctly merge with independently added functionality to
    compute which items are completed and remove them based on a formula result.}
-  {Programming by Demonstration, Collaborative Editing, Incremental Recomputation}
+  {Collaborative Editing, Programming by Demonstration, Incremental Recomputation, Schema Change Control}
 \extablecol{Conference List}{\cite{edwards-2025-schema}}
-  {Manage a list of invited conference speakers.}
-  {Adding speakers to a list through an in-document user interface merges with \\[-0.2em]\quad\footnotesize
+  {Managing a list of invited conference speakers and schema change.}
+  {Adding speakers to a list through an in-document user interface merges with
    refactoring that turns the list into a table and separates name from an email.}
-  {Local-First Collaboration, Programming by Demonstration}
+  {Collaborative Editing, Programming by Demonstration, Schema Change Control}
 \extablecol{Conference Budget}{\cite{edwards-2025-schema}}
-  {Calculate budget based on a speaker list.}
-  {References are updated  when the list is refactored. Only affected formulas\\[-0.2em]\quad\footnotesize
+  {Calculate budget based on a speaker list or a table.}
+  {References are updated  when the list is refactored. Only affected formulas
    are recomputed and the user can view elements on which the result depends.}
-  {Local-First Collaboration, Incremental Recomputation, End-User Debugging}
+  {Collaborative Editing, Incremental Recomputation, Schema Change Control, End-User Debugging}
 \extablecol{Hello World}{\cite{miller-2001-simult}}
   {Normalize the capitalization of two word messages.}
-  {An operation to normalize the text in a list item can be recorded and applied to\\[-0.2em]\quad\footnotesize
+  {An operation to normalize the text in a list item can be recorded and applied to
    all list items or, alternatively, applied directly to all list items.}
@@
   {Compute statistics using two data sources.}
-  {Formula to compute statistics can be reused with a different data source;\\[-0.2em]\quad\footnotesize
+  {Formula to compute statistics can be reused with a different data source;
    error correction is propagated automatically to the copied version of the formula.}
@@
 \label{fig:examples}
-\end{figure}
+\end{figure*}
 

### 2025-07-22 submit/uist25-52/Source/paper.tex
ACM Reference Format
Replace tcolorbox and lettrine
Adjust layout
remove hhline package
remove \setlist
Convert \Verb to \texttt

--- 2025-07-19 paper.tex
@@
 
-\usepackage{hhline,xcolor,colortbl}
+\usepackage{xcolor,colortbl}
 \usepackage{fontawesome}
@@
 \usepackage{tikz}
-\usepackage[breakable,theorems,skins]{tcolorbox}
-\usepackage{lettrine}
+%\usepackage[breakable,theorems,skins]{tcolorbox}
+%\usepackage{lettrine}
+\usepackage{float}
 \usepackage{hyperref}
@@
 
-\DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
-  boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
-  colback=gray!20,colframe=gray!20,
-  enlarge bottom by=0pt,enlarge top by=0pt,
-  arc=0pt,outer arc=0pt]
-\lettrine[lraise=0.3]{\LARGE #1}{~}
-\small \textbf{#2.} #3
-\end{tcolorbox}
-}
-
-\setlist{leftmargin=1.5em}
+% \DeclareRobustCommand{\keyideabox}[3]{\begin{tcolorbox}[breakable,
+%   boxsep=5pt,left=0pt,right=0pt,top=0pt,bottom=0pt,width=\dimexpr\columnwidth\relax,
+%   colback=gray!20,colframe=gray!20,
+%   enlarge bottom by=0pt,enlarge top by=0pt,
+%   arc=0pt,outer arc=0pt]
+% \lettrine[lraise=0.3]{\LARGE #1}{~}
+% \small \textbf{#2.} #3
+% \end{tcolorbox}
+% }
+\DeclareRobustCommand{\keyideabox}[3]
+{\vspace{\dimexpr\baselineskip\relax} \noindent\colorbox{gray!20}{
+\parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
+{\small {#1} \textbf{#2.} #3}
+}}
+
+% \setlist{leftmargin=1.5em}
 \setlength\itemsep{0.5em}
@@
 \includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
-\vspace{-0.5em}
+% \vspace{-0.5em}
 \caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
@@
 Evelyn starts with an empty document, which is represented as a record. She adds a field for
-each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. \diff{She
+each heading and a field named {\small\texttt{speakers}}
+containing a list {\small\texttt{<ul>}}. \diff{She
 uses the command toolbox to issue edit commands that add the first speaker.}
@@
 further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
-the details of another speaker \diff{into the textbox}, adds a new {\small\Verb_<li>_} element
+the details of another speaker \diff{into the textbox}, adds a new {\small\texttt{<li>}} element
 and copies the speaker details from the textbox into the new element in the source view,
@@
 Evelyn opens the history view, selects edits that added the speaker and saves them as the
-{\small\Verb_add-speaker_} interaction. She attaches this interaction as an event handler for the
-{\small\Verb_click_} event of the button.
+{\small\texttt{add-speaker}} interaction. She attaches this interaction as an event handler for the
+{\small\texttt{click}} event of the button.
 
@@
 the initial document. \diff{She uses the command toolbox to create document nodes representing
-formulas. Formulas are regular nodes with a special} {\small\Verb_x-formula_} \diff{tag and
-arguments as child nodes. To count the speakers, she uses the} {\small\Verb_count_} \diff{builtin
+formulas. Formulas are regular nodes with a special} {\small\texttt{x-formula}} \diff{tag and
+arguments as child nodes. To count the speakers, she uses the} {\small\texttt{count}} \diff{builtin
 with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
@@
 &&\\[-1em]
-\seltablecol{\Verb|..|}{Parent}{Refers to a parent of a node}
-\seltablecol{\Verb|field|}{Field}{Refers to record field of a given name}
-\seltablecol{\Verb|\#index|}{Index}{Refers to list element at a given index}
-\seltablecol{\Verb|*|}{Any}{Refers to all children of a list node}
+\seltablecol{\texttt{..}}{Parent}{Refers to a parent of a node}
+\seltablecol{\texttt{field}}{Field}{Refers to record field of a given name}
+\seltablecol{texttt{\textbackslash\#index}}{Index}{Refers to list element at a given index}
+\seltablecol{\texttt{*}}{Any}{Refers to all children of a list node}
 \hline
@@
 \ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
-  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
+  {Ordered list of nodes, addressable by \textit{index}. Renders as \texttt{<tag>} with children.}
 \ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
-  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
+  {Record with children addressable by \textit{field}. Renders as \texttt{<tag>} with children.}
 \ndtablecol{\faExternalLink}{Reference}{selectors}
-  {Reference to another document location. Displays the \Verb|selectors| as a link.}
+  {Reference to another document location. Displays the \texttt{selectors} as a link.}
 \ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
@@
 Note that the affected references in the document may be more specific than the edit target. For
-example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|}, a reference
-{\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. (A reference in the
+example, if we rename {\small\texttt{old}} to {\small\texttt{new}} at {\small\texttt{/foo/*}}, a reference
+{\small\texttt{/foo/3/old}} will become {\small\texttt{/foo/3/new}}. (A reference in the
 document cannot be more general; a more specific edit would have to contain the \ident{Index} selector
@@
 
-To illustrate merging, consider a case where we created a list of work items {\small\Verb|/todo|}.
+To illustrate merging, consider a case where we created a list of work items {\small\texttt{/todo}}.
 In one branch, we add an additional item to the list ($e_2$). In another branch, we wrap the
@@
 \begin{array}{rcl}
-e_2 &=& \srckvd{Append}(\textnormal{\small\Verb|/todo|}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
-e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
-e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+e_2 &=& \srckvd{Append}(\textnormal{\small\texttt{/todo}}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
+e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
+e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
 \end{array}
@@
 \begin{array}{rcl}
-  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\Verb|/todo|}, \srcid{<div>}, \srcid{items})\\
-  e_1' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/*|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
-e_2'  &=& \srckvd{Append}(\textnormal{\small\Verb|/todo/items|}, \#1, \#0, \srcid{<li>Do some work</li>})\\
-e_2'' &=& \srckvd{Add}(\textnormal{\small\Verb|/todo/items/\#1|}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+  e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
+  e_1' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/*}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
+e_2'  &=& \srckvd{Append}(\textnormal{\small\texttt{/todo/items}}, \#1, \#0, \srcid{<li>Do some work</li>})\\
+e_2'' &=& \srckvd{Add}(\textnormal{\small\texttt{/todo/items/\#1}}, \srcid{done}, \srckvd{nil}, \srcid{<input type="checkbox"/>})\\
 \end{array}
@@
 \\[-1em]
-\tttablecol{RenameField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
+\tttablecol{RenameField}{target, old field, new field}{\texttt{/target/old\_field/nested}}{\texttt{/target/new\_field/nested}}
   {Replace Field for matching references.}
-\tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
+\tttablecol{WrapRecord}{target, tag, field}{\texttt{/target/nested}}{\texttt{/target/field/nested}}
   {Insert extra Field selector after matching prefix.}
-\tttablecol{WrapList}{target, index, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
+\tttablecol{WrapList}{target, index, tag}{\texttt{/target/nested}}{\texttt{/target/*/nested}}
   {Insert extra All selector after matching prefix.}
@@
 First, Webnicek records the saved edits in the document itself (by representing individual edits
-as nodes and storing them in a list inside the {\small\Verb_/saved-interactions_} field).
+as nodes and storing them in a list inside the {\small\texttt{/saved-interactions}} field).
 This means that no other implementation mechanism outside of Denicek is needed and also
@@
 Programming by demonstration can be used to define interactive elements in the document. In a
-simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\Verb_click_} event
+simple scenario, illustrated in Fig.~\ref{fig:interactive}~(1), the {\small\texttt{click}} event
 handler is set to a reference to a sequence of edits recorded in the document. Clicking the button
@@
 in Fig.~\ref{fig:interactive}~(2), a button to remove all completed Todo items can be created by
-generalising the {\small\Verb_remove-item_} interaction, which removes the list item at the index 0.
+generalising the {\small\texttt{remove-item}} interaction, which removes the list item at the index 0.
 In addition to the recorded interaction, we manually specify (in the source view) that the edits
-should be applied to all elements selected by the {\small\Verb_/items/*_} selector, instead of
-the original {\small\Verb_/items/0_} selector (prefix) and that the edits should only be
+should be applied to all elements selected by the {\small\texttt{/items/*}} selector, instead of
+the original {\small\texttt{/items/0}} selector (prefix) and that the edits should only be
 applied to elements for which the formula (which tests if the checkbox is checked) specified by a
-relative selector {\small\Verb_./condition/comp/result_} evaluates to true.
+relative selector {\small\texttt{./condition/comp/result}} evaluates to true.
 
@@
 As illustrated in Fig.~\ref{fig:counter}, formulas are represented as document nodes with a
-special tag ({\small\Verb_<x-formula>_}). They are recognized by a formula evaluator and
+special tag ({\small\texttt{<x-formula>}}). They are recognized by a formula evaluator and
 rendered in a special way (4), but they are created using ordinary
@@
 
-To evaluate formulas, the formula evaluator generates edits that turn the {\small\Verb_<x-formula>_} (2)
-record into {\small\Verb_<x-evaluated>_} (3), which keeps the previous
-formula state in the {\small\Verb_formula_} field and the evaluation result in the {\small\Verb_result_} field.
+To evaluate formulas, the formula evaluator generates edits that turn the {\small\texttt{<x-formula>}} (2)
+record into {\small\texttt{<x-evaluated>}} (3), which keeps the previous
+formula state in the {\small\texttt{formula}} field and the evaluation result in the {\small\texttt{result}} field.
 Keeping the previous state of the formula is not necessary, but it enables provenance analysis
@@
 The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
-Denicek generates edits that wrap the formula in {\small\Verb_<x-evaluated>_} and set the result
+Denicek generates edits that wrap the formula in {\small\texttt{<x-evaluated>}} and set the result
 to the computed value. In Webnicek, those evaluated edits are kept in a separate list that is always
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 \centering
@@
 
-Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\Verb_<ul>_} is
-turned into {\small\Verb_<tbody>_} using \ident{UpdateTag} and wrapped inside {\small\Verb_<table>_}
-with a field {\small\Verb_body_} using \ident{WrapRecord}. For the latter, Denicek updates the
-reference accordingly turning the original {\small\Verb_/speaker_} reference in the formula
-into {\small\Verb_/speaker/body_}.
+Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\texttt{<ul>}} is
+turned into {\small\texttt{<tbody>}} using \ident{UpdateTag} and wrapped inside {\small\texttt{<table>}}
+with a field {\small\texttt{body}} using \ident{WrapRecord}. For the latter, Denicek updates the
+reference accordingly turning the original {\small\texttt{/speaker}} reference in the formula
+into {\small\texttt{/speaker/body}}.
 
@@
 Edits produced by document evaluation transform the formula structure (wrapping it in the
-{\small\Verb_<x-evaluated>_} record). However, the reference behaviour of those edits is set
+{\small\texttt{<x-evaluated>}} record). However, the reference behaviour of those edits is set
 to keep references unchanged (as discussed in \S\ref{sec:system-defs}) and so references to
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 \includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
@@
 The implementation leverages the fact that evaluation wraps the previous state of the formula
-(\S\ref{sec:impl-eval}). When a formula is fully evaluated, the {\small\Verb_<x-evaluated>_}
+(\S\ref{sec:impl-eval}). When a formula is fully evaluated, the {\small\texttt{<x-evaluated>}}
 document node contains the result, but also a sub-tree with the full evaluation trace
@@
 \diff{The functionality is illustrated in Fig.~\ref{fig:copypaste}, which uses the builtin operation}
-{\small\Verb_read-csv_} \diff{to load a table from a file. The operation imports the data as a
+{\small\texttt{read-csv}} \diff{to load a table from a file. The operation imports the data as a
 document table with rows that can be addressed by subsequent formulas. The then user writes two
@@
 The structural invariants are violated when constructing values gradually (e.g., when adding a new
-speaker to a table), as well as during evaluation (wrapping {\small\Verb_<x-formula>_}
-in {\small\Verb_<x-evaluated>_}). A system with static types could use a form of edit
+speaker to a table), as well as during evaluation (wrapping {\small\texttt{<x-formula>}}
+in {\small\texttt{<x-evaluated>}}). A system with static types could use a form of edit
 transactions, where structural invariants are reestablished after a sequence of edits.
@@
 of individual edits. This design also limits the expressiveness of selectors. For example,
-an edit targeting nodes with a specific tag (e.g., all {\small\Verb_<h3>_} elements) would also be
+an edit targeting nodes with a specific tag (e.g., all {\small\texttt{<h3>}} elements) would also be
 conditional.
@@
 % design decision that merging should not depend on document state. (If an edit wraps the body of all
-% {\small\Verb_<h3>_} elements, we need to know the document state in order to determine whether
+% {\small\texttt{<h3>}} elements, we need to know the document state in order to determine whether
 % we need to modify the selector pointing to an item at a specific index.)
@@
 represented using nested formulas. Evaluation of formulas proceeds as in Webnicek
-and wraps {\small\Verb_<x-formula>_} in {\small\Verb_<x-evaluated>_}.
+and wraps {\small\texttt{<x-formula>}} in {\small\texttt{<x-evaluated>}}.
 The user interface displays the original formula alongside with the evaluated result, supporting
@@
 Grid cells consist of a reference to the data source they edit and a collection of
-{\small\Verb_<transform>_} nodes that represent individual data transformations constructed via
+{\small\texttt{<transform>}} nodes that represent individual data transformations constructed via
 programming by demonstration. A transformation consists of metadata,
-a sequence of edits and a formula representing the transformation (with {\small\Verb_<x-hole>_}
+a sequence of edits and a formula representing the transformation (with {\small\texttt{<x-hole>}}
 standing for the data table to be transformed).
@@
 To implement conditional transformations, we add a special kind of (virtual) edit
-{\small\Verb_<x-expandable-edit>_} that stores the underlying edit and a condition.
+{\small\texttt{<x-expandable-edit>}} that stores the underlying edit and a condition.
 As in \S\ref{sec:impl-interaction}, when applying the edit, Datnicek finds all rows for which the
@@
 changing a primitive value is an edit, and recommendations in the interactive grid are also edits
-(appending a {\small\Verb_<transform>_} node).
+(appending a {\small\texttt{<transform>}} node).
 
@@
 
-\bibliographystyle{plain}
+%\bibliographystyle{plain}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{paper}

### 2025-07-31 submit/uist25-52/Source/paper.tex
Replace tikz for circled letters
TAPS revisions

@@
 \usepackage{fancyvrb}
-\usepackage{tikz}
+% \usepackage{tikz}
 %\usepackage[breakable,theorems,skins]{tcolorbox}
@@
 
+\usepackage{acmart-taps}
+
 \setlength{\marginparwidth}{1.4cm}
@@
 
-\newcommand*\circled[1]{\textnormal{\footnotesize\sffamily\bfseries\protect\tikz[baseline=(char.base)]{
-  \node[shape=circle,fill=black,text=white,draw,inner sep=1pt] (char) {#1};}}}
+% \newcommand*\circled[1]{\textnormal{\footnotesize\sffamily\bfseries\protect\tikz[baseline=(char.base)]{
+%   \node[shape=circle,fill=black,text=white,draw,inner sep=1pt] (char) {#1};}}}
+
+\newcommand*\circled[1]
+{{\Large \textcircled{\footnotesize {#1}}}}
 
@@
 \DeclareRobustCommand{\keyideabox}[3]
-{\vspace{\dimexpr\baselineskip\relax} \noindent\colorbox{gray!20}{
+{\vspace{\dimexpr\baselineskip\relax} \colorbox{gray!20}{
 \parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
@@
 Those experiences advance the state of the art of end-user programming, but they are hard to
-implement on the basis of established programming languages and system.
+implement on the basis of established programming languages and system.\enlargethispage{12pt}
 
@@
 
-A computational substrate defines the structures with which programs are constructed, how
+A computational\enlargethispage{12pt} substrate defines the structures with which programs are constructed, how
 the program state is represented and how the state evolves during execution \cite{jakubovic-2022-ladder}.
@@
 \begin{array}{rcl}
-e_2 &=& \srckvd{Append}(\textnormal{\small\texttt{/todo}}, \#1, \#0, \srcid{<li>Do some work</li>})\\[0.5em]
+e_2 &=& \srckvd{Append}(\textnormal{\small\texttt{/todo}}, \#1, \#0, \srcid{<li>Do some work</li>})\\
 e_1 &=& \srckvd{WrapRecord}(\textnormal{\small\texttt{/todo}}, \srcid{<div>}, \srcid{items})\\
@@
 %
-\begin{enumerate}[itemsep=0pt,label=(\roman*),topsep=3pt,leftmargin=20pt]
-\item collaborative document editing (\S\ref{sec:impl-collab}),
-\item programming by demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
-\item incremental recomputation (\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}),
-\item schema change control (\S\ref{sec:impl-schema}),
-\item end-user debugging via provenance tracking (\S\ref{sec:impl-provenance}),
-\item concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
+\begin{enumerate}%[itemsep=0pt,label=(\roman*),topsep=3pt,leftmargin=20pt]
+\item[(i)] collaborative document editing (\S\ref{sec:impl-collab}),
+\item[(ii)] programming by demonstration (\S\ref{sec:impl-pbd} and \S\ref{sec:impl-interaction}),
+\item[(iii)] incremental recomputation (\S\ref{sec:impl-eval} and \S\ref{sec:impl-incremental}),
+\item[(iv)] schema change control (\S\ref{sec:impl-schema}),
+\item[(v)] end-user debugging via provenance tracking (\S\ref{sec:impl-provenance}),
+\item[(vi)] concrete programming via managed copy \& paste (\S\ref{sec:impl-copy}).
 \end{enumerate}

### 2025-08-06 submit/uist25-52/Source/paper.tex
Fix keyidea boxes and circled letters
Tweak table widths
fix orphan in intro
Adjust margin fig 3
Page break tweaks, margins, circled hack

@@
 \newcommand*\circled[1]
-{{\Large \textcircled{\footnotesize {#1}}}}
+{{\normalfont \Large \textcircled{\footnotesize {#1}}}}
 
@@
 \DeclareRobustCommand{\keyideabox}[3]
-{\vspace{\dimexpr\baselineskip\relax} \colorbox{gray!20}{
-\parbox{\dimexpr\columnwidth-\marginparsep+1pt\relax}
-{\small {#1} \textbf{#2.} #3}
+{\vspace{\dimexpr0.5\baselineskip\relax} \noindent\setlength\fboxsep{6pt}\colorbox{gray!20}{%
+\parbox{\dimexpr\columnwidth-\marginparsep\relax}
+{\small {#1}\; \textbf{#2.} #3}
 }}
@@
 \item \emph{Collaborative Editing.} Users should be able to locally modify a shared document
-  and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
-\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
-  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
+  and merge their changes in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
+\item \emph{Programming by Demonstration.} Allow users to construct programs by
+  enacting the steps of the desired behavior using concrete examples and generalizing from those~\cite{leiva-2021-rapido,cypher-1993-pbd}.
 \item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
@@
   data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
-\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
-  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
+\item \emph{End-User Debugging.} The user should be able to ask about provenance~\cite{cheney-2009-provenance} to understand
+  why a computation resulted in a particular result and what inputs contributed to it \cite{ko-2009-whyline}.
 \item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
@@
   implemented on top of the Denicek substrate (\S\ref{sec:impl}).
-\item We document important design decisions, alternatives and limitations (\S\ref{sec:discuss}).
-  The analysis shows that the desired functionality requires a careful choice among interconnected
+\item We analyse key design decisions, alternatives and limitations (\S\ref{sec:discuss}), showing
+  that the desired functionality requires a careful choice among interconnected
   design options.
 \item To evaluate how Denicek simplifies the development of program\-ming systems, we
-  build an innovative data science notebook Datnicek and assess its implementation
-  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system (\S\ref{sec:eval}).
+  build an innovative data science notebook and assess its implementation
+  complexity~(\S\ref{sec:case}). We also present heuristic evaluation of the system~(\S\ref{sec:eval}).
 \end{itemize}
@@
 
-\subsubsection*{\circled{A} Adding a Speaker}
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=0em 0em 30mm 0em]{fig/circled.pdf}} Adding a Speaker}
 Evelyn starts with an empty document, which is represented as a record. She adds a field for
@@
 
-\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=5mm 0em 25mm 0em]{fig/circled.pdf}} Creating a User Interface} To simplify adding of
 further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
@@
 
-\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker,
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=10mm 0em 20mm 0em]{fig/circled.pdf}} Abstracting the Interaction} After adding the speaker,
 Evelyn opens the history view, selects edits that added the speaker and saves them as the
@@
 
-\subsubsection*{\circled{D} Refactoring Document Structure} Alfred starts with the initial
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=15mm 0em 15mm 0em]{fig/circled.pdf}} Refactoring Document Structure} Alfred starts with the initial
 version of the document (A) and turns the list into a table. He invokes a series of commands that
@@
 
-\subsubsection*{\circled{E} Merging Edits}
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=20mm 0em 10mm 0em]{fig/circled.pdf}} Merging Edits}
 \diff{Alfred then merges the later edits (B, C) done by Evelyn into his version of the document.}
@@
 
-\subsubsection*{\circled{F} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=25mm 0em 5mm 0em]{fig/circled.pdf}} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
 the initial document. \diff{She uses the command toolbox to create document nodes representing
@@
 
-\subsubsection*{\circled{G} Merging Formulas} When Juliana merges the budget calculation (F) with
+\subsubsection*{\raisebox{-0.2em}{\includegraphics[height=1em,clip,trim=30mm 0em 0mm 0em]{fig/circled.pdf}} Merging Formulas} When Juliana merges the budget calculation (F) with
 the earlier edits (E), references in formulas are automatically updated to point to the \diff{rows of the table}.
@@
 
-\begin{tabular}{|llp{18.08em}|}
+\begin{tabular}{|llp{16em}|}
 \hline
@@
 
-\begin{tabular}{|cl|}
+\begin{tabular}{|cp{23.48em}|}
 \hline
@@
   whose structure is changed by $e_1$, the target reference in $e_2$ is updated in a way that
-  corresponds to the new structure. This is done using the rules shown in Fig.~\ref{fig:updates},
+  corresponds to the new structure. This is done using the rules in Fig.~\ref{fig:updates},
   that apply when transforming references inside a document, although we now also support the case
-  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs} for details).
+  when $e1$ is \ident{Copy} (see \S\ref{app:merge-transform-refs}).
 \end{enumerate}
@@
 }
-\begin{tabular}{|p{27em}|}
+\begin{tabular}{|p{25.5em}|}
 \hline
 \\[-1em]
-\tttablecol{RenameField}{target, old field, new field}{\texttt{/target/old\_field/nested}}{\texttt{/target/new\_field/nested}}
+\tttablecol{RenameField}{target, old, new}{\texttt{/target/old\_field/nested}}{\texttt{/target/new\_field/nested}}
   {Replace Field for matching references.}
 \tttablecol{WrapRecord}{target, tag, field}{\texttt{/target/nested}}{\texttt{/target/field/nested}}
-  {Insert extra Field selector after matching prefix.}
+  {Insert Field selector after matching prefix.}
 \tttablecol{WrapList}{target, index, tag}{\texttt{/target/nested}}{\texttt{/target/*/nested}}
@@
 }
-\begin{tabular}{|p{27em}|}
+\begin{tabular}{|p{25.5em}|}
 \hline
@@
   {\ident{RenameField}, \ident{DeleteField}, \ident{WrapRecord}, \ident{WrapList}, \ident{Copy}}
-\eftablecol{ValueEffect}{Transforms value, modifies list or adds an additional field.}
+\eftablecol{ValueEffect}{Transforms value, modifies list or adds a field.}
   {\ident{Add}, \ident{Append}, \ident{Reorder}, \ident{DeleteItem}, \ident{PrimitiveEdit}}
@@
 \begin{figure}[h]
+\vspace{0.5em}
 \includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
-\vspace{-1em}
 \caption{Using provenance tracking to highlight document parts that contributed to the
@@
 \begin{figure}[t]
-\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
+\includegraphics[width=1\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
 \caption{When an edit changes the document structure, references in formulas are updated accordingly.}
 \label{fig:coevolution}
+\vspace{0.5em}
 \end{figure}
@@
   as in Histogram \cite{petricek-2019-histogram}. This allows Datnicek to keep track of the code
-  structure and potentially makes the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
+  structure and can make the system accessible to non-programmers \cite{mcnutt-2023-projectional}.
   We do not initially aim to implement advanced keyboard-based editing \cite{moon-2022-tylr,beckman-2023-sandblocks}.
@@
   transformations and operations as when using type providers \cite{syme-2013-inforich} or
-  iterative prompting in The Gamma \cite{petricek-2022-thegamma}.
+  when using iterative prompting in The~Gamma \cite{petricek-2022-thegamma}.
 
@@
 \item \emph{Wrangling Code Synthesis.} Interactively constructed data transformations should be
-  convertible into code that can be checked by the data scientist and further edited as in Wrex \cite{drossos-2020-wrex}.
+  convertible into code that can be checked and further edited as in Wrex \cite{drossos-2020-wrex}.
 \end{itemize}
@@
 \begin{acks}
-The authors thank Clemens Nylandsted Klokmose for research guidance and
+The authors thank Clemens Nylandsted Klokmose for research guidance,
 attendees of the IFIP WG 2.16 meeting in Serpiano, members of the Software Architecture Group at HPI
@@
 
-\begin{figure*}
-\newcommand{\extablecol}[5]{
-\small{\bfseries #1} #2\; --\; #3\\
-\quad \footnotesize #4 \\[-0.2em]
-\quad \footnotesize Programming experiences: \emph{#5} \\[0.6em]
+\section{Formative Examples}
+\label{app:examples}
+
+The design the Denicek substrate, we identified six formative examples shown in
+Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
+(Todo and Counter apps) to cases from literature~\cite{edwards-2022-copypaste} and
+problems posed as schema change challenges \cite{edwards-2025-schema}.
+The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
+built (as directly as possible) on top of the substrate and was used to solve implement the
+formative examples.
+
+Many of the formative examples include a small programming challenge, such as adding
+user interface to add a new speaker, a new list item or modify the count. Our aim was for
+the substrate to enable solving  those through programming by demonstration.
+Programming by Demonstration is often used in data wrangling
+\cite{gulwani-2012-spreadsheets,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
+is only a minimalistic illustration of such use, loosely inspired by earlier work \cite{miller-2001-simult}.
+
+\section{Merging Edit Histories}
+Recall that merging takes two edit histories, $E, E_1$ and $E, E_2$,
+transforms edits $E_2$ into $E_2'$ that can be reapplied on top of the first history
+resulting in $E, E_1, E_2'$. The key operation takes two individual edits, $e_1$ and $e_2$
+and produces a sequence of edits $e_2', e_2'', \ldots$ that can be applied after $e_1$,
+and combine the two edits. This section provides details about the two aspects of this operation.
+
+\subsection{Apply to Newly Added}
+\label{app:merge-apply-to-added}
+
+Assume that edits $e_1$ and $e_2$ occurred independently. We want to modify $e_2$ so that
+it can be placed after $e_1$. If the edit $e_2$ added new nodes to the document that the
+edit $e_1$ would affect, we generate an additional edit that apply the transformation of $e_1$
+to the newly added nodes (and only to those).
+
+The only edits that add new document nodes are \ident{Add}, \ident{Append}, \ident{Copy} and
+so we consider this case if the edit $e_2$ is one of those. If so, we check
+whether the target of $e_1$ is within the target of $e_2$, i.e., the list of selectors that
+forms the target of $e_2$ is a prefix of the list of selectors that forms the target of $e_1$.
+
+Along the way, we compute a \emph{more specific prefix}. If the target of $e_1$ contains
+the \ident{All} selector, it can be matched against a specific \ident{Index} selector in the
+target of $e_2$ (if the selector of $e_1$ is more specific than that of $e_2$, the targets
+are not matched). We then replace the original prefix in $e_1$ with the \emph{more specific prefix}
+that contains \ident{Index} selector in places where the original edit contained \ident{All}.
+This way, we obtain $e_1'$ which is a focused version of $e_1$ that applies only to the
+nodes newly added by $e_2$. The edit $e_2$ thus becomes a pair of edits $e_2, e_1'$.
+The final document will contain edits $e_1, e_2, e_1'$ -- that is, it will first apply the edit
+$e_1$ to nodes already in the document, then add new nodes and then apply the transformation
+represented by $e_1$ to the newly added nodes.
+
+\begin{figure}[t]
+\newcommand{\extablecolx}[6]{\hline\\[-0.8em]
+\small{\bfseries #1} #2\\[-0.1em] \small #3\\[0.3em]
+\quad \parbox{7cm}{\footnotesize #4}\\
+\quad \parbox{7cm}{\vspace{0.5em}\footnotesize #6: \emph{#5}} \\[1em]
 }
+\newcommand{\extablecol}[5]{\extablecolx{#1}{#2}{#3}{#4}{#5}{Programming experiences}}
 \begin{tabular}{|l|}
-\hline
-\\[-1em]
 \extablecol{Counter App}{\cite{kiss-2014-7guis}}
@@
   {The current count is represented by a formula that is modified by the buttons.
+  Formula modification is shown once and then repeated on click.
   The user can inspect the evaluation trace to see how the count was modified.}
@@
   {Todo with buttons to add an item and remove all completed items.}
-  {Adding an item must correctly merge with independently added functionality to
+  {Functionality to add an item is shown once and then recorded.
+   Adding an item must correctly merge with independently added functionality to
    compute which items are completed and remove them based on a formula result.}
@@
   {Adding speakers to a list through an in-document user interface merges with
-   refactoring that turns the list into a table and separates name from an email.}
+   refactoring that turns the list into a table and separates name from an email by
+   copying the value and removing a part of the string before/after a comma.}
   {Collaborative Editing, Programming by Demonstration, Schema Change Control}
@@
   {Programming by Demonstration}
-\extablecol{Traffic Accidents}{\cite{edwards-2022-copypaste}}
+\extablecolx{Traffic Accidents}{\cite{edwards-2022-copypaste}}
   {Compute statistics using two data sources.}
-  {Formula to compute statistics can be reused with a different data source;
-   error correction is propagated automatically to the copied version of the formula.}
-  {Concrete Programming, Incremental Recomputation}
+  {Formula to compute statistics can be reused with a different data source. If an error
+   in the formula is corrected, the correction is propagated automatically to the copied
+   version of the formula.}
+  {Concrete Programming, Incremental Recomputation}{Experiences}
 \hline
@@
 \label{fig:examples}
-\end{figure*}
-
-\section{Formative Examples}
-\label{app:examples}
-
-The design the Denicek substrate, we identified six formative examples shown in
-Fig.~\ref{fig:examples}. The examples range from established industry benchmarks
-(Todo and Counter apps) to cases from literature~\cite{edwards-2022-copypaste} and
-problems posed as schema change challenges \cite{edwards-2025-schema}.
-The Denicek substrate then co-evolved with Webnicek, a simple web-based programming environment
-built (as directly as possible) on top of the substrate and was used to solve implement the
-formative examples.
-
-Many of the formative examples include a small programming challenge, such as adding
-user interface to add a new speaker, a new list item or modify the count. Our aim was for
-the substrate to enable solving  those through programming by demonstration.
-Programming by Demonstration is often used in data wrangling
-\cite{gulwani-2012-spreadsheets,drossos-2020-wrex,gulwani-2014-flash}. Our Hello World example
-is only a minimalistic illustration of such use, loosely inspired by earlier work \cite{miller-2001-simult}.
-
-\section{Merging Edit Histories}
-Recall that merging takes two edit histories, $E, E_1$ and $E, E_2$,
-transforms edits $E_2$ into $E_2'$ that can be reapplied on top of the first history
-resulting in $E, E_1, E_2'$. The key operation takes two individual edits, $e_1$ and $e_2$
-and produces a sequence of edits $e_2', e_2'', \ldots$ that can be applied after $e_1$,
-and combine the two edits. This section provides details about the two aspects of this operation.
-
-\subsection{Apply to Newly Added}
-\label{app:merge-apply-to-added}
-
-Assume that edits $e_1$ and $e_2$ occurred independently. We want to modify $e_2$ so that
-it can be placed after $e_1$. If the edit $e_2$ added new nodes to the document that the
-edit $e_1$ would affect, we generate an additional edit that apply the transformation of $e_1$
-to the newly added nodes (and only to those).
-
-The only edits that add new document nodes are \ident{Add}, \ident{Append}, \ident{Copy} and
-so we consider this case if the edit $e_2$ is one of those. If so, we check
-whether the target of $e_1$ is within the target of $e_2$, i.e., the list of selectors that
-forms the target of $e_2$ is a prefix of the list of selectors that forms the target of $e_1$.
-
-Along the way, we compute a \emph{more specific prefix}. If the target of $e_1$ contains
-the \ident{All} selector, it can be matched against a specific \ident{Index} selector in the
-target of $e_2$ (if the selector of $e_1$ is more specific than that of $e_2$, the targets
-are not matched). We then replace the original prefix in $e_1$ with the \emph{more specific prefix}
-that contains \ident{Index} selector in places where the original edit contained \ident{All}.
-This way, we obtain $e_1'$ which is a focused version of $e_1$ that applies only to the
-nodes newly added by $e_2$. The edit $e_2$ thus becomes a pair of edits $e_2, e_1'$.
-The final document will contain edits $e_1, e_2, e_1'$ -- that is, it will first apply the edit
-$e_1$ to nodes already in the document, then add new nodes and then apply the transformation
-represented by $e_1$ to the newly added nodes.
+\end{figure}
 
